### 2017-02-02 paper/AMIA/main.tex
added initial latex files for sequential analysis paper

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{multirow}

\begin{document}

\title{Comparison of clinical trial with the results obtained from EHRs along with missing value analysis}

\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, Rick Krajenta, BS$^{2}$, Melody J. Eide, MD, MPH$^{2,3}$, Christine C. Johnson, PhD, MPH$^{4}$}

\institutes{
$^1$Department of Computer Science, Wayne State University, Detroit, Michigan; 
$^2$Departments of Biostatistics and Research Epidemiology,
$^3$Department of Dermatology, 
$^4$Department of Public Health Science, Henry Ford Health System, Detroit, Michigan\\
}

\maketitle

\noindent{\bf Abstract}
\textit{This study compared the PLCO survey response with the results obtained from electronic health records. Our study also explored the missing data problem and offer advice for dealing with missing data. To extract clinical information, we built an integrated information extraction system that leveraged several publicly available clinical NLP and information extraction tools. Experimental results showed the significant difference between these two results, while extraction results achieved 91.2\% to 100\% accuracy. The structure of missing values was analyzed with visualization tools implemented in R-package. We observed MAR situation for certain variables such as ``quit time'', ``smoking duration'', and ``pack-year smoked''. These results indicate that EHR data contained sufficient information can be used directly for certain research. Besides that, a clinical trial might be suffered by missing values or biasing or noise occurred during the trials. These problems can be resolved with EHRs in combination with properly addressing missing values.}

\section*{Introduction}
Electronic health records (EHRs) are increasingly being used to document and store large amounts of clinical information \cite{jha2009use}. They replaced traditional paper-based systems in many healthcare organizations and became a vast data sources which can facilitate clinical research and studies to ultimately improve healthcare quality \cite{blumenthal2010launching, erickstad2011use, lau2011use}. Much of the available clinical data are in narrative form as a result of transcription of dictations, direct entry by providers, or use of speech recognition applications. Comparing to structured data, free text is a more conventional way in the health care environment to express concepts and events. However, free text is very challenging for searching, summarization, decision-support, or statistical analysis. To reduce medical errors, improve health care quality, and enable secondary use of EHRs, information extraction (IE), which structures and encodes clinical information stored in free text, is necessary. However, traditional natural language processing (NLP) tools are not designed for the fragmented free text found in narrative clinical records; therefore, they do not perform well on this type of data.

In the past decade, several different techniques have been developed to extract information, from simple pattern matching to complete processing methods based on symbolic information and rules or based on statistical methods and machine learning. Many prior natural language processing (NLP) studies in the clinical domain have used regular expressions in designing their NLP solutions \cite{stenner2012paste,piwowar2008identifying,denny2009identifying, mccart2012using,xu2011facilitating,matheny2009detection,zeng2006extracting,huang2007novel}. These regular expressions are typically created by software developers working with domain experts. One of the important regular expression based work is NegEx \cite{chapman2001simple}. This is because significant portion of medical text is negated in electronic health records. NegEx depends on a list of phrases that indicate the absence of a particular medical condition such as disease or symptom to negate the conditions within a window of five words or phrases.

An extended version of NegEx called ConText \cite{harkema2009context} was proposed to determine three contextual features including negation (negated, affirmed), temporality (historical, recent, hypothetical) and experiencer (patient, other). To encode these contextual features, several NLP systems have also been designed to encode detailed information from clinical reports, such as MedLEE \cite{friedman2000broad}, MPLUS \cite{christensen2002mplus}, and MedSyndikate \cite{hahn2002medsyndikate}.

However, health information text extraction (HITEx \cite{zeng2006extracting}) was developed to extract diagnoses and smoking status. Statistics shows that smoking causes nearly one of every five deaths in the United States each year. Therefore, extraction of patients smoking status is necessary for clinician and researcher. Mayo Clinic developed an NLP System named clinical text analysis and knowledge extraction system (cTAKES \cite{savova2010mayo}) to extract clinical information from medical domain. Additional modules have been implemented within cTAKES recently \cite{savova2008mayo} for identification of patient smoking status.

While substantial progress has been made in the development of NLP tools, no information extraction (IE) system is perfect for extracting all kind of variables in different domain, and still suffer from various weaknesses. In this study, we built an integrated system that leveraged several existing NLP components. Experimental results shows that our system performed well in biomedical domain.

In this study, we mainly focus on findings from electronic health records and then compare the result with PLCO clinical trial. We also describe missing patterns observed in our experiment. This paper reports that which information are sufficiently available in EHRs for conducting clinical research. It also states that which of the variables are noisy in clinical trials. These findings are significant in clinician and researcher who are conducting their research with clinical trials or EHRs. To the best of our knowledge, there is no report that compare the results between response from clinical trials and NLP-based extraction from EHRs.

\section*{Methods}
\subsection*{\textit{Data collection}}
Four hundred and fifty thousand free-text EHRs of 11,968 patients were obtained from Henry Ford Health System, who participated in the PLCO clinical trial. Out of this 11,968 patients, 1000 patients were randomly selected for this study. The PLCO trial began enrolling participants in November 1993 and completed recruitment of 154,938 subjects over different health systems in June 2001. Men and women aged from 55 to 74 years participated in the PLCO trial at ten screening centers nationwide with balanced randomization to intervention and control arms. All participants received a posterior anterior view chest x-ray for lung cancer screening. A baseline epidemiological questionnaire and a supplemental questionnaire (SQ) were administered as part of the PLCO surveys. The obtained patient EHRs consist of different types of clinical documents including clinical practice guidelines, diagnostic reports and discharge summaries.

\subsection*{\textit{System overview}}
We built an integrated information extraction system that leveraged several publicly available clinical NLP and information extraction tools. In particular, our system consists of the following components: 1) A sentence boundary detection program, which was developed at OpenNLP \cite{morton2005opennlp}; 2) an existing corpus based section identification program (SecTag \cite{denny2009evaluation}); 3) regular expression based NLP program that utilizes the clinical Text Analysis and Knowledge Extraction System (cTAKES \cite{savova2010mayo}) and contextual feature extraction system (ConText \cite{harkema2009context}); 4) post processing program that maps our results into PLCO compareable format. Figure \ref{fig:extraction} shows the architecture of our system.

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{figures/extraction.eps}
\caption{An architecture of the integrated system for information extraction from free-text clinical narrative used in this study.}
\label{fig:extraction}
\end{figure}

\textbf {Sentence splitter}: Sentence detection is important and necessary task for our study because we used sentence as the basic unit of variable extraction. As sentence splitting is at the core of a many NLP activities, it is provided by most NLP frameworks and libraries. We used an open source Java tools openNLP for identifying the sentence boundary. The OpenNLP sentence detector cannot identify sentence boundaries based on the contents of the sentence. However, this tools can detect that a punctuation character marks the end of a sentence or not. In this sense, a sentence is defined as the longest white space trimmed character sequence between two punctuation marks. This rule is not applicable for the first and last sentence. The first non whitespace character is assumed to be the begin of a sentence, and the last non whitespace character is assumed to be a sentence end.

\textbf {Section identification}:
Electronic health records are often divided into sections, or segments, such as ``past medical history'', ``family history'', ``assessment and plan'', etc. One can gain better understanding of clinical documents by recognition of the section in which a concept lives. For instance, both a ``past medical history'' and the ``family history'' sections can contain a list of diseases, but the context describes very different import to the patient about whom the note was written. We used SecTags for accomplishing the following three tasks: 1) excluding false positive by recognizing experiencer (patient or family member) mentioned in the family history sections; 2) determine scope of the temporal information, e.g. all events mentioned in the past medical history were historical event; 3) extract demographic information about patient from the 1st section.

\textbf {NLP program}:
As follows from Figure 1, a regular expression based NLP program was developed that utilizes the clinical Text Analysis and Knowledge Extraction System (cTAKES), which is an open source NLP and IE framework developed by the Mayo Clinic for information extraction from EHRs. This tool is comprised of several important components: sentence boundary detector, tokenizer, part-of-speech tagger, shallow parser, named entity recognizer, concept mapper, negation and status attribute detector. These components were trained on clinical text to create rich linguistic and semantic annotations of EHRs. In our study, concept mapping is used to annotate several important concepts, such as sign and symptoms, disease and disorder, lab procedure, negation and other clinical concepts.

Although most components of cTAKES have been shown to achieve good performance on a variety of NLP tasks, it has some limitations  \cite{savova2010mayo} related to certainty detection and temporal resolution. For this reason, we used ConText and semantic rules to identify temporal information. ConText is an expansion of NegEx. It relies on the same basic algorithm but applies different term lists and different windows of scope depending on the contextual feature being annotated.

In total, eighteen variables were extracted in this study. Since demographic variables are generally mentioned in the header or 1st section of the EHRs, we used this section for the extraction of age, sex and race. Age was determined by calculating the time period between ``date of birth'' and ``visit date or report date'', which were annotated by cTAKES. Sex and race were extracted by using simple regular expression in conjunction with a list of trigger terms such as male, female, white, african american etc. It was also confirm that the scope of the trigger term is outside the family member or relatives. We applied exactly same patterns for the extraction of ``COPD history'', ``emphysema'', ``asthma'', ``pneumonia'' and ``prior cancer''. At first, we find the triggered term or UMLS concept we are interested in. Then, we checked that the concept was not negated or fall into different experiencer such as family member. Finally, we confirmed the certainty of concept so that it was not uncertain or hypothetical such as ``suspected for pneumonia'', ``pneumonia vaccine'', etc. We also applied the semantic rules to extract ``lung cancer family history'' and ``chest x-ray in the past three years''. In addition to that, we  checked the experiencer for ``lung cancer family history'', and compared the ``visit date or report date'' with the x-ray date.

\begin{table}[h!]
\centering
\caption{Example of information extraction rules.}
\label{tab:rule}
  \begin{tabular}{|l|l|}
  \hline
    \textbf{Rule}  & \textbf{Example} \\ \hline

 quit smoking \textless duration\textgreater\hspace{1mm}ago or quit smoking since \textless time\textgreater & quit smoking 30 years ago or quit smoking \\
 or quit smoking from \textless time\textgreater &  since 1960 or quit smoking from 1880 \\ \hline

 \textless amount\textgreater\hspace{1mm}(cigarettes\textbar pack of cigarettes) per (day\textbar week\textbar month) & 10-15 cigarettes per day or \\
 or \textless amount\textgreater\hspace{1mm}(pack\textbar pack-) year of smoking history & 30 pack year of smoking history \\ \hline

 (smoking\textbar smokes\textbar smoked\textbar) for \textless duration\textgreater\hspace{1mm} or & he has been smoking for 45 years or \\
 (smoking\textbar smokes\textbar smoked\textbar) (since\textbar from) \textless year\textgreater\hspace{1mm} &  he has been smoking since 1965 \\ \hline

  \end{tabular}
\end{table}

Smoking status is extracted using a rule based method \cite{cohen2008five, uzuner2008identifying}, which consists of two steps: hot-spot identification and post-processing rules. At first, we find the hot-spots based on keywords and sections (in many cases, smoking status was found in history and demographic sections). When a document containing one of the terms ``smok'', ``tob'', ``packs'', ``cigar'' etc. was found, a window of text up to 100 characters before and after those terms is considered as a hot-spot. Rules are then used to classify smoking status based on temporal information. After extracting the smoking status, we extracted associated information such as smoking duration, quit time and pack-year history. Several rules were employed to extract clinical variables, some of which are shown in Table~\ref{tab:rule}.

\textbf {Post processing and filter}:
To allow direct comparison between the extracted results from EHRs and PLCO results, post processing of the proposed system's output stage was required. The output of NLP program is a set of pivot text snippets such as ``two packs of cigarettes per day'', ``quit smoking in 1993'', ``smoked from the age of 18'', etc. These text fragments are then parsed to generate numeric values. Finally, the output of post-processing was filtered by keeping the latest information, for instance, a non-smoker can be a smoker after few months. To generate numeric values for statistical analysis, simple regular expression is used together with a list of terms such as ``half'', ``two'', ``one and half'', etc.

\subsection*{\textit{Statistical inference and treatment of missing values}}
Missing data frequently occur in electronic health records, which is a very common problem in data mining and statistical inference. According to Little and Rubin \cite{little2014statistical,rubin1976inference}, missing data can be categorized into three types: i) missing completely at random (MCAR)-the absence of a data element is not associated with any other value in the data set, observed or missing; ii) missing at random (MAR)-this is a less restrictive assumption than MCAR; the probability of data being missing depends only on the observed values in the data set, not on unobserved data. The implied information of missingness can be inferred from observed values; ii) not missing at random (NMAR)-the condition is opposite of MAR. The probability of data being missing depends on unobserved (missing) data. Under this circumstance, the missing information cannot be inferred from observed values.

In order to improve the quality of research, it is important to understand the kind of missing data present in EHRs. However, missing values mechanisms are difficult to identify, because it would require the knowledge of the missing values themselves \cite{little2014statistical}. Naturally, MNAR is very difficult to be detected. This is even more complicated in case of multivariate data, outliers, inhomogeneous data or very skewed data distribution. In this paper, we explained the missingness in terms of the above mechanisms with matrix and scatter plots. Several approaches have been used to deal with missing data. The methods for coping with missing values can be grouped into three main categories \cite{little2014statistical}: inference restricted to complete data, imputation-based approaches, and likelihood-based approaches.

The simplest and most common approach to handling missing values is to omit the cases with missing values and do the analysis based only on the complete cases. Data set obtained by the result of this process can be a biased sample of complete cases because the missing of data is not a random process (i.e. data is not always MCAR). Furthermore, excluding patients with missing data will lead to a loss in statistical power.

In imputation-based methods, the missing values are filled in and the resultant data can be analyzed as a complete data set. Imputation of missing data in the setting of MCAR will increase power but should not change the point estimates from those obtained with a complete case analysis. Therefore, it seems reasonable to perform imputations using EHR data under the assumption of MAR. Commonly imputed values are based on the value of known cases: the mean of the variable in either the whole data set or in select data subsets, or an estimated value from regression procedures on known variables. Multiple imputation methods, i.e., filling with more than one value, have been developed to avoid biasing the variances of imputed variables \cite{enders2006primer}.

There exist approaches that derive a prediction model by inferring the model's parameters from the existing data, rather than imputing data where values are missing. Likelihood-based approaches are an example of these. They developed a model by attempting to find the set of model parameters that make the observed data most likely. The resulting system parameters can then be used as the base for future inferences. The expectation maximization (EM \cite{dempster1977maximum}) algorithm is commonly used for finding maximum likelihood estimates in the face of incomplete data.

\subsection*{\textit{Evaluation methods}}
The golden standard for the evaluation of information extraction methods was created based on the EHRs collected from Henry Ford Health Systems. Each patient comprised of several EHR documents. All EHRs were manually investigated by domain expert to find the values of target variables for each patient. Then, we compared the values of extracted categorical and non-categorical (continuous) variables with gold standard to evaluate the accuracy of IE.

\section*{Results}
We present the results and outcomes of our study regarding the performance of information extraction from EHRs and statistical significance between EHR extraction and PLCO survey response. \\

%We begin presentation of our experimental results with an analysis of the quality of extraction against PLCO survey re-sponses as well as the missing patterns in extracted and survey variables, and conclude with comparing the outcomes of lung cancer risk factors identification and prediction models on both datasets. Table 2 shows the distribution of data for both PLCO and EHRs.\\

\begin{table}[h]
\centering
\caption{Performance of information extraction from electronic medical record prepared with 1000 samples from the entire population.}
\label{tab:result1}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
    \multirow{2}{*}{\textbf{Variables}}  & \multicolumn{2}{c|}{\textbf{Retrieved}}  & \multirow{2}{*}{\textbf{Not Retrieved (\%)}} & \multirow{2}{*}{\textbf{Accuracy}} \\ \cline{2-3}

 & \textbf{Correct} & \textbf{Wrong (\%)}  &  &  \\ \hline

 Age & 998 & 2 (0.20\%) & 0 (0\%) & 0.998\\ \hline
 Sex & 975 & 7 (0.70\%) & 0 (0\%) & 0.975\\ \hline 
 Race & 999 & 1 (0.01\%) & 0 (0\%) & 0.999\\ \hline
 Lung cancer family history & 996 & 4 (0.40\%) & 0 (0\%) &  0.996\\ \hline
 COPD history & 987 & 13 (1.30\%) & 0 (0\%) & 0.987\\ \hline
 Chest x-ray in last 3 years & 1000 & 0 (0\%) & 0 (0\%)  & 1.000\\ \hline
 Emphysema & 994 & 6 (.60\%) & 0 (0\%) &  0.994\\ \hline
 Asthma & 995 & 5 (0.50\%) & 0 (0\%) &  0.995\\ \hline
 Prior cancer & 962 & 38 (3.80\%) & 0 (0\%) &  0.962\\ \hline
 Pneumonia & 982 & 18 (1.80\%) & 0 (0\%) & 0.982\\ \hline
 Hay fever & 999 & 1 (0.10\%) & 0 (0\%) &  0.999\\ \hline
 Smoking status & 912 & 88 (8.80\%) & 0 (0\%)  & 0.912\\ \hline
 Smoker status & 490 & 18 (3.53\%) & 2 (0.40\%) &  0.961\\ \hline
 Age at start smoking & 10 & 1 (5.26\%) & 8 (42.11\%) & 0.526\\ \hline
 Smoking amount (\# cigarettes per day) & 277 & 24 (7.62\%) & 14 (4.44\%) & 0.879\\ \hline
 Smoking duration & 191 & 16 (6.58\%) & 36 (14.81\%) & 0.786\\ \hline
 Quit time of former smoker & 241 & 6 (2.31\%) & 13 (5.00\%) & 0.927\\ \hline
 Pack-years smoked & 212 & 18 (6.84\%) & 33 (12.55\%) & 0.806\\ \hline

  \end{tabular}
\end{table}

%\begin{table}[h]
%\centering
%\caption{Performance of information extraction from electronic medical record prepared with 1000 samples from the entire population.}
%\label{tab:result1}
%  \begin{tabular}{|l|l|l|l|l|l|}
%  \hline
%    \multirow{2}{*}{\textbf{Variables}}  & \multicolumn{2}{c|}{\textbf{Retrieved}}  & \multirow{2}{*}{\textbf{Not Retrieved}}   & \textbf{Missing in}  & \multirow{2}{*}{\textbf{Accuracy}} \\ \cline{2-3}
%    
% & \textbf{Correct} & \textbf{Wrong}  & & \textbf{EHR} &  \\ \hline    
%    
% Age & 998 & 2 & 0 & 0 & 0.998\\ \hline
% Sex & 975 & 7 & 0 & 18 & 0.975\\ \hline 
% Race & 977 & 1 & 0 & 22 & 0.977\\ \hline
% Lung cancer family history & 996 & 4 & 0 & 0 & 0.996\\ \hline
% COPD history & 987 & 13 & 0 & 0 & 0.987\\ \hline
% Chest x-ray in last 3 years & 1000 & 0 & 0 & 0 & 1.000\\ \hline
% Emphysema & 994 & 6 & 0 & 0 & 0.994\\ \hline
% Asthma & 995 & 5 & 0 & 0 & 0.995\\ \hline
% Prior cancer & 962 & 38 & 0 & 0 & 0.962\\ \hline
% Pneumonia & 982 & 18 & 0 & 0 & 0.982\\ \hline
% Hay fever & 999 & 1 & 0 & 0 & 0.999\\ \hline
% Smoking status & 912 & 88 & 0 & 0 & 0.912\\ \hline
% Smoker status & 490 & 18 & 2 & 490 & 0.980\\ \hline
% Smoking start age & 10 & 1 & 8 & 981 & 0.991\\ \hline
% Smoking amount (\# cigarettes per day) & 277 & 24 & 14 & 685 & 0.962\\ \hline
% Smoking duration & 191 & 16 & 36 & 757 & 0.948\\ \hline
% Quit time of former smoker & 241 & 6 & 13 & 740 & 0.981\\ \hline
% Pack-years smoked & 212 & 18 & 33 & 737 & 0.949\\ \hline 
% 
%  \end{tabular}
%\end{table}

\subsection*{\textit{Performance of information extraction from EHRs}}
Table~\ref{tab:result1} shows the performance of our system evaluated with the gold standard on the 1000 patients randomly selected from the 11,968 patients. Performance was the best with chest x-ray identification (100\%) and was good (approximately 99\% accuracy) for age, race, lung cancer family history, emphysema, asthma and hey fever. Results were above 90\% accurate for sex, COPD history, prior cancer, pneumonia, smoking status and smoker status, except smoking duration, smoking amount and pack-year smoked with 78.6\%, 87.9\% and 80.6\% accuracy, respectively. Our system performed worst when identifying ``age at start smoking''. However, the system unable to retrieved 0.40\%, 42.11\%, 4.44\%, 14.81\%, 5\% and 12.55\% information regarding the variables smoker status, age at start smoking, smoking amount, smoking duration, quit time of former smoker and pack-year smoked. The NLP program provides 8.80\%, 0.50\%, 3.8\%, 1.8\%, , 7.62\%, 3.53\% and 6.84\% incorrect result when identifying  smoking status, asthma, prior cancer, pneumonia, smoking amount, smoker status and pack-year smoked, respectively. \\

\begin{table}[h!]
\centering
\caption{Results of information extraction from electronic health record compared with PLCO trial on the 1000 patients subset.}
\label{tab:result2}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
    \textbf{Variables}  & \textbf{Agreed by }  & \textbf{Correct in}   & \textbf{Correct in}  & \textbf{Information Missing} \\
      & \textbf{EHR and PLCO}  & \textbf{EHR}   & \textbf{PLCO}  & \textbf{from EHR} \\ \hline

 Age & 992 (99.2\%) & 6 (0.60\%) & 2 (0.20\%) & 0 (0\%) \\ \hline
 Sex & 964 (96.4\%) & 11 (1.1\%) & 7 (0.70\%) & 18 (1.8\%) \\ \hline 
 Race & 949 (94.9\%) & 28 (2.8\%) & 1 (0.10\%) & 22 (2.2\%) \\ \hline
 Lung cancer family history & 906 (90.6\%) & 25 (2.5\%) & 4 (0.40\%) & 65 (6.5\%) \\ \hline
 COPD history & 873 (87.3\%) & 68 (6.8\%) & 13 (1.3\%) & 46 (4.6\%) \\ \hline
 Chest x-ray in last 3 years & 546 (54.6\%) & 19 (1.9\%) & 26 (2.6\%) & 409 (40.9\%) \\ \hline
 Emphysema & 933 (93.3\%) & 29 (2.9\%) & 6 (0.6\%) & 32 (3.2\%) \\ \hline
 Asthma & 86 (8.6\%) & 0 (0\%) & 5 (0.50\%) & 909 (90.9\%) \\ \hline
 Prior cancer & 677 (67.7\%) & 0 (0\%) & 38 (3.8\%) & 285 (28.5\%) \\ \hline
 Pneumonia & 106 (10.6\%) & 0 (0\%) & 18 (1.8\%) & 876 (87.6\%) \\ \hline
 Hay fever & 13 (1.3\%) & 0 (0\%) & 1 (0.1\%) & 986 (98.6\%) \\ \hline
 Smoking status & 611 (61.1\%) & 99 (9.9\%) & 88 (8.8\%) & 202 (20.2\%) \\ \hline
 Smoker status & 488 (48.8\%) & 2 (0.20\%) & 18 (1.8\%) & 492 (49.2\%) \\ \hline
 Age at start smoking & 10 (1\%) & 0 (0\%) & 9 (0.90\%) & 981 (98.1\%) \\ \hline
 Smoking amount (\# cigarettes per day) & 277 (27.7\%) & 0 (0\%) & 38 (3.8\%) & 685 (68.5\%) \\ \hline
 Smoking duration & 191 (19.10\%) & 0 (0\%) & 52 (5.2\%) & 757 (75.7\%) \\ \hline
 Quit time of former smoker & 241 (24.1\%) & 0 (0\%) & 19 (1.9\%) & 740 (74\%) \\ \hline
 Pack-years smoked & 212 (21.2\%) & 0 (0\%) & 51 (5.1\%) & 737 (73.7\%) \\ \hline

  \end{tabular}
\end{table}

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth]{figures/matrixplot.eps}
\caption{Matrix plots demonstrate the MAR situation among variables.}
\label{fig:matrixplot}
\end{figure}

\subsection*{\textit{Comparing the results between EHR extraction and PLCO survey response}}
To compare the results from the proposed system directly with PLCO survey response, the same 1000 cases were compared. Table~\ref{tab:result2} presents the EHR results compared with PLCO clinical trial response. This results revealed some interesting findings.

First, highest agreement of results by both EHR and PLCO is observed in age value, achieved 99.2\% agreement against gold standard. The main reason is that ``date of birth'' is well formatted and structured across all reports. However, lowest agreement (1\%) is examined for ``age at start smoking''. The common reason might be the ambiguity between patient age and smoking age or difficulties to create an extraction system for identifying ``age at start smoking''. The values extracted from EHRs were agreed well with PLCO responses for the variables ``sex'',``race'',``lung cancer family history'',``COPD history'' and ``emphysema'' in 96.4\%, 94.9\%,90.6\%, 87.3\% and 93.3\% cases, respectively. On the other hand, ``chest x-ray in last 3 years'', ``prior cancer'', ``smoking status'' and ``smoker status'' were moderately agreed by PLCO and EHR extraction.

Second, we observed a certain percentage of disagreement between PLCO and EHR results. Compared to gold standard, EHR results are correct in some cases while PLCO are correct in other cases. The highest disagreement was noticed in ``smoking status'' where EHR results (9.9\%) have higher accuracy compared to PLCO results (8.8\%). Error occurred most frequently in EHR results for many variables due to the assumption that either the value is missing or has a default value. For example, we assigned the default value ``non-smoker'' to the smoking status, when system fails to find any information related to smoking status. Because of this, ``asthma'', ``prior cancer'', ``pneumonia'', ``hay fever'', ``age at start smoking'', ``smoking amount'',``smoking duration'', ``quit time of former smoker'' and ``pack-year smoked'' were only correct in PLCO when disagreed with EHR results. Certain variables such as ``age'', ``sex'', ``race'', ``lung cancer family history'' and ``COPD history'' provided higher percentage of correct values in EHRs compared to PLCO response. Contrarily, PLCO results provided higher percentage of accurate results for the variables ``chest x-ray in last 3 years'' and ``smoker status''.

Finally, highest number of missing values is observed in ``hay fever'', which displayed 98.6\% missing values in EHRs. However, no value is missing for the variable age. Although we observed a significant number of missing values in smoking related variables, around 50\%-70\% values are missing due to the non-smokers or former smokers. Therefore, there are few percentage of actual missing values for the variables ``smoker status'', ``age at start smoking'', ``smoking amount'', ``smoking duration'', ``quit time of former smoker'' and ``pack-year smoked''.

\subsection*{\textit{Missing pattern analysis}}
We utilized all patients participated in PLCO clinical trials for the analysis of missing values. Figure~\ref{fig:matrixplot} shows the MAR situation among variables where the sorting was done by quit time, age and status, respectively. It can be seen that the higher the quit time, the more missing values in the variables smoking amount and pack-year smoked. On the other hand, missing values increase with the decrements of quit time. It was also shown that more missing values were observed for sex and race when age values are decreasing. Furthermore, they were more likely missing from EHR when smoking status is unavailable.

\begin{figure}[h!]
\centering
\includegraphics[width=1.0\textwidth]{figures/marginplot.eps}
\caption{Scatter plots illustrate the MAR situation between variables.}
\label{fig:marginplot}
\end{figure}

The MAR relations were observed more clearly in Figure~\ref{fig:marginplot}. Along the horizontal axis, the red boxplot is for those values of x (i.e. quit time, age) where no values for y (i.e. pack-year smoked, smoking amount) are available, and the blue boxplot for x values where also information for y is available. A comparison of the two boxplots can indicate the missing data mechanism. For example, the quit time for former smokers that provide no information for smoking amount seems to be higher than that where information is available.

\section*{Discussion}
In this study, we investigated errors of extracting information from EHRs to characterize the reasons for disagreement between extracted values and report values. This information have significant value in establishing priorities for the development of an Information Extraction System in Bio-medical domain as well as conducting clinical trials.

The results suggest that automated extraction of clinical information may be more valid for some variables, and less valid for others. In spite of disagreements between information in the reports, automatic information extraction could still be beneficial for quick access to information, depending on the error threshold a user can withstand. For example, EHR data provide nearly the human accuracy for variables such as age, sex, race, lung cancer family history, COPD history and emphysema. Furthermore, less than 1\% disagreement were observed only with PLCO responses, so the information extracted could still be extremely useful.

Numerous errors were observed, where quantitative information extraction errors were the most prevalent among all variables. Several formats were observed in various report with different sentence structures. Furthermore, spelling error, ungrammatical structure, abbreviation and acronym create this extraction process more difficult. For instance, quit time of former smoker needs to be extracted from the clause ``q age 20''. We can calculate it with (current age - 20) by identifying q and age as ``quit time'' and ``age at quited from smoking''.

For medical condition extraction, the most common failure was due to errors in section boundary and ambiguous pronoun. If the section of the current sentence is not identified correctly, we will not be able to extract the medical condition such as lung cancer family history, COPD history, pneumonia, asthma, etc. Many of the false positive were due to ambiguous pronoun that may refer family member instead of patient.

A  few data inconsistencies were observed in some of the variables such as sex and race. In order to determine the status of sex, we needed to determine the presence or absence of UMLS concepts that represent sex, such as ``male'', ``woman'', ``gentleman'', ``lady'', etc. However, we observed information mismatch between or within the same EHR data source.

In order to choose a proper imputation method one must be aware of the missing data mechanism(s). The quality of the imputed values depends on the imputation itself and on the imputation method used. In this study, matrix and scatter plots were used to determine the missing data mechanism(s) among variables. We demonstrate that there were MAR relation between quit time and smoking amount, quit time and smoking duration, quit time and pack-year smoked, age and sex, and age and race.

\section*{Conclusion}
In this paper, we compared result of the integrated systems with PLCO clinical trials and observed that the performance of the systems rivaled the clinical trial, when clinical information appears in electronic health records. However, clinical trial is necessary for certain variables such as ``age of start smoking'', ``education'', ``occupation'', etc. The proposed system consists of several state-of-the-art NLP systems to identify patients information from electronic health records. Experimental result shows 91.2\% to 100\% accuracy in retrieved variables. This study also provides an analysis of some of the
difficulties that could be encountered in extraction of information from free-text clinical reports to automatically extract variables. Each of the
variables had a different profile of errors, suggesting that some variables may be ``better targets'' for information extraction and auto-annotation than others. Finally, we addressed data missingness issue and reviewed different mechanisms through which data may be missing. The absence of data elements is the consequence of physician decisions and reflects aspects of the patient's condition. As the absence of data is usually considered a hindrance to accurate computer-based diagnosis, we need proper treatment of missing data. In this paper, we present missing patterns with matrix and scatter plots, and suggest various ways of dealing with missing values including ``multiple imputation'', ``use of means'', etc.

\section*{Acknowledgments}
We would like to thank the technical staff at Henry Ford Health Systems and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming binary data into free text EHR used for experiments reported in this paper.

\bibliographystyle{vancouver}
\bibliography{references}

\end{document}

### 2017-02-02 paper/AMIA/main.tex
set layout with some old text

@@
 
-\title{Comparison of clinical trial with the results obtained from EHRs along with missing value analysis}
+\title{Leverage machine learning and textual data mining methods for the classification of clinical interview fragments and their sequences}
 
-\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, Rick Krajenta, BS$^{2}$, Melody J. Eide, MD, MPH$^{2,3}$, Christine C. Johnson, PhD, MPH$^{4}$}
+\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Sylvie Naar-King, PhD$^{2}$}
 
 \institutes{
-$^1$Department of Computer Science, Wayne State University, Detroit, Michigan; 
-$^2$Departments of Biostatistics and Research Epidemiology,
-$^3$Department of Dermatology, 
-$^4$Department of Public Health Science, Henry Ford Health System, Detroit, Michigan\\
+$^1$Department of Computer Science,  
+$^2$Pediatric Prevention Research Center, Wayne State University, Detroit, Michigan\\
 }
@@
 \noindent{\bf Abstract}
-\textit{This study compared the PLCO survey response with the results obtained from electronic health records.
-Our study also explored the missing data problem and offer advice for dealing with missing data.
-To extract clinical information, we built an integrated information extraction system that leveraged several publicly available clinical NLP and information extraction tools.
-Experimental results showed the significant difference between these two results, while extraction results achieved 91.2\% to 100\% accuracy.
-The structure of missing values was analyzed with visualization tools implemented in R-package.
-We observed MAR situation for certain variables such as ``quit time'', ``smoking duration'', and ``pack-year smoked''.
-These results indicate that EHR data contained sufficient information can be used directly for certain research.
-Besides that, a clinical trial might be suffered by missing values or biasing or noise occurred during the trials.
-These problems can be resolved with EHRs in combination with properly addressing missing values.}
+\textit{This study examines the effectiveness of state-of-the-art supervised machine learning methods for the task of automatic annotation of fragments of clinical interviews in conjunction with sequential analysis of that annotation.
+We used a collection of motivational interview transcripts consisting of 11,353 utterances, which were manually annotated by two human coders as a golden standard, and experimented with state-of-art classifiers, including Na\"{i}ve Bayes, J48 Decision Tree, Support Vector Machine (SVM), Random Forest (RF), AdaBoost, DiscLDA, Conditional Random Fields (CRF) and Convolutional Neural Network (CNN) in conjunction with lexical, contextual (label of the previous utterance) and semantic (distribution of words in the utterance across the Linguistic Inquiry and Word Count dictionaries) features.
+We found out that, when the number of classes is large, the performance of the CNN and CRFs is inferior to SVM.
+Among all classifiers using only lexical features, SVM achieved the highest classification accuracy of 70.8\%, 61\% and 53.7\%, when interview transcripts were annotated with the codebooks consisting of 17, 20 and 41 codes, respectively.
+Using contextual and semantic features, as well as their combination, in addition to lexical ones improved the accuracy of SVM for annotation of utterances in motivational interview transcripts with a codebook consisting of 17 classes to 71.5\%, 74.2\% and 75.1\%, respectively.
+We also used first order markov model that achieved 68.03\% accuracy for the classification of sequence of behavioral codes extracted from the above dataset.
+Therefore, our results demonstrate the potential of using machine learning  and textual data mining methods in conjunction with semantic and contextual features for automatic annotation of clinical interview transcripts  and their sequences with near-human accuracy.}
 
 \section*{Introduction}
-Electronic health records (EHRs) are increasingly being used to document and store large amounts of clinical information \cite{jha2009use}.
-They replaced traditional paper-based systems in many healthcare organizations and became a vast data sources which can facilitate clinical research and studies to ultimately improve healthcare quality \cite{blumenthal2010launching, erickstad2011use, lau2011use}.
-Much of the available clinical data are in narrative form as a result of transcription of dictations, direct entry by providers, or use of speech recognition applications.
-Comparing to structured data, free text is a more conventional way in the health care environment to express concepts and events.
-However, free text is very challenging for searching, summarization, decision-support, or statistical analysis.
-To reduce medical errors, improve health care quality, and enable secondary use of EHRs, information extraction (IE), which structures and encodes clinical information stored in free text, is necessary.
-However, traditional natural language processing (NLP) tools are not designed for the fragmented free text found in narrative clinical records; therefore, they do not perform well on this type of data.
+In this study, our analysis have two directions: automatic annotation and sequential data analysis.
+Annotation (or labeling) of fragments of clinical text with the categories (or labels, codes) from a predefined codebook is an integral part of qualitative research.
+It can also be viewed as classification of textual fragments into a predefined number of categories (classes).
+Textual annotation has been traditionally performed manually by trained coders, which is a tedious, costly and time-consuming process.
+Furthermore, manual annotation increases the likelihood of errors due to coder fatigue and bias associated with human subjectivity.
+To automate tedious cognitive tasks such as classification, supervised machine learning methods have been recently proposed.
+Although these methods have been shown to be successful at binary (two-class) classification \citep{40,41} (e.g.
+classifying textual fragments as neutral or opinionated), we are not aware of any prior studies reported in the literature that evaluate their performance for textual classification tasks involving large number of classes.
+Such tasks, however, are fairly common in clinical setting (e.g.
+annotation of clinical interviews, assignment of ICD-9/10 codes to patient records).
+To address this limitation, in this paper we propose contextual and semantic features and present the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods in conjunction with lexical and the proposed features for the task of automatic annotation of utterances in clinical interview transcripts with the codebooks consisting of large number of classes.
+Our study provides a guideline for clinical informatics researchers and practitioners, who consider an option of using machine learning methods for automatic annotation of clinical text in their projects.
 
-In the past decade, several different techniques have been developed to extract information, from simple pattern matching to complete processing methods based on symbolic information and rules or based on statistical methods and machine learning.
-Many prior natural language processing (NLP) studies in the clinical domain have used regular expressions in designing their NLP solutions \cite{stenner2012paste,piwowar2008identifying,denny2009identifying, mccart2012using,xu2011facilitating,matheny2009detection,zeng2006extracting,huang2007novel}.
-These regular expressions are typically created by software developers working with domain experts.
-One of the important regular expression based work is NegEx \cite{chapman2001simple}.
-This is because significant portion of medical text is negated in electronic health records.
-NegEx depends on a list of phrases that indicate the absence of a particular medical condition such as disease or symptom to negate the conditions within a window of five words or phrases.
+Annotation of clinical interview transcripts to distinguish different behavior types is an important part of clinical research aimed at designing effective interventions for many conditions and disorders.
+In this paper, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers.
+Childhood obesity is a serious public health concern in the United States and worldwide.
+Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \citep{01}.
+Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \citep{02}.
+Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity.
+One approach to designing intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change \citep{03,04}.
+The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
+These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}.
+The process of establishing causal linkages to identify effective communication strategies for eliciting change talk and commitment language in MI involves a resource-intensive qualitative coding process.
+First, clinical interviews are transcribed and then each utterance is manually annotated with a set of codes from a pre-defined codebook designating specific behavior types.
+Training human coders to reliably and accurately assign codes to textual fragments requires a large investment of manpower, time and money.
+For example, in a recent MI study \citep{07}, training coders to reliability took about four months and, once trained, coders required five hours to code every recorded hour.
+A similar study reported requiring 60 hours of training over six weeks to attain coder reliability, and the actual coding involved two coding passes and six coders \citep{08}.
 
-An extended version of NegEx called ConText \cite{harkema2009context} was proposed to determine three contextual features including negation (negated, affirmed), temporality (historical, recent, hypothetical) and experiencer (patient, other).
-To encode these contextual features, several NLP systems have also been designed to encode detailed information from clinical reports, such as MedLEE \cite{friedman2000broad}, MPLUS \cite{christensen2002mplus}, and MedSyndikate \cite{hahn2002medsyndikate}.
+Automatic annotation of patient utterances in clinical communication is a challenging task, since patients usually come from a variety of cultural and educational backgrounds and
+their language use can be quite different \citep{09}. This problem is exacerbated when the interviews are conducted with children and adolescents due to their tendency to use
+incomplete sentences and frequently change subjects.
 
-However, health information text extraction (HITEx \cite{zeng2006extracting}) was developed to extract diagnoses and smoking status.
-Statistics shows that smoking causes nearly one of every five deaths in the United States each year.
-Therefore, extraction of patients smoking status is necessary for clinician and researcher.
-Mayo Clinic developed an NLP System named clinical text analysis and knowledge extraction system (cTAKES \cite{savova2010mayo}) to extract clinical information from medical domain.
-Additional modules have been implemented within cTAKES recently \cite{savova2008mayo} for identification of patient smoking status.
+Previous quantitative studies of clinical conversation have resulted in creation of Generalized Medical Interaction Analysis System (GMIAS) \citep{10}, which uses a codebook with generic hierarchical categories.
+The small-size codebook in Comprehensive Analysis of the Structure of Encounters System (CASES) \citep{11} was designed to annotate several
+meta-discursive aspects of medical interviews, such as assigning ``ownership'' of topics and partitioning them into distinct segments (speech acts). It was also shown that the
+fragments of transcripts of routine outpatient visits consisting of several speech acts coded using GMIAS and CASES can be annotated as ``information giving'' and ``requesting
+information'' \citep{12}. Other related previous studies focused on categorizing assertions of medical problems in clinical narrative into 5 classes (present, absent, possible,
+hypothetical, conditional and associated with someone else) using SVM \citep{13} and annotating the utterances in hemodialysis phone dialogue with 3 categories using AdaBoost
+classifier \citep{14}. The present study reports the results of comprehensive evaluation of 8 state-of-the-art classifiers (Na\"{i}ve Bayes \citep{15,16,17}, Support Vector Machine
+\citep{18,19}, Conditional Random Fields \citep{21,22}, J48 \citep{23}, AdaBoost \citep{24}, Random Forest \citep{42}, DiscLDA \citep{25} and Convolutional Neural Network
+\citep{38}) for the task of annotating clinical interviews with a codebook, consisting of a large number of classes. We also propose and experimentally evaluate two novel features
+for this task: contextual features based on the label of the preceding textual fragment and semantic features based on the distribution of words in the annotated fragment over a linguistic lexicon.
 
-While substantial progress has been made in the development of NLP tools, no information extraction (IE) system is perfect for extracting all kind of variables in different domain, and still suffer from various weaknesses.
-In this study, we built an integrated system that leveraged several existing NLP components.
-Experimental results shows that our system performed well in biomedical domain.
+Codes automatically assigned to transcripts using classification models presented above allow to abstract away from the raw text  of the interviews and unequivocally establish the casual relationship between provider communication and client change talk (CT).
+Two recent studies \citep{46,47} have examined causality in treatment communication sequences using sequential analysis techniques to assess the probability that CT actually follows specific provider behaviors.
+Both studies found that provider use of MICO communication behaviors was followed by CT more frequently than expected and followed by client counter-CT (CCT; client statements against making changes) less frequently than expected by chance.
+Again, while suggestive of important PPC sequences, these studies were conducted among adults in treatment for alcohol abuse.
+In this project, we used a probabilistic model, in particular, first order markov model for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
 
-In this study, we mainly focus on findings from electronic health records and then compare the result with PLCO clinical trial.
-We also describe missing patterns observed in our experiment.
-This paper reports that which information are sufficiently available in EHRs for conducting clinical research.
-It also states that which of the variables are noisy in clinical trials.
-These findings are significant in clinician and researcher who are conducting their research with clinical trials or EHRs.
-To the best of our knowledge, there is no report that compare the results between response from clinical trials and NLP-based extraction from EHRs.
+\begin{table}[htb!]
+\caption{\textbf{Distribution of utterances over 16 classes in the caregiver dataset.}}    
+\label{table:01}
+\centering
+\begin{tabular}{lp{7cm}rr}
+\hline
+\hline
+Code  & Description & Utterance & \% \\
+\hline
+209 &	Caregiver Change Talk, negative &	297 & 6.82 \\
+212 &	Caregiver Change Talk, positive &	1107 & 25.40 \\
+232 &	Low Uptake, positive &	518 & 11.89 \\
+235 &	High uptake &	231 & 5.30 \\
+301 &	Structure Session &	206 & 4.73 \\
+302 &	General Information, positive &	309 & 7.09 \\
+305 &	Emphasize Autonomy &	148 & 3.40 \\
+306 &	Closed question, Elicit Feedback &	50 & 1.15 \\
+307 &	Support &	108 & 2.48 \\
+308 &	Affirm &	289 & 6.63 \\
+315 &	Reflect, change talk positive, about caregiver &	659 & 15.12 \\
+329 &	Self-disclose &	44 & 1.01 \\
+330 &	Statement, other &	121 & 2.78 \\
+331 &	Open-ended question, elicit change talk positive &	200 & 4.59 \\
+343 &	Open-ended question, target behavior neutral &	33 & 0.76 \\
+344 &	Open-ended question, elicit barriers &	38 & 0.87 \\
+\hline
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[htb!]
+\caption{\textbf{Distribution of sequences over the adolescent interview session}} 
+\label{table:0001}
+\centering
+\begin{tabular}{lrr}
+\hline
+\hline
+Sequence type  & \# of Sequences & \% \\
+\hline
+Successful & 880 & 82.08\% \\
+Unsuccessful & 192 & 17.92\% \\
+Total & 1072 & 100\% \\
+\hline
+\hline
+\end{tabular}
+\end{table}
 
@@
 \subsection*{\textit{Data collection}}
-Four hundred and fifty thousand free-text EHRs of 11,968 patients were obtained from Henry Ford Health System, who participated in the PLCO clinical trial.
-Out of this 11,968 patients, 1000 patients were randomly selected for this study.
-The PLCO trial began enrolling participants in November 1993 and completed recruitment of 154,938 subjects over different health systems in June 2001.
-Men and women aged from 55 to 74 years participated in the PLCO trial at ten screening centers nationwide with balanced randomization to intervention and control arms.
-All participants received a posterior anterior view chest x-ray for lung cancer screening.
-A baseline epidemiological questionnaire and a supplemental questionnaire (SQ) were administered as part of the PLCO surveys.
-The obtained patient EHRs consist of different types of clinical documents including clinical practice guidelines, diagnostic reports and discharge summaries.
+The golden standard for evaluation of machine learning methods was created based on the transcripts of motivational interviews conducted by the clinicians at the Pediatric
+Prevention Research Center (PPRC) of Wayne State University. Each interview is comprised of two parts: conversation of a clinician with an adolescent followed by a conversation of
+a clinician with the adolescent's caregiver. All adolescents in this study were between the ages of 12 and 17 (M = 14.7, SD = 1.63) and most were female (n = 27). Most caregivers
+were biological mothers (n = 36), who were married or living with a partner (n = 25). The median family income was \$16,000--\$21,999 ranging from less than \$1,000 to
+\$50,000--\$74,999. The audio recordings of the interviews were first transcribed and segmented into utterances belonging to adolescents, caregivers, and counselors, preserving the
+sequence of utterances. Transcripts were then manually annotated by trained human coders according to MYSCOPE \citep{07}, a specialized codebook including a large number of
+behavior codes, which was developed by an interdisciplinary team including a clinical psychologist, a nutrition scientist, a communication scientist, a linguist and a community
+health worker specifically for annotating motivational interviews with obese adolescents. A primary coder independently coded interview sessions and a secondary coder co-coded a
+randomly selected 20\% of the transcripts to monitor reliability ($\kappa$ = 0.696) \citep{07}. The MYSCOPE codebook contains a total of 115 different codes that are grouped into
+the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
+include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
+sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
+(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}.
+
+\begin{table*}[htb]
+\caption{\textbf{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}}    
+\label{table:06}
+\centering
+\begin{tabular}{lp{3.6cm}lp{8cm}}
+\hline
+\hline
+Annotation  & Description & Speaker & Text \\
+\hline
+331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do ?
+mm-hmm meaning is that food available for you ?
+\\
+117 &	Low Uptake, positive	& Adolescent &	Yes \\
+301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\
+112 &	Change Talk positive	& Adolescent &	my um aunt \\
+301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\
+112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\
+309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\
+\hline
+\hline
+\end{tabular}
+\end{table*}
+
+To conduct a detailed analysis of performance of classification methods, we used the following two-stage process to create the codebooks with different number of codes for
+adolescent and caregiver sessions.
+In the first stage, we merged conceptually similar behavior codes as well as the codes with similar data distributions, while in the second stage, we eliminated the codes with insufficient data samples.
+In case of the adolescent sessions, we started with 55 adolescent session-specific codes and, after merging the codes with subtle differences (e.g.
+converting valiances of change talk CHT+1, CHT+2 and CHT+3 into CHT+), obtained a codebook with 41 classes.
+We further reduced this codebook to 20 classes after merging 21 classes with similar sample distributions.
+After eliminating the codes that had less than 10 data samples (to ensure that there can be at least one sample of each class in each fold when using 10-fold cross validation experimental design), we obtained a third codebook with 17 codes.
+Using the same approach, we created the codebooks containing 58, 19, and 16 caregiver session-specific codes.
+Table~\ref{table:01} shows the distribution of utterances over 16 classes in the caregiver session transcripts.
+As follows from Table~\ref{table:01}, the distribution of utterances over classes is highly imbalanced even for the codebook of the smallest size, which is fairly common for clinical text.
+To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language.
+Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.
+
+After creating the codebooks, we pre-processed the dataset using the Snowball stemmer available as part of the Weka \citep{26} machine learning toolkit\footnote{
+http://www.cs.waikato.ac.nz/ml/weka/}. We also found out that stopword removal decreased the performance of classification models for our task (e.g., in case of the codebook
+consisting of 17 classes, the accuracy of Na\"{i}ve Bayes decreased from 67\% to 47.10\%, while the accuracy of SVM decreased from 70.76\% to 55.26\%). A likely reason is that,
+although negations are typically considered as stopwords, they are fairly important clues for inferring certain behavior types (e.g., removing the stopword ``not'' completely
+transforms the meaning of a phrase ``not great'').
 
 \subsection*{\textit{System overview}}
-We built an integrated information extraction system that leveraged several publicly available clinical NLP and information extraction tools.
-In particular, our system consists of the following components: 1) A sentence boundary detection program, which was developed at OpenNLP \cite{morton2005opennlp}; 2) an existing corpus based section identification program (SecTag \cite{denny2009evaluation}); 3) regular expression based NLP program that utilizes the clinical Text Analysis and Knowledge Extraction System (cTAKES \cite{savova2010mayo}) and contextual feature extraction system (ConText \cite{harkema2009context}); 4) post processing program that maps our results into PLCO compareable format.
-Figure \ref{fig:extraction} shows the architecture of our system.
+The pipeline consists of two stages: training and testing.
+Prior to the training stage, we preprocess the collected clinical interview transcripts by performing stemming, punctuation removal, word segmentation and tokenization.
+Features are then extracted from the preprocessed data.
+During this stage, previous label and LIWC features are used in conjunction with the lexical features to create the feature vectors.
+After that, classifiers are trained on the feature vectors extracted from the training samples and their associated annotations.
+In the testing stage, after creation of feature vectors, the previously trained classifiers predict the label of each utterance in the testing sample.
+Finally, performance of different classifiers is evaluated by calculating standard metrics such as precision, recall, F-score (F1), kappa measure and accuracy.
+Specifically, we evaluated the performance of the following state-of-the-art supervised machine learning methods.
 
-\begin{figure}[h!]
+\textbf {Markov Model}: In probability theory, a markov model \citep{45} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property).
+Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable.
+For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
+A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
+
+\begin{itemize}
+\item $V = \{CML+, CHT+, CHT-, T-AMB, CCT, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of possible youth and counselor MI behavior codes;
+\item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
+\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
+
+\begin{equation}
+P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
+\end{equation}
+
+\end{itemize}
+
+where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ have been observed in the training data.
+Given a markov model $M$ (such that $S\subseteq V$), a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ have been generated from a markov model $M$ is:
+
+\begin{equation}
+P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
+\end{equation}
+
+Success of a given motivational interview, given a sequence of MI behavior codes corresponding to it, can be predicted based on the following formula:
+
+\begin{equation}
+p(S\rightarrow CT) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})
+\end{equation}
+
+if $p(S\rightarrow CT) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in CT.
+We experimentally determined the value of $\delta$ to 0.3 which provides acceptable accuracy with minimum false positive rate.
+Figure\ref{fig:delta} illustrates the characteristics of proposed model by varying the value of $\delta$.
+
+\begin{figure}[htb!]
     \centering
-\includegraphics[width=0.5\textwidth]{figures/extraction.eps}
-\caption{An architecture of the integrated system for information extraction from free-text clinical narrative used in this study.}
-\label{fig:extraction}
+    \includegraphics[width=0.70\textwidth]{figures/deltadata.jpg}
+    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
+    \label{fig:delta}
 \end{figure}
 
-\textbf {Sentence splitter}: Sentence detection is important and necessary task for our study because we used sentence as the basic unit of variable extraction.
-As sentence splitting is at the core of a many NLP activities, it is provided by most NLP frameworks and libraries.
-We used an open source Java tools openNLP for identifying the sentence boundary.
-The OpenNLP sentence detector cannot identify sentence boundaries based on the contents of the sentence.
-However, this tools can detect that a punctuation character marks the end of a sentence or not.
-In this sense, a sentence is defined as the longest white space trimmed character sequence between two punctuation marks.
-This rule is not applicable for the first and last sentence.
-The first non whitespace character is assumed to be the begin of a sentence, and the last non whitespace character is assumed to be a sentence end.
+\subsection*{\textit{Evaluation methods}}
+To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \citep{36} as an experimental design.
+The performance of different classifiers and feature sets was evaluated in terms of precision, recall, F1 score (F1), kappa measure and accuracy using weighted macro-averaging over folds.
 
-\textbf {Section identification}:
-Electronic health records are often divided into sections, or segments, such as ``past medical history'', ``family history'', ``assessment and plan'', etc.
-One can gain better understanding of clinical documents by recognition of the section in which a concept lives.
-For instance, both a ``past medical history'' and the ``family history'' sections can contain a list of diseases, but the context describes very different import to the patient about whom the note was written.
-We used SecTags for accomplishing the following three tasks: 1) excluding false positive by recognizing experiencer (patient or family member) mentioned in the family history sections; 2) determine scope of the temporal information, e.g.
-all events mentioned in the past medical history were historical event; 3) extract demographic information about patient from the 1st section.
+\section*{Results}
+Experimental evaluation of automatic annotation of clinical interview fragments and their sequences by using machine learning and probabilistic model included several dimensions:
+\begin{itemize}
+\item determining the performance of classifiers on the codebooks of different size;
+\item determining the effectiveness of the proposed contextual and semantic features.
+\item determining the performance of markov model in conjunction with determining PPC sequences that are most likely to translate into CT.
+\end{itemize}
 
-\textbf {NLP program}:
-As follows from Figure 1, a regular expression based NLP program was developed that utilizes the clinical Text Analysis and Knowledge Extraction System (cTAKES), which is an open source NLP and IE framework developed by the Mayo Clinic for information extraction from EHRs.
-This tool is comprised of several important components: sentence boundary detector, tokenizer, part-of-speech tagger, shallow parser, named entity recognizer, concept mapper, negation and status attribute detector.
-These components were trained on clinical text to create rich linguistic and semantic annotations of EHRs.
-In our study, concept mapping is used to annotate several important concepts, such as sign and symptoms, disease and disorder, lab procedure, negation and other clinical concepts.
+Since clinical researchers typically annotate caregiver and adolescent sessions separately, we first created two experimental datasets consisting of only adolescent and only caregiver session transcripts.
+Second, besides evaluating the accuracy of annotating adolescent and caregiver transcripts with the codebooks containing an entire set of codes, we also conducted a series of experiments with the codebooks of smaller sizes created as outlined above.
+Third, besides training and testing NB, SVM, CRF, Decision Tree, Boosting, DiscLDA, Random Forest and CNN classifiers using only lexical features, we also evaluated the effectiveness of the proposed contextual and semantic features.
 
-Although most components of cTAKES have been shown to achieve good performance on a variety of NLP tasks, it has some limitations  \cite{savova2010mayo} related to certainty detection and temporal resolution.
-For this reason, we used ConText and semantic rules to identify temporal information.
-ConText is an expansion of NegEx.
-It relies on the same basic algorithm but applies different term lists and different windows of scope depending on the contextual feature being annotated.
+Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
 
-In total, eighteen variables were extracted in this study.
-Since demographic variables are generally mentioned in the header or 1st section of the EHRs, we used this section for the extraction of age, sex and race.
-Age was determined by calculating the time period between ``date of birth'' and ``visit date or report date'', which were annotated by cTAKES.
-Sex and race were extracted by using simple regular expression in conjunction with a list of trigger terms such as male, female, white, african american etc.
-It was also confirm that the scope of the trigger term is outside the family member or relatives.
-We applied exactly same patterns for the extraction of ``COPD history'', ``emphysema'', ``asthma'', ``pneumonia'' and ``prior cancer''.
-At first, we find the triggered term or UMLS concept we are interested in.
-Then, we checked that the concept was not negated or fall into different experiencer such as family member.
-Finally, we confirmed the certainty of concept so that it was not uncertain or hypothetical such as ``suspected for pneumonia'', ``pneumonia vaccine'', etc.
-We also applied the semantic rules to extract ``lung cancer family history'' and ``chest x-ray in the past three years''.
-In addition to that, we  checked the experiencer for ``lung cancer family history'', and compared the ``visit date or report date'' with the x-ray date.
+\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
+Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
+Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
+From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
+The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
 
-\begin{table}[h!]
+\begin{table}
+\parbox{.45\linewidth}{
 \centering
-\caption{Example of information extraction rules.}
-\label{tab:rule}
-  \begin{tabular}{|l|l|}
+\caption{\textbf{Performance of markov model for the classification of PPC code sequences}}
+\label{table:sequence}
+\begin{tabular}{ll}
 \hline
-    \textbf{Rule}  & \textbf{Example} \\ \hline
+\hline
+Metrics & Value \\
+\hline
+Accuracy &	0.6803 \\
+Precision &	0.7574 \\
+Recall & 0.6803 \\
+F1-Measure & 0.7092 \\
+\hline
+\hline
+\end{tabular}
+}
+\hfill
+\parbox{.45\linewidth}{
+\centering
+\caption{\textbf{Confusion matrix of proposed markov model for the classification of sequence of codes}}
+\label{table:conf}
+\begin{tabular}{llll}
+\hline
+\hline
+& Successful & Fail & Total\\
+\hline
+Successful & 642 & 242 & 884 \\
+Fail & 102 & 90 & 192 \\
+Total & 744 & 332 & 1076 \\
+\hline
+\hline
+\end{tabular}
+}
+\end{table}
 
- quit smoking \textless duration\textgreater\hspace{1mm}ago or quit smoking since \textless time\textgreater & quit smoking 30 years ago or quit smoking \\
- or quit smoking from \textless time\textgreater &  since 1960 or quit smoking from 1880 \\ \hline
+Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
 
- \textless amount\textgreater\hspace{1mm}(cigarettes\textbar pack of cigarettes) per (day\textbar week\textbar month) & 10-15 cigarettes per day or \\
- or \textless amount\textgreater\hspace{1mm}(pack\textbar pack-) year of smoking history & 30 pack year of smoking history \\ \hline
-
- (smoking\textbar smokes\textbar smoked\textbar) for \textless duration\textgreater\hspace{1mm} or & he has been smoking for 45 years or \\
- (smoking\textbar smokes\textbar smoked\textbar) (since\textbar from) \textless year\textgreater\hspace{1mm} &  he has been smoking since 1965 \\ \hline
-
+\begin{table}[htb!]
+\caption{\textbf{Top 5 most likely successful and unsuccessful interview sessions}}    
+\label{table:topsec}
+\centering
+\begin{tabular}{lll}
+\hline
+\hline
+No. & Successful Sequences &  Unsuccessful Sequences\\
+\hline
+1 & 311\rightarrow112\rightarrow301\rightarrow500 & 343\rightarrow109\rightarrow312\rightarrow400 \\
+2 & 310\rightarrow112\rightarrow311\rightarrow500 & 311\rightarrow109\rightarrow312\rightarrow400 \\
+3 & 311\rightarrow117\rightarrow309\rightarrow500 & 327\rightarrow120\rightarrow312\rightarrow400 \\
+4 & 312\rightarrow112\rightarrow311\rightarrow500 & 314\rightarrow103\rightarrow314\rightarrow400 \\
+5 & 331\rightarrow112\rightarrow311\rightarrow500 & 344\rightarrow109\rightarrow326\rightarrow400 \\
+\hline
+\hline
 \end{tabular}
@@
 
-Smoking status is extracted using a rule based method \cite{cohen2008five, uzuner2008identifying}, which consists of two steps: hot-spot identification and post-processing rules.
-At first, we find the hot-spots based on keywords and sections (in many cases, smoking status was found in history and demographic sections).
-When a document containing one of the terms ``smok'', ``tob'', ``packs'', ``cigar'' etc.
-was found, a window of text up to 100 characters before and after those terms is considered as a hot-spot.
-Rules are then used to classify smoking status based on temporal information.
-After extracting the smoking status, we extracted associated information such as smoking duration, quit time and pack-year history.
-Several rules were employed to extract clinical variables, some of which are shown in Table~\ref{tab:rule}.
-
-\textbf {Post processing and filter}:
-To allow direct comparison between the extracted results from EHRs and PLCO results, post processing of the proposed system's output stage was required.
-The output of NLP program is a set of pivot text snippets such as ``two packs of cigarettes per day'', ``quit smoking in 1993'', ``smoked from the age of 18'', etc.
-These text fragments are then parsed to generate numeric values.
-Finally, the output of post-processing was filtered by keeping the latest information, for instance, a non-smoker can be a smoker after few months.
-To generate numeric values for statistical analysis, simple regular expression is used together with a list of terms such as ``half'', ``two'', ``one and half'', etc.
-
-\subsection*{\textit{Statistical inference and treatment of missing values}}
-Missing data frequently occur in electronic health records, which is a very common problem in data mining and statistical inference.
-According to Little and Rubin \cite{little2014statistical,rubin1976inference}, missing data can be categorized into three types: i) missing completely at random (MCAR)-the absence of a data element is not associated with any other value in the data set, observed or missing; ii) missing at random (MAR)-this is a less restrictive assumption than MCAR; the probability of data being missing depends only on the observed values in the data set, not on unobserved data.
-The implied information of missingness can be inferred from observed values; ii) not missing at random (NMAR)-the condition is opposite of MAR.
-The probability of data being missing depends on unobserved (missing) data.
-Under this circumstance, the missing information cannot be inferred from observed values.
-
-In order to improve the quality of research, it is important to understand the kind of missing data present in EHRs.
-However, missing values mechanisms are difficult to identify, because it would require the knowledge of the missing values themselves \cite{little2014statistical}.
-Naturally, MNAR is very difficult to be detected.
-This is even more complicated in case of multivariate data, outliers, inhomogeneous data or very skewed data distribution.
-In this paper, we explained the missingness in terms of the above mechanisms with matrix and scatter plots.
-Several approaches have been used to deal with missing data.
-The methods for coping with missing values can be grouped into three main categories \cite{little2014statistical}: inference restricted to complete data, imputation-based approaches, and likelihood-based approaches.
-
-The simplest and most common approach to handling missing values is to omit the cases with missing values and do the analysis based only on the complete cases.
-Data set obtained by the result of this process can be a biased sample of complete cases because the missing of data is not a random process (i.e.
-data is not always MCAR).
-Furthermore, excluding patients with missing data will lead to a loss in statistical power.
-
-In imputation-based methods, the missing values are filled in and the resultant data can be analyzed as a complete data set.
-Imputation of missing data in the setting of MCAR will increase power but should not change the point estimates from those obtained with a complete case analysis.
-Therefore, it seems reasonable to perform imputations using EHR data under the assumption of MAR.
-Commonly imputed values are based on the value of known cases: the mean of the variable in either the whole data set or in select data subsets, or an estimated value from regression procedures on known variables.
-Multiple imputation methods, i.e., filling with more than one value, have been developed to avoid biasing the variances of imputed variables \cite{enders2006primer}.
-
-There exist approaches that derive a prediction model by inferring the model's parameters from the existing data, rather than imputing data where values are missing.
-Likelihood-based approaches are an example of these.
-They developed a model by attempting to find the set of model parameters that make the observed data most likely.
-The resulting system parameters can then be used as the base for future inferences.
-The expectation maximization (EM \cite{dempster1977maximum}) algorithm is commonly used for finding maximum likelihood estimates in the face of incomplete data.
-
-\subsection*{\textit{Evaluation methods}}
-The golden standard for the evaluation of information extraction methods was created based on the EHRs collected from Henry Ford Health Systems.
-Each patient comprised of several EHR documents.
-All EHRs were manually investigated by domain expert to find the values of target variables for each patient.
-Then, we compared the values of extracted categorical and non-categorical (continuous) variables with gold standard to evaluate the accuracy of IE.
-
-\section*{Results}
-We present the results and outcomes of our study regarding the performance of information extraction from EHRs and statistical significance between EHR extraction and PLCO survey response. \\
-
-%We begin presentation of our experimental results with an analysis of the quality of extraction against PLCO survey re-sponses as well as the missing patterns in extracted and survey variables, and conclude with comparing the outcomes of lung cancer risk factors identification and prediction models on both datasets.
-Table 2 shows the distribution of data for both PLCO and EHRs.\\
-
-\begin{table}[h]
-\centering
-\caption{Performance of information extraction from electronic medical record prepared with 1000 samples from the entire population.}
-\label{tab:result1}
-  \begin{tabular}{|l|l|l|l|l|l|}
-  \hline
-    \multirow{2}{*}{\textbf{Variables}}  & \multicolumn{2}{c|}{\textbf{Retrieved}}  & \multirow{2}{*}{\textbf{Not Retrieved (\%)}} & \multirow{2}{*}{\textbf{Accuracy}} \\ \cline{2-3}
-
- & \textbf{Correct} & \textbf{Wrong (\%)}  &  &  \\ \hline
-
- Age & 998 & 2 (0.20\%) & 0 (0\%) & 0.998\\ \hline
- Sex & 975 & 7 (0.70\%) & 0 (0\%) & 0.975\\ \hline 
- Race & 999 & 1 (0.01\%) & 0 (0\%) & 0.999\\ \hline
- Lung cancer family history & 996 & 4 (0.40\%) & 0 (0\%) &  0.996\\ \hline
- COPD history & 987 & 13 (1.30\%) & 0 (0\%) & 0.987\\ \hline
- Chest x-ray in last 3 years & 1000 & 0 (0\%) & 0 (0\%)  & 1.000\\ \hline
- Emphysema & 994 & 6 (.60\%) & 0 (0\%) &  0.994\\ \hline
- Asthma & 995 & 5 (0.50\%) & 0 (0\%) &  0.995\\ \hline
- Prior cancer & 962 & 38 (3.80\%) & 0 (0\%) &  0.962\\ \hline
- Pneumonia & 982 & 18 (1.80\%) & 0 (0\%) & 0.982\\ \hline
- Hay fever & 999 & 1 (0.10\%) & 0 (0\%) &  0.999\\ \hline
- Smoking status & 912 & 88 (8.80\%) & 0 (0\%)  & 0.912\\ \hline
- Smoker status & 490 & 18 (3.53\%) & 2 (0.40\%) &  0.961\\ \hline
- Age at start smoking & 10 & 1 (5.26\%) & 8 (42.11\%) & 0.526\\ \hline
- Smoking amount (\# cigarettes per day) & 277 & 24 (7.62\%) & 14 (4.44\%) & 0.879\\ \hline
- Smoking duration & 191 & 16 (6.58\%) & 36 (14.81\%) & 0.786\\ \hline
- Quit time of former smoker & 241 & 6 (2.31\%) & 13 (5.00\%) & 0.927\\ \hline
- Pack-years smoked & 212 & 18 (6.84\%) & 33 (12.55\%) & 0.806\\ \hline
-
-  \end{tabular}
-\end{table}
-
-%\begin{table}[h]
-%\centering
-%\caption{Performance of information extraction from electronic medical record prepared with 1000 samples from the entire population.}
-%\label{tab:result1}
-%  \begin{tabular}{|l|l|l|l|l|l|}
-%  \hline
-%    \multirow{2}{*}{\textbf{Variables}}  & \multicolumn{2}{c|}{\textbf{Retrieved}}  & \multirow{2}{*}{\textbf{Not Retrieved}}   & \textbf{Missing in}  & \multirow{2}{*}{\textbf{Accuracy}} \\ \cline{2-3}
-%    
-% & \textbf{Correct} & \textbf{Wrong}  & & \textbf{EHR} &  \\ \hline    
-%    
-% Age & 998 & 2 & 0 & 0 & 0.998\\ \hline
-% Sex & 975 & 7 & 0 & 18 & 0.975\\ \hline 
-% Race & 977 & 1 & 0 & 22 & 0.977\\ \hline
-% Lung cancer family history & 996 & 4 & 0 & 0 & 0.996\\ \hline
-% COPD history & 987 & 13 & 0 & 0 & 0.987\\ \hline
-% Chest x-ray in last 3 years & 1000 & 0 & 0 & 0 & 1.000\\ \hline
-% Emphysema & 994 & 6 & 0 & 0 & 0.994\\ \hline
-% Asthma & 995 & 5 & 0 & 0 & 0.995\\ \hline
-% Prior cancer & 962 & 38 & 0 & 0 & 0.962\\ \hline
-% Pneumonia & 982 & 18 & 0 & 0 & 0.982\\ \hline
-% Hay fever & 999 & 1 & 0 & 0 & 0.999\\ \hline
-% Smoking status & 912 & 88 & 0 & 0 & 0.912\\ \hline
-% Smoker status & 490 & 18 & 2 & 490 & 0.980\\ \hline
-% Smoking start age & 10 & 1 & 8 & 981 & 0.991\\ \hline
-% Smoking amount (\# cigarettes per day) & 277 & 24 & 14 & 685 & 0.962\\ \hline
-% Smoking duration & 191 & 16 & 36 & 757 & 0.948\\ \hline
-% Quit time of former smoker & 241 & 6 & 13 & 740 & 0.981\\ \hline
-% Pack-years smoked & 212 & 18 & 33 & 737 & 0.949\\ \hline 
-% 
-%  \end{tabular}
-%\end{table}
-
-\subsection*{\textit{Performance of information extraction from EHRs}}
-Table~\ref{tab:result1} shows the performance of our system evaluated with the gold standard on the 1000 patients randomly selected from the 11,968 patients.
-Performance was the best with chest x-ray identification (100\%) and was good (approximately 99\% accuracy) for age, race, lung cancer family history, emphysema, asthma and hey fever.
-Results were above 90\% accurate for sex, COPD history, prior cancer, pneumonia, smoking status and smoker status, except smoking duration, smoking amount and pack-year smoked with 78.6\%, 87.9\% and 80.6\% accuracy, respectively.
-Our system performed worst when identifying ``age at start smoking''.
-However, the system unable to retrieved 0.40\%, 42.11\%, 4.44\%, 14.81\%, 5\% and 12.55\% information regarding the variables smoker status, age at start smoking, smoking amount, smoking duration, quit time of former smoker and pack-year smoked.
-The NLP program provides 8.80\%, 0.50\%, 3.8\%, 1.8\%, , 7.62\%, 3.53\% and 6.84\% incorrect result when identifying  smoking status, asthma, prior cancer, pneumonia, smoking amount, smoker status and pack-year smoked, respectively.
-\\
-
-\begin{table}[h!]
-\centering
-\caption{Results of information extraction from electronic health record compared with PLCO trial on the 1000 patients subset.}
-\label{tab:result2}
-  \begin{tabular}{|l|l|l|l|l|l|}
-  \hline
-    \textbf{Variables}  & \textbf{Agreed by }  & \textbf{Correct in}   & \textbf{Correct in}  & \textbf{Information Missing} \\
-      & \textbf{EHR and PLCO}  & \textbf{EHR}   & \textbf{PLCO}  & \textbf{from EHR} \\ \hline
-
- Age & 992 (99.2\%) & 6 (0.60\%) & 2 (0.20\%) & 0 (0\%) \\ \hline
- Sex & 964 (96.4\%) & 11 (1.1\%) & 7 (0.70\%) & 18 (1.8\%) \\ \hline 
- Race & 949 (94.9\%) & 28 (2.8\%) & 1 (0.10\%) & 22 (2.2\%) \\ \hline
- Lung cancer family history & 906 (90.6\%) & 25 (2.5\%) & 4 (0.40\%) & 65 (6.5\%) \\ \hline
- COPD history & 873 (87.3\%) & 68 (6.8\%) & 13 (1.3\%) & 46 (4.6\%) \\ \hline
- Chest x-ray in last 3 years & 546 (54.6\%) & 19 (1.9\%) & 26 (2.6\%) & 409 (40.9\%) \\ \hline
- Emphysema & 933 (93.3\%) & 29 (2.9\%) & 6 (0.6\%) & 32 (3.2\%) \\ \hline
- Asthma & 86 (8.6\%) & 0 (0\%) & 5 (0.50\%) & 909 (90.9\%) \\ \hline
- Prior cancer & 677 (67.7\%) & 0 (0\%) & 38 (3.8\%) & 285 (28.5\%) \\ \hline
- Pneumonia & 106 (10.6\%) & 0 (0\%) & 18 (1.8\%) & 876 (87.6\%) \\ \hline
- Hay fever & 13 (1.3\%) & 0 (0\%) & 1 (0.1\%) & 986 (98.6\%) \\ \hline
- Smoking status & 611 (61.1\%) & 99 (9.9\%) & 88 (8.8\%) & 202 (20.2\%) \\ \hline
- Smoker status & 488 (48.8\%) & 2 (0.20\%) & 18 (1.8\%) & 492 (49.2\%) \\ \hline
- Age at start smoking & 10 (1\%) & 0 (0\%) & 9 (0.90\%) & 981 (98.1\%) \\ \hline
- Smoking amount (\# cigarettes per day) & 277 (27.7\%) & 0 (0\%) & 38 (3.8\%) & 685 (68.5\%) \\ \hline
- Smoking duration & 191 (19.10\%) & 0 (0\%) & 52 (5.2\%) & 757 (75.7\%) \\ \hline
- Quit time of former smoker & 241 (24.1\%) & 0 (0\%) & 19 (1.9\%) & 740 (74\%) \\ \hline
- Pack-years smoked & 212 (21.2\%) & 0 (0\%) & 51 (5.1\%) & 737 (73.7\%) \\ \hline
-
-  \end{tabular}
-\end{table}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=1.0\textwidth]{figures/matrixplot.eps}
-\caption{Matrix plots demonstrate the MAR situation among variables.}
-\label{fig:matrixplot}
-\end{figure}
-
-\subsection*{\textit{Comparing the results between EHR extraction and PLCO survey response}}
-To compare the results from the proposed system directly with PLCO survey response, the same 1000 cases were compared.
-Table~\ref{tab:result2} presents the EHR results compared with PLCO clinical trial response.
-This results revealed some interesting findings.
-
-First, highest agreement of results by both EHR and PLCO is observed in age value, achieved 99.2\% agreement against gold standard.
-The main reason is that ``date of birth'' is well formatted and structured across all reports.
-However, lowest agreement (1\%) is examined for ``age at start smoking''.
-The common reason might be the ambiguity between patient age and smoking age or difficulties to create an extraction system for identifying ``age at start smoking''.
-The values extracted from EHRs were agreed well with PLCO responses for the variables ``sex'',``race'',``lung cancer family history'',``COPD history'' and ``emphysema'' in 96.4\%, 94.9\%,90.6\%, 87.3\% and 93.3\% cases, respectively.
-On the other hand, ``chest x-ray in last 3 years'', ``prior cancer'', ``smoking status'' and ``smoker status'' were moderately agreed by PLCO and EHR extraction.
-
-Second, we observed a certain percentage of disagreement between PLCO and EHR results.
-Compared to gold standard, EHR results are correct in some cases while PLCO are correct in other cases.
-The highest disagreement was noticed in ``smoking status'' where EHR results (9.9\%) have higher accuracy compared to PLCO results (8.8\%).
-Error occurred most frequently in EHR results for many variables due to the assumption that either the value is missing or has a default value.
-For example, we assigned the default value ``non-smoker'' to the smoking status, when system fails to find any information related to smoking status.
-Because of this, ``asthma'', ``prior cancer'', ``pneumonia'', ``hay fever'', ``age at start smoking'', ``smoking amount'',``smoking duration'', ``quit time of former smoker'' and ``pack-year smoked'' were only correct in PLCO when disagreed with EHR results.
-Certain variables such as ``age'', ``sex'', ``race'', ``lung cancer family history'' and ``COPD history'' provided higher percentage of correct values in EHRs compared to PLCO response.
-Contrarily, PLCO results provided higher percentage of accurate results for the variables ``chest x-ray in last 3 years'' and ``smoker status''.
-
-Finally, highest number of missing values is observed in ``hay fever'', which displayed 98.6\% missing values in EHRs.
-However, no value is missing for the variable age.
-Although we observed a significant number of missing values in smoking related variables, around 50\%-70\% values are missing due to the non-smokers or former smokers.
-Therefore, there are few percentage of actual missing values for the variables ``smoker status'', ``age at start smoking'', ``smoking amount'', ``smoking duration'', ``quit time of former smoker'' and ``pack-year smoked''.
-
-\subsection*{\textit{Missing pattern analysis}}
-We utilized all patients participated in PLCO clinical trials for the analysis of missing values.
-Figure~\ref{fig:matrixplot} shows the MAR situation among variables where the sorting was done by quit time, age and status, respectively.
-It can be seen that the higher the quit time, the more missing values in the variables smoking amount and pack-year smoked.
-On the other hand, missing values increase with the decrements of quit time.
-It was also shown that more missing values were observed for sex and race when age values are decreasing.
-Furthermore, they were more likely missing from EHR when smoking status is unavailable.
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=1.0\textwidth]{figures/marginplot.eps}
-\caption{Scatter plots illustrate the MAR situation between variables.}
-\label{fig:marginplot}
-\end{figure}
-
-The MAR relations were observed more clearly in Figure~\ref{fig:marginplot}.
-Along the horizontal axis, the red boxplot is for those values of x (i.e.
-quit time, age) where no values for y (i.e.
-pack-year smoked, smoking amount) are available, and the blue boxplot for x values where also information for y is available.
-A comparison of the two boxplots can indicate the missing data mechanism.
-For example, the quit time for former smokers that provide no information for smoking amount seems to be higher than that where information is available.
-
 \section*{Discussion}
-In this study, we investigated errors of extracting information from EHRs to characterize the reasons for disagreement between extracted values and report values.
-This information have significant value in establishing priorities for the development of an Information Extraction System in Bio-medical domain as well as conducting clinical trials.
-
-The results suggest that automated extraction of clinical information may be more valid for some variables, and less valid for others.
-In spite of disagreements between information in the reports, automatic information extraction could still be beneficial for quick access to information, depending on the error threshold a user can withstand.
-For example, EHR data provide nearly the human accuracy for variables such as age, sex, race, lung cancer family history, COPD history and emphysema.
-Furthermore, less than 1\% disagreement were observed only with PLCO responses, so the information extracted could still be extremely useful.
-
-Numerous errors were observed, where quantitative information extraction errors were the most prevalent among all variables.
-Several formats were observed in various report with different sentence structures.
-Furthermore, spelling error, ungrammatical structure, abbreviation and acronym create this extraction process more difficult.
-For instance, quit time of former smoker needs to be extracted from the clause ``q age 20''.
-We can calculate it with (current age - 20) by identifying q and age as ``quit time'' and ``age at quited from smoking''.
-
-For medical condition extraction, the most common failure was due to errors in section boundary and ambiguous pronoun.
-If the section of the current sentence is not identified correctly, we will not be able to extract the medical condition such as lung cancer family history, COPD history, pneumonia, asthma, etc.
-Many of the false positive were due to ambiguous pronoun that may refer family member instead of patient.
-
-A  few data inconsistencies were observed in some of the variables such as sex and race.
-In order to determine the status of sex, we needed to determine the presence or absence of UMLS concepts that represent sex, such as ``male'', ``woman'', ``gentleman'', ``lady'', etc.
-However, we observed information mismatch between or within the same EHR data source.
-
-In order to choose a proper imputation method one must be aware of the missing data mechanism(s).
-The quality of the imputed values depends on the imputation itself and on the imputation method used.
-In this study, matrix and scatter plots were used to determine the missing data mechanism(s) among variables.
-We demonstrate that there were MAR relation between quit time and smoking amount, quit time and smoking duration, quit time and pack-year smoked, age and sex, and age and race.
+From the sequential analysis, we observed that markov model achieved near-human accuracy to categorise the sequence of behavior codes.
+It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112.
+However, unsuccessful interviews are responded with the behavior code 109.
 
 \section*{Conclusion}
-In this paper, we compared result of the integrated systems with PLCO clinical trials and observed that the performance of the systems rivaled the clinical trial, when clinical information appears in electronic health records.
-However, clinical trial is necessary for certain variables such as ``age of start smoking'', ``education'', ``occupation'', etc.
-The proposed system consists of several state-of-the-art NLP systems to identify patients information from electronic health records.
-Experimental result shows 91.2\% to 100\% accuracy in retrieved variables.
-This study also provides an analysis of some of the
-difficulties that could be encountered in extraction of information from free-text clinical reports to automatically extract variables. Each of the
-variables had a different profile of errors, suggesting that some variables may be ``better targets'' for information extraction and auto-annotation than others.
-Finally, we addressed data missingness issue and reviewed different mechanisms through which data may be missing.
-The absence of data elements is the consequence of physician decisions and reflects aspects of the patient's condition.
-As the absence of data is usually considered a hindrance to accurate computer-based diagnosis, we need proper treatment of missing data.
-In this paper, we present missing patterns with matrix and scatter plots, and suggest various ways of dealing with missing values including ``multiple imputation'', ``use of means'', etc.
+In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
+We also propose novel features and report the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods for text
+classification using those features, to help clinical researchers and practitioners assess the feasibility of using these methods for the task of automatic annotation of clinical
+text using the codebooks of realistic size. We found out that Support Vector Machine using only lexical features consistently outperforms all other classifiers on caregiver and
+adolescent datasets according to most metrics. Adding contextual and semantic features further improves the performance of SVM on both datasets, achieving close to human accuracy
+when the codebooks consisting of 16 and 17 classes are used to annotate caregiver and adolescent transcripts, respectively.
+
+This work has important practical implications. First, it can facilitate researchers to establish causal relationship between different communication strategies and desired
+behavioral outcomes without having to repeatedly wade through pages of interview transcripts. Second, since automatic annotation is significantly faster than manual, it can
+dramatically accelerate the pace of research in behavioral sciences.
+Third, information that can directly inform and increase the efficiency of clinical practice for a successful interview.
+Although all experiments were conducted on interview transcripts, the proposed methods and features are not
+specific to a particular domain of Motivational Interviewing, and thus there is also no prima facie reason to believe that they will not be effective for annotation of any other
+type of clinical conversation.
 

### 2017-02-27 src/AppMain.java
update introduction

@@ -1,14 +1,14 @@
 public class AppMain {
 	public static void main(String[] args) throws Exception {
-		Model model = new Model();
-		model.evaluateModelByKFolds(5, false, 2);
+		//Model model = new Model();
+		//model.evaluateModelByKFolds(5, false, 2);
 		//model.printTopKSuccessAndFailureSequence(20,false);
 		//model.provideFrequencyDistributionOfSequence(false);
-		System.out.println("\nFinished execution.");
+		//System.out.println("\nFinished execution.");
 		
-		//preprocess dmc files
-		//WordtoText.wordToText("c:/dmc/rawfiles", "c:/dmc/input");
-		//UtilityClass.preprocessMITranscripts("c:/dmc/input", "c:/dmc/output/");
+		// Preprocess dmc files
+		WordtoText.wordToText("c:/dmc/rawfiles", "c:/dmc/input");
+		UtilityClass.preprocessMITranscripts("c:/dmc/input", "c:/dmc/output/");
 		
 		//CRFClassifier crfModel = new CRFClassifier();
 		//crfModel.evaluateModel(5);

### 2017-02-27 paper/AMIA/main.tex
update introduction section of DMC paper for AMIA 2017

@@
 
-\title{Leverage machine learning and textual data mining methods for the classification of clinical interview fragments and their sequences}
+\title{Designing an effective intervention of motivational interview using sequence classification methods}
 
@@
 \noindent{\bf Abstract}
-\textit{This study examines the effectiveness of state-of-the-art supervised machine learning methods for the task of automatic annotation of fragments of clinical interviews in conjunction with sequential analysis of that annotation.
-We used a collection of motivational interview transcripts consisting of 11,353 utterances, which were manually annotated by two human coders as a golden standard, and experimented with state-of-art classifiers, including Na\"{i}ve Bayes, J48 Decision Tree, Support Vector Machine (SVM), Random Forest (RF), AdaBoost, DiscLDA, Conditional Random Fields (CRF) and Convolutional Neural Network (CNN) in conjunction with lexical, contextual (label of the previous utterance) and semantic (distribution of words in the utterance across the Linguistic Inquiry and Word Count dictionaries) features.
-We found out that, when the number of classes is large, the performance of the CNN and CRFs is inferior to SVM.
-Among all classifiers using only lexical features, SVM achieved the highest classification accuracy of 70.8\%, 61\% and 53.7\%, when interview transcripts were annotated with the codebooks consisting of 17, 20 and 41 codes, respectively.
-Using contextual and semantic features, as well as their combination, in addition to lexical ones improved the accuracy of SVM for annotation of utterances in motivational interview transcripts with a codebook consisting of 17 classes to 71.5\%, 74.2\% and 75.1\%, respectively.
-We also used first order markov model that achieved 68.03\% accuracy for the classification of sequence of behavioral codes extracted from the above dataset.
-Therefore, our results demonstrate the potential of using machine learning  and textual data mining methods in conjunction with semantic and contextual features for automatic annotation of clinical interview transcripts  and their sequences with near-human accuracy.}
+\textit{Coming soon!}
 
 \section*{Introduction}
-In this study, our analysis have two directions: automatic annotation and sequential data analysis.
-Annotation (or labeling) of fragments of clinical text with the categories (or labels, codes) from a predefined codebook is an integral part of qualitative research.
-It can also be viewed as classification of textual fragments into a predefined number of categories (classes).
-Textual annotation has been traditionally performed manually by trained coders, which is a tedious, costly and time-consuming process.
-Furthermore, manual annotation increases the likelihood of errors due to coder fatigue and bias associated with human subjectivity.
-To automate tedious cognitive tasks such as classification, supervised machine learning methods have been recently proposed.
-Although these methods have been shown to be successful at binary (two-class) classification \citep{40,41} (e.g.
-classifying textual fragments as neutral or opinionated), we are not aware of any prior studies reported in the literature that evaluate their performance for textual classification tasks involving large number of classes.
-Such tasks, however, are fairly common in clinical setting (e.g.
-annotation of clinical interviews, assignment of ICD-9/10 codes to patient records).
-To address this limitation, in this paper we propose contextual and semantic features and present the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods in conjunction with lexical and the proposed features for the task of automatic annotation of utterances in clinical interview transcripts with the codebooks consisting of large number of classes.
-Our study provides a guideline for clinical informatics researchers and practitioners, who consider an option of using machine learning methods for automatic annotation of clinical text in their projects.
 
-Annotation of clinical interview transcripts to distinguish different behavior types is an important part of clinical research aimed at designing effective interventions for many conditions and disorders.
-In this paper, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers.
+In the recent times, more and more data from various domains being produced in the form of event sequences.
+Thus, sequence classification has become an important task in sequence mining.
+Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types.
+Usually, sequences don't have explicit features and suffer with high dimension of feature space.
+Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications.
+The sequence classification methods can be divided into three large categories: feature-based classification, distance based method and model based classifications.
+In feature-based classification, a sequence is transformed into a single vector of features.
+A supervised learning algorithm such as an support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier.
+An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications.
+On the other hand, distance based method measure the similarity between sequences to determine the quality of the classification significance.
+The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data.
+The third category of sequence classification is model based classifiers such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM) or other statistical models.
+Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.
+
+Sequence classification has a broad range of real word applications such as information retrieval, genomic analysis, health informatics, finance and abnormal detection.
+In genomic research, sequence classification is widely used to detect the function of a new protein \cite{deshpande2002evaluation}.
+In health informatics, ECG can be used as a multi-dimensional time series data to classify an individual as healthy or patient with heart disease \cite{wei2006semi}.
+Another example is to classify rumour stance in Twitter \cite{lukasik2016hawkes}.
+In information retrieval, sequence classification is employed to build  models for classifying documents into different topic categories \cite{sebastiani2002machine} such as sports, politics, news, finance and style.
+Sequence classification is also used in anomaly detection such as abnormal access to systems \cite{lane1999temporal}, malware detection \cite{drew2017polymorphic} and money laundering \cite{liu2008sequence}.
+Other interesting applications include classifying query log sequences to distinguish web-robots from human users \cite{duskin2009distinguishing, tan2004discovery}.
+In computer vision, sequence classification is applied for classifying images \cite{li2000image} and videos \cite{zhang2016tensor}.
+
+In this paper, our target area of application is motivational interview.
+In particular, this paper describes our work on the classification of utterance sequences.
+Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment verses negative change talk or commitment, is an important part of clinical research aimed at designing effective interventions for many conditions and disorders.
+In this study, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers.
 Childhood obesity is a serious public health concern in the United States and worldwide.
@@
 These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}.
-The process of establishing causal linkages to identify effective communication strategies for eliciting change talk and commitment language in MI involves a resource-intensive qualitative coding process.
-First, clinical interviews are transcribed and then each utterance is manually annotated with a set of codes from a pre-defined codebook designating specific behavior types.
-Training human coders to reliably and accurately assign codes to textual fragments requires a large investment of manpower, time and money.
-For example, in a recent MI study \citep{07}, training coders to reliability took about four months and, once trained, coders required five hours to code every recorded hour.
-A similar study reported requiring 60 hours of training over six weeks to attain coder reliability, and the actual coding involved two coding passes and six coders \citep{08}.
 
-Automatic annotation of patient utterances in clinical communication is a challenging task, since patients usually come from a variety of cultural and educational backgrounds and
-their language use can be quite different \citep{09}. This problem is exacerbated when the interviews are conducted with children and adolescents due to their tendency to use
-incomplete sentences and frequently change subjects.
-
-Previous quantitative studies of clinical conversation have resulted in creation of Generalized Medical Interaction Analysis System (GMIAS) \citep{10}, which uses a codebook with generic hierarchical categories.
-The small-size codebook in Comprehensive Analysis of the Structure of Encounters System (CASES) \citep{11} was designed to annotate several
-meta-discursive aspects of medical interviews, such as assigning ``ownership'' of topics and partitioning them into distinct segments (speech acts). It was also shown that the
-fragments of transcripts of routine outpatient visits consisting of several speech acts coded using GMIAS and CASES can be annotated as ``information giving'' and ``requesting
-information'' \citep{12}. Other related previous studies focused on categorizing assertions of medical problems in clinical narrative into 5 classes (present, absent, possible,
-hypothetical, conditional and associated with someone else) using SVM \citep{13} and annotating the utterances in hemodialysis phone dialogue with 3 categories using AdaBoost
-classifier \citep{14}. The present study reports the results of comprehensive evaluation of 8 state-of-the-art classifiers (Na\"{i}ve Bayes \citep{15,16,17}, Support Vector Machine
-\citep{18,19}, Conditional Random Fields \citep{21,22}, J48 \citep{23}, AdaBoost \citep{24}, Random Forest \citep{42}, DiscLDA \citep{25} and Convolutional Neural Network
-\citep{38}) for the task of annotating clinical interviews with a codebook, consisting of a large number of classes. We also propose and experimentally evaluate two novel features
-for this task: contextual features based on the label of the preceding textual fragment and semantic features based on the distribution of words in the annotated fragment over a linguistic lexicon.
-
-Codes automatically assigned to transcripts using classification models presented above allow to abstract away from the raw text  of the interviews and unequivocally establish the casual relationship between provider communication and client change talk (CT).
-Two recent studies \citep{46,47} have examined causality in treatment communication sequences using sequential analysis techniques to assess the probability that CT actually follows specific provider behaviors.
-Both studies found that provider use of MICO communication behaviors was followed by CT more frequently than expected and followed by client counter-CT (CCT; client statements against making changes) less frequently than expected by chance.
-Again, while suggestive of important PPC sequences, these studies were conducted among adults in treatment for alcohol abuse.
-In this project, we used a probabilistic model, in particular, first order markov model for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
-
-\begin{table}[htb!]
-\caption{\textbf{Distribution of utterances over 16 classes in the caregiver dataset.}}    
-\label{table:01}
-\centering
-\begin{tabular}{lp{7cm}rr}
-\hline
-\hline
-Code  & Description & Utterance & \% \\
-\hline
-209 &	Caregiver Change Talk, negative &	297 & 6.82 \\
-212 &	Caregiver Change Talk, positive &	1107 & 25.40 \\
-232 &	Low Uptake, positive &	518 & 11.89 \\
-235 &	High uptake &	231 & 5.30 \\
-301 &	Structure Session &	206 & 4.73 \\
-302 &	General Information, positive &	309 & 7.09 \\
-305 &	Emphasize Autonomy &	148 & 3.40 \\
-306 &	Closed question, Elicit Feedback &	50 & 1.15 \\
-307 &	Support &	108 & 2.48 \\
-308 &	Affirm &	289 & 6.63 \\
-315 &	Reflect, change talk positive, about caregiver &	659 & 15.12 \\
-329 &	Self-disclose &	44 & 1.01 \\
-330 &	Statement, other &	121 & 2.78 \\
-331 &	Open-ended question, elicit change talk positive &	200 & 4.59 \\
-343 &	Open-ended question, target behavior neutral &	33 & 0.76 \\
-344 &	Open-ended question, elicit barriers &	38 & 0.87 \\
-\hline
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}[htb!]
-\caption{\textbf{Distribution of sequences over the adolescent interview session}} 
-\label{table:0001}
-\centering
-\begin{tabular}{lrr}
-\hline
-\hline
-Sequence type  & \# of Sequences & \% \\
-\hline
-Successful & 880 & 82.08\% \\
-Unsuccessful & 192 & 17.92\% \\
-Total & 1072 & 100\% \\
-\hline
-\hline
-\end{tabular}
-\end{table}
+Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes.
+In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
+Currently, a very similar work \cite{jaber2016multi} was done on education data for sequence classifications.
+Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification.
+Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.
 
@@
 \subsection*{\textit{Data collection}}
-The golden standard for evaluation of machine learning methods was created based on the transcripts of motivational interviews conducted by the clinicians at the Pediatric
-Prevention Research Center (PPRC) of Wayne State University. Each interview is comprised of two parts: conversation of a clinician with an adolescent followed by a conversation of
-a clinician with the adolescent's caregiver. All adolescents in this study were between the ages of 12 and 17 (M = 14.7, SD = 1.63) and most were female (n = 27). Most caregivers
-were biological mothers (n = 36), who were married or living with a partner (n = 25). The median family income was \$16,000--\$21,999 ranging from less than \$1,000 to
-\$50,000--\$74,999. The audio recordings of the interviews were first transcribed and segmented into utterances belonging to adolescents, caregivers, and counselors, preserving the
-sequence of utterances. Transcripts were then manually annotated by trained human coders according to MYSCOPE \citep{07}, a specialized codebook including a large number of
-behavior codes, which was developed by an interdisciplinary team including a clinical psychologist, a nutrition scientist, a communication scientist, a linguist and a community
-health worker specifically for annotating motivational interviews with obese adolescents. A primary coder independently coded interview sessions and a secondary coder co-coded a
-randomly selected 20\% of the transcripts to monitor reliability ($\kappa$ = 0.696) \citep{07}. The MYSCOPE codebook contains a total of 115 different codes that are grouped into
-the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
+The MYSCOPE \citep{07} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups.
+The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
 include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
@@
 
-To conduct a detailed analysis of performance of classification methods, we used the following two-stage process to create the codebooks with different number of codes for
-adolescent and caregiver sessions.
-In the first stage, we merged conceptually similar behavior codes as well as the codes with similar data distributions, while in the second stage, we eliminated the codes with insufficient data samples.
-In case of the adolescent sessions, we started with 55 adolescent session-specific codes and, after merging the codes with subtle differences (e.g.
-converting valiances of change talk CHT+1, CHT+2 and CHT+3 into CHT+), obtained a codebook with 41 classes.
-We further reduced this codebook to 20 classes after merging 21 classes with similar sample distributions.
-After eliminating the codes that had less than 10 data samples (to ensure that there can be at least one sample of each class in each fold when using 10-fold cross validation experimental design), we obtained a third codebook with 17 codes.
-Using the same approach, we created the codebooks containing 58, 19, and 16 caregiver session-specific codes.
-Table~\ref{table:01} shows the distribution of utterances over 16 classes in the caregiver session transcripts.
-As follows from Table~\ref{table:01}, the distribution of utterances over classes is highly imbalanced even for the codebook of the smallest size, which is fairly common for clinical text.
 To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language.
 Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.
-
-After creating the codebooks, we pre-processed the dataset using the Snowball stemmer available as part of the Weka \citep{26} machine learning toolkit\footnote{
-http://www.cs.waikato.ac.nz/ml/weka/}. We also found out that stopword removal decreased the performance of classification models for our task (e.g., in case of the codebook
-consisting of 17 classes, the accuracy of Na\"{i}ve Bayes decreased from 67\% to 47.10\%, while the accuracy of SVM decreased from 70.76\% to 55.26\%). A likely reason is that,
-although negations are typically considered as stopwords, they are fairly important clues for inferring certain behavior types (e.g., removing the stopword ``not'' completely
-transforms the meaning of a phrase ``not great'').
 
@@
 The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
 
-\begin{table}
-\parbox{.45\linewidth}{
+\begin{table}[h]
 \centering
-\caption{\textbf{Performance of markov model for the classification of PPC code sequences}}
-\label{table:sequence}
-\begin{tabular}{ll}
+\caption{Performance of markov models for the classification of normal PPC code sequence.}
+\label{tab:result_norm_seq}
+  \begin{tabular}{|l|l|l|l|l|l|}
   \hline
+   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline
+
+ \multirow{2}{*}{General Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
+ & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
+ \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989} & 0.8059\\ \cline{2-6}
+ & Second order & 0.7400 & \textbf{0.8449} & 0.7822  & \textbf{0.8449}\\ \hline
+
+  \end{tabular}
+\end{table}
+
+\begin{table}[h]
+\centering
+\caption{Performance of markov models for the classification of alternate PPC code sequence.}
+\label{tab:result_alt_seq}
+  \begin{tabular}{|l|l|l|l|l|l|}
   \hline
-Metrics & Value \\
-\hline
-Accuracy &	0.6803 \\
-Precision &	0.7574 \\
-Recall & 0.6803 \\
-F1-Measure & 0.7092 \\
-\hline
-\hline
+   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline
+
+ \multirow{2}{*}{General Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
+ & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
+ \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9253 & 0.9392\\ \cline{2-6}
+ & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
+
   \end{tabular}
-}
-\hfill
-\parbox{.45\linewidth}{
-\centering
-\caption{\textbf{Confusion matrix of proposed markov model for the classification of sequence of codes}}
-\label{table:conf}
-\begin{tabular}{llll}
-\hline
-\hline
-& Successful & Fail & Total\\
-\hline
-Successful & 642 & 242 & 884 \\
-Fail & 102 & 90 & 192 \\
-Total & 744 & 332 & 1076 \\
-\hline
-\hline
-\end{tabular}
-}
 \end{table}
@@
 Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-
-\begin{table}[htb!]
-\caption{\textbf{Top 5 most likely successful and unsuccessful interview sessions}}    
-\label{table:topsec}
-\centering
-\begin{tabular}{lll}
-\hline
-\hline
-No. & Successful Sequences &  Unsuccessful Sequences\\
-\hline
-1 & 311\rightarrow112\rightarrow301\rightarrow500 & 343\rightarrow109\rightarrow312\rightarrow400 \\
-2 & 310\rightarrow112\rightarrow311\rightarrow500 & 311\rightarrow109\rightarrow312\rightarrow400 \\
-3 & 311\rightarrow117\rightarrow309\rightarrow500 & 327\rightarrow120\rightarrow312\rightarrow400 \\
-4 & 312\rightarrow112\rightarrow311\rightarrow500 & 314\rightarrow103\rightarrow314\rightarrow400 \\
-5 & 331\rightarrow112\rightarrow311\rightarrow500 & 344\rightarrow109\rightarrow326\rightarrow400 \\
-\hline
-\hline
-\end{tabular}
-\end{table}
 

### 2017-02-27 paper/AMIA/main.tex
update introduction and conclusion sections

@@
 Childhood obesity is a serious public health concern in the United States and worldwide.
-Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \citep{01}.
-Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \citep{02}.
+Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
+Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
 Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity.
-One approach to designing intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change \citep{03,04}.
+One approach to designing an effective intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
 The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
-These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}.
+These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.
 
@@
 In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
-Currently, a very similar work \cite{jaber2016multi} was done on education data for sequence classifications.
+To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue.
+A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues.
+Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns.
 Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification.
@@
 \subsection*{\textit{Data collection}}
-The MYSCOPE \citep{07} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups.
+The MYSCOPE \cite{} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups.
 The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
@@
 sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
-(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}.
+(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}. \\
 
@@
 
-\textbf {Markov Model}: In probability theory, a markov model \citep{45} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property).
+\textbf {Markov Model}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property).
 Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable.
@@
     \centering
-    \includegraphics[width=0.70\textwidth]{figures/deltadata.jpg}
+    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
     \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
@@
 \subsection*{\textit{Evaluation methods}}
-To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \citep{36} as an experimental design.
+To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \cite{} as an experimental design.
 The performance of different classifiers and feature sets was evaluated in terms of precision, recall, F1 score (F1), kappa measure and accuracy using weighted macro-averaging over folds.
@@
 Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
-
-\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
-Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
-Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
-From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
-The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
+\\
 
@@
 
+\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
+Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
+Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
+From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
+The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
+
+Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
+\\
+
 \begin{table}[h]
@@
 
-Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
-Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-
 \section*{Discussion}
@@
 In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
-We also propose novel features and report the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods for text
-classification using those features, to help clinical researchers and practitioners assess the feasibility of using these methods for the task of automatic annotation of clinical
-text using the codebooks of realistic size. We found out that Support Vector Machine using only lexical features consistently outperforms all other classifiers on caregiver and
-adolescent datasets according to most metrics. Adding contextual and semantic features further improves the performance of SVM on both datasets, achieving close to human accuracy
-when the codebooks consisting of 16 and 17 classes are used to annotate caregiver and adolescent transcripts, respectively.
+We also propose novel features and report the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods for text classification using those features, to help clinical researchers and practitioners assess the feasibility of using these methods for the task of automatic annotation of clinical text using the codebooks of realistic size.
+We found out that Support Vector Machine using only lexical features consistently outperforms all other classifiers on caregiver and adolescent datasets according to most metrics.
+Adding contextual and semantic features further improves the performance of SVM on both datasets, achieving close to human accuracy when the codebooks consisting of 16 and 17 classes are used to annotate caregiver and adolescent transcripts, respectively.
 
-This work has important practical implications. First, it can facilitate researchers to establish causal relationship between different communication strategies and desired
-behavioral outcomes without having to repeatedly wade through pages of interview transcripts. Second, since automatic annotation is significantly faster than manual, it can
-dramatically accelerate the pace of research in behavioral sciences.
+This work has important practical implications.
+First, it can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
+Second, since automatic annotation is significantly faster than manual, it can dramatically accelerate the pace of research in behavioral sciences.
 Third, information that can directly inform and increase the efficiency of clinical practice for a successful interview.
-Although all experiments were conducted on interview transcripts, the proposed methods and features are not
-specific to a particular domain of Motivational Interviewing, and thus there is also no prima facie reason to believe that they will not be effective for annotation of any other
-type of clinical conversation.
+Although all experiments were conducted on interview transcripts, the proposed methods and features are not specific to a particular domain of Motivational Interviewing, and thus there is also no prima facie reason to believe that they will not be effective for annotation of any other type of clinical conversation.
+
+Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the sequence classification.
 
 \section*{Acknowledgments}
-We would like to thank the technical staff at Henry Ford Health Systems and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming binary data into free text EHR used for experiments reported in this paper.
+We would like to thank the technical staff at Pediatric Prevention Research Center and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming data into free text EHR used for experiments reported in this paper.
 

### 2017-02-27 paper/AMIA/main.tex
update markov chain subsection

@@
 Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns.
-Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification.
+Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
+In addition, a k-order Markov Model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.
 Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.
@@
 
-\subsection*{\textit{System overview}}
-The pipeline consists of two stages: training and testing.
-Prior to the training stage, we preprocess the collected clinical interview transcripts by performing stemming, punctuation removal, word segmentation and tokenization.
-Features are then extracted from the preprocessed data.
-During this stage, previous label and LIWC features are used in conjunction with the lexical features to create the feature vectors.
-After that, classifiers are trained on the feature vectors extracted from the training samples and their associated annotations.
-In the testing stage, after creation of feature vectors, the previously trained classifiers predict the label of each utterance in the testing sample.
-Finally, performance of different classifiers is evaluated by calculating standard metrics such as precision, recall, F-score (F1), kappa measure and accuracy.
-Specifically, we evaluated the performance of the following state-of-the-art supervised machine learning methods.
+\subsection*{\textit{Sequence classification techniques}}
+Generally, a sequence is an ordered list of events.
+In our project, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.
+Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$.
+Here, L is fixed and known in advances.
+In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.
 
-\textbf {Markov Model}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property).
+\textbf {Markov Chain}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property).
 Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable.
@@
 Figure\ref{fig:delta} illustrates the characteristics of proposed model by varying the value of $\delta$.
+
+The Markov Model we have discussed so far is referred to as the first order Markov Model, since we only look one code in the past to compute the transition probability.
+A generalization of the first order markov model is the $k^{th}$ order markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols.
+Thus, $k^{th}$ order markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
+In our experiment, we also used second order markov model.
+Hence, our model have $N^2$ states with transitional probability matrix of size $N^2 \times N$.
+
+\textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
 
@@
 
-Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the sequence classification.
+Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
 

### 2017-02-28 paper/AMIA/main.tex
update part of abstract and methods sections

@@
 \noindent{\bf Abstract}
-\textit{Coming soon!}
+\textit{Childhood obesity is a serious public health concern in the United States and worldwide.
+Therefore, there is a need for identification of successful and unsuccessful motivational interviews to facilitate the development of effective interventions for childhood obesity.
+This paper describes our work on the classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment.
+Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences.
+The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively.
+}
 
 \section*{Introduction}
-
 In the recent times, more and more data from various domains being produced in the form of event sequences.
@@
 Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes.
-In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
+In this study, probabilistic models are employed for the analysis of sequence data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
 To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue.
@@
 \subsection*{\textit{Data collection}}
-The MYSCOPE \cite{} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups.
-The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
-include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
-sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
-(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}. \\
+The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances.
+A fragment of an adolescent session transcript is presented in Table~\ref{table:anno_examp}.
+The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
+These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found.
+A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences).
+In our experiment, these two subsets are utilized to train two Markov chain or HMM.\\
 
 \begin{table*}[htb]
-\caption{\textbf{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}}    
-\label{table:06}
+\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
+\label{table:anno_examp}
 \centering
@@
 
-To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language.
-Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.
-
 \subsection*{\textit{Sequence classification techniques}}
 Generally, a sequence is an ordered list of events.
-In our project, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.
+In our study, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.
 Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$.
@@
 \textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
-    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
-    \label{fig:delta}
-\end{figure}
+HMM (Hidden Markov Model) is a very
+powerful tool to statistically model a process that varies in time.
+It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces the time set of observations.
+A HMM can be fully specified by (1) N, the number of states in the model; (2) M, the number of distinct observation symbols per state; (3) { } A ij = a , the state transition probability distribution; (4) { ( )} B j = b k , the observation symbol probability distribution; and (5) { } Π = π i , the initial state distribution[10].
+The Baum-Welch reestimation method was implemented to train a hidden Markov model for each country using the training set.
+To identify the country of an unknown melody, the Viterbi algorithm was used to decode the sequence and compute its log probabilities respectively using HMM’s trained for different countries.
+We then assign the melody to the country with the highest log probability.
 
@@
 
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
+    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
+    \label{fig:delta}
+\end{figure}
+
 Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
-\\
+
+\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
+Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
+Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
+From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
+The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
 
@@
 
-\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
-Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
-Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
-From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
-The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
-
 Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
 Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-\\
+
+In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the dataset.
+However, the number of states in higher order Markov chains grows exponentially with the order of the chain.
+Consequently, it is harder to accurately estimate all the transition probabilities from the training set.
+That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates.
+In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain.
+However, this cannot be easily done for many applications areas.
+To overcome these problems many variants of the Markov chains have been developed.
+In the next section we discuss two such variants: the Interpolated Markov Models [SDKW98] and the Selective Markov Models [DK01].
+The basic idea of these approaches is to combine portions of different order Markov chains so that to improve the overall classification accuracy.\\
 
@@
 \section*{Conclusion}
+It has been shown how Hidden Markov Models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
 In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
-We also propose novel features and report the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods for text classification using those features, to help clinical researchers and practitioners assess the feasibility of using these methods for the task of automatic annotation of clinical text using the codebooks of realistic size.
-We found out that Support Vector Machine using only lexical features consistently outperforms all other classifiers on caregiver and adolescent datasets according to most metrics.
-Adding contextual and semantic features further improves the performance of SVM on both datasets, achieving close to human accuracy when the codebooks consisting of 16 and 17 classes are used to annotate caregiver and adolescent transcripts, respectively.
-
-This work has important practical implications.
-First, it can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
-Second, since automatic annotation is significantly faster than manual, it can dramatically accelerate the pace of research in behavioral sciences.
-Third, information that can directly inform and increase the efficiency of clinical practice for a successful interview.
-Although all experiments were conducted on interview transcripts, the proposed methods and features are not specific to a particular domain of Motivational Interviewing, and thus there is also no prima facie reason to believe that they will not be effective for annotation of any other type of clinical conversation.
-
+This work can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
+Information that can directly inform and increase the efficiency of clinical practice for a successful interview.
 Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.

### 2017-02-28 paper/AMIA/main.tex
completed abstract writing

@@
 \textit{Childhood obesity is a serious public health concern in the United States and worldwide.
-Therefore, there is a need for identification of successful and unsuccessful motivational interviews to facilitate the development of effective interventions for childhood obesity.
-This paper describes our work on the classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment.
+Therefore, there is a need for identification of successful motivational interviews to facilitate the development of effective interventions for childhood obesity.
+This paper presents our work on the classification of utterance sequences into successful and unsuccessful behavior outcomes.
 Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences.
 The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively.
-}
+This shows that the behavior codes carry some inherited information to distinguish motivational interviews.
+We expect that the result will improve if we use a more discriminable data set.
+Furthermore, the results suggest us a new way to think about designing the motivational interview.}
 
@@
 A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences).
-In our experiment, these two subsets are utilized to train two Markov chain or HMM.\\
+In our experiment, these two subsets are utilized to train two Markov chain or HMM.
 
@@
 
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
+    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
+    \label{fig:delta}
+\end{figure}
+
 The Markov Model we have discussed so far is referred to as the first order Markov Model, since we only look one code in the past to compute the transition probability.
@@
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
-    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
-    \label{fig:delta}
-\end{figure}
-
 Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
@@
 The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
+
+Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
+\\
 
@@
 \end{table}
-
-Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
-Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
 

### 2017-02-28 paper/AMIA/main.tex
update conclusion

@@
 \section*{Conclusion}
-It has been shown how Hidden Markov Models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
-In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
+It has been shown how Markov chain and Hidden Markov Models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
+Experimental results shows that the behavior codes carry some inherited information to distinguish motivational interviews.
 This work can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
-Information that can directly inform and increase the efficiency of clinical practice for a successful interview.
+It provides information that can directly inform and increase the efficiency of clinical practice for a successful interview.
+This work also suggest some successful patterns for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
 Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
+We also have plan to use large data set in conjunction with some advanced features for the evaluation of sequence classifiers.
 

### 2017-02-28 paper/AMIA/main.tex
completed description of Hidden Markov Model method

@@
 A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences).
-In our experiment, these two subsets are utilized to train two Markov chain or HMM.
+In our experiment, these two subsets are utilized to train Markov chain and HMM.
 
@@
 \textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-HMM (Hidden Markov Model) is a very
+Hidden Markov Model (HMM) is a very
 powerful tool to statistically model a process that varies in time.
 It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces the time set of observations.
-A HMM can be fully specified by (1) N, the number of states in the model; (2) M, the number of distinct observation symbols per state; (3) { } A ij = a , the state transition probability distribution; (4) { ( )} B j = b k , the observation symbol probability distribution; and (5) { } Π = π i , the initial state distribution[10].
-The Baum-Welch reestimation method was implemented to train a hidden Markov model for each country using the training set.
-To identify the country of an unknown melody, the Viterbi algorithm was used to decode the sequence and compute its log probabilities respectively using HMM’s trained for different countries.
-We then assign the melody to the country with the highest log probability.
+A HMM can be fully specified by the following quintuple:
+
+\begin{center}
+$\lambda = (N, M, A, B, \pi)$
+\end{center}
+
+\begin{itemize}
+\item N is the number of states for the model
+\item M is the number of distinct observations symbols per state, i.e. the discrete alphabet size
+\item A is the $N\times N$ state transition probability distribution given in the form of a matrix $A = \{a_{ij}\}$
+\item B is the $N\times M$ observation symbol probability distribution given in the form of a matrix $B = \{b_j(k)\}$
+\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
+\end{itemize}
+
+If we opt out the structure parameters M and N we have the more often used compact notation:
+
+\begin{center}
+$\lambda = (A, B, \pi)$
+\end{center}
+
+This model is also called hidden for a reason.
+We are not required to tell it what those aforementioned states should be.
+We just have to tell it how many states we think are needed to model what we want.
+The model should be able to figure a suitable sequence of states to support our observations.
+The Baum-Welch algorithm was implemented to train a hidden Markov model for estimating parameters of each models using the training set.
+To identify the label of an unknown code sequence, the Viterbi algorithm was used to decode the sequence and compute its log probabilities respectively using HMM's trained for different classes.
+Then, the class with the highest log probability was assigned to the given behavior sequences.
+Like Markov chain, we also employed second order hidden markov model for classification of utterance sequences.
 
@@
 
-\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
+\subsection*{\textit{Performance of Markov models for the classification of PPC code sequences}}
 Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
@@
 It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
-
-Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
-Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
 \\
@@
 
-In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the dataset.
-However, the number of states in higher order Markov chains grows exponentially with the order of the chain.
-Consequently, it is harder to accurately estimate all the transition probabilities from the training set.
-That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates.
-In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain.
-However, this cannot be easily done for many applications areas.
-To overcome these problems many variants of the Markov chains have been developed.
-In the next section we discuss two such variants: the Interpolated Markov Models [SDKW98] and the Selective Markov Models [DK01].
-The basic idea of these approaches is to combine portions of different order Markov chains so that to improve the overall classification accuracy.\\
+Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
+Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
+Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
+\\
 
@@
 \end{table}
+
+In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the dataset.
+However, the number of states in higher order Markov chains grows exponentially with the order of the chain.
+Consequently, it is harder to accurately estimate all the transition probabilities from the training set.
+That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates.
+In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain.
+However, this cannot be easily done for many applications areas.
+To overcome these problems hidden Markov model is used with existing library available in online.
+The basic idea of these approaches is to combine portions of different order Markov chains so that to improve the overall classification accuracy.
 

### 2017-02-28 paper/AMIA/main.tex
update format of table 1

@@
 
-\begin{table*}[htb]
+\begin{table}[h]
 \caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
@@
 \centering
-\begin{tabular}{lp{3.6cm}lp{8cm}}
+\begin{tabular}{|l|p{3.6cm}|l|p{8cm}|}
 \hline
-\hline
-Annotation  & Description & Speaker & Text \\
-\hline
+Annotation  & Description & Speaker & Text \\\hline
 331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do ?
 mm-hmm meaning is that food available for you ?
-\\
-117 &	Low Uptake, positive	& Adolescent &	Yes \\
-301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\
-112 &	Change Talk positive	& Adolescent &	my um aunt \\
-301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\
-112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\
-309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\
-\hline
-\hline
+\\\hline
+117 &	Low Uptake, positive	& Adolescent &	Yes \\\hline
+301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\\hline
+112 &	Change Talk positive	& Adolescent &	my um aunt \\\hline
+301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\\hline
+112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\\hline
+309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\\hline
 \end{tabular}
-\end{table*}
+\end{table}
 

### 2017-02-28 paper/AMIA/main.tex
update evaluation methods of the paper

@@
 The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data.
-The third category of sequence classification is model based classifiers such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM) or other statistical models.
+The third category of sequence classification is model based classifiers such as hidden Markov model \cite{rabiner1989tutorial} (HMM) or other statistical models.
 Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.
@@
 Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns.
-Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
-In addition, a k-order Markov Model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.
-Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.
+Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
+In addition, a k-order Markov model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.
+Therefore, we utilize Markov chain and hidden Markov model for the classification of utterance sequences.
 
@@
 \subsection*{\textit{Data collection}}
-The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances.
-A fragment of an adolescent session transcript is presented in Table~\ref{table:anno_examp}.
+Table~\ref{tab:data_dist} shows the statistics of used data sets.
+The experimental data sets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances.
+A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
 The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
 These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found.
-A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one dataset that includes all sequences of positive change talk or commitment language (882 sequences) and the other dataset which includes utterances ended with negative change talk or commitment language (192 sequences).
+A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences).
 In our experiment, these two subsets are utilized to train Markov chain and HMM.
+\\
+
+\begin{table}[h]
+\centering
+\caption{Dataset statistics.}
+\label{tab:data_dist}
+  \begin{tabular}{|l|l|l|l|l|l|}
+  \hline
+   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{Successful sequences}  & \textbf{Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
+ Normal sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 3.5 \\\hline
+Alternate sequences & 18000 & 15000 (82.09\%) & 3000 (17.91\%) & 4.5 \\\hline 
+  \end{tabular}
+\end{table}
 
@@
 \caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
-\label{table:anno_examp}
+\label{tab:anno_examp}
 \centering
@@
 Annotation  & Description & Speaker & Text \\\hline
-331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do ?
-mm-hmm meaning is that food available for you ?
+331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do?
+mm-hmm meaning is that food available for you?
 \\\hline
@@
 
-\textbf {Markov Chain}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property).
+\textbf {Markov Chain}: In probability theory, a Markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the Markov property).
 Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable.
-For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
-A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
+For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
+A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
 
@@
 where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ have been observed in the training data.
-Given a markov model $M$ (such that $S\subseteq V$), a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ have been generated from a markov model $M$ is:
+Given a Markov model $M$ (such that $S\subseteq V$), a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ have been generated from a Markov model $M$ is:
 
@@
     \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
-    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
+    \caption{\textbf{Characteristics of proposed Markov model by varying the value of $\delta$}}
     \label{fig:delta}
@@
 
-The Markov Model we have discussed so far is referred to as the first order Markov Model, since we only look one code in the past to compute the transition probability.
-A generalization of the first order markov model is the $k^{th}$ order markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols.
-Thus, $k^{th}$ order markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
-In our experiment, we also used second order markov model.
+The Markov model we have discussed so far is referred to as the first order Markov model, since we only look one code in the past to compute the transition probability.
+A generalization of the first order Markov model is the $k^{th}$ order Markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols.
+Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
+In our experiment, we also used second order Markov model.
 Hence, our model have $N^2$ states with transitional probability matrix of size $N^2 \times N$.
 
-\textbf {Hidden Markov Model}: Hidden Markov Models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-Hidden Markov Model (HMM) is a very
+\textbf {Hidden Markov Model (HMM)}: hidden Markov models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
+Hidden Markov model is a very
 powerful tool to statistically model a process that varies in time.
@@
 Then, the class with the highest log probability was assigned to the given behavior sequences.
-Like Markov chain, we also employed second order hidden markov model for classification of utterance sequences.
+Like Markov chain, we also employed second order hidden Markov model for classification of utterance sequences.
 
 \subsection*{\textit{Evaluation methods}}
-To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \cite{} as an experimental design.
-The performance of different classifiers and feature sets was evaluated in terms of precision, recall, F1 score (F1), kappa measure and accuracy using weighted macro-averaging over folds.
+The performance of classifiers are reported on the basis of 5-fold cross validation by using weighted macro-averaging over folds.
+Precision, recall, F-measure and accuracy were used to evaluate the performance of classifiers.
+A true positive (TP) was counted when classifier correctly classified the sequence into actual class; a false positive (FP) was counted when classifier incorrectly classified other classes to the considered class; a false negative (FN) when the considered class incorrectly classified to other classes.
+Precision of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences identified as that particular class by the classifier.
+In other words, Precision = TP / (TP + FP).
+Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (labeled data set).
+In more specifically, Recall = TP / (TP + FN).
+Similarly, Accuracy of a class was also defined as the ratio between the numbers of correctly classified sequences and total number of sequences for that particular class in the gold standard.
+However, F-measure can be computed as the harmonic mean of precision and recall, or 2 x Precision x Recall / (Precision + Recall).
 
@@
 \item determining the effectiveness of the proposed contextual and semantic features.
-\item determining the performance of markov model in conjunction with determining PPC sequences that are most likely to translate into CT.
+\item determining the performance of Markov model in conjunction with determining PPC sequences that are most likely to translate into CT.
 \end{itemize}
 
-Since clinical researchers typically annotate caregiver and adolescent sessions separately, we first created two experimental datasets consisting of only adolescent and only caregiver session transcripts.
-Second, besides evaluating the accuracy of annotating adolescent and caregiver transcripts with the codebooks containing an entire set of codes, we also conducted a series of experiments with the codebooks of smaller sizes created as outlined above.
-Third, besides training and testing NB, SVM, CRF, Decision Tree, Boosting, DiscLDA, Random Forest and CNN classifiers using only lexical features, we also evaluated the effectiveness of the proposed contextual and semantic features.
-
-Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
-
-\subsection*{\textit{Performance of Markov models for the classification of PPC code sequences}}
+\subsection*{\textit{Performance of models for the classification of normal PPC code sequences}}
 Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
 Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
-From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
+From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
 The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
-\\
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
 
@@
 \centering
-\caption{Performance of markov models for the classification of normal PPC code sequence.}
+\caption{Performance of Markov models for the classification of normal PPC code sequence.}
 \label{tab:result_norm_seq}
@@
 
- \multirow{2}{*}{General Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
+ \multirow{2}{*}{General Markov chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
  & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
@@
 
-Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Since clinical researchers typically annotate caregiver and adolescent sessions separately, we first created two experimental data sets consisting of only adolescent and only caregiver session transcripts.
+Second, besides evaluating the accuracy of annotating adolescent and caregiver transcripts with the codebooks containing an entire set of codes, we also conducted a series of experiments with the codebooks of smaller sizes created as outlined above.
+Third, besides training and testing NB, SVM, CRF, Decision Tree, Boosting, DiscLDA, Random Forest and CNN classifiers using only lexical features, we also evaluated the effectiveness of the proposed contextual and semantic features.
+Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
+
+\subsection*{\textit{Performance of models for the classification of alternate PPC code sequences}}
+Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
 Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
 Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
+Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
 Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
@@
 \centering
-\caption{Performance of markov models for the classification of alternate PPC code sequence.}
+\caption{Performance of Markov models for the classification of alternate PPC code sequence.}
 \label{tab:result_alt_seq}
@@
 
- \multirow{2}{*}{General Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
+ \multirow{2}{*}{General Markov chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
  & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
@@
 
-In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the dataset.
+In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the data set.
 However, the number of states in higher order Markov chains grows exponentially with the order of the chain.
@@
 \section*{Discussion}
-From the sequential analysis, we observed that markov model achieved near-human accuracy to categorise the sequence of behavior codes.
+From the sequential analysis, we observed that Markov model achieved near-human accuracy to categories the sequence of behavior codes.
 It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112.
 However, unsuccessful interviews are responded with the behavior code 109.
+Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
+Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
+From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
+The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
+
+From the sequential analysis, we observed that Markov model achieved near-human accuracy to categories the sequence of behavior codes.
+It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112.
+However, unsuccessful interviews are responded with the behavior code 109.
+Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
+Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
+From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
+The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
+It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
 
 \section*{Conclusion}
-It has been shown how Markov chain and Hidden Markov Models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
+It has been shown how Markov chain and hidden Markov models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
 Experimental results shows that the behavior codes carry some inherited information to distinguish motivational interviews.

### 2017-03-01 src/AppMain.java
update dataset statistics

@@ -1,14 +1,14 @@
 public class AppMain {
 	public static void main(String[] args) throws Exception {
-		//Model model = new Model();
-		//model.evaluateModelByKFolds(5, false, 2);
+		Model model = new Model();
+		model.evaluateModelByKFolds(5, false, 0);
 		//model.printTopKSuccessAndFailureSequence(20,false);
 		//model.provideFrequencyDistributionOfSequence(false);
-		//System.out.println("\nFinished execution.");
+		System.out.println("\nFinished execution.");
 		
 		// Preprocess dmc files
-		WordtoText.wordToText("c:/dmc/rawfiles", "c:/dmc/input");
-		UtilityClass.preprocessMITranscripts("c:/dmc/input", "c:/dmc/output/");
+		//WordtoText.wordToText("c:/dmc/rawfiles", "c:/dmc/input");
+		//UtilityClass.preprocessMITranscripts("c:/dmc/input", "c:/dmc/output/");
 		
 		//CRFClassifier crfModel = new CRFClassifier();
 		//crfModel.evaluateModel(5);

### 2017-03-01 src/Model.java

@@ -43,6 +43,12 @@ public class Model {
 		Collections.shuffle(data);
 		
 		System.out.println("Total sequences: " + data.size());
+		double seq_len = 0;
+		for(int t=0; t<data.size(); t++){
+			seq_len += data.get(t).getStates().size();
+		}
+		seq_len = seq_len/data.size();
+		System.out.println("Avg. sequence length: " + seq_len);
 		
 		String finalPrint = "";
 		double maxF1Measure = 0, optP = 0, optR = 0, optAcc = 0, optErrRateofFail = 0, optDelta = 0;

### 2017-03-01 paper/AMIA/main.tex
update dataset statistics
update data collection and part of result

@@
 \subsection*{\textit{Data collection}}
-Table~\ref{tab:data_dist} shows the statistics of used data sets.
 The experimental data sets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances.
-A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
+Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequences from adolescent session and excluded sequences from caregiver session in the MI transcript.
 The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
@@
 A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences).
-In our experiment, these two subsets are utilized to train Markov chain and HMM.
+Alternate sequences are the possible combination of normal sequence.
+For example, there are 6 possible alternate sequences for the following sequences: (C1, A1, A2, C2, C3, C4, A3), where adolescent and counselor codes are represented by A and C, respectively.
+Therefore, (C1, A1, C2, A3), (C1, A1, C3, A3), (C1, A1, C4, A3), (C1, A2, C2, A3), (C1, A2, C3, A3) and (C1, A2, C4, A3) were used as alternate sequences for the above sequence.
+A total of 18753 alternate sequences have been observed which further partitioned into two subsets of 18139 successful sequences and 614 unsuccessful sequences.
+In our experiment, normal and alternate sequences are employed with two subsets (successful and unsuccessful) to train Markov chain and HMM.
+Dataset statistics and a fragment of an adolescent session transcript are presented in Table~\ref{tab:data_dist} and Table~\ref{tab:anno_examp}.
 \\
@@
 \centering
-\caption{Dataset statistics.}
+\caption{General statistics of dataset.}
 \label{tab:data_dist}
@@
   \hline
-   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{Successful sequences}  & \textbf{Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
- Normal sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 3.5 \\\hline
-Alternate sequences & 18000 & 15000 (82.09\%) & 3000 (17.91\%) & 4.5 \\\hline 
+   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
+ Normal sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
+Alternate sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
   \end{tabular}
@@
 \section*{Results}
-Experimental evaluation of automatic annotation of clinical interview fragments and their sequences by using machine learning and probabilistic model included several dimensions:
-\begin{itemize}
-\item determining the performance of classifiers on the codebooks of different size;
-\item determining the effectiveness of the proposed contextual and semantic features.
-\item determining the performance of Markov model in conjunction with determining PPC sequences that are most likely to translate into CT.
-\end{itemize}
+Experimental evaluation of the classification of utterance sequences by using probabilistic models are reported with two set of experiments: first set of experiment was focused on normal sequences, while the second set of experiment was performed on alternate sequences.
 
 \subsection*{\textit{Performance of models for the classification of normal PPC code sequences}}
-Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
 Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.

### 2017-03-01 paper/AMIA/main.tex
update data collection subsection

@@
 A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences).
-Alternate sequences are the possible combination of normal sequence.
+Alternate sequences are the possible combination of normal sequence where two consecutive codes can not be come from same speaker.
 For example, there are 6 possible alternate sequences for the following sequences: (C1, A1, A2, C2, C3, C4, A3), where adolescent and counselor codes are represented by A and C, respectively.
 Therefore, (C1, A1, C2, A3), (C1, A1, C3, A3), (C1, A1, C4, A3), (C1, A2, C2, A3), (C1, A2, C3, A3) and (C1, A2, C4, A3) were used as alternate sequences for the above sequence.
-A total of 18753 alternate sequences have been observed which further partitioned into two subsets of 18139 successful sequences and 614 unsuccessful sequences.
+A total of 18753 alternate sequences have been observed which further partitioned into two subsets of 18139 successful and 614 unsuccessful sequences.
 In our experiment, normal and alternate sequences are employed with two subsets (successful and unsuccessful) to train Markov chain and HMM.

### 2017-03-02 paper/AMIA/main.tex
update results and discussion sections

@@
 if $p(S\rightarrow CT) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in CT.
-We experimentally determined the value of $\delta$ to 0.3 which provides acceptable accuracy with minimum false positive rate.
-Figure\ref{fig:delta} illustrates the characteristics of proposed model by varying the value of $\delta$.
+To fit the model, we experimentally determined the value of $\delta$ where model provides highest F-measure with lower bias.
+Figure\ref{fig:delta} [this is a temporary figure, I will change it later] illustrates an example of the characteristics of Markov model by varying the value of $\delta$.
 
@@
 \subsection*{\textit{Performance of models for the classification of normal PPC code sequences}}
-Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
-From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
-The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\
+Performance summary for the classification of normal sequences is presented in Table~\ref{tab:result_norm_seq}.
+It follows that first order hidden Markov model works very well and achieves highest precision 0.7980 with F-measure 0.7989.
+However, our designed second order Markov model shows lowest performance in terms of precision and F-measure.
+It was also noticed that the lowest recall 0.7686 achieved in first order Markov model.
+On the other hand, second order hidden Markov model achieves the highest recall 0.8449 and accuracy 0.8449.
+From Table~\ref{tab:result_norm_seq}, we observed that performance degrades from first order to second order.
+Precision and F-measures drop 6.74\% \& 2.39\% in Markov chain and 7.27\% \& 2.09\% in HMM, respectively, when using second order models.
+In contrast, recall improves 2.64\%--4.84\% in second order Markov model and HMM.
+Overall, HMM performed better than general Markov model in normal utterance sequences.
 
@@
 \centering
-\caption{Performance of Markov models for the classification of normal PPC code sequence.}
+\caption{Performance of Markov models for the classification of normal PPC code sequence. Highest performances for each metrics are highlighted in bold faces.}
 \label{tab:result_norm_seq}
@@
 
-Since clinical researchers typically annotate caregiver and adolescent sessions separately, we first created two experimental data sets consisting of only adolescent and only caregiver session transcripts.
-Second, besides evaluating the accuracy of annotating adolescent and caregiver transcripts with the codebooks containing an entire set of codes, we also conducted a series of experiments with the codebooks of smaller sizes created as outlined above.
-Third, besides training and testing NB, SVM, CRF, Decision Tree, Boosting, DiscLDA, Random Forest and CNN classifiers using only lexical features, we also evaluated the effectiveness of the proposed contextual and semantic features.
-Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.
-
 \subsection*{\textit{Performance of models for the classification of alternate PPC code sequences}}
-Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
-Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
-Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
-Two Markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses.
-Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.
+Table~\ref{tab:result_alt_seq} shows the classification performance of models applied on alternate sequences.
+The results demonstrate that second order hidden Markov model performed better than Markov chain in terms of recall and achieved 97.13\% recall.
+Second order general Markov model shows the highest precision 0.9778 as well as highest F-measure 0.9736.
+On the other hand, first order hidden Markov model displays lowest performance in all metrics compared to all other models.
+It was also examined that every models perform better in second order compared to first order.
+General Markov chain achieves 1\%, 1\% and 1\% improvement for precision, recall and F-measure, respectively, from first order to second order.
+Similarly, hidden Markov model achieves 1\%, 1\% and 1\% improvement for precision, recall and F-measure, respectively, from first order to second order.
+Overall, second order Markov model performs better for alternate utterance sequences.
 \\
@@
 \centering
-\caption{Performance of Markov models for the classification of alternate PPC code sequence.}
+\caption{Performance of Markov models for the classification of alternate PPC code sequence. Highest performances for each metrics are highlighted in bold faces.}
 \label{tab:result_alt_seq}
@@
 
-In general, these higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the data set.
+\section*{Discussion}
+The purpose of this study was to facilitate the design of motivational interview by examining the performance of sequence classification in the field of health informatics.
+Studying the results across the different classification schemes we can make some key observations.
+
+First, higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the data set.
 However, the number of states in higher order Markov chains grows exponentially with the order of the chain.
@@
 In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain.
-However, this cannot be easily done for many applications areas.
-To overcome these problems hidden Markov model is used with existing library available in online.
-The basic idea of these approaches is to combine portions of different order Markov chains so that to improve the overall classification accuracy.
+However, this cannot be easily done for many applications areas include our case.
+To skip these problems we restricted our study upto second order.
+Shorter average length of utterance sequence is another reason for keeping the order lower.
 
-\section*{Discussion}
-From the sequential analysis, we observed that Markov model achieved near-human accuracy to categories the sequence of behavior codes.
-It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112.
-However, unsuccessful interviews are responded with the behavior code 109.
-Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
-Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
-From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
-The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
+Second, overall performance of hidden Markov model is better than general Markov model in both type of sequences.
+We observed that hidden Markov model achieved near-human accuracy to categories the sequences of behavior codes.
+This means that hidden states capture the propagated information passes through the sequence for determining final outcome.
 
-From the sequential analysis, we observed that Markov model achieved near-human accuracy to categories the sequence of behavior codes.
-It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112.
-However, unsuccessful interviews are responded with the behavior code 109.
-Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis.
-Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}.
-From Table~\ref{table:sequence}, it follows that our designed Markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092.
-The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}.
-It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.
+Third, alternate sequences provide higher performance for all Markov models.
+Even, it provides highest precision and F-measure for alternate sequences.
+This indicates that each pair of alternate behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview.
+And this is obvious because if we know the previous two behavior codes (e.g.
+provider's question and adolescent's answer) then it is easier to know the expected answer for the current question.
+
+Finally, it was seen that successful interviews are more frequently responded by the adolescent with positive behavior changes which lead to make a positive commitment for loosing weight.
+Whereas, unsuccessful interviews are responded with negative behavior changes which lead to a negative commitment or no commitment.
 

### 2017-03-02 src/Model.java

@@ -15,7 +15,7 @@ public class Model {
 	public static ArrayList<CodeSequence> testData = new ArrayList<>();
 	public static ArrayList<CodeSequence> data = new ArrayList<>();
 	public static HashMap<String,String> codemap = new HashMap<>();
-	public static double delta = -1, sumRatio400 = 0, sumRatio500=0;
+	public static double delta = -4, sumRatio400 = 0, sumRatio500=0;
 	public static double fpRate = 0;
 	public static int tp=0, fp=0, tn=0, fn=0;
 	
@@ -53,9 +53,9 @@ public class Model {
 		String finalPrint = "";
 		double maxF1Measure = 0, optP = 0, optR = 0, optAcc = 0, optErrRateofFail = 0, optDelta = 0;
 		
-		for(int j=0; j < 1; j++){
+		for(int j=0; j < 5; j++){
 			System.out.println("\n\nIteration: " + j);
-			delta = delta + 0.1;
+			delta = delta + 1;
 			//delta = 0.2;
 			results = 0;
 			prec = 0;

### 2017-03-02 paper/AMIA/main.tex
update discussion sections
update data collection subsection

@@
 \end{table}
+
+Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor start with an open-ended question and get positive feedback at the end. \\
 
@@
 We observed that hidden Markov model achieved near-human accuracy to categories the sequences of behavior codes.
-This means that hidden states capture the propagated information passes through the sequence for determining final outcome.
+This means that hidden states capture the information passes through the codes of the utterance sequence for determining final outcome.
 

### 2017-03-03 paper/AMIA/main.tex
update discussion sections
update paper title

@@
 
-\title{Designing an effective intervention of motivational interview using sequence classification methods}
+\title{Sequence classifications to facilitate designing an effective intervention of motivational interview}
 
@@
 Third, alternate sequences provide higher performance for all Markov models.
-Even, it provides highest precision and F-measure for alternate sequences.
+Even, it provides highest precision and F-measure for general Markov model.
 This indicates that each pair of alternate behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview.

### 2017-03-03 src/AppMain.java

@@ -2,7 +2,7 @@ public class AppMain {
 	public static void main(String[] args) throws Exception {
 		Model model = new Model();
 		model.evaluateModelByKFolds(5, false, 0);
-		//model.printTopKSuccessAndFailureSequence(20,false);
+		model.printTopKSuccessAndFailureSequence(20,false);
 		//model.provideFrequencyDistributionOfSequence(false);
 		System.out.println("\nFinished execution.");
 		

### 2017-03-03 src/Model.java

@@ -15,7 +15,7 @@ public class Model {
 	public static ArrayList<CodeSequence> testData = new ArrayList<>();
 	public static ArrayList<CodeSequence> data = new ArrayList<>();
 	public static HashMap<String,String> codemap = new HashMap<>();
-	public static double delta = -4, sumRatio400 = 0, sumRatio500=0;
+	public static double delta = -1, sumRatio400 = 0, sumRatio500=0;
 	public static double fpRate = 0;
 	public static int tp=0, fp=0, tn=0, fn=0;
 	
@@ -53,9 +53,9 @@ public class Model {
 		String finalPrint = "";
 		double maxF1Measure = 0, optP = 0, optR = 0, optAcc = 0, optErrRateofFail = 0, optDelta = 0;
 		
-		for(int j=0; j < 5; j++){
+		for(int j=0; j < 1; j++){
 			System.out.println("\n\nIteration: " + j);
-			delta = delta + 1;
+			delta = delta + 0.1;
 			//delta = 0.2;
 			results = 0;
 			prec = 0;

### 2017-03-03 paper/AMIA/main.tex
added another subsection for common patterns

@@
 It was also examined that every models perform better in second order compared to first order.
-General Markov chain achieves 1\%, 1\% and 1\% improvement for precision, recall and F-measure, respectively, from first order to second order.
-Similarly, hidden Markov model achieves 1\%, 1\% and 1\% improvement for precision, recall and F-measure, respectively, from first order to second order.
+General Markov chain achieves 0.01\%, 2.72\% and 1.37\% improvement for precision, recall and F-measure, respectively, from first order to second order.
+Similarly, hidden Markov model achieves 3.32\%, 4.30\% and 4.82\% improvement for precision, recall and F-measure, respectively, from first order to second order.
 Overall, second order Markov model performs better for alternate utterance sequences.
@@
 
+  \end{tabular}
+\end{table}
+
+\subsection*{\textit{Some common patterns}}
+Table~\ref{tab:common_patterns} shows some successful and unsuccessful patterns which appeared more frequently in motivational interview.
+It can be seen that most successful patterns start with summary of the discussion or open-ended or closed ended question.
+After that, if adolescent demonstrate positive behavior change, counselor reflect that positive attitude immediately so that adolescent can stick with the decision and show his own desire for positive behavior change.
+On the other hand, general information of counselor can lead to a negative behavior change of adolescent, while showing positive attitude in the previous step.
+Because adolescent changes their topic more frequently when general question was asked.
+
+\begin{table}[h]
+\centering
+\caption{Some common patterns in motivational interview for successful and unsuccessful behavior outcome.}
+\label{tab:common_patterns}
+  \begin{tabular}{|l|l|l|l|l|l|}
+  \hline
+   \textbf{Type} & \textbf{Pattern} \\ \hline      
+successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
+& commitment language positive $\rightarrow $ [positive commitment] \\\hline
+successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
+& commitment language positive $\rightarrow $ [positive commitment] \\\hline
+successful & 306 CQ-EF: Closed question, Elicit Feedback $\rightarrow $ 120 HUP-O: High Uptake, other \\
+& $\rightarrow $ 313 R-CML+: Reflect, commitment language positive $\rightarrow $ [positive commitment] \\\hline
+unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
+& General Information, positive $\rightarrow $ [negative commitment] \\\hline
+unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
+& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
+unsuccessful & 305 EA: Emphasize Autonomy $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
+& General Information, positive $\rightarrow $ [negative commitment] \\\hline
   \end{tabular}

### 2017-03-03 paper/AMIA/main.tex
added new figure for figure 1

@@
 To fit the model, we experimentally determined the value of $\delta$ where model provides highest F-measure with lower bias.
-Figure\ref{fig:delta} [this is a temporary figure, I will change it later] illustrates an example of the characteristics of Markov model by varying the value of $\delta$.
+Figure~\ref{fig:delta} illustrates an example of the characteristics of Markov model by varying the value of $\delta$.
+From the figure, it was observed that F-score value is not changing significantly with respect to $\delta $ from -5 to -1.
+This is because a bias is involved.
+However, F-score sharply decreased after $\delta = -0.9$.
+Thus, -0.9 will the optimal value of $\delta $, where F-measure is highest with lower bias.
 
@@
     \centering
-    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
+    \includegraphics[width=0.70\textwidth]{figures/delta.png}
     \caption{\textbf{Characteristics of proposed Markov model by varying the value of $\delta$}}

### 2017-03-03 src/AppMain.java

@@ -2,7 +2,7 @@ public class AppMain {
 	public static void main(String[] args) throws Exception {
 		Model model = new Model();
 		model.evaluateModelByKFolds(5, false, 0);
-		model.printTopKSuccessAndFailureSequence(20,false);
+		//model.printTopKSuccessAndFailureSequence(20,false);
 		//model.provideFrequencyDistributionOfSequence(false);
 		System.out.println("\nFinished execution.");
 		

### 2017-03-03 src/Model.java

@@ -15,7 +15,7 @@ public class Model {
 	public static ArrayList<CodeSequence> testData = new ArrayList<>();
 	public static ArrayList<CodeSequence> data = new ArrayList<>();
 	public static HashMap<String,String> codemap = new HashMap<>();
-	public static double delta = -1, sumRatio400 = 0, sumRatio500=0;
+	public static double delta = 0.9, sumRatio400 = 0, sumRatio500=0;
 	public static double fpRate = 0;
 	public static int tp=0, fp=0, tn=0, fn=0;
 	
@@ -88,7 +88,7 @@ public class Model {
 				optDelta = delta;
 			}
 	
-			System.out.println("Average false positive for failure: " + new DecimalFormat("#.####").format(sumRatio500/(folds)) + " for delta: " + delta);
+			System.out.println("Average false positive for failure: " + new DecimalFormat("#.####").format(sumRatio500/(folds)) + " for delta: " + delta + ", F1-score: " + (f1m/folds));
 			//System.out.println("Confushion Matrix: " + " tp=" + tp + " fp=" + fp + " tn=" + tn + " fn=" + fn +  " for delta: " + delta);
 			sumRatio500 = 0;
 			

### 2017-03-03 paper/AMIA/main.tex
finished first draft of DMC AMIA 2017 Paper

@@
 Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types.
-Usually, sequences don't have explicit features and suffer with high dimension of feature space.
+Usually, sequences don't have explicit features and suffer from the high dimension of feature space.
 Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications.
-The sequence classification methods can be divided into three large categories: feature-based classification, distance based method and model based classifications.
+The sequence classification methods can be divided into three large categories: feature-based classification, distance based method, and model based classifications.
 In feature-based classification, a sequence is transformed into a single vector of features.
-A supervised learning algorithm such as an support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier.
+A supervised learning algorithm such as a support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier.
 An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications.
@@
 In health informatics, ECG can be used as a multi-dimensional time series data to classify an individual as healthy or patient with heart disease \cite{wei2006semi}.
-Another example is to classify rumour stance in Twitter \cite{lukasik2016hawkes}.
+Another example is to classify rumor stance in Twitter \cite{lukasik2016hawkes}.
 In information retrieval, sequence classification is employed to build models for classifying documents into different topic categories \cite{sebastiani2002machine} such as sports, politics, news, finance and style.
@@
 
-In this paper, our target area of application is motivational interview.
+In this paper, our target area of application is the motivational interview.
 In particular, this paper describes our work on the classification of utterance sequences.
-Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment verses negative change talk or commitment, is an important part of clinical research aimed at designing effective interventions for many conditions and disorders.
-In this study, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers.
+Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many conditions and disorders.
+In this study, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents (teens) and their caregivers.
 Childhood obesity is a serious public health concern in the United States and worldwide.
 Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
-Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
-Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity.
-One approach to designing an effective intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
-The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
+Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
+Therefore, there is a need for informatics-based methods to facilitate the development of effective interventions for childhood obesity.
+One approach to designing an effective intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
+The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
 These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.
@@
 In this study, probabilistic models are employed for the analysis of sequence data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
-To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue.
-A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues.
+To the best of our knowledge, there have been very limited studies on sequence classification in a motivational interview such as patient-provider dialogue.
+A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exhibiting cues.
 Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns.
 Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
-In addition, a k-order Markov model is applied on genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.
+In addition, a k-order Markov model is applied to genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.
 Therefore, we utilize Markov chain and hidden Markov model for the classification of utterance sequences.
@@
 The experimental data sets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances.
-Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequences from adolescent session and excluded sequences from caregiver session in the MI transcript.
+Since clinical researchers tried to increase the desire and ability of an adolescent to change their current behavior to target behavior, we used sequences from an adolescent session and excluded sequences from caregiver session in the MI transcript.
 The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
@@
 A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences).
-Alternate sequences are the possible combination of normal sequence where two consecutive codes can not be come from same speaker.
+Alternate sequences are the possible combination of a normal sequence where two consecutive codes can not come from the same speaker.
 For example, there are 6 possible alternate sequences for the following sequences: (C1, A1, A2, C2, C3, C4, A3), where adolescent and counselor codes are represented by A and C, respectively.
 Therefore, (C1, A1, C2, A3), (C1, A1, C3, A3), (C1, A1, C4, A3), (C1, A2, C2, A3), (C1, A2, C3, A3) and (C1, A2, C4, A3) were used as alternate sequences for the above sequence.
-A total of 18753 alternate sequences have been observed which further partitioned into two subsets of 18139 successful and 614 unsuccessful sequences.
+A total of 18753 alternate sequences has been observed which further partitioned into two subsets of 18139 successful and 614 unsuccessful sequences.
 In our experiment, normal and alternate sequences are employed with two subsets (successful and unsuccessful) to train Markov chain and HMM.
@@
 Generally, a sequence is an ordered list of events.
-In our study, event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.
+In our study, an event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.
 Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$.
@@
 \begin{equation}
-p(S\rightarrow CT) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})
+p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})
 \end{equation}
 
-if $p(S\rightarrow CT) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in CT.
+if $p(S\rightarrow CML+) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in positive commitment.
 To fit the model, we experimentally determined the value of $\delta$ where model provides highest F-measure with lower bias.
@@
 However, F-score sharply decreased after $\delta = -0.9$.
-Thus, -0.9 will the optimal value of $\delta $, where F-measure is highest with lower bias.
+Thus, -0.9 will be the optimal value of $\delta $, where F-measure is highest with lower bias.
 
@@
 
-The Markov model we have discussed so far is referred to as the first order Markov model, since we only look one code in the past to compute the transition probability.
-A generalization of the first order Markov model is the $k^{th}$ order Markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols.
+The Markov model we have discussed so far is referred to as the first-order Markov model, since we only look one code in the past to compute the transition probability.
+A generalization of the first-order Markov model is the $k^{th}$ order Markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols.
 Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
 In our experiment, we also used second order Markov model.
-Hence, our model have $N^2$ states with transitional probability matrix of size $N^2 \times N$.
+Hence, our model has $N^2$ states with transitional probability matrix of size $N^2 \times N$.
 
@@
 It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces the time set of observations.
-A HMM can be fully specified by the following quintuple:
+An HMM can be fully specified by the following quintuple:
 
@@
 \subsection*{\textit{Evaluation methods}}
-The performance of classifiers are reported on the basis of 5-fold cross validation by using weighted macro-averaging over folds.
+The performance of classifiers is reported on the basis of 5-fold cross-validation by using weighted macro-averaging over folds.
 Precision, recall, F-measure and accuracy were used to evaluate the performance of classifiers.
 A true positive (TP) was counted when classifier correctly classified the sequence into actual class; a false positive (FP) was counted when classifier incorrectly classified other classes to the considered class; a false negative (FN) when the considered class incorrectly classified to other classes.
-Precision of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences identified as that particular class by the classifier.
+The precision of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences identified as that particular class by the classifier.
 In other words, Precision = TP / (TP + FP).
@@
 In more specifically, Recall = TP / (TP + FN).
-Similarly, Accuracy of a class was also defined as the ratio between the numbers of correctly classified sequences and total number of sequences for that particular class in the gold standard.
+Similarly, Accuracy of a class was also defined as the ratio between the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard.
 However, F-measure can be computed as the harmonic mean of precision and recall, or 2 x Precision x Recall / (Precision + Recall).
@@
 \section*{Results}
-Experimental evaluation of the classification of utterance sequences by using probabilistic models are reported with two set of experiments: first set of experiment was focused on normal sequences, while the second set of experiment was performed on alternate sequences.
+Experimental evaluation of the classification of utterance sequences by using probabilistic models is reported with two set of experiments: the first set of experiment was focused on normal sequences, while the second set of experiment was performed on alternate sequences.
 
@@
 It follows that first order hidden Markov model works very well and achieves highest precision 0.7980 with F-measure 0.7989.
-However, our designed second order Markov model shows lowest performance in terms of precision and F-measure.
+However, our designed second order Markov model shows the lowest performance in terms of precision and F-measure.
 It was also noticed that the lowest recall 0.7686 achieved in first order Markov model.
@@
 Overall, HMM performed better than general Markov model in normal utterance sequences.
+\\
 
@@
 Second order general Markov model shows the highest precision 0.9778 as well as highest F-measure 0.9736.
-On the other hand, first order hidden Markov model displays lowest performance in all metrics compared to all other models.
-It was also examined that every models perform better in second order compared to first order.
+On the other hand, first order hidden Markov model displays the lowest performance in all metrics compared to all other models.
+It was also examined that every model perform better in second order compared to first order.
 General Markov chain achieves 0.01\%, 2.72\% and 1.37\% improvement for precision, recall and F-measure, respectively, from first order to second order.
@@
 \subsection*{\textit{Some common patterns}}
-Table~\ref{tab:common_patterns} shows some successful and unsuccessful patterns which appeared more frequently in motivational interview.
-It can be seen that most successful patterns start with summary of the discussion or open-ended or closed ended question.
-After that, if adolescent demonstrate positive behavior change, counselor reflect that positive attitude immediately so that adolescent can stick with the decision and show his own desire for positive behavior change.
-On the other hand, general information of counselor can lead to a negative behavior change of adolescent, while showing positive attitude in the previous step.
-Because adolescent changes their topic more frequently when general question was asked.
+Table~\ref{tab:common_patterns} shows some successful and unsuccessful patterns which appeared more frequently in a motivational interview.
+It can be seen that most successful patterns start with a summary of the discussion or open-ended or closed-ended question.
+After that, if adolescent demonstrates positive behavior change, counselor reflect that positive attitude immediately so that adolescent can stick with the decision and show his own desire for positive behavior change.
+On the other hand, general information of counselor can lead to a negative behavior change of adolescent, while showing a positive attitude in the previous step.
+Because adolescent changes their topic more frequently when general questions were asked.
 
@@
 However, this cannot be easily done for many applications areas include our case.
-To skip these problems we restricted our study upto second order.
-Shorter average length of utterance sequence is another reason for keeping the order lower.
+To skip these problems we restricted our study up to second order.
+The shorter average length of utterance sequence is another reason for keeping the order lower.
 
-Second, overall performance of hidden Markov model is better than general Markov model in both type of sequences.
+Second, the overall accuracy of hidden Markov model is better than general Markov model in both types of sequences.
 We observed that hidden Markov model achieved near-human accuracy to categories the sequences of behavior codes.
-This means that hidden states capture the information passes through the codes of the utterance sequence for determining final outcome.
+This means that hidden states capture the information passes through the codes of the utterance sequence for determining the final outcome.
 
 Third, alternate sequences provide higher performance for all Markov models.
-Even, it provides highest precision and F-measure for general Markov model.
+Even it provides the highest precision and F-measure for general Markov model.
 This indicates that each pair of alternate behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview.
-And this is obvious because if we know the previous two behavior codes (e.g.
-provider's question and adolescent's answer) then it is easier to know the expected answer for the current question.
+And this is obvious because if we know the previous two behavior codes, for instance, provider's question and adolescent's response then it is easier to know the expected response for the current question of the counselor.
 
-Finally, it was seen that successful interviews are more frequently responded by the adolescent with positive behavior changes which lead to make a positive commitment for loosing weight.
+Finally, it was seen that successful interviews are more frequently responded by the adolescent with positive behavior changes which lead to making a positive commitment for loosing weight.
 Whereas, unsuccessful interviews are responded with negative behavior changes which lead to a negative commitment or no commitment.
@@
 It has been shown how Markov chain and hidden Markov models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
-Experimental results shows that the behavior codes carry some inherited information to distinguish motivational interviews.
-This work can facilitate researchers to establish causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
+Experimental results show that the behavior codes carry some inherited information to distinguish motivational interviews.
+This work can facilitate researchers to establish a causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
 It provides information that can directly inform and increase the efficiency of clinical practice for a successful interview.
-This work also suggest some successful patterns for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
-Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
-We also have plan to use large data set in conjunction with some advanced features for the evaluation of sequence classifiers.
+This work also explains some successful patterns for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
+The direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
+We also have the plan to use large data set in conjunction with some advanced features for the evaluation of sequence classifiers.
 

### 2017-03-07 paper/AMIA/main.tex
first revision by Alex

@@
 
-\title{Sequence classifications to facilitate designing an effective intervention of motivational interview}
+\title{Predicting Success of Clinical Interviews via Sequential Analysis of Patient-Provider Interactions}
 
-\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Sylvie Naar-King, PhD$^{2}$}
+\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Mind Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
 
@@
 \noindent{\bf Abstract}
-\textit{Childhood obesity is a serious public health concern in the United States and worldwide.
-Therefore, there is a need for identification of successful motivational interviews to facilitate the development of effective interventions for childhood obesity.
-This paper presents our work on the classification of utterance sequences into successful and unsuccessful behavior outcomes.
-Markov chain and hidden Markov models were employed and compared with a different order for the classification of utterance sequences.
-The classification performances achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using 1st order Markov chain and second order hidden Markov model, respectively.
-This shows that the behavior codes carry some inherited information to distinguish motivational interviews.
-We expect that the result will improve if we use a more discriminable data set.
-Furthermore, the results suggest us a new way to think about designing the motivational interview.}
+\textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics.
+In this paper, we focus on the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication sequences for triggering a particular type of patient response and dynamically predicting this type of response based on observed interactions.
+Such analysis is necessary to facilitate the development of effective interventions to address many public health conditions, such as childhood obesity.
+Our proposed probabilistic method for predicting a patient behavioral response in communication sequences can be utilized with both Markov chains or Hidden Markov Models.
+It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively.
+Our results indicate that individual short length patient-provider communication sequences are highly indicative of the progression and trajectory of clinical interviews and can be used to predict the overall success of motivational interviews.
+Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of more effective behavioral interventions.}
 
 \section*{Introduction}
-In the recent times, more and more data from various domains being produced in the form of event sequences.
-Thus, sequence classification has become an important task in sequence mining.
-Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types.
-Usually, sequences don't have explicit features and suffer from the high dimension of feature space.
-Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications.
-The sequence classification methods can be divided into three large categories: feature-based classification, distance based method, and model based classifications.
-In feature-based classification, a sequence is transformed into a single vector of features.
-A supervised learning algorithm such as a support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier.
-An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications.
-On the other hand, distance based method measure the similarity between sequences to determine the quality of the classification significance.
-The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data.
-The third category of sequence classification is model based classifiers such as hidden Markov model \cite{rabiner1989tutorial} (HMM) or other statistical models.
-Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.
+Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics.
+The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification.
+In general, sequence classification methods can be divided into three  categories: feature-based, distance metric-based and model-based method.
+Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
+Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
+Distance-based methods measure the similarity between sequences to determine the quality of classification.
+The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
+The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
 
-Sequence classification has a broad range of real word applications such as information retrieval, genomic analysis, health informatics, finance and abnormal detection.
-In genomic research, sequence classification is widely used to detect the function of a new protein \cite{deshpande2002evaluation}.
-In health informatics, ECG can be used as a multi-dimensional time series data to classify an individual as healthy or patient with heart disease \cite{wei2006semi}.
-Another example is to classify rumor stance in Twitter \cite{lukasik2016hawkes}.
-In information retrieval, sequence classification is employed to build models for classifying documents into different topic categories \cite{sebastiani2002machine} such as sports, politics, news, finance and style.
-Sequence classification is also used in anomaly detection such as abnormal access to systems \cite{lane1999temporal}, malware detection \cite{drew2017polymorphic} and money laundering \cite{liu2008sequence}.
-Other interesting applications include classifying query log sequences to distinguish web-robots from human users \cite{duskin2009distinguishing, tan2004discovery}.
-In computer vision, sequence classification is applied for classifying images \cite{li2000image} and videos \cite{zhang2016tensor}.
+Sequence classification has a broad range of real word applications from genomics and health informatics to finance and abnormality detection.
+In genomics research, sequence classification is widely used to detect the function of new proteins \cite{deshpande2002evaluation}.
+In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}.
+Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware detection \cite{drew2017polymorphic}.
 
-In this paper, our target area of application is the motivational interview.
-In particular, this paper describes our work on the classification of utterance sequences.
-Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many conditions and disorders.
+In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews.
+In particular, we propose a method to identify the most effective communication sequences for eliciting a specific type of patient response and .
+Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many behavioral conditions and disorders.
 In this study, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents (teens) and their caregivers.

### 2017-03-08 paper/AMIA/main.tex
Intro updated by Alex

@@
 
-\title{Predicting Success of Clinical Interviews via Sequential Analysis of Patient-Provider Interactions}
+\title{Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences}
 
@@
 \textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics.
-In this paper, we focus on the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication sequences for triggering a particular type of patient response and dynamically predicting this type of response based on observed interactions.
-Such analysis is necessary to facilitate the development of effective interventions to address many public health conditions, such as childhood obesity.
-Our proposed probabilistic method for predicting a patient behavioral response in communication sequences can be utilized with both Markov chains or Hidden Markov Models.
-It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively.
-Our results indicate that individual short length patient-provider communication sequences are highly indicative of the progression and trajectory of clinical interviews and can be used to predict the overall success of motivational interviews.
-Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of more effective behavioral interventions.}
+In this paper, we focus on automating the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication exchanges to elicit a particular type of patient response and dynamically predicting this type of response in an observed patient-provider communication.
+Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity.
+It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively,  to model communication sequences.
+Our results indicate that individual short length patient-provider communication sequences are highly indicative of the overall progression and future trajectory of a clinical interview and can be used to predict its overall success.
+Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of effective behavioral interventions.}
 
@@
 
-Sequence classification has a broad range of real word applications from genomics and health informatics to finance and abnormality detection.
-In genomics research, sequence classification is widely used to detect the function of new proteins \cite{deshpande2002evaluation}.
+Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection.
+In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}.
 In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}.
-Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware detection \cite{drew2017polymorphic}.
+Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.
 
 In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews.
-In particular, we propose a method to identify the most effective communication sequences for eliciting a specific type of patient response and .
-Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment versus negative change talk or commitment is an important part of clinical research aimed at designing effective interventions for many behavioral conditions and disorders.
-In this study, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents (teens) and their caregivers.
+Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
 Childhood obesity is a serious public health concern in the United States and worldwide.
@@
 Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
-Therefore, there is a need for informatics-based methods to facilitate the development of effective interventions for childhood obesity.
-One approach to designing an effective intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
+One approach to designing an effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
 The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
 These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.
+However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
+Since manual annotation and examination of transcripts is a very time-consuming process, development of new obesity interventions can take years.
+Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions.
 
-Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes.
-In this study, probabilistic models are employed for the analysis of sequence data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.
-To the best of our knowledge, there have been very limited studies on sequence classification in a motivational interview such as patient-provider dialogue.
-A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exhibiting cues.
-Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns.
-Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
-In addition, a k-order Markov model is applied to genomic research \cite{yakhnenko2005discriminatively} to classify protein and text sequence data.
-Therefore, we utilize Markov chain and hidden Markov model for the classification of utterance sequences.
+While our previous work \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing 115 patient and provider behavior codes \cite{carcone2013provider}, in the work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
+change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
+
+While sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to establish the relationship between physician and patient behaviors associated with patients' exhibiting cues, there have been only a handful of previous studies on modeling annotated sequences of patient-provider communication exchanges and predicting a desired patient behavior in the context of clinical interviews.
 
@@
 \subsection*{\textit{Data collection}}
-The experimental data sets for this work were constructed based on the transcripts of 37 motivational interview sessions, which include a total of 11,353 segmented and annotated utterances.
-Since clinical researchers tried to increase the desire and ability of an adolescent to change their current behavior to target behavior, we used sequences from an adolescent session and excluded sequences from caregiver session in the MI transcript.
-The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 different behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
-These utterances are then scanned from first to last and divided into successful and unsuccessful sequences when a commitment or change talk is found.
+The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances.
+Each motivational interview consists of counselor-adolescent and counselor-caregiver sessions.
+Since the ultimate goal of a motivational interview is to increase the desire and ability of an adolescent for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions.
+The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 distinct behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
+These utterances are then scanned from first to last and divided into successful and unsuccessful communication sequences.
+A successful communication sequence results in positive change talk or commitment languages, while an unsuccessful sequence results in or change talk is found.
 A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences).
@@
 \subsection*{\textit{Sequence classification techniques}}
+
+ Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
+model patient-provider interactions using first- and second-order Markov Chains and Hidden Markov Models (HMM)
+popular probabilistic methods for modeling discrete observation sequences with finite vocabulary,
 Generally, a sequence is an ordered list of events.

### 2017-03-08 paper/AMIA/main.tex
Methods updated by Alex

@@
 In this paper, we focus on automating the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication exchanges to elicit a particular type of patient response and dynamically predicting this type of response in an observed patient-provider communication.
-Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity.
-It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov chain and Hidden Markov Model, respectively,  to model communication sequences.
+Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov Chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity.
+It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov Chain and Hidden Markov Model, respectively,  to model communication sequences.
 Our results indicate that individual short length patient-provider communication sequences are highly indicative of the overall progression and future trajectory of a clinical interview and can be used to predict its overall success.
@@
 Each motivational interview consists of counselor-adolescent and counselor-caregiver sessions.
-Since the ultimate goal of a motivational interview is to increase the desire and ability of an adolescent for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions.
+Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions.
 The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 distinct behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
-These utterances are then scanned from first to last and divided into successful and unsuccessful communication sequences.
-A successful communication sequence results in positive change talk or commitment languages, while an unsuccessful sequence results in or change talk is found.
-A total of 1072 sequences have been observed which further partitioned into two subsets based on the outcome of motivational interviews: one data set that includes all sequences of positive change talk or commitment language (880 sequences) and the other data set which includes utterances ended with negative change talk or commitment language (192 sequences).
-Alternate sequences are the possible combination of a normal sequence where two consecutive codes can not come from the same speaker.
-For example, there are 6 possible alternate sequences for the following sequences: (C1, A1, A2, C2, C3, C4, A3), where adolescent and counselor codes are represented by A and C, respectively.
-Therefore, (C1, A1, C2, A3), (C1, A1, C3, A3), (C1, A1, C4, A3), (C1, A2, C2, A3), (C1, A2, C3, A3) and (C1, A2, C4, A3) were used as alternate sequences for the above sequence.
-A total of 18753 alternate sequences has been observed which further partitioned into two subsets of 18139 successful and 614 unsuccessful sequences.
-In our experiment, normal and alternate sequences are employed with two subsets (successful and unsuccessful) to train Markov chain and HMM.
-Dataset statistics and a fragment of an adolescent session transcript are presented in Table~\ref{tab:data_dist} and Table~\ref{tab:anno_examp}.
+Utterances were divided into successful and unsuccessful communication sequences.
+Successful communication sequences result in either positive change talk or commitment language statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk, negative commitment language or the ones in which neither change talk nor commitment language statements occur.
+Out of 1072 observed sequences, 880 were positive and 192 were negative.
+Alternate sequences are the possible combination of a normal sequence where two consecutive utterances cannot belong to the same speaker.
+For example, there are 6 possible alternate sequences ({C1, A1, C2, A3}, {C1, A1, C3, A3}, {C1, A1, C4, A3}, {C1, A2, C2, A3}, {C1, A2, C3, A3} and {C1, A2, C4, A3}) for the normal sequence {C1, A1, A2, C2, C3, C4, A3}, where adolescent and counselor codes are represented by A and C, respectively.
+A total of 18753 alternate sequences have been observed, out of which 18139 were successful and 614 were unsuccessful ones.
+For each type of sequence (normal and alternate), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences.
+Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
 \\
@@
 \centering
-\caption{General statistics of dataset.}
+\caption{Statistics of experimental dataset.}
 \label{tab:data_dist}
@@
 
-Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor start with an open-ended question and get positive feedback at the end. \\
+Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\
 
@@
 
- Although hidden Markov model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk} for sequence classification.
-model patient-provider interactions using first- and second-order Markov Chains and Hidden Markov Models (HMM)
-popular probabilistic methods for modeling discrete observation sequences with finite vocabulary,
-Generally, a sequence is an ordered list of events.
-In our study, an event is a behavior code represented as a symbolic value such as 117 (low uptake, positive), 331 (open-ended question), etc.
-Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$.
-Here, L is fixed and known in advances.
-In our experiment, known labels are ``successful'' and ``unsuccessful'' motivational interview.
+Generally, a sequence is a temporally ordered set of events.
+In our study, an event is a behavior code represented as a symbolic value, such as 117 (low uptake, positive), 331 (open-ended question), etc.
+Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$, where $L$ is fixed and known in advance.
+In our case, the labels are ``successful'' and ``unsuccessful'' (motivational interview).
+
+We model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are probabilistic methods for modeling discrete observation sequences with finite vocabulary.
+Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.
 
@@
 \begin{itemize}
-\item $V = \{CML+, CHT+, CHT-, T-AMB, CCT, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of possible youth and counselor MI behavior codes;
+\item $V = \{CML+, CHT+, CHT-, T-AMB, CCT, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
 \item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
@@
 
-where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ have been observed in the training data.
-Given a Markov model $M$ (such that $S\subseteq V$), a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ have been generated from a Markov model $M$ is:
+where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data.
+Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
 
@@
 
-Success of a given motivational interview, given a sequence of MI behavior codes corresponding to it, can be predicted based on the following formula:
+The likelihood of success of a given motivational interview at a given time point, given a sequence of MI behavior codes observed prior to that point, is estimated based on the following formula:
 
@@
 
-if $p(S\rightarrow CML+) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in positive commitment.
-To fit the model, we experimentally determined the value of $\delta$ where model provides highest F-measure with lower bias.
-Figure~\ref{fig:delta} illustrates an example of the characteristics of Markov model by varying the value of $\delta$.
-From the figure, it was observed that F-score value is not changing significantly with respect to $\delta $ from -5 to -1.
-This is because a bias is involved.
-However, F-score sharply decreased after $\delta = -0.9$.
-Thus, -0.9 will be the optimal value of $\delta $, where F-measure is highest with lower bias.
+if $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language.
+Estimating the model on the training data involves experimentally determining the value of $\delta$, resulting in the highest prediction F-measure with lower bias.
+Figure~\ref{fig:delta} illustrates the accuracy of Markov model-based method by varying the value of $\delta$.
+As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias.
+However, F-score sharply decreases after $\delta = -0.9$.
+Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.
 
@@
     \includegraphics[width=0.70\textwidth]{figures/delta.png}
-    \caption{\textbf{Characteristics of proposed Markov model by varying the value of $\delta$}}
+    \caption{\textbf{Accuracy of the proposed Markov model-based method by varying the value of $\delta$}.}
     \label{fig:delta}
@@
 
-The Markov model we have discussed so far is referred to as the first-order Markov model, since we only look one code in the past to compute the transition probability.
-A generalization of the first-order Markov model is the $k^{th}$ order Markov chain in which the transition probability for a particular symbol $x_i$ is computed by looking at the k preceding symbols.
+The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code to compute the state transition probability matrix.
+A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular symbol $x_i$ is computed by looking at the k preceding symbols.
 Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
@@
 
-\textbf {Hidden Markov Model (HMM)}: hidden Markov models are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-Hidden Markov model is a very
+\textbf {Hidden Markov Model (HMM)}: HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Hidden Markov model is a very
 powerful tool to statistically model a process that varies in time.
-It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces the time set of observations.
+It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations.
 An HMM can be fully specified by the following quintuple:
@@
 \begin{itemize}
-\item N is the number of states for the model
-\item M is the number of distinct observations symbols per state, i.e. the discrete alphabet size
+\item N is the number of hidden states in the model
+\item M is the number of distinct observations symbols per state, i.e. the discrete vocabulary size
 \item A is the $N\times N$ state transition probability distribution given in the form of a matrix $A = \{a_{ij}\}$

### 2017-03-08 paper/AMIA/main.tex
up to Table 4

@@
 \begin{equation}
-p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})
+p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})\label{eq:class}
 \end{equation}
@@
 The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code to compute the state transition probability matrix.
-A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular symbol $x_i$ is computed by looking at the k preceding symbols.
+A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular state (symbol) $x_i$ is computed by looking at the k preceding states (symbols).
 Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
 In our experiment, we also used second order Markov model.
-Hence, our model has $N^2$ states with transitional probability matrix of size $N^2 \times N$.
+Hence, our model has $N^2$ states with transition probability matrix of size $N^2 \times N$.
 
-\textbf {Hidden Markov Model (HMM)}: HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Hidden Markov model is a very
-powerful tool to statistically model a process that varies in time.
-It can be seen as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations.
+\textbf {Hidden Markov Model (HMM)} is a powerful tool for statistical modeling of processes varying in time.
+HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
+HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations.
 An HMM can be fully specified by the following quintuple:
@@
 \begin{itemize}
-\item N is the number of hidden states in the model
-\item M is the number of distinct observations symbols per state, i.e. the discrete vocabulary size
-\item A is the $N\times N$ state transition probability distribution given in the form of a matrix $A = \{a_{ij}\}$
-\item B is the $N\times M$ observation symbol probability distribution given in the form of a matrix $B = \{b_j(k)\}$
+\item N is a number of hidden states in the model
+\item M is a number of distinct observations symbols per state, i.e. the discrete vocabulary size
+\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
+\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
 \item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
@@
 
-If we opt out the structure parameters M and N we have the more often used compact notation:
+We exclude the structure parameters M and N to designate HMM using a compact notation:
 
@@
 
-This model is also called hidden for a reason.
-We are not required to tell it what those aforementioned states should be.
-We just have to tell it how many states we think are needed to model what we want.
-The model should be able to figure a suitable sequence of states to support our observations.
-The Baum-Welch algorithm was implemented to train a hidden Markov model for estimating parameters of each models using the training set.
-To identify the label of an unknown code sequence, the Viterbi algorithm was used to decode the sequence and compute its log probabilities respectively using HMM's trained for different classes.
-Then, the class with the highest log probability was assigned to the given behavior sequences.
-Like Markov chain, we also employed second order hidden Markov model for classification of utterance sequences.
+The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model figures out a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
+The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations.
+After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.
+As in the case of MC, we also experimented with second-order HMM.
 
@@
 \section*{Results}
-Experimental evaluation of the classification of utterance sequences by using probabilistic models is reported with two set of experiments: the first set of experiment was focused on normal sequences, while the second set of experiment was performed on alternate sequences.
+Experimental evaluation of the classification of utterance sequences by using the proposed probabilistic models is reported with two set of experiments: the first set of experiment was focused on normal sequences, while the second set of experiment was performed on alternate sequences.
 
-\subsection*{\textit{Performance of models for the classification of normal PPC code sequences}}
+\subsection*{\textit{Predictive performance for normal PPC code sequences.}}
 Performance summary for the classification of normal sequences is presented in Table~\ref{tab:result_norm_seq}.
-It follows that first order hidden Markov model works very well and achieves highest precision 0.7980 with F-measure 0.7989.
-However, our designed second order Markov model shows the lowest performance in terms of precision and F-measure.
-It was also noticed that the lowest recall 0.7686 achieved in first order Markov model.
-On the other hand, second order hidden Markov model achieves the highest recall 0.8449 and accuracy 0.8449.
-From Table~\ref{tab:result_norm_seq}, we observed that performance degrades from first order to second order.
-Precision and F-measures drop 6.74\% \& 2.39\% in Markov chain and 7.27\% \& 2.09\% in HMM, respectively, when using second order models.
-In contrast, recall improves 2.64\%--4.84\% in second order Markov model and HMM.
-Overall, HMM performed better than general Markov model in normal utterance sequences.
+It follows that first-order HMM demonstrates the best performance achieving the highest precision 0.7980 with 0.7989 F-measure.
+However, second-order Markov Chain demonstrates the lowest performance in terms of precision and F-measure.
+It can be seen that the lowest recall of 0.7686 is achieved by the first-order MC.
+On the other hand, second-order HMM achieves the highest recall 0.8449 and accuracy 0.8449.
+From Table~\ref{tab:result_norm_seq}, it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM.
+Precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
+In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models.
+Overall, HMM performed better than MC for normal sequences.
 \\
@@
 \centering
-\caption{Performance of Markov models for the classification of normal PPC code sequence. Highest performances for each metrics are highlighted in bold faces.}
+\caption{Performance of MC and HMM for predicting success of normal patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
 \label{tab:result_norm_seq}
@@
 
-\subsection*{\textit{Performance of models for the classification of alternate PPC code sequences}}
+\subsection*{\textit{Predictive performance for alternate PPC code sequences}}
 Table~\ref{tab:result_alt_seq} shows the classification performance of models applied on alternate sequences.
@@
 \centering
-\caption{Performance of Markov models for the classification of alternate PPC code sequence. Highest performances for each metrics are highlighted in bold faces.}
+\caption{Performance of MC and HMM for predicting success of normal patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
 \label{tab:result_alt_seq}
@@
 
-\subsection*{\textit{Some common patterns}}
+\subsection*{\textit{Common patterns}}
 Table~\ref{tab:common_patterns} shows some successful and unsuccessful patterns which appeared more frequently in a motivational interview.
@@
 \centering
-\caption{Some common patterns in motivational interview for successful and unsuccessful behavior outcome.}
+\caption{Common patterns in successful and unsuccessful motivational interviews.}
 \label{tab:common_patterns}
@@
 \section*{Conclusion}
-It has been shown how Markov chain and hidden Markov models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
+In this paper, we proposed Markov chain and hidden Markov models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
 Experimental results show that the behavior codes carry some inherited information to distinguish motivational interviews.

### 2017-03-08 paper/AMIA/main.tex
updated by Alex until discussion

@@
 Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov Chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity.
-It achieved 75.32\% and 79.89\% F-measure for normal sequences and 97.36\% and 96.99\% F-measure for alternate sequences using first and second order Markov Chain and Hidden Markov Model, respectively,  to model communication sequences.
+It achieved 75.32\% and 79.89\% F-measure for natural sequences and 97.36\% and 96.99\% F-measure for alternating sequences using first and second order Markov Chain and Hidden Markov Model, respectively,  to model communication sequences.
 Our results indicate that individual short length patient-provider communication sequences are highly indicative of the overall progression and future trajectory of a clinical interview and can be used to predict its overall success.
@@
 Out of 1072 observed sequences, 880 were positive and 192 were negative.
-Alternate sequences are the possible combination of a normal sequence where two consecutive utterances cannot belong to the same speaker.
-For example, there are 6 possible alternate sequences ({C1, A1, C2, A3}, {C1, A1, C3, A3}, {C1, A1, C4, A3}, {C1, A2, C2, A3}, {C1, A2, C3, A3} and {C1, A2, C4, A3}) for the normal sequence {C1, A1, A2, C2, C3, C4, A3}, where adolescent and counselor codes are represented by A and C, respectively.
-A total of 18753 alternate sequences have been observed, out of which 18139 were successful and 614 were unsuccessful ones.
-For each type of sequence (normal and alternate), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences.
+alternating sequences are the possible combination of a natural sequence where two consecutive utterances cannot belong to the same speaker.
+For example, there are 6 possible alternating sequences ({C1, A1, C2, A3}, {C1, A1, C3, A3}, {C1, A1, C4, A3}, {C1, A2, C2, A3}, {C1, A2, C3, A3} and {C1, A2, C4, A3}) for the natural sequence {C1, A1, A2, C2, C3, C4, A3}, where adolescent and counselor codes are represented by A and C, respectively.
+A total of 18753 alternating sequences have been observed, out of which 18139 were successful and 614 were unsuccessful ones.
+For each type of sequence (natural and alternating), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences.
 Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
@@
    \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
- Normal sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
-Alternate sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
+ natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
+alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
   \end{tabular}
@@
 \section*{Results}
-Experimental evaluation of the classification of utterance sequences by using the proposed probabilistic models is reported with two set of experiments: the first set of experiment was focused on normal sequences, while the second set of experiment was performed on alternate sequences.
+Experimental evaluation of the classification of utterance sequences by using the proposed probabilistic models is reported with two set of experiments: the first set of experiment was focused on natural sequences, while the second set of experiment was performed on alternating sequences.
 
-\subsection*{\textit{Predictive performance for normal PPC code sequences.}}
-Performance summary for the classification of normal sequences is presented in Table~\ref{tab:result_norm_seq}.
+\subsection*{\textit{Predictive performance for natural PPC code sequences}}
+Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}.
 It follows that first-order HMM demonstrates the best performance achieving the highest precision 0.7980 with 0.7989 F-measure.
@@
 In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models.
-Overall, HMM performed better than MC for normal sequences.
+Overall, HMM performed better than MC for natural sequences.
 \\
@@
 \centering
-\caption{Performance of MC and HMM for predicting success of normal patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
+\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
 \label{tab:result_norm_seq}
@@
 
- \multirow{2}{*}{General Markov chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
+ \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
  & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
@@
 
-\subsection*{\textit{Predictive performance for alternate PPC code sequences}}
-Table~\ref{tab:result_alt_seq} shows the classification performance of models applied on alternate sequences.
-The results demonstrate that second order hidden Markov model performed better than Markov chain in terms of recall and achieved 97.13\% recall.
-Second order general Markov model shows the highest precision 0.9778 as well as highest F-measure 0.9736.
-On the other hand, first order hidden Markov model displays the lowest performance in all metrics compared to all other models.
-It was also examined that every model perform better in second order compared to first order.
-General Markov chain achieves 0.01\%, 2.72\% and 1.37\% improvement for precision, recall and F-measure, respectively, from first order to second order.
-Similarly, hidden Markov model achieves 3.32\%, 4.30\% and 4.82\% improvement for precision, recall and F-measure, respectively, from first order to second order.
-Overall, second order Markov model performs better for alternate utterance sequences.
-\\
+\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
+Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences.
+These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
+Similar to the case of natural sequences, first-order MC has the lowest performance compared to all other models across all metrics.
+Also, as in the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models.
+However, different from natural sequences, second-order MC demonstrates the highest precision 0.9778 as well as the highest F-measure of 0.9736.
+We also observed that second-order MC and HMM outperform first-order MC and HMM.
+Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
+Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 4.82\%, respectively.
 
@@
 \centering
-\caption{Performance of MC and HMM for predicting success of normal patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
+\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
 \label{tab:result_alt_seq}
@@
 
- \multirow{2}{*}{General Markov chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
+ \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
  & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
@@
 \subsection*{\textit{Common patterns}}
-Table~\ref{tab:common_patterns} shows some successful and unsuccessful patterns which appeared more frequently in a motivational interview.
-It can be seen that most successful patterns start with a summary of the discussion or open-ended or closed-ended question.
-After that, if adolescent demonstrates positive behavior change, counselor reflect that positive attitude immediately so that adolescent can stick with the decision and show his own desire for positive behavior change.
-On the other hand, general information of counselor can lead to a negative behavior change of adolescent, while showing a positive attitude in the previous step.
-Because adolescent changes their topic more frequently when general questions were asked.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question.
+After that, if adolescents show positive behavior change, the counselor immediately reflects on that positive tendency, to reinforce adolescent's desire to adhere to positive behavior change.
+On the other hand, general information by the counselor can lead to a negative behavior change in adolescents, even in the cases when adolescents were showing positive tendency in their previous communication.
+This observation can be explained by adolescents quickly loosing focus when presented with general information.
 
@@
 
-Third, alternate sequences provide higher performance for all Markov models.
+Third, alternating sequences provide higher performance for all Markov models.
 Even it provides the highest precision and F-measure for general Markov model.
-This indicates that each pair of alternate behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview.
+This indicates that each pair of alternating behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview.
 And this is obvious because if we know the previous two behavior codes, for instance, provider's question and adolescent's response then it is easier to know the expected response for the current question of the counselor.
@@
 \section*{Conclusion}
-In this paper, we proposed Markov chain and hidden Markov models could be used to build classifiers for the sequence of behavior codes in motivational interviews.
+In this paper, we proposed a method for based on Markov Chain and Hidden Markov Model could be used to build classifiers for the sequence of behavior codes in motivational interviews.
 Experimental results show that the behavior codes carry some inherited information to distinguish motivational interviews.

### 2017-03-09 paper/AMIA/main.tex
Added hidden state figure and table for emission matrix

@@
 Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
-Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 4.82\%, respectively.
+Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 4.82\%, respectively.\\
 
@@
  & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
+  \end{tabular}
+\end{table}
+
+\subsection*{\textit{Impact of hidden states}}
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+
+\begin{table}[h]
+\centering
+\caption{Emission matrix}
+\label{tab:emission_matrix}
+  \begin{tabular}{|l|l|}
+  \hline
+   \textbf{Hidden} & \textbf{Log probability of each symbol (301, 117, 327, 343, 101, 305, 311, 331, 310, 328, 112,)} \\ 
+   \textbf{states} & \textbf{ 326, 302, 308, 314, 306, 119, 313, 333, 307, 325, 344, 109, 106, 332, 120, 345, 309, 312, 114,} \\
+    & \textbf{ 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
+S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
+& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
+& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
+S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+S-3 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+S-4 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+
+U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
+& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
+& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
+U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, \\
+& -2.415, -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, \\
+& -8, -8, -3.794, -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
+U-3 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, \\
+& -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, \\
+& -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
+U-4 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, \\
+& -2.415, -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, \\
+& -8, -8, -3.794, -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
 
@@
 \end{table}
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
+    \caption{\textbf{Impact of hidden states in the performance of HMM}.}
+    \label{fig:hidden-states}
+\end{figure}
 
@@
 This observation can be explained by adolescents quickly loosing focus when presented with general information.
+\\
 
@@
 \label{tab:common_patterns}
-  \begin{tabular}{|l|l|l|l|l|l|}
+  \begin{tabular}{|l|l|}
   \hline

### 2017-03-09 paper/AMIA/main.tex
added subsection impact of hidden states

@@
 Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
-Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 4.82\%, respectively.\\
+Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\
 
@@
  & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
- \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9253 & 0.9392\\ \cline{2-6}
+ \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353 & 0.9392\\ \cline{2-6}
  & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
@@
 \subsection*{\textit{Impact of hidden states}}
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+Since the number of hidden states may impact the classification
+performance, we use an optimum number of hidden states in HMM's.
+Figure~\ref{fig:hidden-states} shows the performance of HMMs with a different number of hidden states.
+The results illustrate that, in general, the state number didn't impact the classification performance significantly.
+All four models show their best performance when state number in between 4-8.
+Table~\ref{tab:emission_matrix} shows an example of the distribution of PPC codes over the optimum number of hidden states (4) for successful (S) and unsuccessful (U) HMMs.
+Log probability of each code reflects the state significance.
+In other words, these probabilities indicate that which hidden states are close to adolescent commitment language and which are not?
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
+    \caption{\textbf{Performance of HMM's with different sequence type, order, and number of hidden states used in the experiment}.}
+    \label{fig:hidden-states}
+\end{figure}
 
@@
 \centering
-\caption{Emission matrix}
+\caption{Distribution of PPC codes over the optimal hidden states for successful (S) and unsuccessful (U) HMMs.}
 \label{tab:emission_matrix}
@@
   \hline
-   \textbf{Hidden} & \textbf{Log probability of each symbol (301, 117, 327, 343, 101, 305, 311, 331, 310, 328, 112,)} \\ 
+   \textbf{Hidden} & \textbf{Log probability of each symbol (301, 117, 327, 343, 101, 305, 311, 331, 310, 328, 112,} \\ 
    \textbf{states} & \textbf{ 326, 302, 308, 314, 306, 119, 313, 333, 307, 325, 344, 109, 106, 332, 120, 345, 309, 312, 114,} \\
@@
 \end{table}
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
-    \caption{\textbf{Impact of hidden states in the performance of HMM}.}
-    \label{fig:hidden-states}
-\end{figure}
 

### 2017-03-09 paper/AMIA/main.tex
update table 5

--- 2017-03-08 paper/AMIA/main.tex
@@
 \label{tab:data_dist}
-  \begin{tabular}{|l|l|l|l|l|l|}
+  \begin{tabular}{|l|l|l|l|l|}
   \hline
@@
 \begin{equation}
-p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})\label{eq:class}
+p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
 \end{equation}
@@
 HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations.
-An HMM can be fully specified by the following quintuple:
-
-\begin{center}
-$\lambda = (N, M, A, B, \pi)$
-\end{center}
+An HMM can be fully specified by the following quintuple: $\lambda = (N, M, A, B, \pi)$
 
@@
 
-We exclude the structure parameters M and N to designate HMM using a compact notation:
-
-\begin{center}
-$\lambda = (A, B, \pi)$
-\end{center}
-
+We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$.
 The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model figures out a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
@@
 Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
-Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 4.82\%, respectively.
+Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\
 
@@
  & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
- \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9253 & 0.9392\\ \cline{2-6}
+ \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353 & 0.9392\\ \cline{2-6}
  & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
-
   \end{tabular}
 \end{table}
+
+\subsection*{\textit{Impact of hidden states}}
+Since the number of hidden states may impact the classification
+performance, we use an optimum number of hidden states in HMMs.
+Figure~\ref{fig:hidden-states} shows the performance of HMMs with a different number of hidden states.
+The results illustrate that, in general, the state number didn't impact the classification performance significantly.
+All four models show their best performance when state number in between 4-8.
+Table~\ref{tab:emission_matrix} shows an example of the distribution of PPC codes over the states for successful (S) and unsuccessful (U) HMMs.
+Log probability of each code reflects the state significance.
+In other words, these probabilities indicate that which PPC codes are close to adolescent commitment language and which are not?
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
+    \caption{\textbf{Performance of HMMs with different sequence type, order, and number of hidden states used in the experiment}.}
+    \label{fig:hidden-states}
+\end{figure}
+
+\begin{table}[h]
+\centering
+\caption{Distribution of PPC codes over the states for successful (S) and unsuccessful (U) HMMs. Significant PPC codes are given in short form, which have log probability higher than -2.}
+\label{tab:emission_matrix}
+  \begin{tabular}{|l|l|}
+  \hline
+   \textbf{State} & \textbf{Log probability of each symbol (301,LUP+, 327,OQ-TBN,101,EA,R-CT+,OQ-ECT+,AR,SUM-S,} \\ 
+   \textbf{} & \textbf{R-CT-S,RBA-S,G-INFO+,AF,314,CQ-EF,119,R-CML+, OQ-ECML+, 307, 325, 344,CHT-,106,} \\
+    & \textbf{OQ-ECT-, HUP-O, 345, 309, R-CT-S, 114, 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
+S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
+& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
+& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
+S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+
+U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
+& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
+& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
+U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, -6.451, \\
+& -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, -8, -8, \\
+& -8, -8, -8, -8, -8 \\\hline
+  \end{tabular}
+\end{table}
+
+%\begin{table}[h]
+%\centering
+%\caption{Distribution of PPC codes over the optimal hidden states for successful (S) and unsuccessful (U) HMMs.
+Significant PPC codes are given in short form, which have log probability higher than -2.}
+%\label{tab:emission_matrix}
+%  \begin{tabular}{|l|l|}
+%  \hline
+%   \textbf{Hidden} & \textbf{Log probability of each symbol (301,LUP+, 327,OQ-TBN,101,EA,R-CT+,OQ-ECT+,AR,SUM-S,} \\ 
+%   \textbf{states} & \textbf{R-CT-S,RBA-S,G-INFO+,AF,314,CQ-EF,119,R-CML+, OQ-ECML+, 307, 325, 344,CHT-,106,} \\
+%    & \textbf{OQ-ECT-, HUP-O, 345, 309, R-CT-S, 114, 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
+%S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
+%& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
+%& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
+%S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+%& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+%& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+%S-3 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+%& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+%& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+%S-4 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
+%& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
+%& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+%
+%U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
+%& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
+%& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
+%U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, -6.451, \\
+%& -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, -8, -8, \\
+%& -8, -8, -8, -8, -8 \\\hline
+%U-3 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, -6.451, \\
+%& -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, -8, \\
+%& -8, -8, -8, -8, -8, -8 \\\hline
+%U-4 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, \\
+%& -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, \\
+%& -8, -8, -8, -8, -8, -8, -8 \\\hline
+%
+%  \end{tabular}
+%\end{table}
 

### 2017-03-10 paper/AMIA/main.tex
final version by Alex

--- 2017-03-09 paper/AMIA/main.tex
@@
 
-\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Mind Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
+\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
 
@@
 $^1$Department of Computer Science,  
-$^2$Pediatric Prevention Research Center, Wayne State University, Detroit, Michigan\\
+$^2$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
 }
@@
 \noindent{\bf Abstract}
-\textit{The problem of analyzing temporally ordered sequences of past events to make predictions related to the observed sequence of events arises in many data analytics scenarios in healthcare informatics.
-In this paper, we focus on automating the analysis of patient-provider communication sequences in the context of clinical interviews and propose the methods for identifying the most effective communication exchanges to elicit a particular type of patient response and dynamically predicting this type of response in an observed patient-provider communication.
-Our proposed method for predicting the likelihood of desired patient behavioral response in communication sequences can be used in conjunction with either Markov Chain or Hidden Markov Model and can significantly decrease the effort  required to develop effective interventions to address many public health conditions, such as childhood obesity.
-It achieved 75.32\% and 79.89\% F-measure for natural sequences and 97.36\% and 96.99\% F-measure for alternating sequences using first and second order Markov Chain and Hidden Markov Model, respectively,  to model communication sequences.
-Our results indicate that individual short length patient-provider communication sequences are highly indicative of the overall progression and future trajectory of a clinical interview and can be used to predict its overall success.
-Our results have broad implications for qualitative public health research by providing a formal mechanism to facilitate the development of effective behavioral interventions.}
+\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics.
+In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose two methods, based on Markov chain and Hidden Markov Model, for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.
+Our method achieved 75.32\% and 79.89\% F-measure for natural sequences and 97.36\% and 96.99\% F-measure for alternating sequences in predicting the outcome of motivational interviews with obese adolescents using first and second order Markov Chain and Hidden Markov Model, respectively.
+The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}
 
@@
 Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
-One approach to designing an effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
+One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
 The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
@@
 However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
-Since manual annotation and examination of transcripts is a very time-consuming process, development of new obesity interventions can take years.
-Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions.
+Since manual annotation and examination of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years.
+Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.
 
-While our previous work \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing 115 patient and provider behavior codes \cite{carcone2013provider}, in the work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
+Our research addresses this problem from multiple angles.
+In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}.
+In this work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
 change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
-
-While sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to establish the relationship between physician and patient behaviors associated with patients' exhibiting cues, there have been only a handful of previous studies on modeling annotated sequences of patient-provider communication exchanges and predicting a desired patient behavior in the context of clinical interviews.
+While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting a desired patient behavior in a context of motivational interviews.
 
@@
 The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances.
-Each motivational interview consists of counselor-adolescent and counselor-caregiver sessions.
+Each transcript consists of a counselor-adolescent and a counselor-caregiver session.
 Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions.
-The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider} with 115 distinct behavior codes, which are grouped into the youth, caregiver, and counselor code groups.
+The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups.
 Utterances were divided into successful and unsuccessful communication sequences.
-Successful communication sequences result in either positive change talk or commitment language statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk, negative commitment language or the ones in which neither change talk nor commitment language statements occur.
+Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
 Out of 1072 observed sequences, 880 were positive and 192 were negative.
-alternating sequences are the possible combination of a natural sequence where two consecutive utterances cannot belong to the same speaker.
-For example, there are 6 possible alternating sequences ({C1, A1, C2, A3}, {C1, A1, C3, A3}, {C1, A1, C4, A3}, {C1, A2, C2, A3}, {C1, A2, C3, A3} and {C1, A2, C4, A3}) for the natural sequence {C1, A1, A2, C2, C3, C4, A3}, where adolescent and counselor codes are represented by A and C, respectively.
-A total of 18753 alternating sequences have been observed, out of which 18139 were successful and 614 were unsuccessful ones.
+Alternating communication sequences are sub-sequences of a natural communication sequence, in which two consecutive utterances cannot belong to the same speaker (i.e.
+a counselor behavior code has to be followed by an adolescent behavior code and vice versa).
+For example, 6 alternating sequences (\{C1, A1, C2, A3\}, \{C1, A1, C3, A3\}, \{C1, A1, C4, A3\}, \{C1, A2, C2, A3\}, \{C1, A2, C3, A3\} and \{C1, A2, C4, A3\}) can be created from a natural sequence \{C1, A1, A2, C2, C3, C4, A3\}, where adolescent and counselor codes are represented by A and C, respectively.
+A total of 18753 alternating sequences have been observed in experimental dataset, out of which 18139 were successful and 614 were unsuccessful ones.
 For each type of sequence (natural and alternating), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences.
@@
 \centering
-\caption{Statistics of experimental dataset.}
+\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
 \label{tab:data_dist}
-  \begin{tabular}{|l|l|l|l|l|l|}
+  \begin{tabular}{|l|l|l|l|l|}
   \hline
    \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
- natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
-alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
+Natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
+Alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
   \end{tabular}
@@
 
-\subsection*{\textit{Sequence classification techniques}}
+\subsection*{\textit{Prediction method}}
 
 Generally, a sequence is a temporally ordered set of events.
-In our study, an event is a behavior code represented as a symbolic value, such as 117 (low uptake, positive), 331 (open-ended question), etc.
-Given a sequence of behavior codes represented as $S_i = \{x_1, x_2,...,x_n\}$ and a set of class labels $L = \{l_1, l_2,...,l_m\}$, the task of sequence classification is to learn a sequence classifier C, which is a function mapping a sequence $S_i$ to a class label $l_i \in L$, written as, $C : S_i \to l_i, l_i \in L$, where $L$ is fixed and known in advance.
-In our case, the labels are ``successful'' and ``unsuccessful'' (motivational interview).
+In this study, an event is a behavior code that also has a symbolic representation, such as 117 $LUP+$ (low uptake, positive), 331 $OQ-ECT+$ (open-ended question, elicit change talk positive), etc.
+Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some part of a motivational interview, the task of predicting interview success can be viewed as sequence classification.
+Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are ``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.
 
-We model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are probabilistic methods for modeling discrete observation sequences with finite vocabulary.
+Our proposed prediction method consists of two steps.
+In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary.
 Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.
 
-\textbf {Markov Chain}: In probability theory, a Markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the Markov property).
-Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable.
-For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
+\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property).
+Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable.
+For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
 A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
@@
 
-The likelihood of success of a given motivational interview at a given time point, given a sequence of MI behavior codes observed prior to that point, is estimated based on the following formula:
+In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
 
 \begin{equation}
-p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})\label{eq:class}
+p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
 \end{equation}
 
-if $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language.
-Estimating the model on the training data involves experimentally determining the value of $\delta$, resulting in the highest prediction F-measure with lower bias.
-Figure~\ref{fig:delta} illustrates the accuracy of Markov model-based method by varying the value of $\delta$.
+If $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language.
+Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias.
+Figure~\ref{fig:delta} illustrates the accuracy of Markov Chain-based prediction method by varying the value of $\delta$.
 As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias.
@@
 
-The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code to compute the state transition probability matrix.
+The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
 A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular state (symbol) $x_i$ is computed by looking at the k preceding states (symbols).
@@
 
-We exclude the structure parameters M and N to designate HMM using a compact notation:
-
-\begin{center}
-$\lambda = (A, B, \pi)$
-\end{center}
-
-The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model figures out a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
+We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$.
+The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
 The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations.
@@
 
-\subsection*{\textit{Evaluation methods}}
-The performance of classifiers is reported on the basis of 5-fold cross-validation by using weighted macro-averaging over folds.
-Precision, recall, F-measure and accuracy were used to evaluate the performance of classifiers.
-A true positive (TP) was counted when classifier correctly classified the sequence into actual class; a false positive (FP) was counted when classifier incorrectly classified other classes to the considered class; a false negative (FN) when the considered class incorrectly classified to other classes.
-The precision of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences identified as that particular class by the classifier.
-In other words, Precision = TP / (TP + FP).
-Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (labeled data set).
-In more specifically, Recall = TP / (TP + FN).
-Similarly, Accuracy of a class was also defined as the ratio between the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard.
-However, F-measure can be computed as the harmonic mean of precision and recall, or 2 x Precision x Recall / (Precision + Recall).
+\subsection*{\textit{Evaluation metrics}}
+Performance of the proposed method was evaluated in terms of precision, recall, F-measure and accuracy using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds.
+A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
+Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)).
+Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e.
+Recall = TP / (TP + FN)).
+F-measure is computed as the harmonic mean of precision and recall (i.e.
+2 x Precision x Recall / (Precision + Recall)).
+Accuracy of a class is defined as the ratio of the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard.
 
 \section*{Results}
-Experimental evaluation of the classification of utterance sequences by using the proposed probabilistic models is reported with two set of experiments: the first set of experiment was focused on natural sequences, while the second set of experiment was performed on alternating sequences.
+Experimental evaluation of the proposed method is conducted on both natural sequences and alternating sequences.
 
@@
 Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}.
-It follows that first-order HMM demonstrates the best performance achieving the highest precision 0.7980 with 0.7989 F-measure.
-However, second-order Markov Chain demonstrates the lowest performance in terms of precision and F-measure.
-It can be seen that the lowest recall of 0.7686 is achieved by the first-order MC.
-On the other hand, second-order HMM achieves the highest recall 0.8449 and accuracy 0.8449.
-From Table~\ref{tab:result_norm_seq}, it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM.
-Precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
+Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}.
+First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences.
+Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449) and accuracy (0.8449).
+Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM.
+In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
 In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models.
-Overall, HMM performed better than MC for natural sequences.
 \\
@@
 These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
-Similar to the case of natural sequences, first-order MC has the lowest performance compared to all other models across all metrics.
-Also, as in the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models.
-However, different from natural sequences, second-order MC demonstrates the highest precision 0.9778 as well as the highest F-measure of 0.9736.
+Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models.
+However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736).
 We also observed that second-order MC and HMM outperform first-order MC and HMM.
@@
 \subsection*{\textit{Impact of hidden states}}
-Since the number of hidden states may impact the classification
-performance, we use an optimum number of hidden states in HMM's.
-Figure~\ref{fig:hidden-states} shows the performance of HMMs with a different number of hidden states.
-The results illustrate that, in general, the state number didn't impact the classification performance significantly.
-All four models show their best performance when state number in between 4-8.
-Table~\ref{tab:emission_matrix} shows an example of the distribution of PPC codes over the optimum number of hidden states (4) for successful (S) and unsuccessful (U) HMMs.
-Log probability of each code reflects the state significance.
-In other words, these probabilities indicate that which hidden states are close to adolescent commitment language and which are not?
+While MC is a non-parametric method, HMM has an important parameter, the number of hidden states.
+A natural question is how does this parameter affect the performance of HMM-based prediction  method.
+Figure~\ref{fig:hidden-states} provides an answer to this question by illustrating the performance of the proposed method using HMMs of different order in terms of F-measure on different types of sequences.
+As follows from this Figure, in general, the number of states does not significantly impact the classification performance.
+All four models show their best performance when the number of states is between 4 and 8.
 
-\begin{figure}[htb!]
+Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\
+
+\begin{figure}[!htb]
     \centering
     \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
-    \caption{\textbf{Performance of HMM's with different sequence type, order, and number of hidden states used in the experiment}.}
+    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
     \label{fig:hidden-states}
@@
 
-\begin{table}[h]
+\begin{table}[!htb]
 \centering
-\caption{Distribution of PPC codes over the optimal hidden states for successful (S) and unsuccessful (U) HMMs.}
-\label{tab:emission_matrix}
+\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
+\label{tab:emission_matrix_s}
   \begin{tabular}{|l|l|}
   \hline
-   \textbf{Hidden} & \textbf{Log probability of each symbol (301, 117, 327, 343, 101, 305, 311, 331, 310, 328, 112,} \\ 
-   \textbf{states} & \textbf{ 326, 302, 308, 314, 306, 119, 313, 333, 307, 325, 344, 109, 106, 332, 120, 345, 309, 312, 114,} \\
-    & \textbf{ 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
-S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
-& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
-& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
-S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
-& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
-& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
-S-3 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
-& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
-& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
-S-4 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
-& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
-& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
+   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
+1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
+2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
+  \end{tabular}
+\end{table}
 
-U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
-& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
-& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
-U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, \\
-& -2.415, -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, \\
-& -8, -8, -3.794, -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
-U-3 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, \\
-& -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, \\
-& -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
-U-4 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, \\
-& -2.415, -3.1, -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, \\
-& -8, -8, -3.794, -8, -8, -8, -8, -8, -8, -8, -8 \\\hline
-
+\begin{table}[!htb]
+\centering
+\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
+\label{tab:emission_matrix_u}
+  \begin{tabular}{|l|l|}
+  \hline
+   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
+1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
+2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
   \end{tabular}
@@
 It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question.
-After that, if adolescents show positive behavior change, the counselor immediately reflects on that positive tendency, to reinforce adolescent's desire to adhere to positive behavior change.
-On the other hand, general information by the counselor can lead to a negative behavior change in adolescents, even in the cases when adolescents were showing positive tendency in their previous communication.
-This observation can be explained by adolescents quickly loosing focus when presented with general information.
+After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
+On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing positive tendency in their previous communication.
+This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation.
+Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews.
 \\
@@
 \section*{Discussion}
-The purpose of this study was to facilitate the design of motivational interview by examining the performance of sequence classification in the field of health informatics.
-Studying the results across the different classification schemes we can make some key observations.
+By analyzing the experimental results of different communication sequence outcome prediction schemes proposed in this paper, we arrived at the following conclusions.
+First, using higher-order models results in better prediction accuracy compared to lower-order models, since higher-order models utilize larger context and can better capture different nuances of patient-provider communication.
+On the other hand, the number of states in higher-order Markov models grows exponentially with their order.
+Therefore, accurate estimation of transition probabilities requires much larger training set.
+Using smaller datasets will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates.
+Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing.
+In this study, we found out that using second-order Markov models is a reasonable trade-off between efficiency and accuracy.
 
-First, higher order models have better classification accuracy as compared to lower order models because they are able to capture longer ordering constraints present in the data set.
-However, the number of states in higher order Markov chains grows exponentially with the order of the chain.
-Consequently, it is harder to accurately estimate all the transition probabilities from the training set.
-That is, there are many high-order states that are not frequently observed in the training set, leading to inaccurate probability estimates.
-In principle, this problem can be solved if the size of the training set is increased at the same rate as the state-size of the higher order Markov chain.
-However, this cannot be easily done for many applications areas include our case.
-To skip these problems we restricted our study up to second order.
-The shorter average length of utterance sequence is another reason for keeping the order lower.
+Second, the overall predictive performance of HMM is substantially better than MC for both types of sequences.
+In particular, HMM-based method achieves near-human accuracy for predicting the success of motivational interviews.
+This indicates that hidden states in HMM are able to capture the structure of discourse in motivational interviews by grouping together the codes that correspond to different cognitive states of adolescents during motivational interviews, which reflect the overall progression of the interviews.
+This allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
 
-Second, the overall accuracy of hidden Markov model is better than general Markov model in both types of sequences.
-We observed that hidden Markov model achieved near-human accuracy to categories the sequences of behavior codes.
-This means that hidden states capture the information passes through the codes of the utterance sequence for determining the final outcome.
+Third, converting natural PPC communication sequences to alternating ones results in better performance for all configurations of the proposed method.
+This indicates that alternating sequences emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the corresponding conditional probabilities.
+Better estimates of parameters in Markov models of successful and unsuccessful interviews are propagated to the next step, where they are aggregated into predictions for the entire sequence.
+This allows to achieve a dramatic improvement in the prediction accuracy of the method.
 
-Third, alternating sequences provide higher performance for all Markov models.
-Even it provides the highest precision and F-measure for general Markov model.
-This indicates that each pair of alternating behavior codes forward their mutual information to the next step more accurately which improve performance for distinguishing outcomes of the motivational interview.
-And this is obvious because if we know the previous two behavior codes, for instance, provider's question and adolescent's response then it is easier to know the expected response for the current question of the counselor.
-
-Finally, it was seen that successful interviews are more frequently responded by the adolescent with positive behavior changes which lead to making a positive commitment for loosing weight.
-Whereas, unsuccessful interviews are responded with negative behavior changes which lead to a negative commitment or no commitment.
+Fourth, the proposed method can be used to identify the most effective communication strategies at eliciting a particular type of behavioral response.
+Awareness of these strategies by researchers can significantly decrease the time and effort required to develop effective interventions to address many public health conditions, such as childhood obesity, and tailor these interventions to particular patient cohorts.
+Awareness of these strategies by the counselors can lead to greater success rate of motivational interviews.
 
 \section*{Conclusion}
-In this paper, we proposed a method for based on Markov Chain and Hidden Markov Model could be used to build classifiers for the sequence of behavior codes in motivational interviews.
-Experimental results show that the behavior codes carry some inherited information to distinguish motivational interviews.
-This work can facilitate researchers to establish a causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
-It provides information that can directly inform and increase the efficiency of clinical practice for a successful interview.
-This work also explains some successful patterns for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language.
-The direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the classification of utterance sequences.
-We also have the plan to use large data set in conjunction with some advanced features for the evaluation of sequence classifiers.
+In this paper, we proposed a method based on Markov Chain and Hidden Markov Model for predicting the success of motivational interviews.
+We found out that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
+Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the  interviews without resource intensive manual qualitative analysis of interview transcripts.
+Our proposed method can also help identify specific patterns that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.
+This work also has broad implications for qualitative public health research by providing a formal theoretically-grounded computational mechanism to facilitate the development of effective behavioral interventions.
 
 \section*{Acknowledgments}
-We would like to thank the technical staff at Pediatric Prevention Research Center and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming data into free text EHR used for experiments reported in this paper.
+This study was supported by an R21 grant DK108071 from the NIH.
+We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help with transcribing the recordings of motivational interviews.
+\pagebreak
 

### 2017-03-10 paper/AMIA/main.tex
Modified the authors order and their contribution

@@
 \usepackage{multirow}
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
 
@@
 
-\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
+\author{Alexander Kotov, PhD$^{1}$\footnote[1]{Authors provided equal contribution.
+\label{footnote1}}, Mehedi Hasan, BS$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
 

### 2017-08-13 mchain-python/obesity-mchain-best.py
added python version of markov chain model

@@ -1,0 +1,123 @@
+import numpy as np
+import utility
+import argparse
+import random
+
+#############################################################################################
+# Parse command line arguments
+parser = argparse.ArgumentParser(description='Train HMM Model.')
+# parser.add_argument('-data',
+#                     default='/home/mehedi/teana/data-source/seq-analysis/hmm/unbalanced/cht-cml/allsequence.txt',
+#                     help='File location containing training sequence.')
+# parser.add_argument('-output_directory', default='/home/mehedi/teana/data-source/seq-analysis/hmm/',
+#                     help='Directory to save results.')
+
+parser.add_argument('--data',
+                    default='/home/mehedi/teana/data-source/seq-analysis/hmm/balanced/cht-cml/under-sampling-new.txt',
+                    help='File location containing training sequence.')
+parser.add_argument('--output_directory', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/parameter-tuning/results/best hmm cv borderline2/',
+                    help='Directory to save results.')
+parser.add_argument('--hidden_states', default=5, type=int, help='number of hidden states')
+parser.add_argument('--codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
+                    help='File location containing codebook.')
+parser.add_argument('--sampling', default='oversampling',  type=str, help='dropout rate parameter.')
+
+args = parser.parse_args()
+
+############################################################################################
+# Load up training data
+data_filename = args.data
+codebook_filename = args.codebook
+output_directory = args.output_directory
+sampling = args.sampling
+hidden_states = args.hidden_states
+training_filename = "train.txt"
+testing_filename = "test.txt"
+
+# get results fro k folds
+kFolds = 10
+macro_results = []
+micro_results = []
+codebook = utility.loadCodeBook(codebook_filename)
+foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
+
+for k in np.arange(0, kFolds):
+    # get train and test data
+    utility.createUnderOrOverSample(sampling, foldData[k][0], testing_filename, max_len, codebook)
+    utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
+
+    # determine codebook size and number of hidden states
[…]

### 2017-08-13 mchain-python/runcode.sh

@@ -1,0 +1,5 @@
+
+for (( i=1; i <= 10; ++i ))
+do
+    python obesity-mchain-best.py
+done

### 2017-08-13 mchain-python/utility.py

@@ -1,0 +1,330 @@
+import numpy as np
+import random
+import re
+import operator
+from hmmlearn import hmm
+from keras.preprocessing.sequence import pad_sequences
+from imblearn.over_sampling import SMOTE
+from collections import Counter
+from imblearn.under_sampling import ClusterCentroids
+
+###################################################################
+# read codebook from file
+def loadCodeBook(codebook_filename):
+    codebook = []
+    with open(codebook_filename, "r") as filestream:
+        for line in filestream:
+            codebook.append(line.replace("\n", ""))
+    return codebook
+
+
+###########################################################################################
+def loadCodeBookFromTrainingFile(fileLocation):
+    all_code_map = {}
+    success_code_map = {}
+    unsuccess_code_map = {}
+    with open(fileLocation, "r") as filestream:
+        for line in filestream:
+            l = re.sub(r"\s+", "", line).split(",")
+            if l[len(l) - 1] == "500":
+                for i in range(0, len(l) - 1):
+                    success_code_map[l[i]] = l[i]
+		    all_code_map[l[i]] = l[i]
+            elif l[len(l) - 1] == "400":
+                for i in range(0, len(l) - 1):
+                    unsuccess_code_map[l[i]] = l[i]
+                    all_code_map[l[i]] = l[i]
+
+    success_sorted_map = sorted(success_code_map.items(), key=operator.itemgetter(0))
+    unsuccess_sorted_map = sorted(unsuccess_code_map.items(), key=operator.itemgetter(0))
+    all_sorted_map = sorted(all_code_map.items(), key=operator.itemgetter(0))
+
+    success_map = []
+    unsuccess_map = []
+    all_map = []
+    
+    for key, value in all_sorted_map:
+        all_map.append(key)
+    for key, value in success_sorted_map:
+        success_map.append(key)
+    for key, value in unsuccess_sorted_map:
+        unsuccess_map.append(key)
+
+    return all_map, success_map, unsuccess_map
+
+
+###########################################################################################
+def loadData(fileLocation, codebook, flag):
+    code_to_int = dict((c, i) for i, c in enumerate(codebook))
+    seq = []
+    seq_label = []
+    lengths = []
+    map = {}
+    with open(fileLocation, "r") as filestream:
+        for line in filestream:
+            l = re.sub(r"\s+", "", line).split(",")
+            if (flag == 1 and l[len(l) - 1] == "500") or (flag == 0 and l[len(l) - 1] == "400"):
[…]

### 2017-08-14 mchain-python/obesity-mchain-best.py
modified code to get 10 folds cross validation results with Markov chain model

@@ -5,32 +5,17 @@ import random
 
 #############################################################################################
 # Parse command line arguments
-parser = argparse.ArgumentParser(description='Train HMM Model.')
-# parser.add_argument('-data',
-#                     default='/home/mehedi/teana/data-source/seq-analysis/hmm/unbalanced/cht-cml/allsequence.txt',
-#                     help='File location containing training sequence.')
-# parser.add_argument('-output_directory', default='/home/mehedi/teana/data-source/seq-analysis/hmm/',
-#                     help='Directory to save results.')
-
-parser.add_argument('--data',
-                    default='/home/mehedi/teana/data-source/seq-analysis/hmm/balanced/cht-cml/under-sampling-new.txt',
-                    help='File location containing training sequence.')
-parser.add_argument('--output_directory', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/parameter-tuning/results/best hmm cv borderline2/',
-                    help='Directory to save results.')
-parser.add_argument('--hidden_states', default=5, type=int, help='number of hidden states')
+parser = argparse.ArgumentParser(description='Train Markov Model.')
 parser.add_argument('--codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
                     help='File location containing codebook.')
-parser.add_argument('--sampling', default='oversampling',  type=str, help='dropout rate parameter.')
+parser.add_argument('--sampling', default='oversampling', type=str, help='dropout rate parameter.')
 
 args = parser.parse_args()
 
 ############################################################################################
 # Load up training data
-data_filename = args.data
 codebook_filename = args.codebook
-output_directory = args.output_directory
 sampling = args.sampling
-hidden_states = args.hidden_states
 training_filename = "train.txt"
 testing_filename = "test.txt"
 
@@ -40,70 +25,82 @@ macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
 foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
+n_order = 1
 
 for k in np.arange(0, kFolds):
     # get train and test data
     utility.createUnderOrOverSample(sampling, foldData[k][0], testing_filename, max_len, codebook)
     utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
 
-    # determine codebook size and number of hidden states
[…]

### 2017-08-14 mchain-python/utility.py

@@ -8,6 +8,7 @@ from imblearn.over_sampling import SMOTE
 from collections import Counter
 from imblearn.under_sampling import ClusterCentroids
 
+
 ###################################################################
 # read codebook from file
 def loadCodeBook(codebook_filename):
@@ -29,7 +30,7 @@ def loadCodeBookFromTrainingFile(fileLocation):
             if l[len(l) - 1] == "500":
                 for i in range(0, len(l) - 1):
                     success_code_map[l[i]] = l[i]
-		    all_code_map[l[i]] = l[i]
+                    all_code_map[l[i]] = l[i]
             elif l[len(l) - 1] == "400":
                 for i in range(0, len(l) - 1):
                     unsuccess_code_map[l[i]] = l[i]
@@ -42,7 +43,7 @@ def loadCodeBookFromTrainingFile(fileLocation):
     success_map = []
     unsuccess_map = []
     all_map = []
-    
+
     for key, value in all_sorted_map:
         all_map.append(key)
     for key, value in success_sorted_map:
@@ -64,14 +65,14 @@ def loadData(fileLocation, codebook, flag):
         for line in filestream:
             l = re.sub(r"\s+", "", line).split(",")
             if (flag == 1 and l[len(l) - 1] == "500") or (flag == 0 and l[len(l) - 1] == "400"):
-                for i in range(0, len(l)-1):
+                for i in range(0, len(l) - 1):
                     seq.append([code_to_int[l[i]]])
                     map[l[i]] = l[i]
                 lengths.append(len(l) - 1)
                 seq_label.append(l[len(l) - 1])
             elif flag == 2:
                 var_len = 0
-                for i in range(0, len(l)-1):
+                for i in range(0, len(l) - 1):
                     if l[i] in code_to_int.keys():
                         seq.append([code_to_int[l[i]]])
                         map[l[i]] = l[i]
@@ -230,6 +231,7 @@ def AUC(y_true, y_pred):
     FPR = np.float(FP) / (FP + TN)
     return ((1 + TPR - FPR) / 2)
 
+
 #######################################################################
 # create startified folds for cross validation
 def createStartifiedFolds(codebook, kFolds=10):
@@ -285,9 +287,10 @@ def createStartifiedFolds(codebook, kFolds=10):
         folds.append([test, train])
     return folds, max_len
 
+
 ###################################################################
 def writeSampledSequences(X, y, codebook, outputdata_filename):
-    int_to_code = dict((i+1, c) for i, c in enumerate(codebook))
[…]

### 2017-08-14 mchain-python/obesity-mchain-tran.py
fix error to calculate transition matrix in Markov Chain

@@ -2,6 +2,8 @@ import numpy as np
 import utility
 import argparse
 import random
+import math
+from collections import Counter
 
 #############################################################################################
 # Parse command line arguments
@@ -25,7 +27,7 @@ macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
 foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
-n_order = 1
+n_order = 2
 
 for k in np.arange(0, kFolds):
     # get train and test data
@@ -33,10 +35,10 @@ for k in np.arange(0, kFolds):
     utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
 
     # set dictionary for successful sequences
-    successful_dict = utility.loadDictionary(training_filename, n_order, "500")
+    successful_dict = utility.loadTransitionDictionary(training_filename, n_order, "500")
 
     # set dictionary for unsuccessful sequences
-    unsuccessful_dict = utility.loadDictionary(training_filename, n_order, "400")
+    unsuccessful_dict = utility.loadTransitionDictionary(training_filename, n_order, "400")
 
     # set likelihood of the given test sequence(s) and then classify
     prediction_labels = []
@@ -56,34 +58,36 @@ for k in np.arange(0, kFolds):
                     prediction_labels.append("400")
                 continue
 
-            successful_prob = 0.5
-            unsuccessful_prob = 0.5
+            successful_prob = math.log(0.5)
+            unsuccessful_prob = math.log(0.5)
 
             # get probability of generating sequence from successful transcript
             words[len(words)-1] = "500"
             for i in xrange(0, len(words) - n_order - 1):
-                current_tuple = tuple([words[j] for j in xrange(i, i + n_order + 1)])
+                current_tuple = tuple([words[j] for j in xrange(i, i + n_order)])
                 if current_tuple in successful_dict.keys():
-                    transition_prob = 1.0
-                    if current_tuple in unsuccessful_dict.keys():
-                        transition_prob = float(successful_dict[current_tuple]) / (successful_dict[current_tuple] +
-                                                                                   unsuccessful_dict[current_tuple])
-                    successful_prob *= transition_prob
+                    next_states = successful_dict[current_tuple]
+                    next_states_counter = Counter(next_states)
+                    transition_prob = -5.0
+                    if words[i + n_order] in next_states_counter:
+                        transition_prob = math.log(float(next_states_counter[words[i + n_order]]) / len(next_states))
[…]

### 2017-08-14 mchain-python/utility.py

@@ -299,8 +299,6 @@ def writeSampledSequences(X, y, codebook, outputdata_filename):
             if val > 0:
                 if val in int_to_code.keys():
                     seq.append(str(int_to_code[val]))
-                else:
-                    print "Error code: ", val
         seq.append(str(y[i]))
         f.write(",".join(seq) + "\n")
     f.close()
@@ -335,7 +333,7 @@ def createUnderOrOverSample(method, given_data, outputdata_filename, max_len, co
 
 #######################################################################
 # load transition dictionary
-def loadTransitionDictionary(training_filename, n_order, label):
+def loadDictionary(training_filename, n_order, label):
     transition_dict = {}
     with open(training_filename, "r") as file_stream:
         for line in file_stream:
@@ -351,4 +349,25 @@ def loadTransitionDictionary(training_filename, n_order, label):
                         transition_dict[current_tuple] += 1
                     else:
                         transition_dict[current_tuple] = 1
+    return transition_dict
+
+
+#######################################################################
+# load transition dictionary
+def loadTransitionDictionary(training_filename, n_order, label):
+    transition_dict = {}
+    with open(training_filename, "r") as file_stream:
+        for line in file_stream:
+            words = line.replace("\n", "").split(",")
+            actual_label = words[len(words) - 1]
+            if len(words) <= n_order:
+                continue
+
+            if label == actual_label:
+                for i in xrange(0, len(words) - n_order - 1):
+                    current_tuple = tuple([words[j] for j in xrange(i, i + n_order)])
+                    if current_tuple in transition_dict.keys():
+                        transition_dict[current_tuple].append(words[i + n_order])
+                    else:
+                        transition_dict[current_tuple] = [words[i + n_order]]
     return transition_dict

### 2017-08-15 biosketch/mehedi-nihbiosketch.tex
added biosketch for me

%!TEX TS-program = xelatex
\documentclass{nihbiosketch}

%\usepackage{draftwatermark}  % delete this in your document!
%\SetWatermarkText{Sample}    % delete this in your document!
%\SetWatermarkLightness{0.9}  % delete this in your document!

%------------------------------------------------------------------------------

\name{Hunt, Morgan Casey}
\eracommons{huntmc}
\position{Associate Professor of Psychology}

\begin{document}
%------------------------------------------------------------------------------

\begin{education}
University of California, Berkeley  & B.S           & 05/1990  & Psychology \\
University of Vermont               & Ph.D.         & 05/1996  & Experimental Psychology \\
University of California, Berkeley  & Postdoctoral  & 08/1998  & Public Health and Epidemiology \\
\end{education}

\section{Personal Statement}

\begin{statement}
I have the expertise, leadership, training, expertise and motivation necessary
to successfully carry out the proposed research project.  I have a broad
background in psychology, with specific training and expertise in ethnographic
and survey research and secondary data analysis on psychological aspects of
drug addiction.  My research includes neuropsychological changes associated
with addiction.  As PI or co-Investigator on several university- and NIH-
funded grants, I laid the groundwork for the proposed research by developing
effective measures of disability, depression, and other psychosocial factors
relevant to the aging substance abuser, and by establishing strong ties with
community providers that will make it possible to recruit and track
participants over time as documented in the following publications.  In
addition, I successfully administered the projects (e.g.\ staffing, research
protections, budget), collaborated with other researchers, and produced several
peer-reviewed publications from each project.  As a result of these previous
experiences, I am aware of the importance of frequent communication among
project members and of constructing a realistic research plan, timeline, and
budget.  The current application builds logically on my prior work. During
2005--2006 my career was disrupted due to family obligations. However, upon
returning to the field I immediately resumed my research projects and
collaborations and successfully competed for NIH support.

\begin{enumerate}

\item Merryle, R.J. \& Hunt, M.C. (2004). Independent living, physical
        disability and substance abuse among the elderly. Psychology and Aging,
        23(4), 10--22.

\item Hunt, M.C., Jensen, J.L. \& Crenshaw, W. (2007). Substance abuse and
        mental health among community-dwelling elderly. International Journal
        of Geriatric Psychiatry, 24(9), 1124--1135.

\item Hunt, M.C., Wiechelt, S.A. \& Merryle, R. (2008). Predicting the
        substance-abuse treatment needs of an aging population.  American
        Journal of Public Health, 45(2), 236--245. PMCID: PMC9162292

\item Hunt, M.C., Newlin, D.B. \& Fishbein, D. (2009). Brain imaging in
        methamphetamine abusers across the life-span. Gerontology, 46(3),
        122--145.

\end{enumerate}

\end{statement}

%------------------------------------------------------------------------------
\section{Positions and Honors}

\subsection*{Positions and Employment}
\begin{datetbl}
1998--2000  & Fellow, Division of Intramural Research, National Institute of Drug Abuse, Bethesda, MD \\
2000--2002  & Lecturer, Department of Psychology, Middlebury College, Middlebury, VT \\
2001--      & Consultant, Coastal Psychological Services, San Francisco, CA  \\
2002--2005  & Assistant Professor, Department of Psychology, Washington University, St.\ Louis, MO \\
2007--      & Associate Professor, Department of Psychology, Washington University, St.\ Louis, MO\\
\end{datetbl}

\subsection*{Other Experience and Professional Memberships}
\begin{datetbl}
1995--           & Member, American Psychological Association \\
1998--           & Member, Gerontological Society of America \\
1998--           & Member, American Geriatrics Society \\
2000--           & Associate Editor, Psychology and Aging \\ 
2003--           & Board of Advisors, Senior Services of Eastern Missouri \\
2003--05         & NIH Peer Review Committee: Psychobiology of Aging, ad hoc reviewer \\
2007--11         & NIH Risk, Adult Addictions Study Section, members \\
\end{datetbl}

\subsection*{Honors}
\begin{datetbl}
2003            & Outstanding Young Faculty Award, Washington University, St.\ Louis, MO \\
2004            & Excellence in Teaching, Washington University, St.\ Louis, MO \\
2009            & Award for Best in Interdisciplinary Ethnography, International Ethnographic Society \\
\end{datetbl}

%------------------------------------------------------------------------------

\section{Contribution to Science}

\begin{enumerate}

\item My early publications directly addressed the fact that substance abuse is
    often overlooked in older adults. However, because many older adults were
    raised during an era of increased drug and alcohol use, there are reasons
    to believe that this will become an increasing issue as the population
    ages.   These publications found that older adults appear in a variety of
    primary care settings or seek mental health providers to deal with emerging
    addiction problems.  These publications document this emerging problem but
    guide primary care providers and geriatric mental health providers to
    recognize symptoms, assess the nature of the problem and apply the
    necessary interventions.   By providing evidence and simple clinical
    approaches, this body of work has changed the standards of care for
    addicted older adults and will continue to provide assistance in relevant
    medical settings well into the future.  I served as the primary
    investigator or co-investigator in all of these studies.

\begin{enumerate}

\item Gryczynski, J., Shaft, B.M., Merryle, R., \& Hunt, M.C. (2002). Community
        based participatory research with late-life addicts. American Journal
        of Alcohol and Drug Abuse, 15(3), 222--238.

\item Shaft, B.M., Hunt, M.C., Merryle, R., \& Venturi, R. (2003). Policy
        implications of genetic transmission of alcohol and drug abuse in
        female nonusers. International Journal of Drug Policy, 30(5), 46--58.

\item Hunt, M.C., Marks, A.E., Shaft, B.M., Merryle, R., \& Jensen, J.L.
        (2004). Early-life family and community characteristics and late-life
        substance abuse. Journal of Applied Gerontology, 28(2),26--37.

\item Hunt, M.C., Marks, A.E., Venturi, R., Crenshaw, W. \& Ratonian, A.
        (2007). Community-based intervention strategies for reducing alcohol
        and drug abuse in the elderly.  Addiction, 104(9), 1436--1606. PMCID:
        PMC9000292

\end{enumerate}

\item In addition to the contributions described above, with a team of
    collaborators, I directly documented the effectiveness of various
    intervention models for older substance abusers and demonstrated the
    importance of social support networks.   These studies emphasized
    contextual factors in the etiology and maintenance of addictive disorders
    and the disruptive potential of networks in substance abuse treatment. This
    body of work also discusses the prevalence of alcohol, amphetamine, and
    opioid abuse in older adults and how networking approaches can be used to
    mitigate the effects of these disorders.

\begin{enumerate}

\item Hunt, M.C., Merryle, R. \& Jensen, J.L. (2005). The effect of social
        support networks on morbidity among elderly substance abusers. Journal
        of the American Geriatrics Society, 57(4), 15--23.

\item Hunt, M.C., Pour, B., Marks, A.E., Merryle, R. \& Jensen, J.L. (2005).
        Aging out of methadone treatment. American Journal of Alcohol and Drug
        Abuse, 15(6), 134--149.

\item Merryle, R. \& Hunt, M.C. (2007). Randomized clinical trial of cotinine
        in older nicotine addicts. Age and Ageing, 38(2), 9--23. PMCID:
        PMC9002364

\end{enumerate}

\item Methadone maintenance has been used to treat narcotics addicts for many
    years but I led research that  has shown that over the long-term, those in
    methadone treatment view themselves negatively and they gradually begin to
    view treatment as an intrusion into normal life.   Elderly narcotics users
    were shown in carefully constructed ethnographic studies to be especially
    responsive to tailored social support networks that allow them to
    eventually reduce their maintenance doses and move into other forms of
    therapy.  These studies also demonstrate the policy and commercial
    implications associated with these findings.

\begin{enumerate}

\item Hunt, M.C. \& Jensen, J.L. (2003). Morbidity among elderly substance
        abusers. Journal of the Geriatrics, 60(4), 45--61.

\item Hunt, M.C. \& Pour, B. (2004). Methadone treatment and personal
        assessment. Journal Drug Abuse, 45(5), 15--26.

\item  Merryle, R. \& Hunt, M.C. (2005). The use of various nicotine delivery
        systems by older nicotine addicts. Journal of Ageing, 54(1), 24--41.
        PMCID: PMC9112304

\item Hunt, M.C., Jensen, J.L. \& Merryle, R. (2008). The aging addict:
        ethnographic profiles of the elderly drug user.  NY, NY: W. W. Norton
        \& Company.

\end{enumerate}

\end{enumerate}

\subsection*{Complete List of Published Work in MyBibliography:} 
\url{http://www.ncbi.nlm.nih.gov/myncbi/browse/collection/45972964/}

%------------------------------------------------------------------------------

\section{Research Support}

\subsection*{Ongoing Research Support}

\grantinfo{R01 DA942367}{Hunt (PI)}{09/01/08--08/31/16}
{Health trajectories and behavioral interventions among older substance abusers}
{The goal of this study is to compare the effects of two substance abuse interventions on health 
outcomes in an urban population of older opiate addicts.}
{Role: PI}

\bigskip

\grantinfo{R01 MH922731}{Merryle (PI)}{12/15/07--11/30/15}
{Physical disability, depression and substance abuse in the elderly}
{The goal of this study is to identify disability and depression trajectories and demographic factors 
associated with substance abuse in an independently-living elderly population.}
{Role: Co-Investigator}

\bigskip

\grantinfo{Faculty Resources Grant, Washington University}{08/15/09--08/14/15}
{Opiate Addiction Database}
{The goal of this project is to create an integrated database of demographic, social and biomedical 
information for homeless opiate abusers in two urban Missouri locations, using a number of state and 
local data sources.}
{Role: PI}

%------------------------------------------------------------------------------

\subsection*{Completed Research Support}

\grantinfo{R21 AA998075}{Hunt (PI)}{01/01/11--12/31/13}
{Community-based intervention for alcohol abuse}
{The goal of this project was to assess a community-based strategy for reducing alcohol abuse among 
older individuals.}
{Role: PI}

\end{document}

### 2017-08-15 paper-amia-summit-2018/main.tex
added initial files for AMIA Summit 2018

--- 2017-03-10 paper/AMIA/main.tex
@@
 Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias.
-Figure~\ref{fig:delta} illustrates the accuracy of Markov Chain-based prediction method by varying the value of $\delta$.
+Figure~\ref{fig:delta} illustrates the performance of Markov Chain-based prediction method by varying the value of $\delta$.
 As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias.
@@
     \includegraphics[width=0.70\textwidth]{figures/delta.png}
-    \caption{\textbf{Accuracy of the proposed Markov model-based method by varying the value of $\delta$}.}
+    \caption{\textbf{F-Measure of the proposed Markov model-based method by varying the value of $\delta$}.}
     \label{fig:delta}
@@
 \subsection*{\textit{Evaluation metrics}}
-Performance of the proposed method was evaluated in terms of precision, recall, F-measure and accuracy using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds.
+Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds.
 A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
@@
 2 x Precision x Recall / (Precision + Recall)).
-Accuracy of a class is defined as the ratio of the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard.
 
@@
 First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences.
-Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449) and accuracy (0.8449).
-Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM.
+Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449).
+Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM.
 In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
@@
   \hline
-   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline
+   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline
 
- \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
- & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
- \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989} & 0.8059\\ \cline{2-6}
- & Second order & 0.7400 & \textbf{0.8449} & 0.7822  & \textbf{0.8449}\\ \hline
+ \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532\\\cline{2-5}
+ & Second order & 0.6889 & 0.7889 & 0.7352\\ \hline
+ \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989}\\ \cline{2-5}
+ & Second order & 0.7400 & \textbf{0.8449} & 0.7822\\ \hline
 
@@
 These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
-Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models.
+Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models.
 However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736).
@@
   \hline
-   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline
+   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline
 
- \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
- & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
- \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353 & 0.9392\\ \cline{2-6}
- & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
+ \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604\\\cline{2-5}
+ & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736}\\ \hline
+ \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353\\ \cline{2-5}
+ & Second order & 0.9704 & \textbf{0.9713} & 0.9699\\ \hline
   \end{tabular}

### 2017-08-15 paper-amia-summit-2018/main_back.tex
added initial files for AMIA Summit 2018

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}

\begin{document}

\title{Title of Your Submission}

\author{Firstname A. Lastname, Degrees$^{1}$, Firstname B. Lastname, Degrees$^{2}$}

\institutes{
    $^1$Institution, City, State, Country (if applicable); $^2$Institution, City, State, Country (if applicable)\\
}

\maketitle

\noindent{\bf Abstract}

\textit{Abstract text goes here, justified and in italics.  The abstract would normally be one paragraph long.  See Table 1. for appropriate abstract length by submission type.}

\section*{Introduction}
This template should be used as a starting point for AMIA submissions.  
It is important to review the AMIA Call for Participation (CFP) where types of submissions considered and general requirements for each submission type are listed. All submissions must conform to the format and presentation requirements described in the CFP and at the submission site.

\section*{Another Major Heading and References}
This sentence has two reference citations\cite{ref1,ref2}.

More text of an additional paragraph, with a figure reference (Figure ~\ref{fig1}) and a figure inside a Word text box below.  Figures need to be placed as close to the corresponding text as possible and not extend beyond one page.\\
\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{figures/cnn.eps}
\caption{Total allergy alerts, overridden alerts, or drug order cancelled.}
\label{fig1}
\end{figure}

This is additional text added just to show the one-column formatting.  This is additional text added just to show the one-column formatting.  This is additional text added just to show the one-column formatting.  This is additional text added just to show the one-column formatting.  This is additional text added just to show the one-column formatting.  This is additional text added just to show the one-column formatting.  This is additional text added just to show the one-column formatting.

This paragraph contains a reference to a table just below (Table 1).  All tables need to be placed as close to the corresponding text as possible, But each individual table should be on one page and not extend to multiple pages unless labeled as ``ÄúContinued"Äù.

\begin{table}[h]
\centering
\caption{Submission type, abstract length, and page length maximum for AMIA submissions.}
  \begin{tabular}{|l|l|l|}
  \hline
    \textbf{Submission Type}    & \textbf{Abstract Length}  & \textbf{Page Length Maximum**} \\ \hline
    Paper  & 125-150 words  & Ten   \\ \hline
    Student Paper  & 125-150 words  & Ten \\ \hline
    Poster  &50-75 words*   & One \\ \hline
    Podium  Abstract & 50-75 words*  & Two \\ \hline
    Panel   &150-200 words  & Three \\ \hline
    System Demonstrations    &150-200 words  & One \\ \hline
  \end{tabular}
\end{table}
*: All podium abstract and poster submissions must have a brief (50-75 words) abstract. The abstract does NOT have to be part of the document, but must be entered on the submission website in the Abstract box in Step 2.

**: \textcolor{red}{If your submission is longer than what is specified below, it will be rejected without review.}

This is another paragraph.

\section*{Conclusion}
Your conclusion goes at the end, followed by References, which must follow the Vancouver Style (see: www.icmje.org/index.html). References begin below with a header that is centered. Only the first word of an article title is capitalized in the References.

\makeatletter
\renewcommand{\@biblabel}[1]{\hfill #1.}
\makeatother

\bibliographystyle{unsrt}
\begin{thebibliography}{1}
\setlength\itemsep{-0.1em}

\bibitem{ref1}
Pryor TA, Gardner RM, Clayton RD, Warner HR. The HELP system. J Med Sys. 1983;7:87-101.
\bibitem{ref2}
Gardner RM, Golubjatnikov OK, Laub RM, Jacobson JT, Evans RS. Computer-critiqued blood ordering using the HELP system. Comput Biomed Res 1990;23:514-28.

\end{thebibliography}

\end{document}

### 2017-08-15 paper-amia-summit-2018/main_back_2.26.tex
added initial files for AMIA Summit 2018

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{multirow}

\begin{document}

\title{Leverage machine learning methods for the classification of clinical interview fragment sequences}

\author{Mehedi Hasan, BS$^{1}$, Alexander Kotov, PhD$^{1}$, April Carcone, PhD$^{2}$, Sylvie Naar-King, PhD$^{2}$}

\institutes{
$^1$Department of Computer Science,  
$^2$Pediatric Prevention Research Center, Wayne State University, Detroit, Michigan\\
}

\maketitle

\noindent{\bf Abstract}
\textit{This study examines the effectiveness of state-of-the-art supervised machine learning methods for the task of automatic annotation of fragments of clinical interviews in conjunction with sequential analysis of that annotation. We used a collection of motivational interview transcripts consisting of 11,353 utterances, which were manually annotated by two human coders as a golden standard, and experimented with state-of-art classifiers, including Na\"{i}ve Bayes, J48 Decision Tree, Support Vector Machine (SVM), Random Forest (RF), AdaBoost, DiscLDA, Conditional Random Fields (CRF) and Convolutional Neural Network (CNN) in conjunction with lexical, contextual (label of the previous utterance) and semantic (distribution of words in the utterance across the Linguistic Inquiry and Word Count dictionaries) features. We found out that, when the number of classes is large, the performance of the CNN and CRFs is inferior to SVM. Among all classifiers using only lexical features, SVM achieved the highest classification accuracy of 70.8\%, 61\% and 53.7\%, when interview transcripts were annotated with the codebooks consisting of 17, 20 and 41 codes, respectively. Using contextual and semantic features, as well as their combination, in addition to lexical ones improved the accuracy of SVM for annotation of utterances in motivational interview transcripts with a codebook consisting of 17 classes to 71.5\%, 74.2\% and 75.1\%, respectively. We also used first order markov model that achieved 68.03\% accuracy for the classification of sequence of behavioral codes extracted from the above dataset. Therefore, our results demonstrate the potential of using machine learning  and textual data mining methods in conjunction with semantic and contextual features for automatic annotation of clinical interview transcripts  and their sequences with near-human accuracy.}

\section*{Introduction}
In this study, our analysis have two directions: automatic annotation and sequential data analysis. Annotation (or labeling) of fragments of clinical text with the categories (or labels, codes) from a predefined codebook is an integral part of qualitative research. It can also be viewed as classification of textual fragments into a predefined number of categories (classes). Textual annotation has been traditionally performed manually by trained coders, which is a tedious, costly and time-consuming process. Furthermore, manual annotation increases the likelihood of errors due to coder fatigue and bias associated with human subjectivity. To automate tedious cognitive tasks such as classification, supervised machine learning methods have been recently proposed. Although these methods have been shown to be successful at binary (two-class) classification \citep{40,41} (e.g. classifying textual fragments as neutral or opinionated), we are not aware of any prior studies reported in the literature that evaluate their performance for textual classification tasks involving large number of classes. Such tasks, however, are fairly common in clinical setting (e.g. annotation of clinical interviews, assignment of ICD-9/10 codes to patient records). To address this limitation, in this paper we propose contextual and semantic features and present the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods in conjunction with lexical and the proposed features for the task of automatic annotation of utterances in clinical interview transcripts with the codebooks consisting of large number of classes. Our study provides a guideline for clinical informatics researchers and practitioners, who consider an option of using machine learning methods for automatic annotation of clinical text in their projects.

Annotation of clinical interview transcripts to distinguish different behavior types is an important part of clinical research aimed at designing effective interventions for many conditions and disorders. In this paper, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \citep{01}. Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \citep{02}. Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity. One approach to designing intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change \citep{03,04}. The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}. The process of establishing causal linkages to identify effective communication strategies for eliciting change talk and commitment language in MI involves a resource-intensive qualitative coding process. First, clinical interviews are transcribed and then each utterance is manually annotated with a set of codes from a pre-defined codebook designating specific behavior types. Training human coders to reliably and accurately assign codes to textual fragments requires a large investment of manpower, time and money. For example, in a recent MI study \citep{07}, training coders to reliability took about four months and, once trained, coders required five hours to code every recorded hour. A similar study reported requiring 60 hours of training over six weeks to attain coder reliability, and the actual coding involved two coding passes and six coders \citep{08}.

Automatic annotation of patient utterances in clinical communication is a challenging task, since patients usually come from a variety of cultural and educational backgrounds and
their language use can be quite different \citep{09}. This problem is exacerbated when the interviews are conducted with children and adolescents due to their tendency to use
incomplete sentences and frequently change subjects.

Previous quantitative studies of clinical conversation have resulted in creation of Generalized Medical Interaction Analysis System (GMIAS) \citep{10}, which uses a codebook with generic hierarchical categories. The small-size codebook in Comprehensive Analysis of the Structure of Encounters System (CASES) \citep{11} was designed to annotate several
meta-discursive aspects of medical interviews, such as assigning ``ownership'' of topics and partitioning them into distinct segments (speech acts). It was also shown that the
fragments of transcripts of routine outpatient visits consisting of several speech acts coded using GMIAS and CASES can be annotated as ``information giving'' and ``requesting
information'' \citep{12}. Other related previous studies focused on categorizing assertions of medical problems in clinical narrative into 5 classes (present, absent, possible,
hypothetical, conditional and associated with someone else) using SVM \citep{13} and annotating the utterances in hemodialysis phone dialogue with 3 categories using AdaBoost
classifier \citep{14}. The present study reports the results of comprehensive evaluation of 8 state-of-the-art classifiers (Na\"{i}ve Bayes \citep{15,16,17}, Support Vector Machine
\citep{18,19}, Conditional Random Fields \citep{21,22}, J48 \citep{23}, AdaBoost \citep{24}, Random Forest \citep{42}, DiscLDA \citep{25} and Convolutional Neural Network
\citep{38}) for the task of annotating clinical interviews with a codebook, consisting of a large number of classes. We also propose and experimentally evaluate two novel features
for this task: contextual features based on the label of the preceding textual fragment and semantic features based on the distribution of words in the annotated fragment over a linguistic lexicon.

Codes automatically assigned to transcripts using classification models presented above allow to abstract away from the raw text  of the interviews and unequivocally establish the casual relationship between provider communication and client change talk (CT). Two recent studies \citep{46,47} have examined causality in treatment communication sequences using sequential analysis techniques to assess the probability that CT actually follows specific provider behaviors. Both studies found that provider use of MICO communication behaviors was followed by CT more frequently than expected and followed by client counter-CT (CCT; client statements against making changes) less frequently than expected by chance. Again, while suggestive of important PPC sequences, these studies were conducted among adults in treatment for alcohol abuse. In this project, we used a probabilistic model, in particular, first order markov model for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.

\section*{Methods}
\subsection*{\textit{Data collection}}
The golden standard for evaluation of machine learning methods was created based on the transcripts of motivational interviews conducted by the clinicians at the Pediatric
Prevention Research Center (PPRC) of Wayne State University. Each interview is comprised of two parts: conversation of a clinician with an adolescent followed by a conversation of
a clinician with the adolescent's caregiver. All adolescents in this study were between the ages of 12 and 17 (M = 14.7, SD = 1.63) and most were female (n = 27). Most caregivers
were biological mothers (n = 36), who were married or living with a partner (n = 25). The median family income was \$16,000--\$21,999 ranging from less than \$1,000 to
\$50,000--\$74,999. The audio recordings of the interviews were first transcribed and segmented into utterances belonging to adolescents, caregivers, and counselors, preserving the
sequence of utterances. Transcripts were then manually annotated by trained human coders according to MYSCOPE \citep{07}, a specialized codebook including a large number of
behavior codes, which was developed by an interdisciplinary team including a clinical psychologist, a nutrition scientist, a communication scientist, a linguist and a community
health worker specifically for annotating motivational interviews with obese adolescents. A primary coder independently coded interview sessions and a secondary coder co-coded a
randomly selected 20\% of the transcripts to monitor reliability ($\kappa$ = 0.696) \citep{07}. The MYSCOPE codebook contains a total of 115 different codes that are grouped into
the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}.

\begin{table*}[htb]
\caption{\textbf{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}}    
\label{table:06}
\centering
\begin{tabular}{lp{3.6cm}lp{8cm}}
\hline
\hline
Annotation  & Description & Speaker & Text \\
\hline
331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do ? mm-hmm meaning is that food available for you ? \\
117 &	Low Uptake, positive	& Adolescent &	Yes \\
301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\
112 &	Change Talk positive	& Adolescent &	my um aunt \\
301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\
112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\
309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\
\hline
\hline
\end{tabular}
\end{table*}

To conduct a detailed analysis of performance of classification methods, we used the following two-stage process to create the codebooks with different number of codes for
adolescent and caregiver sessions. In the first stage, we merged conceptually similar behavior codes as well as the codes with similar data distributions, while in the second stage, we eliminated the codes with insufficient data samples. In case of the adolescent sessions, we started with 55 adolescent session-specific codes and, after merging the codes with subtle differences (e.g. converting valiances of change talk CHT+1, CHT+2 and CHT+3 into CHT+), obtained a codebook with 41 classes. We further reduced this codebook to 20 classes after merging 21 classes with similar sample distributions. After eliminating the codes that had less than 10 data samples (to ensure that there can be at least one sample of each class in each fold when using 10-fold cross validation experimental design), we obtained a third codebook with 17 codes. Using the same approach, we created the codebooks containing 58, 19, and 16 caregiver session-specific codes. Table~\ref{table:01} shows the distribution of utterances over 16 classes in the caregiver session transcripts. As follows from Table~\ref{table:01}, the distribution of utterances over classes is highly imbalanced even for the codebook of the smallest size, which is fairly common for clinical text. To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language. Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.

After creating the codebooks, we pre-processed the dataset using the Snowball stemmer available as part of the Weka \citep{26} machine learning toolkit\footnote{
http://www.cs.waikato.ac.nz/ml/weka/}. We also found out that stopword removal decreased the performance of classification models for our task (e.g., in case of the codebook
consisting of 17 classes, the accuracy of Na\"{i}ve Bayes decreased from 67\% to 47.10\%, while the accuracy of SVM decreased from 70.76\% to 55.26\%). A likely reason is that,
although negations are typically considered as stopwords, they are fairly important clues for inferring certain behavior types (e.g., removing the stopword ``not'' completely
transforms the meaning of a phrase ``not great'').

\subsection*{\textit{System overview}}
The pipeline consists of two stages: training and testing. Prior to the training stage, we preprocess the collected clinical interview transcripts by performing stemming, punctuation removal, word segmentation and tokenization. Features are then extracted from the preprocessed data. During this stage, previous label and LIWC features are used in conjunction with the lexical features to create the feature vectors. After that, classifiers are trained on the feature vectors extracted from the training samples and their associated annotations. In the testing stage, after creation of feature vectors, the previously trained classifiers predict the label of each utterance in the testing sample. Finally, performance of different classifiers is evaluated by calculating standard metrics such as precision, recall, F-score (F1), kappa measure and accuracy. Specifically, we evaluated the performance of the following state-of-the-art supervised machine learning methods.

\textbf {Markov Model}: In probability theory, a markov model \citep{45} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property). Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable. For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, T-AMB, CCT, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of possible youth and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:

\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}

\end{itemize}

where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ have been observed in the training data. Given a markov model $M$ (such that $S\subseteq V$), a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ have been generated from a markov model $M$ is:

\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}

Success of a given motivational interview, given a sequence of MI behavior codes corresponding to it, can be predicted based on the following formula:

\begin{equation}
p(S\rightarrow CT) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})
\end{equation}

if $p(S\rightarrow CT) > \delta $, where $\delta$ is experimentally defined threshold, the interview transcript (or a portion of it) is predicted to result in CT. We experimentally determined the value of $\delta$ to 0.3 which provides acceptable accuracy with minimum false positive rate. Figure\ref{fig:delta} illustrates the characteristics of proposed model by varying the value of $\delta$.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/deltadata.jpg}
    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
    \label{fig:delta}
\end{figure}

\subsection*{\textit{Evaluation methods}}
To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \citep{36} as an experimental design. The performance of different classifiers and feature sets was evaluated in terms of precision, recall, F1 score (F1), kappa measure and accuracy using weighted macro-averaging over folds.

\section*{Results}
Experimental evaluation of automatic annotation of clinical interview fragments and their sequences by using machine learning and probabilistic model included several dimensions:
\begin{itemize}
\item determining the performance of classifiers on the codebooks of different size;
\item determining the effectiveness of the proposed contextual and semantic features.
\item determining the performance of markov model in conjunction with determining PPC sequences that are most likely to translate into CT.
\end{itemize}

Since clinical researchers typically annotate caregiver and adolescent sessions separately, we first created two experimental datasets consisting of only adolescent and only caregiver session transcripts. Second, besides evaluating the accuracy of annotating adolescent and caregiver transcripts with the codebooks containing an entire set of codes, we also conducted a series of experiments with the codebooks of smaller sizes created as outlined above. Third, besides training and testing NB, SVM, CRF, Decision Tree, Boosting, DiscLDA, Random Forest and CNN classifiers using only lexical features, we also evaluated the effectiveness of the proposed contextual and semantic features.

Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.

\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. 
From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\

\begin{table}[h]
\centering
\caption{Performance of markov models for the classification of normal PPC code sequence.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

 \multirow{2}{*}{General Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
 & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989} & 0.8059\\ \cline{2-6}
 & Second order & 0.7400 & \textbf{0.8449} & 0.7822  & \textbf{0.8449}\\ \hline

  \end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Performance of markov models for the classification of alternate PPC code sequence.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

 \multirow{2}{*}{General Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9253 & 0.9392\\ \cline{2-6}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline

  \end{tabular}
\end{table}

Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.

\section*{Discussion}
From the sequential analysis, we observed that markov model achieved near-human accuracy to categorise the sequence of behavior codes. It was also found that successful interviews are more frequently responded by the adolescent with PPC code 112. However, unsuccessful interviews are responded with the behavior code 109.

\section*{Conclusion}
In this work, we suggest some successful sequences of codes for the practical use by the clinician during the interview session which will translate into positive change talk and commitment language. We also propose novel features and report the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods for text
classification using those features, to help clinical researchers and practitioners assess the feasibility of using these methods for the task of automatic annotation of clinical
text using the codebooks of realistic size. We found out that Support Vector Machine using only lexical features consistently outperforms all other classifiers on caregiver and
adolescent datasets according to most metrics. Adding contextual and semantic features further improves the performance of SVM on both datasets, achieving close to human accuracy
when the codebooks consisting of 16 and 17 classes are used to annotate caregiver and adolescent transcripts, respectively.

This work has important practical implications. First, it can facilitate researchers to establish causal relationship between different communication strategies and desired
behavioral outcomes without having to repeatedly wade through pages of interview transcripts. Second, since automatic annotation is significantly faster than manual, it can
dramatically accelerate the pace of research in behavioral sciences. Third, information that can directly inform and increase the efficiency of clinical practice for a successful interview. Although all experiments were conducted on interview transcripts, the proposed methods and features are not
specific to a particular domain of Motivational Interviewing, and thus there is also no prima facie reason to believe that they will not be effective for annotation of any other
type of clinical conversation.

\section*{Acknowledgments}
We would like to thank the technical staff at Henry Ford Health Systems and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming binary data into free text EHR used for experiments reported in this paper.

\bibliographystyle{vancouver}
\bibliography{references}

\end{document}

### 2017-08-15 biosketch/mehedi-nihbiosketch.tex
finished biographical sketch for the submission of paper in AMIA summit 2018

@@
 
-%\usepackage{draftwatermark}  % delete this in your document!
-%\SetWatermarkText{Sample}    % delete this in your document!
-%\SetWatermarkLightness{0.9}  % delete this in your document!
-
 %------------------------------------------------------------------------------
 
-\name{Hunt, Morgan Casey}
-\eracommons{huntmc}
-\position{Associate Professor of Psychology}
+\name{Hasan, Md Mehedi}
+\eracommons{MDHASAN}
+\position{Graduate Teaching Assistant of Computer Science}
 
@@
 \begin{education}
-University of California, Berkeley  & B.S           & 05/1990  & Psychology \\
-University of Vermont               & Ph.D.         & 05/1996  & Experimental Psychology \\
-University of California, Berkeley  & Postdoctoral  & 08/1998  & Public Health and Epidemiology \\
+Bangladesh University of Engineering and Technology, Dhaka, Bangladesh     & B.S.         & 03/2009  & Computer Science and Engineering \\
+Wayne State University, Detroit, MI     & Ph.D.         & 05/2019  & Computer Science \\
 \end{education}
 
-\section{Personal Statement}
+\section{A. Personal Statement}
 
 \begin{statement}
-I have the expertise, leadership, training, expertise and motivation necessary
-to successfully carry out the proposed research project.  I have a broad
-background in psychology, with specific training and expertise in ethnographic
-and survey research and secondary data analysis on psychological aspects of
-drug addiction.  My research includes neuropsychological changes associated
-with addiction.  As PI or co-Investigator on several university- and NIH-
-funded grants, I laid the groundwork for the proposed research by developing
-effective measures of disability, depression, and other psychosocial factors
-relevant to the aging substance abuser, and by establishing strong ties with
-community providers that will make it possible to recruit and track
-participants over time as documented in the following publications.  In
-addition, I successfully administered the projects (e.g.\ staffing, research
-protections, budget), collaborated with other researchers, and produced several
-peer-reviewed publications from each project.  As a result of these previous
-experiences, I am aware of the importance of frequent communication among
-project members and of constructing a realistic research plan, timeline, and
-budget.  The current application builds logically on my prior work. During
-2005--2006 my career was disrupted due to family obligations. However, upon
-returning to the field I immediately resumed my research projects and
-collaborations and successfully competed for NIH support.
+I have the expertise and motivation necessary to successfully carry out the proposed research project.
+I have a background in computer science, with specific expertise in health informatics, machine learning, and textual data analytics.
+My research includes automatic annotation of clinical text, information extraction from electronic health records for improving the study of cancer risk and outcomes.
+As a graduate reserach assistant on several university- and NIH-funded grants, I successfully collaborated with other researchers, and produced the following publications.
 
@@
 
-\item Merryle, R.J. \& Hunt, M.C. (2004). Independent living, physical
-        disability and substance abuse among the elderly. Psychology and Aging,
-        23(4), 10--22.
+\item Hasan, M., Kotov, A., Carcone, A.I., Dong, M., Naar, S.
+and Hartlieb, K.B.
+(2016).
+A study of the effectiveness of machine learning methods for classification of clinical interview fragments into a large number of categories.
+Journal of biomedical informatics, 62, 21-31.
 
-\item Hunt, M.C., Jensen, J.L. \& Crenshaw, W. (2007). Substance abuse and
-        mental health among community-dwelling elderly. International Journal
-        of Geriatric Psychiatry, 24(9), 1124--1135.
+\item Hasan, M., Kotov, A., Mohan, A., Lu, S.
+and Stieg, P.M.
+(2016).
+Feedback or Research: Separating Pre-purchase from Post-purchase Consumer Reviews.
+In European Conference on Information Retrieval, 682-688.
+Springer International Publishing.
 
-\item Hunt, M.C., Wiechelt, S.A. \& Merryle, R. (2008). Predicting the
-        substance-abuse treatment needs of an aging population.  American
-        Journal of Public Health, 45(2), 236--245. PMCID: PMC9162292
-
-\item Hunt, M.C., Newlin, D.B. \& Fishbein, D. (2009). Brain imaging in
-        methamphetamine abusers across the life-span. Gerontology, 46(3),
-        122--145.
+\item Kotov, A., Hasan, M., Carcone, A., Dong, M., Naar-King, S.
+and BroganHartlieb, K.
+(2015).
+Interpretable probabilistic latent variable models for automatic annotation of clinical text.
+In AMIA Annual Symposium Proceedings, 2015, 785.
+American Medical Informatics Association.
 
@@
 %------------------------------------------------------------------------------
-\section{Positions and Honors}
+\section{B. Positions and Honors}
 
@@
 \begin{datetbl}
-1998--2000  & Fellow, Division of Intramural Research, National Institute of Drug Abuse, Bethesda, MD \\
-2000--2002  & Lecturer, Department of Psychology, Middlebury College, Middlebury, VT \\
-2001--      & Consultant, Coastal Psychological Services, San Francisco, CA  \\
-2002--2005  & Assistant Professor, Department of Psychology, Washington University, St.\ Louis, MO \\
-2007--      & Associate Professor, Department of Psychology, Washington University, St.\ Louis, MO\\
+2015--2017  & Graduate Research Assistant, Department of Computer Science, Wayne State University, Detroit, MI \\
+2017--2017  & Part-time Faculty, Department of Computer Science, Wayne State University, Detroit, MI \\
+2017--  & Graduate Teaching Assistant, Department of Computer Science, Wayne State University, Detroit, MI \\
 \end{datetbl}
@@
 \begin{datetbl}
-1995--           & Member, American Psychological Association \\
-1998--           & Member, Gerontological Society of America \\
-1998--           & Member, American Geriatrics Society \\
-2000--           & Associate Editor, Psychology and Aging \\ 
-2003--           & Board of Advisors, Senior Services of Eastern Missouri \\
-2003--05         & NIH Peer Review Committee: Psychobiology of Aging, ad hoc reviewer \\
-2007--11         & NIH Risk, Adult Addictions Study Section, members \\
+2010--2014 		 & System Analyst and Software Developer, KB Group UK, Dhaka, Bangladesh. \\
+2015--2016 		 & Research Volunteer, Henry Ford Health Systems, Detroit, MI. \\
+2016--2016		 & Software Engineer Summer Intern, Title Source Inc., Detroit, MI. \\
+2017--           & Reviewer, Annual Symposium of American Medical Informatics Association. \\
+2017--           & Member, Association for Computing Machinery. \\
 \end{datetbl}
@@
 \begin{datetbl}
-2003            & Outstanding Young Faculty Award, Washington University, St.\ Louis, MO \\
-2004            & Excellence in Teaching, Washington University, St.\ Louis, MO \\
-2009            & Award for Best in Interdisciplinary Ethnography, International Ethnographic Society \\
+2003            & Excellence in H.S.C results, Islamic Foundation, Ministry of Religious Affairs, Bangladesh. \\
+2004			& University Merit Scholarship, Bangladesh University of Engineering and Technology. \\
+2014            & Thomas C. Rumble Fellowship, Wayne State University, Detroit, MI. \\
 \end{datetbl}
@@
 
-\section{Contribution to Science}
+\section{C. Contribution to Science}
 
@@
 
-\item My early publications directly addressed the fact that substance abuse is
-    often overlooked in older adults. However, because many older adults were
-    raised during an era of increased drug and alcohol use, there are reasons
-    to believe that this will become an increasing issue as the population
-    ages.   These publications found that older adults appear in a variety of
-    primary care settings or seek mental health providers to deal with emerging
-    addiction problems.  These publications document this emerging problem but
-    guide primary care providers and geriatric mental health providers to
-    recognize symptoms, assess the nature of the problem and apply the
-    necessary interventions.   By providing evidence and simple clinical
-    approaches, this body of work has changed the standards of care for
-    addicted older adults and will continue to provide assistance in relevant
-    medical settings well into the future.  I served as the primary
-    investigator or co-investigator in all of these studies.
-
-\begin{enumerate}
-
-\item Gryczynski, J., Shaft, B.M., Merryle, R., \& Hunt, M.C. (2002). Community
-        based participatory research with late-life addicts. American Journal
-        of Alcohol and Drug Abuse, 15(3), 222--238.
-
-\item Shaft, B.M., Hunt, M.C., Merryle, R., \& Venturi, R. (2003). Policy
-        implications of genetic transmission of alcohol and drug abuse in
-        female nonusers. International Journal of Drug Policy, 30(5), 46--58.
-
-\item Hunt, M.C., Marks, A.E., Shaft, B.M., Merryle, R., \& Jensen, J.L.
-        (2004). Early-life family and community characteristics and late-life
-        substance abuse. Journal of Applied Gerontology, 28(2),26--37.
-
-\item Hunt, M.C., Marks, A.E., Venturi, R., Crenshaw, W. \& Ratonian, A.
-        (2007). Community-based intervention strategies for reducing alcohol
-        and drug abuse in the elderly.  Addiction, 104(9), 1436--1606. PMCID:
-        PMC9000292
-
-\end{enumerate}
-
-\item In addition to the contributions described above, with a team of
-    collaborators, I directly documented the effectiveness of various
-    intervention models for older substance abusers and demonstrated the
-    importance of social support networks.   These studies emphasized
-    contextual factors in the etiology and maintenance of addictive disorders
-    and the disruptive potential of networks in substance abuse treatment. This
-    body of work also discusses the prevalence of alcohol, amphetamine, and
-    opioid abuse in older adults and how networking approaches can be used to
-    mitigate the effects of these disorders.
-
-\begin{enumerate}
-
-\item Hunt, M.C., Merryle, R. \& Jensen, J.L. (2005). The effect of social
-        support networks on morbidity among elderly substance abusers. Journal
-        of the American Geriatrics Society, 57(4), 15--23.
-
-\item Hunt, M.C., Pour, B., Marks, A.E., Merryle, R. \& Jensen, J.L. (2005).
-        Aging out of methadone treatment. American Journal of Alcohol and Drug
-        Abuse, 15(6), 134--149.
-
-\item Merryle, R. \& Hunt, M.C. (2007). Randomized clinical trial of cotinine
-        in older nicotine addicts. Age and Ageing, 38(2), 9--23. PMCID:
-        PMC9002364
-
-\end{enumerate}
-
-\item Methadone maintenance has been used to treat narcotics addicts for many
-    years but I led research that  has shown that over the long-term, those in
-    methadone treatment view themselves negatively and they gradually begin to
-    view treatment as an intrusion into normal life.   Elderly narcotics users
-    were shown in carefully constructed ethnographic studies to be especially
-    responsive to tailored social support networks that allow them to
-    eventually reduce their maintenance doses and move into other forms of
-    therapy.  These studies also demonstrate the policy and commercial
-    implications associated with these findings.
-
-\begin{enumerate}
-
-\item Hunt, M.C. \& Jensen, J.L. (2003). Morbidity among elderly substance
-        abusers. Journal of the Geriatrics, 60(4), 45--61.
-
-\item Hunt, M.C. \& Pour, B. (2004). Methadone treatment and personal
-        assessment. Journal Drug Abuse, 45(5), 15--26.
-
-\item  Merryle, R. \& Hunt, M.C. (2005). The use of various nicotine delivery
-        systems by older nicotine addicts. Journal of Ageing, 54(1), 24--41.
-        PMCID: PMC9112304
-
-\item Hunt, M.C., Jensen, J.L. \& Merryle, R. (2008). The aging addict:
-        ethnographic profiles of the elderly drug user.  NY, NY: W. W. Norton
-        \& Company.
-
-\end{enumerate}
+Manual encoding has been used to annotate clinical text of motivational interview transcripts for many years, which is time consuming and resource intensive.
+However, my reserach shown that state-of-the-art machine learning methods can be used as a cheap alternate for the annotation that can facilitate researchers to establish a causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
 
@@
 \subsection*{Complete List of Published Work in MyBibliography:} 
-\url{http://www.ncbi.nlm.nih.gov/myncbi/browse/collection/45972964/}
+\url{https://www.ncbi.nlm.nih.gov/sites/myncbi/1H1Fbtu2l5kQn/bibliography/53130053/public/?sort=date}
 
@@
 
-\section{Research Support}
+\section{D. Additional Information: Research Support and/or Scholastic Performance}
 
@@
 
-\grantinfo{R01 DA942367}{Hunt (PI)}{09/01/08--08/31/16}
-{Health trajectories and behavioral interventions among older substance abusers}
-{The goal of this study is to compare the effects of two substance abuse interventions on health 
-outcomes in an urban population of older opiate addicts.}
-{Role: PI}
-
-\bigskip
-
-\grantinfo{R01 MH922731}{Merryle (PI)}{12/15/07--11/30/15}
-{Physical disability, depression and substance abuse in the elderly}
-{The goal of this study is to identify disability and depression trajectories and demographic factors 
-associated with substance abuse in an independently-living elderly population.}
-{Role: Co-Investigator}
-
-\bigskip
-
-\grantinfo{Faculty Resources Grant, Washington University}{08/15/09--08/14/15}
-{Opiate Addiction Database}
-{The goal of this project is to create an integrated database of demographic, social and biomedical 
-information for homeless opiate abusers in two urban Missouri locations, using a number of state and 
-local data sources.}
-{Role: PI}
+\grantinfo{R21 DK108071}{Idalski Carcone \& Kotov (PIs)}{2016--2018}
+{Automated Coding of eCoaching Exchanges to Promote Healthier Eating.}
+{The goal of the study is to automate qualitative communication coding using machine-learning computer science technology.}
+{Role: Graduate Research Assistant}
 
@@
 
-\grantinfo{R21 AA998075}{Hunt (PI)}{01/01/11--12/31/13}
-{Community-based intervention for alcohol abuse}
-{The goal of this project was to assess a community-based strategy for reducing alcohol abuse among 
-older individuals.}
-{Role: PI}
+\grantinfo{Ford Motor Company}{Kotov (PI)}{2014--2015}
+{Extraction of actionable insights from online automotive consumer reviews.}
+{The goal of this project was to extract actionable insights from online automotive consumer reviews.}
+{Role: Graduate Research Assistant}
+
+\bigskip
+
+\grantinfo{R21 DK100760}{Idalski Carcone (PI)}{2014--2016}
+{Patient-Provider Communication to Promote Health Behavior Change in African American Adolescents.}
+{The goal of this study was to identify provider communication behaviors that predict motivation for behavior change and are related to changes in weight-related health behaviors.}
+{Role: Graduate Research Assistant}
+
+\bigskip
+
+\grantinfo{NHLBI 1U01HL097889-01}{Naar--King \& Jen (PIs)}{2014-2016}
+{Provider Communication Behaviors that Predict Motivation to Change in Black Adolescents with Obesity.}
+{The goal of this research was to identify communication behaviors used by weight loss counselors that mostly strongly predicted Black adolescents’ motivational statements.}
+{Role: Graduate Research Assistant}
+
+\bigskip
+
+\grantinfo{Office of Vice President for Research at Wayne State University}{2014--2016}{}
+{Electronic Health Records Information Extraction for Improving the Study of Cancer Risk and Outcomes.}
+{The goal of the research was to extract information from Electronic Health Records for improving the study of cancer risk and outcomes.}
+{Role: Graduate Research Assistant}
 

### 2017-08-23 deeplearning/obesity-lstm-model.py
added changes for markov chain

@@ -4,7 +4,7 @@ import numpy
 import argparse
 from sklearn.model_selection import train_test_split
 from keras.models import Sequential
-from keras.layers import LSTM, Dense, Dropout
+from keras.layers import LSTM, Dense, Embedding, Dropout
 from keras.layers.convolutional import Conv1D
 from keras.layers.convolutional import MaxPooling1D
 from keras.regularizers import l1_l2
@@ -37,7 +37,6 @@ codebook_filename = args.codebook
 model_path = args.model_path
 
 codebook = utility.loadCodeBook(codebook_filename)
-print codebook
 X, y, seq_len = utility.readSequenceFromFile(data_filename, codebook)
 
 # test_X, test_y, max_len = utility.readSequenceFromFile(testing_filename, codebook, seq_len, False)
@@ -56,6 +55,7 @@ def getKFoldsResults(kFolds=10):
 
         # create and fit the model
         batch_size = 1
+
         model = Sequential()
         model.add(
             Conv1D(filters=32, kernel_size=3, padding='same', activation='relu', input_shape=(X_train.shape[1], 1)))

### 2017-08-23 hmm-python/utility.py

@@ -86,7 +86,7 @@ def getHMMModel(n_states, n_observations, sequences, seq_lengths):
     emission_probability = emission_probability / n_observations
 
     # create model and set initial values
-    model = hmm.MultinomialHMM(n_components=n_states)
+    model = hmm.MultinomialHMM(n_components=n_states, random_state=42)
     model.startprob_ = start_probability
     model.transmat_ = transition_probability
     model.emissionprob_ = emission_probability
@@ -137,6 +137,7 @@ def getMacroAveragePerformance(actual, predicted):
 
     return accuracy, precision / 2, recall / 2, f_measure / 2
 
+
 #############################################################################################
 def getMicroAveragePerformance(actual, predicted):
     precision = 0.0
@@ -180,6 +181,7 @@ def getMicroAveragePerformance(actual, predicted):
 
     return accuracy, precision, recall, f_measure
 
+
 #############################################################################################
 def createTrainAndTestFile(data, kFolds, training_filename, testing_filename):
     foldSize = len(data) / kFolds
@@ -202,3 +204,17 @@ def readAllData(data_filename):
         for line in filestream:
             data.append(line)
     return data
+
+
+#############################################################################################
+def AUC(y_true, y_pred):
+    not_y_pred = np.logical_not(y_pred)
+    y_int1 = y_true * y_pred
+    y_int0 = np.logical_not(y_true) * not_y_pred
+    TP = np.sum(y_pred * y_int1)
+    FP = np.sum(y_pred) - TP
+    TN = np.sum(not_y_pred * y_int0)
+    FN = np.sum(not_y_pred) - TN
+    TPR = np.float(TP) / (TP + FN)
+    FPR = np.float(FP) / (FP + TN)
+    return ((1 + TPR - FPR) / 2)

### 2017-08-23 mchain-python/obesity-mchain-tran.py

@@ -28,6 +28,7 @@ micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
 foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
 n_order = 2
+sampling = "under"
 
 for k in np.arange(0, kFolds):
     # get train and test data

### 2017-08-23 mchain-python/obesity-mchain-best.py
clean code

@@ -8,7 +8,8 @@ import random
 parser = argparse.ArgumentParser(description='Train Markov Model.')
 parser.add_argument('--codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
                     help='File location containing codebook.')
-parser.add_argument('--sampling', default='oversampling', type=str, help='dropout rate parameter.')
+parser.add_argument('--sampling', default='under', type=str, help='sampling parameter.')
+parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
 
 args = parser.parse_args()
 
@@ -16,6 +17,7 @@ args = parser.parse_args()
 # Load up training data
 codebook_filename = args.codebook
 sampling = args.sampling
+n_order = args.order
 training_filename = "train.txt"
 testing_filename = "test.txt"
 
@@ -25,7 +27,6 @@ macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
 foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
-n_order = 1
 
 for k in np.arange(0, kFolds):
     # get train and test data
@@ -110,7 +111,7 @@ for k in np.arange(0, kFolds):
 print "\nMacro average results: ", (np.mean(macro_results, axis=0))
 print "\nMicro average results: ", (np.mean(micro_results, axis=0))
 
-f = open("best.txt", "a")
+f = open("results.txt", "a")
 f.write(str(n_order) + ",")
 for x in np.mean(macro_results, axis=0):
     f.write(str(x) + ",")

### 2017-08-23 mchain-python/obesity-mchain-tran.py

@@ -10,7 +10,8 @@ from collections import Counter
 parser = argparse.ArgumentParser(description='Train Markov Model.')
 parser.add_argument('--codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
                     help='File location containing codebook.')
-parser.add_argument('--sampling', default='oversampling', type=str, help='dropout rate parameter.')
+parser.add_argument('--sampling', default='under', type=str, help='sampling parameter.')
+parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
 
 args = parser.parse_args()
 
@@ -18,6 +19,7 @@ args = parser.parse_args()
 # Load up training data
 codebook_filename = args.codebook
 sampling = args.sampling
+n_order = args.order
 training_filename = "train.txt"
 testing_filename = "test.txt"
 
@@ -27,8 +29,7 @@ macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
 foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
-n_order = 2
-sampling = "under"
+
 
 for k in np.arange(0, kFolds):
     # get train and test data
@@ -115,7 +116,7 @@ for k in np.arange(0, kFolds):
 print "\nMacro average results: ", (np.mean(macro_results, axis=0))
 print "\nMicro average results: ", (np.mean(micro_results, axis=0))
 
-f = open("best.txt", "a")
+f = open("results.txt", "a")
 f.write(str(n_order) + ",")
 for x in np.mean(macro_results, axis=0):
     f.write(str(x) + ",")

### 2017-08-23 hmm-python/hmm-test.py
clean code

@@ -1,62 +1,0 @@
-import numpy as np
-from hmmlearn import hmm
-
-states = ["Rainy", "Sunny"]
-n_states = len(states)
-
-observations = ["walk", "shop", "clean"]
-n_observations = len(observations)
-
-start_probability = np.ones(n_states)
-start_probability = start_probability/n_states
-
-#transition_probability = np.np.zeros(n_states*n_states).reshape(n_states, n_states)
-#transition_probability = np.identity(n_states)
-transition_probability = np.ones(n_states*n_states).reshape(n_states, n_states)
-transition_probability = transition_probability/n_states
-
-#emission_probability = np.zeros(n_states*n_observations).reshape(n_states, n_observations)
-emission_probability = np.ones(n_states*n_observations).reshape(n_states, n_observations)
-emission_probability = emission_probability/n_observations
-
-print "\nBefore model fitting..."
-print start_probability
-print transition_probability
-print emission_probability
-
-# create model and set initial values
-model = hmm.MultinomialHMM(n_components=n_states)
-model.startprob_=start_probability
-model.transmat_=transition_probability
-model.emissionprob_=emission_probability
-
-# prepare data
-sequences = np.array([[[0], [2], [1], [1], [2], [0]], [[0], [1], [1], [0], [2]], [[2], [1], [1]], [[1], [1], [1], [0]]])
-seq_lengths = np.array([6, 5, 3, 4])
-data = sequences[0]
-for i in range(1, len(sequences)):
-  data = np.concatenate([data, sequences[i]])
-sequences = data
-
-print sequences
-print seq_lengths
-
-# fit model
-model = model.fit(sequences, seq_lengths)
-
-print "\nAfter model fitting..."
-print model.startprob_
-print model.transmat_
-print model.emissionprob_
-
-# get log likelihood for the given sequence(s)
-test_seq = np.array([[0], [2], [1], [1], [2], [0]])
-logL = model.score(test_seq)
-print "\nLog likelihood of the sequence: ", logL, "\n"
-
-# decode sequences
-logprob, state_sequences = model.decode(test_seq, algorithm="viterbi")
-print "\nLog probability of the produced state sequence: ", logprob, "\n"
-print("Observed sequences:", ", ".join(map(lambda x: observations[int(x)], test_seq)))
-print("State sequences:", ", ".join(map(lambda x: states[int(x)], state_sequences)))
-

### 2017-08-23 hmm-python/obesity-hmm-folds.py

@@ -1,95 +1,0 @@
-import numpy as np
-import utility
-import argparse
-
-#############################################################################################
-# Parse command line arguments
-parser = argparse.ArgumentParser(description='Train HMM Model.')
-# parser.add_argument('-data',
-#                     default='/home/mehedi/teana/data-source/seq-analysis/hmm/unbalanced/cht-cml/allsequence.txt',
-#                     help='File location containing training sequence.')
-# parser.add_argument('-output_directory', default='/home/mehedi/teana/data-source/seq-analysis/hmm/',
-#                     help='Directory to save results.')
-
-parser.add_argument('-data',
-                    default='/home/mehedi/teana/data-source/seq-analysis/hmm/balanced/cht-cml/over-sampling-reg-old.txt',
-                    help='File location containing training sequence.')
-parser.add_argument('-output_directory', default='/home/mehedi/teana/data-source/seq-analysis/hmm/',
-                    help='Directory to save results.')
-
-args = parser.parse_args()
-
-############################################################################################
-# Load up training data
-data_filename = args.data
-output_directory = args.output_directory
-training_filename = output_directory + "train.txt"
-testing_filename = output_directory + "test.txt"
-
-# read entire data to list
-data = utility.readAllData(data_filename)
-
-# get results fro k folds
-kFolds = 10
-macro_results = []
-micro_results = []
-for k in np.arange(0, kFolds):
-    # create training and testing file
-    utility.createTrainAndTestFile(data, kFolds, training_filename, testing_filename)
-
-    # determine codebook size and number of hidden states
-    n_states = 5
-    success_codebook, unsuccess_codebook = utility.loadCodeBookFromTrainingFile(training_filename)
-
-    # fit successful model
-    n_observations = len(success_codebook)
-    sequences, seq_labels, seq_lengths = utility.loadData(training_filename, success_codebook, 1)
-    success_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
-
-    # fit unsuccessful model
-    n_observations = len(unsuccess_codebook)
-    sequences, seq_labels, seq_lengths = utility.loadData(training_filename, unsuccess_codebook, 0)
-    unsuccess_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
-
-    # get log likelihood for the given test sequence(s)
[…]

### 2017-08-23 hmm-python/obesity-hmm.py

@@ -1,88 +1,0 @@
-import numpy as np
-import utility
-import argparse
-
-#############################################################################################
-# Parse command line arguments
-parser = argparse.ArgumentParser(description='Train HMM Model.')
-parser.add_argument('-training_data', default='/home/mehedi/teana/data-source/seq-analysis/hmm/balanced/cht-cml/train_shuffled.txt'
-                    , help='File location containing training sequence.')
-parser.add_argument('-testing_data', default='/home/mehedi/teana/data-source/seq-analysis/hmm/balanced/cht-cml/test.txt',
-                    help='File location containing testing sequence.')
-parser.add_argument('-codebook', default='/home/mehedi/teana/data-source/seq-analysis/hmm/codebook.txt',
-                    help='File location containing codebook.')
-parser.add_argument('-output_directory', default='/home/mehedi/teana/data-source/seq-analysis/hmm/',
-                    help='Directory to save results.')
-
-args = parser.parse_args()
-
-############################################################################################
-# Load up training data
-training_filename = args.training_data
-testing_filename = args.testing_data
-codebook_filename = args.codebook
-output_directory = args.output_directory
-
-kFolds = 5
-results = []
-for k in np.arange(0, kFolds):
-
-    # determine codebook size and number of hidden states
-    n_states = 5
-    success_codebook, unsuccess_codebook = utility.loadCodeBookFromTrainingFile(training_filename)
-
-    print "\n Success: ", success_codebook,
-    print "\n Unsuccess: ", unsuccess_codebook
-
-    # fit successful model
-    n_observations = len(success_codebook)
-    sequences, seq_labels, seq_lengths = utility.loadData(training_filename, success_codebook, 1)
-    success_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
-
-    print "\nAfter model fitting..."
-    print success_model.startprob_
-    print success_model.transmat_
-    print success_model.emissionprob_
-
-    # fit unsuccessful model
-    n_observations = len(unsuccess_codebook)
-    sequences, seq_labels, seq_lengths = utility.loadData(training_filename, unsuccess_codebook, 0)
-    unsuccess_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
-
-    print "\nAfter model fitting..."
-    print unsuccess_model.startprob_
-    print unsuccess_model.transmat_
[…]

### 2017-08-23 hmm-python/utility.py

@@ -3,7 +3,10 @@ import random
 import re
 import operator
 from hmmlearn import hmm
-
+from keras.preprocessing.sequence import pad_sequences
+from imblearn.over_sampling import SMOTE
+from collections import Counter
+from imblearn.under_sampling import ClusterCentroids
 
 ###################################################################
 # read codebook from file
@@ -17,6 +20,7 @@ def loadCodeBook(codebook_filename):
 
 ###########################################################################################
 def loadCodeBookFromTrainingFile(fileLocation):
+    all_code_map = {}
     success_code_map = {}
     unsuccess_code_map = {}
     with open(fileLocation, "r") as filestream:
@@ -25,21 +29,28 @@ def loadCodeBookFromTrainingFile(fileLocation):
             if l[len(l) - 1] == "500":
                 for i in range(0, len(l) - 1):
                     success_code_map[l[i]] = l[i]
+		    all_code_map[l[i]] = l[i]
             elif l[len(l) - 1] == "400":
                 for i in range(0, len(l) - 1):
                     unsuccess_code_map[l[i]] = l[i]
+                    all_code_map[l[i]] = l[i]
 
     success_sorted_map = sorted(success_code_map.items(), key=operator.itemgetter(0))
     unsuccess_sorted_map = sorted(unsuccess_code_map.items(), key=operator.itemgetter(0))
+    all_sorted_map = sorted(all_code_map.items(), key=operator.itemgetter(0))
 
     success_map = []
     unsuccess_map = []
+    all_map = []
+    
+    for key, value in all_sorted_map:
+        all_map.append(key)
     for key, value in success_sorted_map:
         success_map.append(key)
     for key, value in unsuccess_sorted_map:
         unsuccess_map.append(key)
 
-    return success_map, unsuccess_map
+    return all_map, success_map, unsuccess_map
 
 
 ###########################################################################################
@@ -53,14 +64,14 @@ def loadData(fileLocation, codebook, flag):
         for line in filestream:
             l = re.sub(r"\s+", "", line).split(",")
             if (flag == 1 and l[len(l) - 1] == "500") or (flag == 0 and l[len(l) - 1] == "400"):
-                for i in range(0, len(l) - 1):
+                for i in range(0, len(l)-1):
                     seq.append([code_to_int[l[i]]])
                     map[l[i]] = l[i]
                 lengths.append(len(l) - 1)
                 seq_label.append(l[len(l) - 1])
[…]

### 2017-08-23 hmm-python/obesity-hmm-best.py
clean code

@@ -1,0 +1,111 @@
+import numpy as np
+import utility
+import argparse
+import random
+
+#############################################################################################
+# Parse command line arguments
+parser = argparse.ArgumentParser(description='Train HMM Model.')
+parser.add_argument('--hidden_states', default=5, type=int, help='number of hidden states')
+parser.add_argument('--codebook',
+                    default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
+                    help='File location containing codebook.')
+parser.add_argument('--sampling', default='over', type=str, help='dropout rate parameter.')
+
+args = parser.parse_args()
+
+############################################################################################
+# Load up training data
+codebook_filename = args.codebook
+sampling = args.sampling
+hidden_states = args.hidden_states
+training_filename = "train.txt"
+testing_filename = "test.txt"
+
+# get results fro k folds
+kFolds = 10
+macro_results = []
+micro_results = []
+codebook = utility.loadCodeBook(codebook_filename)
+foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
+
+for k in np.arange(0, kFolds):
+    # get train and test data
+    utility.createUnderOrOverSample(sampling, foldData[k][0], testing_filename, max_len, codebook)
+    utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
+
+    # determine codebook size and number of hidden states
+    all_codebook, success_codebook, unsuccess_codebook = utility.loadCodeBookFromTrainingFile(training_filename)
+
+    # fit successful model
+    n_states = 5
+    n_observations = len(success_codebook)
+    sequences, seq_labels, seq_lengths = utility.loadData(training_filename, success_codebook, 1)
+    success_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
+
+    # fit unsuccessful model
+    n_states = 22
+    n_observations = len(unsuccess_codebook)
+    sequences, seq_labels, seq_lengths = utility.loadData(training_filename, unsuccess_codebook, 0)
+    unsuccess_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
+
+    # get log likelihood for the given test sequence(s)
+    sequences_for_success_model, seq_labels, success_seq_lengths = utility.loadData(testing_filename, success_codebook,
+                                                                                    2)
[…]

### 2017-08-23 hmm-python/runcode.sh

@@ -1,0 +1,5 @@
+
+for (( i=1; i <= 60; ++i ))
+do
+    python obesity-hmm-best.py
+done

### 2017-08-23 hmm-python/utility.py

@@ -318,7 +318,7 @@ def createUnderOrOverSample(method, given_data, outputdata_filename, max_len, co
     # perform over or under sampling
     X_d = []
     y_res = []
-    if method == "oversampling":
+    if method == "over":
         sm = SMOTE(kind='borderline2')
         X_res, y_res = sm.fit_sample(X_norm, y_norm)
     else:

### 2017-08-23 deeplearning/obesity-lstm-model.py
added source code of deep learning for sequence classification

@@ -1,107 +1,0 @@
-# LSTM with Variable Length Input Sequences to Output as successful or unsuccessful
-import utility
-import numpy
-import argparse
-from sklearn.model_selection import train_test_split
-from keras.models import Sequential
-from keras.layers import LSTM, Dense, Embedding, Dropout
-from keras.layers.convolutional import Conv1D
-from keras.layers.convolutional import MaxPooling1D
-from keras.regularizers import l1_l2
-from keras.callbacks import EarlyStopping, ModelCheckpoint
-from keras.models import load_model
-
-#############################################################################################
-# Parse command line arguments
-parser = argparse.ArgumentParser(description='Train LSTM Sequential Model.')
-parser.add_argument('-data',
-                    default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/obesity-newfile/balanced/cht-cml/over-sampling-reg-new.txt'
-                    , help='File location containing training sequence.')
-parser.add_argument('-codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/obesity-newfile/codebook-newfile.txt',
-                    help='File location containing codebook.')
-
-# parser.add_argument('-data',
-#                     default='/home/mehedi/teana/data-source/seq-analysis/over-sampling-reg-old-test-post.txt'
-#                     , help='File location containing training sequence.')
-# parser.add_argument('-codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook.txt',
-#                    help='File location containing codebook.')
-parser.add_argument('-model_path', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/obesity-newfile/model.h5',
-                    help='Directory to save model.')
-
-args = parser.parse_args()
-
-#############################################################################################
-# Load up training data
-data_filename = args.data
-codebook_filename = args.codebook
-model_path = args.model_path
-
-codebook = utility.loadCodeBook(codebook_filename)
-X, y, seq_len = utility.readSequenceFromFile(data_filename, codebook)
-
-# test_X, test_y, max_len = utility.readSequenceFromFile(testing_filename, codebook, seq_len, False)
-# display test results one by one
-# utility.showResultsForTestData(trained_model, codebook, testing_filename, seq_len)
-
-##############################################################################################
-# get results for K folds
[…]

### 2017-08-23 deeplearning/obesity_gru_lstm.py

@@ -1,0 +1,239 @@
+# ***************************************************************
+# Author: Md Mehedi Hasan
+# Wayne State University
+# email: mehedi@wayne.edu
+# ***************************************************************
+
+# Import necessary packages
+import utility
+import numpy as np
+import math
+import argparse
+import random
+from keras.preprocessing.sequence import pad_sequences
+from keras.utils import np_utils
+import tensorflow as tf
+from sklearn.metrics import accuracy_score
+
+# Parse command line arguments
+parser = argparse.ArgumentParser(description='Train RNN Model.')
+parser.add_argument('--codebook', default='codebook.txt',
+                    help='File location containing codebook or vocabulary.')
+parser.add_argument('--hidden_units', default=20, type=int, help='no. of hidden units in hidden layer.')
+parser.add_argument('--batch_size', default=4, type=int, help='batch size of the training data.')
+parser.add_argument('--embed_dimension', default=45, type=int, help='embedding dimension parameter.')
+parser.add_argument('--keep_rate', default=1.0, type=float, help='keep rate or (1 - dropout) rate parameter.')
+parser.add_argument('--alpha', default=0.50, type=float, help='alpha for loss optimization.')
+parser.add_argument('--beta', default=0.03, type=float, help='beta parameter.')
+parser.add_argument('--sampling', default='under', type=str, help='sampling parameter.')
+parser.add_argument('--n_epoch', default=100, type=int, help='no. of epoch parameter.')
+parser.add_argument('--min_change', default=0.0015, type=float, help='minimum change parameter.')
+args = parser.parse_args()
+
+# Load up training data
+codebook_filename = args.codebook
+hidden_units = args.hidden_units
+batch_size = args.batch_size
+embed_dimension = args.embed_dimension
+keep_rate = args.keep_rate
+alpha = args.alpha
+beta = args.beta
+sampling = args.sampling
+training_epochs = args.n_epoch
+min_change = args.min_change
+
+# Set initial value of some parameters
+n_steps = 130
+n_classes = 2
+learning_rate = 0.00005
+wait_epoch = 5
+graph_level_seed = 42
+operation_level_seed = 72
+
+# Set randomness deterministic
+np.random.seed(operation_level_seed)
+random.seed(operation_level_seed)
+
+# Read codebook or vocabulary
+codebook = utility.loadCodeBook(codebook_filename)
+
+
[…]

### 2017-08-23 deeplearning/over-under-sampling.py

@@ -1,69 +1,0 @@
-import utility
-import numpy
-import argparse
-from keras.preprocessing.sequence import pad_sequences
-from imblearn.over_sampling import SMOTE
-from imblearn.under_sampling import ClusterCentroids
-from collections import Counter
-
-#############################################################################################
-# parse command line arguments
-parser = argparse.ArgumentParser(description='Train LSTM Sequential Model.')
-parser.add_argument('-data',
-                    default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/balanced/cht-cml/allsequence.txt'
-                    , help='File location containing training sequence.')
-parser.add_argument('-codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook.txt',
-                    help='File location containing codebook.')
-
-args = parser.parse_args()
-#############################################################################################
-# Load up training data
-data_filename = args.data
-codebook_filename = args.codebook
-
-#############################################################################################
-# normalize data
-def normalizeData(dataX, dataY, codebook, max_len):
-    # convert list of lists to array and pad sequences if needed
-    X = pad_sequences(dataX, maxlen=max_len, dtype='float32')
-
-    # normalize
-    X = X / (float(len(codebook)) - 2)
-
-    # get codebook size
-    codebookSize = len(codebook) - 2
-
-    # get label as 0 and 1
-    dataY[:] = [x - codebookSize for x in dataY]
-
-    # one hot encode the output variable
-    y = numpy.array(dataY)
-
-    return X, y, max_len
-
-
-#############################################################################################
-codebook = utility.loadCodeBook(codebook_filename)
-X, y, seq_len = utility.readSequenceFromFile(data_filename, codebook)
-X_norm, y_norm, max_len = normalizeData(X, y, codebook, seq_len)
-
-# perform over or under sampling
-sm = SMOTE(random_state=72, kind='regular')
-#sm = ClusterCentroids(random_state=42)
-X_res, y_res = sm.fit_sample(X_norm, y_norm)
-
-# save data to a csv file
-out = open('/home/mehedi/teana/data-source/seq-analysis/over-sampling-reg-old-test.csv', 'w')
-idx = 0
-for row in X_res:
-    for column in row:
-        out.write('%f,' % column)
-    out.write('%d' % y_res[idx])
-    idx += 1
-    out.write('\n')
-out.close()
-
-# display new sample size
-print Counter(y_res)
-
-#############################################################################################

### 2017-08-23 deeplearning/post-process-sample-data.py

@@ -1,32 +1,0 @@
-import utility
-import numpy
-import argparse
-
-#############################################################################################
-# parse command line arguments
-parser = argparse.ArgumentParser(description='Train LSTM Sequential Model.')
-parser.add_argument('-inputdata',
-                    default='/home/mehedi/Desktop/Link to teana/data-source/seq-analysis/over-sampling-reg-old-test.csv'
-                    , help='File location containing input data sequence.')
-parser.add_argument('-outputdata',
-                    default='/home/mehedi/Desktop/Link to teana/data-source/seq-analysis/over-sampling-reg-old-test-post.txt'
-                    , help='File location containing out data sequence.')
-parser.add_argument('-codebook', default='/home/mehedi/Desktop/Link to teana/data-source/seq-analysis/deepLearn/codebook.txt',
-                    help='File location containing codebook.')
-
-args = parser.parse_args()
-
-#############################################################################################
-inputdata_filename = args.inputdata
-outputdata_filename = args.outputdata
-codebook_filename = args.codebook
-codebook = utility.loadCodeBook(codebook_filename)
-
-# load data and denormalize
-X = numpy.genfromtxt(inputdata_filename, delimiter=",")
-X = utility.denormalizeData(X, codebook)
-
-# write data into file
-utility.writeSequenceFromPaddedSequence(X, codebook, outputdata_filename)
-
-#############################################################################################

### 2017-08-23 deeplearning/preprocess.py

@@ -1,16 +1,0 @@
-import utility
-
-
-# rawFilePath = "/home/mehedi/teana/data-source/seq-analysis/rawFile/"
-# codeMapping = "/home/mehedi/teana/data-source/seq-analysis/codemap-newfile.txt"
-# utility.createSequence(rawFilePath, codeMapping)
-
-#inputFile = "/home/mehedi/teana/data-source/seq-analysis/deepLearn/balanced/cht-cml/train.txt"
-#outputFile = "/home/mehedi/teana/data-source/seq-analysis/deepLearn/balanced/cht-cml/train_shuffled.txt"
-#utility.writeShuffledData(inputFile, outputFile)
-
-inputFile = "/home/mehedi/teana/data-source/seq-analysis/hmm/obesity-newfile/balanced/allsequence.txt"
-outputFile = "/home/mehedi/teana/data-source/seq-analysis/hmm/obesity-newfile/balanced/allsequence_bal.txt"
-utility.writeShuffledData(inputFile, outputFile)
-#utility.writeBalancedData(inputFile, outputFile, 742)
-

### 2017-08-23 deeplearning/seq-gru-model.py

@@ -1,77 +1,0 @@
-# LSTM with Variable Length Input Sequences to One Character Output
-import numpy
-from keras.models import Sequential
-from keras.layers import Dense
-from keras.layers import GRU
-from keras.utils import np_utils
-from keras.preprocessing.sequence import pad_sequences
-
-# fix random seed for reproducibility
-numpy.random.seed(7)
-
-# define the raw dataset
-alphabet = "ABCDEFGHIJKLMNOPQRSTUVWXYZ"
-
-# create mapping of characters to integers (0-25) and the reverse
-char_to_int = dict((c, i) for i, c in enumerate(alphabet))
-int_to_char = dict((i, c) for i, c in enumerate(alphabet))
-
-# prepare the dataset of input to output pairs encoded as integers
-num_inputs = 1000
-max_len = 5
-dataX = []
-dataY = []
-for i in range(num_inputs):
-    start = numpy.random.randint(len(alphabet) - 2)
-    end = numpy.random.randint(start, min(start + max_len, len(alphabet) - 1))
-    sequence_in = alphabet[start:end + 1]
-    sequence_out = numpy.random.randint(2)
-
-    if i % 2 == 0 or i % 5 == 0:
-        if sequence_in[-1] >= 'M':
-            sequence_out = 1
-
-        if sequence_in[-1] < 'M':
-            sequence_out = 0
-
-    dataX.append([char_to_int[char] for char in sequence_in])
-    dataY.append(sequence_out)
-    #print sequence_in, '->', sequence_out
-
-# convert list of lists to array and pad sequences if needed
-X = pad_sequences(dataX, maxlen=max_len, dtype='float32')
-
-# reshape X to be [samples, time steps, features]
-X = numpy.reshape(X, (X.shape[0], max_len, 1))
-
-# normalize
-X = X / float(len(alphabet))
-
-# one hot encode the output variable
-y = np_utils.to_categorical(dataY)
-
-# create and fit the model
-batch_size = 64
-model = Sequential()
-model.add(GRU(32, input_shape=(X.shape[1], 1)))
-model.add(Dense(y.shape[1], activation='softmax'))
-model.compile(loss='categorical_crossentropy', optimizer='adam', metrics=['accuracy'])
-print("\nModel fitting...")
-model.fit(X, y, epochs=500, batch_size=batch_size, verbose=0, shuffle=False)
-
-# summarize performance of the model
-scores = model.evaluate(X, y, verbose=0)
[…]

### 2017-08-23 deeplearning/seq-lstm-model.py

@@ -1,167 +1,0 @@
-#####################################################################
-# Stateless
-#####################################################################
-# LSTM with Variable Length Input Sequences to One Character Output
-import numpy
-from keras.models import Sequential
-from keras.layers import Dense
-from keras.layers import LSTM
-from keras.utils import np_utils
-from keras.preprocessing.sequence import pad_sequences
-
-# fix random seed for reproducibility
-numpy.random.seed(7)
-
-# define the raw dataset
-alphabet = "ABCDEFGHIJKLMNOPQRSTUVWXYZ"
-
-# create mapping of characters to integers (0-25) and the reverse
-char_to_int = dict((c, i) for i, c in enumerate(alphabet))
-int_to_char = dict((i, c) for i, c in enumerate(alphabet))
-
-# prepare the dataset of input to output pairs encoded as integers
-num_inputs = 1000
-max_len = 5
-dataX = []
-dataY = []
-for i in range(num_inputs):
-    start = numpy.random.randint(len(alphabet) - 2)
-    end = numpy.random.randint(start, min(start + max_len, len(alphabet) - 1))
-    sequence_in = alphabet[start:end + 1]
-    sequence_out = numpy.random.randint(2)
-
-    if i % 2 == 0 or i % 5 == 0:
-        if sequence_in[-1] >= 'M':
-            sequence_out = 1
-
-        if sequence_in[-1] < 'M':
-            sequence_out = 0
-
-    dataX.append([char_to_int[char] for char in sequence_in])
-    dataY.append(sequence_out)
-    #print sequence_in, '->', sequence_out
-
-# convert list of lists to array and pad sequences if needed
-X = pad_sequences(dataX, maxlen=max_len, dtype='float32')
-
-# reshape X to be [samples, time steps, features]
-X = numpy.reshape(X, (X.shape[0], max_len, 1))
-
-# normalize
-X = X / float(len(alphabet))
-
-# one hot encode the output variable
-y = np_utils.to_categorical(dataY)
-
-# create and fit the model
-batch_size = 64
-model = Sequential()
-model.add(LSTM(32, input_shape=(X.shape[1], 1)))
-model.add(Dense(y.shape[1], activation='softmax'))
-model.compile(loss='categorical_crossentropy', optimizer='adam', metrics=['accuracy'])
-print("\nModel fitting...")
-model.fit(X, y, epochs=500, batch_size=batch_size, verbose=0, shuffle=False)
-
[…]

### 2017-08-23 deeplearning/utility.py

@@ -6,6 +6,11 @@ import random
 import operator
 from keras.utils import np_utils
 from keras.preprocessing.sequence import pad_sequences
+from imblearn.over_sampling import SMOTE
+from collections import Counter
+from imblearn.under_sampling import ClusterCentroids
+
+random.seed(42)
 
 ###################################################################
 # read codebook from file
@@ -113,6 +118,7 @@ def getMacroAveragePerformance(actual, predicted):
 
     return accuracy / 2, precision / 2, recall / 2, f_measure / 2
 
+
 ###################################################################
 # evaluate model with F1, Precision and Recall
 def getMicroAveragePerformance(actual, predicted):
@@ -157,13 +163,14 @@ def getMicroAveragePerformance(actual, predicted):
         if (local_precision + local_recall) > 0:
             local_f_measure = (float(2 * local_precision * local_recall) / (local_precision + local_recall))
 
-        precision += (local_precision*(float(tp + fn)/total_sample))
-        recall += (local_recall*(float(tp + fn)/total_sample))
-        f_measure += (local_f_measure*(float(tp + fn)/total_sample))
+        precision += (local_precision * (float(tp + fn) / total_sample))
+        recall += (local_recall * (float(tp + fn) / total_sample))
+        f_measure += (local_f_measure * (float(tp + fn) / total_sample))
 
         accuracy = (float(tp + tn) / (tp + fp + tn + fn))
 
     return accuracy, precision, recall, f_measure
+
 
 ###################################################################
 def denormalizeData(X, codebook):
@@ -232,50 +239,20 @@ def showResultsForTestData(model, codebook, training_filename, seq_len):
 
 
 ######################################################################
-# pre-process data from given text files
-def createSequence(rawFileLocation, codeMappingFile):
-    count_pos = 0
-    count_neg = 0
-    codebook_dict = getDictionaryForCodeBook(codeMappingFile)
+def getCodeMapping(dataFileLocation):
+    codebook_dict = {}
+    with open(dataFileLocation, "r") as filestream:
+        for line in filestream:
+            line = line.replace("\n", "").strip()
+            currentline = line.split(",")
+            for x in currentline:
+                if x in codebook_dict.keys():
+                    codebook_dict[x] += 1
+                else:
+                    codebook_dict[x] = 1
 
-    f = open("/home/mehedi/teana/data-source/seq-analysis/hmm/obesity-newfile/allsequence.txt", "w")
-    # codebook_dict = {}
[…]

### 2017-08-23 deeplearning/obesity_gru_lstm.py
correct comments

@@ -30,7 +30,7 @@ parser.add_argument('--n_epoch', default=100, type=int, help='no. of epoch param
 parser.add_argument('--min_change', default=0.0015, type=float, help='minimum change parameter.')
 args = parser.parse_args()
 
-# Load up training data
+# Set arguments value
 codebook_filename = args.codebook
 hidden_units = args.hidden_units
 batch_size = args.batch_size

### 2017-08-23 Parameters/batch-size/batch.r
added results of parameter tuning for deep learning

@@ -1,0 +1,12 @@
+#library
+library(ggplot2)
+library(reshape2)
+
+# melt data
+r1 <- read.csv("/home/mehedi/Desktop/Parameters/batch-size/results.csv")
+
+#plot graph
+plot(r1$x1, r1$x3, col='blue', type="l", ylim=c(0.50, 0.61), ylab = "Loss", xlab = "Batch size", lwd=1)
+lines(r1$x1, r1$x5, col='green', type="l", lwd=1)
+lines(r1$x1, r1$x7, col='red', type="l", lwd=1)
+legend('bottomright',c("val_loss", "test_loss", 'train_loss'), col=c('blue', 'green', 'red'), lwd=1)

### 2017-08-23 Parameters/dropout/dropout.r

@@ -1,0 +1,12 @@
+#library
+library(ggplot2)
+library(reshape2)
+
+# melt data
+r1 <- read.csv("/home/mehedi/Desktop/Parameters/dropout/results.csv")
+
+#plot graph
+plot(r1$x1, r1$x3, col='blue', type="l", ylim=c(0.47, 0.70), ylab = "Loss", xlab = "Keep rate", lwd=1)
+lines(r1$x1, r1$x5, col='green', type="l", lwd=1)
+lines(r1$x1, r1$x7, col='red', type="l", lwd=1)
+legend('topright',c("val_loss", "test_loss", 'train_loss'), col=c('blue', 'green', 'red'), lwd=1)

### 2017-08-23 Parameters/embed/embed.r

@@ -1,0 +1,12 @@
+#library
+library(ggplot2)
+library(reshape2)
+
+# melt data
+r1 <- read.csv("/home/mehedi/Desktop/Parameters/embed/results.csv")
+
+#plot graph
+plot(r1$x1, r1$x3, col='blue', type="l", ylim=c(0.53, 0.62), ylab = "Loss", xlab = "Embedding dimension", lwd=1)
+lines(r1$x1, r1$x5, col='green', type="l", lwd=1)
+lines(r1$x1, r1$x7, col='red', type="l", lwd=1)
+legend('topright',c("val_loss", "test_loss", 'train_loss'), col=c('blue', 'green', 'red'), lwd=1)

### 2017-08-23 Parameters/epoch/epoch.r

@@ -1,0 +1,12 @@
+#library
+library(ggplot2)
+library(reshape2)
+
+# melt data
+r1 <- read.csv("/home/mehedi/Desktop/Parameters/epoch/results.csv")
+
+#plot graph
+plot(r1$x1, r1$x3, col='blue', type="l", ylim=c(0.45, 0.70), ylab = "Loss", xlab = "Epoch", lwd=1)
+lines(r1$x1, r1$x5, col='green', type="l", lwd=1)
+lines(r1$x1, r1$x7, col='red', type="l", lwd=1)
+legend('topright',c("val_loss", "test_loss", 'train_loss'), col=c('blue', 'green', 'red'), lwd=1)

### 2017-08-23 Parameters/hidden-states/hiddenunits.r

@@ -1,0 +1,12 @@
+#library
+library(ggplot2)
+library(reshape2)
+
+# melt data
+r1 <- read.csv("/home/mehedi/Desktop/Parameters/hidden-states/results.csv")
+
+#plot graph
+plot(r1$x1, r1$x3, col='blue', type="l", ylim=c(0.50, 0.70), ylab = "Loss", xlab = "# of hidden units", lwd=1)
+lines(r1$x1, r1$x5, col='green', type="l", lwd=1)
+lines(r1$x1, r1$x7, col='red', type="l", lwd=1)
+legend('topright',c("val_loss", "test_loss", 'train_loss'), col=c('blue', 'green', 'red'), lwd=1)

### 2017-08-23 Parameters/regularization/regularization.r

@@ -1,0 +1,12 @@
+#library
+library(ggplot2)
+library(reshape2)
+
+# melt data
+r1 <- read.csv("/home/mehedi/Desktop/Parameters/regularization/results.csv")
+
+#plot graph
+plot(r1$x1, r1$x3, col='blue', type="l", ylim=c(0.51, 0.59), ylab = "Loss", xlab = expression(paste(beta, " for L2 regularization")), lwd=1)
+lines(r1$x1, r1$x5, col='green', type="l", lwd=1)
+lines(r1$x1, r1$x7, col='red', type="l", lwd=1)
+legend('topright',c("val_loss", "test_loss", 'train_loss'), col=c('blue', 'green', 'red'), lwd=1)

### 2017-08-23 split-data/create_data_folds.py
added data splitters

@@ -1,0 +1,28 @@
+import utility
+import argparse
+
+# Parse command line arguments
+parser = argparse.ArgumentParser(description='Create train and test data.')
+parser.add_argument('--codebook', default='codebook.txt', help='File location containing codebook.')
+
+# Read parameters
+args = parser.parse_args()
+codebook_filename = args.codebook
+
+# create K-folds
+def createFolds(codebook, kFolds=10):
+    foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
+
+    for i in range(0, kFolds):
+        # get train and test data
+        utility.createUnderAndOverSample(foldData[i][0], "folds/fold" + str(i + 1) + "/test.txt", max_len,
+                                        codebook)
+        utility.createUnderAndOverSample(foldData[i][1], "folds/fold" + str(i + 1) + "/train.txt", max_len,
+                                        codebook)
+        utility.splitTrainAndValidationSet("under/folds/fold" + str(i + 1) + "/train.txt")
+        utility.splitTrainAndValidationSet("over/folds/fold" + str(i + 1) + "/train.txt")
+
+
+# create folds with over and under sampling
+codebook = utility.loadCodeBook(codebook_filename)
+createFolds(codebook, kFolds=10)

### 2017-08-23 split-data/utility.py

@@ -1,0 +1,477 @@
+# Various function as helper
+import numpy
+import os
+import re
+import random
+import operator
+from keras.utils import np_utils
+from keras.preprocessing.sequence import pad_sequences
+from imblearn.over_sampling import SMOTE
+from collections import Counter
+from imblearn.under_sampling import ClusterCentroids
+from sklearn.model_selection import train_test_split
+
+
+###################################################################
+# read codebook from file
+def loadCodeBook(codebook_filename):
+    codebook = []
+    with open(codebook_filename, "r") as filestream:
+        for line in filestream:
+            codebook.append(line.replace("\n", ""))
+    return codebook
+
+
+###################################################################
+# read all sequences and return X, y
+def readSequenceFromFile(sequence_file, codebook, seq_len=5, is_train=True):
+    max_len = seq_len
+    dataX = []
+    dataY = []
+
+    # create mapping of characters to integers (0-25) and the reverse
+    code_to_int = dict((c, i) for i, c in enumerate(codebook))
+
+    # prepare the dataset of input to output pairs encoded as integers
+    with open(sequence_file, "r") as filestream:
+        for line in filestream:
+            currentline = line.split(",")
+            dataX.append([code_to_int[item] for item in currentline[0:-1]])
+            dataY.append(code_to_int[str(currentline[-1].strip())])
+            if len(currentline) > max_len and is_train:
+                max_len = len(currentline)
+
+    return dataX, dataY, max_len
+
+
+###################################################################
+# normalize and hot encode
+def normalizeData(dataX, dataY, codebook, max_len):
+    # convert list of lists to array and pad sequences if needed
+    X = pad_sequences(dataX, maxlen=max_len, dtype='float32')
+
+    # reshape X to be [samples, time steps, features]
+    X = numpy.reshape(X, (X.shape[0], max_len, 1))
+
+    # normalize
+    X = X / float(len(codebook))
+
+    # get codebook size
+    codebookSize = len(codebook)
+
+    # get label as 0 and 1
+    dataY[:] = [x - codebookSize + 2 for x in dataY]
+
+    # one hot encode the output variable
+    y = np_utils.to_categorical(dataY)
+
+    return X, y, max_len
+
+
+###################################################################
+# evaluate model with F1, Precision and Recall
+def getMacroAveragePerformance(actual, predicted):
+    precision = 0.0
+    recall = 0.0
[…]

### 2017-08-23 mchain-python/obesity-mchain-best.py
fixed error

@@ -1,121 +1,0 @@
-import numpy as np
-import utility
-import argparse
-import random
-
-#############################################################################################
-# Parse command line arguments
-parser = argparse.ArgumentParser(description='Train Markov Model.')
-parser.add_argument('--codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
-                    help='File location containing codebook.')
-parser.add_argument('--sampling', default='under', type=str, help='sampling parameter.')
-parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
-
-args = parser.parse_args()
-
-############################################################################################
-# Load up training data
-codebook_filename = args.codebook
-sampling = args.sampling
-n_order = args.order
-training_filename = "train.txt"
-testing_filename = "test.txt"
-
-# get results fro k folds
-kFolds = 10
-macro_results = []
-micro_results = []
-codebook = utility.loadCodeBook(codebook_filename)
-foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
-
-for k in np.arange(0, kFolds):
-    # get train and test data
-    utility.createUnderOrOverSample(sampling, foldData[k][0], testing_filename, max_len, codebook)
-    utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
-
-    # set dictionary for successful sequences
-    successful_dict = utility.loadDictionary(training_filename, n_order, "500")
-
-    # set dictionary for unsuccessful sequences
-    unsuccessful_dict = utility.loadDictionary(training_filename, n_order, "400")
-
-    # set likelihood of the given test sequence(s) and then classify
-    prediction_labels = []
-    seq_labels = []
-
-    with open(testing_filename, "r") as file_stream:
-        for line in file_stream:
-            words = line.replace("\n", "").split(",")
-            actual_label = words[len(words)-1]
-            seq_labels.append(actual_label)
-            if len(words) <= n_order:
-                print "Short sequence: use random guess"
-                guess = random.randint(0, 1)
-                if guess > 0:
-                    prediction_labels.append("500")
-                else:
-                    prediction_labels.append("400")
-                continue
-
-            successful_prob = 0.5
-            unsuccessful_prob = 0.5
-
-            # get probability of generating sequence from successful transcript
-            words[len(words)-1] = "500"
-            for i in xrange(0, len(words) - n_order - 1):
[…]

### 2017-08-23 mchain-python/obesity-mchain-tran.py

@@ -8,9 +8,8 @@ from collections import Counter
 #############################################################################################
 # Parse command line arguments
 parser = argparse.ArgumentParser(description='Train Markov Model.')
-parser.add_argument('--codebook', default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
-                    help='File location containing codebook.')
-parser.add_argument('--sampling', default='under', type=str, help='sampling parameter.')
+parser.add_argument('--codebook', default='codebook.txt', help='File location containing codebook.')
+parser.add_argument('--sampling', default='over', type=str, help='sampling parameter.')
 parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
 
 args = parser.parse_args()

### 2017-08-23 mchain-python/utility.py

@@ -320,7 +320,7 @@ def createUnderOrOverSample(method, given_data, outputdata_filename, max_len, co
     # perform over or under sampling
     X_d = []
     y_res = []
-    if method == "oversampling":
+    if method == "over":
         sm = SMOTE(kind='borderline2')
         X_res, y_res = sm.fit_sample(X_norm, y_norm)
     else:

### 2017-08-23 hmm-python/obesity-hmm-best.py
fixed error

@@ -7,23 +7,20 @@ import random
 # Parse command line arguments
 parser = argparse.ArgumentParser(description='Train HMM Model.')
 parser.add_argument('--hidden_states', default=5, type=int, help='number of hidden states')
-parser.add_argument('--codebook',
-                    default='/home/mehedi/teana/data-source/seq-analysis/deepLearn/codebook_improve_new.txt',
-                    help='File location containing codebook.')
+parser.add_argument('--codebook', default='codebook.txt', help='File location containing codebook.')
 parser.add_argument('--sampling', default='over', type=str, help='dropout rate parameter.')
 
-args = parser.parse_args()
-
 ############################################################################################
-# Load up training data
+# Read parameters
+args = parser.parse_args()
 codebook_filename = args.codebook
 sampling = args.sampling
 hidden_states = args.hidden_states
+
+# initialize some variables
+kFolds = 10
 training_filename = "train.txt"
 testing_filename = "test.txt"
-
-# get results fro k folds
-kFolds = 10
 macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)

### 2017-08-24 hmm-python/obesity-hmm-best.py
read data from file for hmm and markov chain

@@ -24,12 +24,11 @@ testing_filename = "test.txt"
 macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
-foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
 
 for k in np.arange(0, kFolds):
     # get train and test data
-    utility.createUnderOrOverSample(sampling, foldData[k][0], testing_filename, max_len, codebook)
-    utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
+    training_filename = sampling + "/folds/fold" + str(k + 1) + "/train.txt"
+    testing_filename = sampling + "/folds/fold" + str(k + 1) + "/test.txt"
 
     # determine codebook size and number of hidden states
     all_codebook, success_codebook, unsuccess_codebook = utility.loadCodeBookFromTrainingFile(training_filename)

### 2017-08-24 mchain-python/obesity-mchain-tran.py

@@ -12,28 +12,25 @@ parser.add_argument('--codebook', default='codebook.txt', help='File location co
 parser.add_argument('--sampling', default='over', type=str, help='sampling parameter.')
 parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
 
-args = parser.parse_args()
-
 ############################################################################################
-# Load up training data
+# Read parameters
+args = parser.parse_args()
 codebook_filename = args.codebook
 sampling = args.sampling
 n_order = args.order
-training_filename = "train.txt"
-testing_filename = "test.txt"
+
 
 # get results fro k folds
 kFolds = 10
 macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
-foldData, max_len = utility.createStartifiedFolds(codebook, kFolds)
 
 
 for k in np.arange(0, kFolds):
     # get train and test data
-    utility.createUnderOrOverSample(sampling, foldData[k][0], testing_filename, max_len, codebook)
-    utility.createUnderOrOverSample(sampling, foldData[k][1], training_filename, max_len, codebook)
+    training_filename = sampling + "/folds/fold" + str(k + 1) + "/train.txt"
+    testing_filename = sampling + "/folds/fold" + str(k + 1) + "/test.txt"
 
     # set dictionary for successful sequences
     successful_dict = utility.loadTransitionDictionary(training_filename, n_order, "500")

### 2017-08-24 hmm-python/obesity-hmm-best.py
update results for 10 folds cross-validation

@@ -1,26 +1,31 @@
+#**************************
+# Author: Md Mehedi Hsan  *
+# Wayne State University  *
+# email: mehedi@wayne.edu *
+#**************************
+
 import numpy as np
 import utility
 import argparse
 import random
 
-#############################################################################################
 # Parse command line arguments
 parser = argparse.ArgumentParser(description='Train HMM Model.')
-parser.add_argument('--hidden_states', default=5, type=int, help='number of hidden states')
+parser.add_argument('--succ_hidden_states', default=20, type=int, help='number of hidden states for successful sequences')
+parser.add_argument('--unsucc_hidden_states', default=20, type=int, help='number of hidden states for unsuccessful sequences')
+parser.add_argument('--folds', default=10, type=int, help='k-folds parameter.')
 parser.add_argument('--codebook', default='codebook.txt', help='File location containing codebook.')
-parser.add_argument('--sampling', default='over', type=str, help='dropout rate parameter.')
+parser.add_argument('--sampling', default='over', type=str, help='sampling parameter.')
 
-############################################################################################
 # Read parameters
 args = parser.parse_args()
+succ_hidden_states = args.succ_hidden_states
+unsucc_hidden_states = args.unsucc_hidden_states
+kFolds = args.folds
 codebook_filename = args.codebook
 sampling = args.sampling
-hidden_states = args.hidden_states
 
 # initialize some variables
-kFolds = 10
-training_filename = "train.txt"
-testing_filename = "test.txt"
 macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
@@ -34,16 +39,14 @@ for k in np.arange(0, kFolds):
     all_codebook, success_codebook, unsuccess_codebook = utility.loadCodeBookFromTrainingFile(training_filename)
 
     # fit successful model
-    n_states = 5
     n_observations = len(success_codebook)
     sequences, seq_labels, seq_lengths = utility.loadData(training_filename, success_codebook, 1)
-    success_model = utility.getHMMModel(n_states, n_observations, sequences, seq_lengths)
+    success_model = utility.getHMMModel(succ_hidden_states, n_observations, sequences, seq_lengths)
 
     # fit unsuccessful model
-    n_states = 22
     n_observations = len(unsuccess_codebook)
     sequences, seq_labels, seq_lengths = utility.loadData(training_filename, unsuccess_codebook, 0)
[…]

### 2017-08-24 mchain-python/obesity-mchain-tran.py

@@ -1,3 +1,9 @@
+#**************************
+# Author: Md Mehedi Hsan  *
+# Wayne State University  *
+# email: mehedi@wayne.edu *
+#**************************
+
 import numpy as np
 import utility
 import argparse
@@ -5,23 +11,23 @@ import random
 import math
 from collections import Counter
 
-#############################################################################################
 # Parse command line arguments
 parser = argparse.ArgumentParser(description='Train Markov Model.')
+parser.add_argument('--folds', default=10, type=int, help='k-folds parameter.')
+parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
 parser.add_argument('--codebook', default='codebook.txt', help='File location containing codebook.')
 parser.add_argument('--sampling', default='over', type=str, help='sampling parameter.')
-parser.add_argument('--order', default=1, type=int, help='order of the markov model.')
 
-############################################################################################
+
 # Read parameters
 args = parser.parse_args()
 codebook_filename = args.codebook
+kFolds = args.folds
 sampling = args.sampling
 n_order = args.order
 
 
 # get results fro k folds
-kFolds = 10
 macro_results = []
 micro_results = []
 codebook = utility.loadCodeBook(codebook_filename)
@@ -94,7 +100,7 @@ for k in np.arange(0, kFolds):
                 prediction_labels.append("400")
 
     # store results in macro average
-        prediction_labels = np.array(prediction_labels)
+    prediction_labels = np.array(prediction_labels)
     accuracy, precision, recall, f_measure = utility.getMacroAveragePerformance(seq_labels, prediction_labels)
     print "\nResults for fold", (
         k + 1), ": Accuracy:", accuracy, "Precision:", precision, "Recall:", recall, "F1:", f_measure
@@ -108,10 +114,11 @@ for k in np.arange(0, kFolds):
     fold_result = [k, accuracy, precision, recall, f_measure]
     micro_results.append(fold_result)
 
-############################################################################################
+# Print micro and macro averaged results
 print "\nMacro average results: ", (np.mean(macro_results, axis=0))
 print "\nMicro average results: ", (np.mean(micro_results, axis=0))
 
+# Write results into file
 f = open("results.txt", "a")
 f.write(str(n_order) + ",")
 for x in np.mean(macro_results, axis=0):

### 2017-08-24 deeplearning/obesity_gru_lstm.py
update readme file

@@ -19,6 +19,7 @@ from sklearn.metrics import accuracy_score
 parser = argparse.ArgumentParser(description='Train RNN Model.')
 parser.add_argument('--codebook', default='codebook.txt',
                     help='File location containing codebook or vocabulary.')
+parser.add_argument('--folds', default=10, type=int, help='k-folds parameter.')
 parser.add_argument('--hidden_units', default=20, type=int, help='no. of hidden units in hidden layer.')
 parser.add_argument('--batch_size', default=4, type=int, help='batch size of the training data.')
 parser.add_argument('--embed_dimension', default=45, type=int, help='embedding dimension parameter.')
@@ -32,6 +33,7 @@ args = parser.parse_args()
 
 # Set arguments value
 codebook_filename = args.codebook
+kFolds = args.folds
 hidden_units = args.hidden_units
 batch_size = args.batch_size
 embed_dimension = args.embed_dimension
@@ -234,6 +236,6 @@ def getKFoldsResults(kFolds=10):
 
 
 # Run the model and print kFolds result
-macro_results, micro_results = getKFoldsResults(kFolds=10)
+macro_results, micro_results = getKFoldsResults(kFolds)
 print "Macro results: ", (np.mean(macro_results, axis=0))
 print "Micro results: ", (np.mean(micro_results, axis=0))

### 2017-08-24 hmm-python/runcode.sh
renamed python files

@@ -1,5 +1,5 @@
 
 for (( i=1; i <= 60; ++i ))
 do
-    python obesity-hmm-best.py
+    python obesity-hmm.py
 done

### 2017-08-24 mchain-python/runcode.sh

@@ -1,5 +1,5 @@
 
 for (( i=1; i <= 10; ++i ))
 do
-    python obesity-mchain-best.py
+    python obesity-markov-chain.py
 done

### 2017-08-28 paper-amia-summit-2018/amia_summit_2018.tex
start editing paper
update results and renamed paper title

--- 2017-08-15 paper-amia-summit-2018/main.tex
@@
-\documentclass{amia}
+\documentclass{amia_summit_2018}
 \usepackage{graphicx}
@@
 
-\title{Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences}
+\title{Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model of Patient-Provider Communication Sequences}
 
-\author{Alexander Kotov, PhD$^{1}$\footnote[1]{Authors provided equal contribution.
-\label{footnote1}}, Mehedi Hasan, BS$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
+\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution.
+\label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
 
@@
   \hline
-   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
-Natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
-Alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
+   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
+Successful sequences & 880 & 82.09\% & 5.83 \\\hline
+Unsuccessful sequences & 880 & 82.09\% & 5.83 \\\hline
   \end{tabular}
@@
 
+\textbf {Recurrent Neural Networks (RNNs)} are a type of Neural Network where the hidden state of one time step is computed by combining the current input with the hidden state of the previous time step....
+
 \subsection*{\textit{Evaluation metrics}}
@@
 
-\subsection*{\textit{Predictive performance for natural PPC code sequences}}
-Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}.
+\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
+Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}.
 Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}.
@@
 \caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
-\label{tab:result_norm_seq}
+\label{tab:result_under_sampled}
   \begin{tabular}{|l|l|l|l|l|l|}
   \hline
-   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline
+   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
 
- \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532\\\cline{2-5}
- & Second order & 0.6889 & 0.7889 & 0.7352\\ \hline
- \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989}\\ \cline{2-5}
- & Second order & 0.7400 & \textbf{0.8449} & 0.7822\\ \hline
-
+ Markov Chain $1^{st}$ Order & 0.7013 & 0.7037 & 0.7013 & 0.7003\\ \hline
+ Markov Chain $2^{nd}$ Order & 0.5905 & 0.5917 & 0.5905 & 0.5890\\ \hline
+ Hidden Markov Model & 0.5439 & 0.5778 & 0.5439 & 0.4865\\ \hline
+ LSTM RNN & 0.8770 & 0.8812 & 0.8770 & 0.8766\\ \hline
+ LSTM RNN - TR & 0.8824 & 0.8890 & 0.8824 & 0.8817\\ \hline
+ GRU RNN & 0.8844 & 0.8897 & 0.8844 & 0.8840\\ \hline
+ GRU RNN - TR & \textbf{0.8858} & \textbf{0.8917} & \textbf{0.8858} & \textbf{0.8853}\\ \hline 
   \end{tabular}
@@
 
-\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
-Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences.
+\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
+Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on alternating sequences.
 These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
@@
 \caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
-\label{tab:result_alt_seq}
+\label{tab:result_over_sampled}
   \begin{tabular}{|l|l|l|l|l|l|}
   \hline
-   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline
+   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
 
- \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604\\\cline{2-5}
- & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736}\\ \hline
- \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353\\ \cline{2-5}
- & Second order & 0.9704 & \textbf{0.9713} & 0.9699\\ \hline
+ Markov Chain $1^{st}$ Order & 0.7930 & 0.8127 & 0.7930 & 0.7896\\ \hline
+ Markov Chain $2^{nd}$ Order & 0.7325 & 0.7461 & 0.7325 & 0.7288\\ \hline
+ Hidden Markov Model & 0.7763 & 0.8062 & 0.7763 & 0.7706\\ \hline
+ LSTM RNN & 0.8703 & 0.8782 & 0.8703 & 0.8696\\ \hline
+ LSTM RNN - TR & 0.8713 & 0.8789 & 0.8713 & 0.8706\\ \hline
+ GRU RNN & 0.8711 & 0.8785 & 0.8711 & 0.8705\\ \hline
+ GRU RNN - TR & \textbf{0.8722} & \textbf{0.8790} & \textbf{0.8722} & \textbf{0.8716}\\ \hline
   \end{tabular}

### 2017-08-28 paper-amia-summit-2018/amia_summit_2018.tex
update abstract

@@
 
-\title{Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model of Patient-Provider Communication Sequences}
+\title{Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences}
 
@@
 \textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics.
-In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose two methods, based on Markov chain and Hidden Markov Model, for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.
-Our method achieved 75.32\% and 79.89\% F-measure for natural sequences and 97.36\% and 96.99\% F-measure for alternating sequences in predicting the outcome of motivational interviews with obese adolescents using first and second order Markov Chain and Hidden Markov Model, respectively.
+In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.
+Our method achieved 70.03\%, 48.65\%, 88.53\% F1-score for under-sampled sequences and 78.96\%, 77.06\%, 87.16\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively.
 The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}

### 2017-08-28 paper-amia-summit-2018/amia_summit_2018.tex
update data collection, evaluation metrics and conclusion

@@
 \subsection*{\textit{Data collection}}
-The experimental dataset for this work was constructed from the transcripts of 37 motivational interviews, which include a total of 11,353 segmented and annotated utterances.
+The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances.
 Each transcript consists of a counselor-adolescent and a counselor-caregiver session.
@@
 Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
-Out of 1072 observed sequences, 880 were positive and 192 were negative.
-Alternating communication sequences are sub-sequences of a natural communication sequence, in which two consecutive utterances cannot belong to the same speaker (i.e.
-a counselor behavior code has to be followed by an adolescent behavior code and vice versa).
-For example, 6 alternating sequences (\{C1, A1, C2, A3\}, \{C1, A1, C3, A3\}, \{C1, A1, C4, A3\}, \{C1, A2, C2, A3\}, \{C1, A2, C3, A3\} and \{C1, A2, C4, A3\}) can be created from a natural sequence \{C1, A1, A2, C2, C3, C4, A3\}, where adolescent and counselor codes are represented by A and C, respectively.
-A total of 18753 alternating sequences have been observed in experimental dataset, out of which 18139 were successful and 614 were unsuccessful ones.
-For each type of sequence (natural and alternating), one probabilistic model was trained using successful sequences and one model was train using unsuccessful sequences.
+Out of 4169 observed sequences, 3427 were positive and 742 were negative.
+For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences.
 Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
@@
    \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
-Successful sequences & 880 & 82.09\% & 5.83 \\\hline
-Unsuccessful sequences & 880 & 82.09\% & 5.83 \\\hline
+Successful sequences & 3427 & 82.20\% & 10.05 \\\hline
+Unsuccessful sequences & 742 & 17.80\% & 9.89 \\\hline
   \end{tabular}
@@
 
-Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\
+As can be seen in Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences.
+Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data.
+Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%.
+Therefore, it is important to handle imbalance data properly.
+Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels.
+In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set.
+Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
+In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}.
+On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
+Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end.
+\\
 
@@
 \begin{itemize}
-\item $V = \{CML+, CHT+, CHT-, T-AMB, CCT, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
+\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
 \item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
@@
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.70\textwidth]{figures/delta.png}
-    \caption{\textbf{F-Measure of the proposed Markov model-based method by varying the value of $\delta$}.}
-    \label{fig:delta}
-\end{figure}
-
 The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
@@
 \subsection*{\textit{Evaluation metrics}}
-Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds.
+Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folda cross-validation and weighted macro-averaging of these metrics over the folds.
 A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
@@
 2 x Precision x Recall / (Precision + Recall)).
+However, Accuracy is simply computed as the ratio of correctly classified sequences and total number of sequences.
 
@@
 Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}.
-Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}.
+Three major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
 First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences.
@@
 \section*{Conclusion}
-In this paper, we proposed a method based on Markov Chain and Hidden Markov Model for predicting the success of motivational interviews.
+In this paper, we proposed Recurrent Neural Networks with two baseline methods Markov Chain and Hidden Markov Model for predicting the success of motivational interviews.
 We found out that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
 Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource intensive manual qualitative analysis of interview transcripts.
-Our proposed method can also help identify specific patterns that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.
+Our proposed method can also help to identify specific patterns that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.
 This work also has broad implications for qualitative public health research by providing a formal theoretically-grounded computational mechanism to facilitate the development of effective behavioral interventions.

### 2017-08-30 paper-amia-summit-2018/amia_summit_2018.tex
update data collection, evaluation metrics and conclusion
update methods

@@
 
-Our proposed prediction method consists of two steps.
+Our proposed baseline prediction method consists of two steps.
 In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary.
+In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model.
 Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.
+The latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification.
+In our experiment, we also employed two especial variant of RNN models: long short-term memory (LSTM) and gated recurrent unit (GRU).
 
@@
 
-If $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language.
-Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias.
-Figure~\ref{fig:delta} illustrates the performance of Markov Chain-based prediction method by varying the value of $\delta$.
-As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias.
-However, F-score sharply decreases after $\delta = -0.9$.
-Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.
+If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.
 
@@
 After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.
-As in the case of MC, we also experimented with second-order HMM.
 
@@
 \section*{Results}
-Experimental evaluation of the proposed method is conducted on both natural sequences and alternating sequences.
+Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.
 
@@
 
-\subsection*{\textit{Impact of hidden states}}
-While MC is a non-parametric method, HMM has an important parameter, the number of hidden states.
+%\subsection*{\textit{Impact of hidden states}}
+%While MC is a non-parametric method, HMM has an important parameter, the number of hidden states.
 A natural question is how does this parameter affect the performance of HMM-based prediction  method.
@@
 All four models show their best performance when the number of states is between 4 and 8.
-
-Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\
-
-\begin{figure}[!htb]
-    \centering
-    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
-    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
-    \label{fig:hidden-states}
-\end{figure}
-
-\begin{table}[!htb]
-\centering
-\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
-\label{tab:emission_matrix_s}
-  \begin{tabular}{|l|l|}
-  \hline
-   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
-1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
-2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
-  \end{tabular}
-\end{table}
-
-\begin{table}[!htb]
-\centering
-\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
-\label{tab:emission_matrix_u}
-  \begin{tabular}{|l|l|}
-  \hline
-   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
-1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
-2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
-  \end{tabular}
-\end{table}
+%
+%Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\
+%
+%\begin{figure}[!htb]
+%    \centering
+%    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
+%    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
+%    \label{fig:hidden-states}
+%\end{figure}
+%
+%\begin{table}[!htb]
+%\centering
+%\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
+%\label{tab:emission_matrix_s}
+%  \begin{tabular}{|l|l|}
+%  \hline
+%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
+%1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
+%2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
+%  \end{tabular}
+%\end{table} 
+%
+%\begin{table}[!htb]
+%\centering
+%\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
+%\label{tab:emission_matrix_u}
+%  \begin{tabular}{|l|l|}
+%  \hline
+%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
+%1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
+%2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
+%  \end{tabular}
+%\end{table}
 

### 2017-08-30 paper-amia-summit-2018/amia_summit_2018.tex
update description of methods specially, Recurrent Neural Network

@@
 \subsection*{\textit{Prediction method}}
-
 Generally, a sequence is a temporally ordered set of events.
@@
 A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
-
 \begin{itemize}
@@
 \item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
-
 \begin{equation}
@@
 \end{equation}
-
 \end{itemize}
-
 where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data.
 Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
-
 \begin{equation}
@@
 \end{equation}
-
 In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
-
 \begin{equation}
@@
 \end{equation}
-
 If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.
@@
 An HMM can be fully specified by the following quintuple: $\lambda = (N, M, A, B, \pi)$
-
 \begin{itemize}
@@
 \end{itemize}
-
 We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$.
@@
 
-\textbf {Recurrent Neural Networks (RNNs)} are a type of Neural Network where the hidden state of one time step is computed by combining the current input with the hidden state of the previous time step....
+\textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
+In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.
+The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed.
+However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}.
+In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM.
+Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}.
+There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}.
+GRUs are simpler than LSTM units and experimentally outperform than other models in many cases.
+We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (equation~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
+\begin{equation}
+z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
+\label{eq:firstgru}
+\end{equation}
+\begin{equation}
+r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
+\end{equation}
+\begin{equation}
+\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
+\end{equation}
+\begin{equation}
+h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
+\label{eq:lastgru}
+\end{equation}  
+\begin{equation}
+total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
+\label{eq:loss}
+\end{equation}  
+The architecture of proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
+We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence.
+Then, we applied a softmax layer at each time step for predicting the target.
+This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
+As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs.
+LSTM and GRU are trained on 70\% of the data and validated on 20\%.
+The remaining 10\% of the data is used as a test set for reporting the performance of the model.
+Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
+We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss.
 
 \subsection*{\textit{Evaluation metrics}}
-Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folda cross-validation and weighted macro-averaging of these metrics over the folds.
+Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds.
 A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
added figure for proposed rnn model

@@
 
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
+    \caption{\textbf{Proposed RNN model with target replication}.}
+    \label{fig:rnn-model}
+\end{figure}
+
 \textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
@@
 GRUs are simpler than LSTM units and experimentally outperform than other models in many cases.
-We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (equation~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
+We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
 \begin{equation}
@@
 We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss.
+Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization.
 

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
added figure for proposed rnn model

@@
 \section*{Results}
-Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.
-
-\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}.
-Three major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
-First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences.
-Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449).
-Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM.
-In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
-In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models.
-\\
+Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences. \\
 
@@
 
-\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
-Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on alternating sequences.
-These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
-Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models.
-However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736).
-We also observed that second-order MC and HMM outperform first-order MC and HMM.
-Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
-Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\
+\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
+Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}.
+Three major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
+First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences.
+Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449).
+Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM.
+In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
+In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models.
+\\
 
@@
 \end{table}
+
+\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
+Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on alternating sequences.
+These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
+Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models.
+However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736).
+We also observed that second-order MC and HMM outperform first-order MC and HMM.
+Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
+Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.
 

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
remove comments
update introduction

@@
 The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
-The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov chain and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+
+However, Markov chain cannot handle long-term dependencies in sequences.
+Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step.
+Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps.
+On the other hand, Recurrent Neural Networks have no such limitations.
+RNNs are capable to capture long-term dependencies with their unlimited memory.
+It was shown that RNNs are very effective in sequence modeling with temporal data.
+Recently, two variant of RNNs are proposed: Long Short-Term Memory (LSTM)\cite{graves2013speech} and Gated Reccurent Unit (GRU)\cite{chung2014empirical}, which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten,lipton2015learning, choi2016doctor}.
 
@@
 In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}.
-In this work, we propose probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
+In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
 change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
@@
 
-As can be seen in Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences.
+As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences.
 Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data.
@@
 Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.
-The latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification.
-In our experiment, we also employed two especial variant of RNN models: long short-term memory (LSTM) and gated recurrent unit (GRU).
+However, the latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification.
+Therefore, we employed two special variant of RNN models in our experiemnt: long short-term memory (LSTM) and gated recurrent unit (GRU).
 
@@
 
-\begin{figure}[!htb]
-    \centering
-    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
-    \caption{\textbf{Proposed RNN model with target replication}.}
-    \label{fig:rnn-model}
-\end{figure}
-
 \textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
@@
 \end{equation}  
-The architecture of proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
+    \caption{\textbf{Proposed RNN model with target replication}.}
+    \label{fig:rnn-model}
+\end{figure}
+The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
 We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence.
@@
 Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization.
+The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
 
@@
 
-%\subsection*{\textit{Impact of hidden states}}
-%While MC is a non-parametric method, HMM has an important parameter, the number of hidden states.
-A natural question is how does this parameter affect the performance of HMM-based prediction  method.
-Figure~\ref{fig:hidden-states} provides an answer to this question by illustrating the performance of the proposed method using HMMs of different order in terms of F-measure on different types of sequences.
-As follows from this Figure, in general, the number of states does not significantly impact the classification performance.
-All four models show their best performance when the number of states is between 4 and 8.
-%
-%Tables~\ref{tab:emission_matrix_s} and \ref{tab:emission_matrix_u} show patient-provider communication (PPC) codes with the highest probability in emission probability distributions associated with different states of HMMs trained on successful and unsuccessful interviews.\\
-%
-%\begin{figure}[!htb]
-%    \centering
-%    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
-%    \caption{\textbf{Performance of the proposed method using HMMs of different order by varying the number of hidden states}.}
-%    \label{fig:hidden-states}
-%\end{figure}
-%
-%\begin{table}[!htb]
-%\centering
-%\caption{PPC codes with the highest probability associated with different states of an HMM trained on successful interviews.}
-%\label{tab:emission_matrix_s}
-%  \begin{tabular}{|l|l|}
-%  \hline
-%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline 
-%1 & SS, EA, R-CT+, OQ-ECT+, SUM-S, AF, R-CML+ \\\hline           
-%2 & LUP+, EA, OQ-ECT+, OQ-ECML+  \\\hline    
-%  \end{tabular}
-%\end{table} 
-%
-%\begin{table}[!htb]
-%\centering
-%\caption{PPC codes with the highest probability associated with different states of an HMM trained on unsuccessful interviews.}
-%\label{tab:emission_matrix_u}
-%  \begin{tabular}{|l|l|}
-%  \hline
-%   \textbf{State} & \textbf{Most likely PPC codes} \\\hline     
-%1 & OQ-TBN, EA, R-CT+, AR, SUM-S, RBA-S, G-INFO+, R-CML+, OQ-ECT- \\\hline
-%2 & LUP+, OQ-ECT+, G-INFO+, CQ-EF, CHT-, HUP-O \\\hline
-%  \end{tabular}
-%\end{table}
-
 \subsection*{\textit{Common patterns}}
@@
 & commitment language positive $\rightarrow $ [positive commitment] \\\hline
-successful & 306 CQ-EF: Closed question, Elicit Feedback $\rightarrow $ 120 HUP-O: High Uptake, other \\
-& $\rightarrow $ 313 R-CML+: Reflect, commitment language positive $\rightarrow $ [positive commitment] \\\hline
 unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
@@
 & 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
-unsuccessful & 305 EA: Emphasize Autonomy $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
-& General Information, positive $\rightarrow $ [negative commitment] \\\hline
   \end{tabular}

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
fix reference format

@@
 The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
-The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov chain and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
 
@@
 Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps.
-On the other hand, Recurrent Neural Networks have no such limitations.
+On the other hand, Recurrent Neural Networks (RNNs) have no such limitations.
 RNNs are capable to capture long-term dependencies with their unlimited memory.
 It was shown that RNNs are very effective in sequence modeling with temporal data.
-Recently, two variant of RNNs are proposed: Long Short-Term Memory (LSTM)\cite{graves2013speech} and Gated Reccurent Unit (GRU)\cite{chung2014empirical}, which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten,lipton2015learning, choi2016doctor}.
+Recently, two variant of RNNs are proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
 
@@
 Our proposed baseline prediction method consists of two steps.
-In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain (MC) and Hidden Markov Model (HMM), which are popular probabilistic models for discrete observation sequences with finite vocabulary.
+In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary.
 In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model.
@@
 
-\textbf {Recurrent Neural Networks (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
+\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
 In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
fix spelling error

@@
 The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
-The third type of sequence classification methods, first create a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+The third type of sequence classification methods, first creates a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
 
@@
 Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step.
-Therfore, markov chain and HMMs are only capable of passing short-term dependency among sequence steps.
+Therefore, Markov chain and HMMs are only capable of passing short-term dependency among sequence steps.
 On the other hand, Recurrent Neural Networks (RNNs) have no such limitations.
@@
 It was shown that RNNs are very effective in sequence modeling with temporal data.
-Recently, two variant of RNNs are proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
+Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
 
@@
 change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
-While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting a desired patient behavior in a context of motivational interviews.
+While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.
 
@@
 For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences.
-Statistics of experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
+Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
 \\
@@
 Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.
-However, the latest advances in deep learning technologis shows that deep learning, in particular RNNs, provide better results than conventional machine learning methods for the task of sequence classification.
-Therefore, we employed two special variant of RNN models in our experiemnt: long short-term memory (LSTM) and gated recurrent unit (GRU).
+However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification.
+Therefore, we employed two special variants of RNN models in our experiemnts: long short-term memory (LSTM) and gated recurrent unit (GRU).
 
@@
 \item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
-\item $E \subseteq V \times V$ is a set of edges corresponding to posssible transitions from one MI behavior code to the other in a sequence;
+\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
 \item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
@@
 The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
-We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from input sequence.
+We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence.
 Then, we applied a softmax layer at each time step for predicting the target.
 This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
-As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determine the relative importance of intermediate outputs.
+As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
 LSTM and GRU are trained on 70\% of the data and validated on 20\%.
@@
 A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
-Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)).
+The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)).
 Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e.
@@
 2 x Precision x Recall / (Precision + Recall)).
-However, Accuracy is simply computed as the ratio of correctly classified sequences and total number of sequences.
+However, Accuracy is simply computed as the ratio of correctly classified sequences and a total number of sequences.
 

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
update results

@@
 As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
-LSTM and GRU are trained on 70\% of the data and validated on 20\%.
-The remaining 10\% of the data is used as a test set for reporting the performance of the model.
-Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
+Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
 We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss.
@@
 Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds.
+However, LSTM and GRU are trained on 70\% of the data and validated on 20\%.
+The remaining 10\% of the data is used as a test set for reporting the performance of the model.
 A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
@@
 \centering
-\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
+\caption{Performance of MC, HMM, and RNN for predicting success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
 \label{tab:result_under_sampled}
@@
 \subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}.
-Three major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
-First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences.
-Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449).
-Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM.
-In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models.
-In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models.
+We applied small learning rate 0.0005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used.
+Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}.
+Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
+First, recurrent neural networks outperforms over probabilistic models, and achieved 18.50\%-39.88\% higher F1-score.
+Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858.
+Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication.
+Fourth, among probabilistic models, MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
+Fifth, it follows that second-order MC have lower acuracy, precision, recall and F-measure than first-order MC.
+In particular, acuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models.
+At last, accuracy and recall illustrate exactly the same values for all models.
 \\
@@
 \centering
-\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
+\caption{Performance of MC, HMM, and RNN for predicting success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
 \label{tab:result_over_sampled}
@@
 \subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
-Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on alternating sequences.
-These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall.
-Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models.
-However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736).
-We also observed that second-order MC and HMM outperform first-order MC and HMM.
-Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively.
-Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.
+For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy.
+Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences.
+These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still have 2.41\% lower F1-score than first-order Markov chain.
+Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models.
+However, different from under-sampled sequences, predictive performance of probabilistic models increases whereas recurrent neural networks performance decreases.
+We also observed that all models have largest value in precision compared to all other performance metrics.
 

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
update results

@@
 Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}.
-There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}.
+There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU.
 GRUs are simpler than LSTM units and experimentally outperform than other models in many cases.
@@
 \subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-We applied small learning rate 0.0005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used.
+We applied small learning rate 0.00005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used.
 Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}.

### 2017-09-01 paper-amia-summit-2018/amia_summit_2018.tex
update counselor and adolescent dialogue

@@
 Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end.
-\\
+It was noticed that same context represents different behavior codes which can be seen at row 7 and 9 in Table~\ref{tab:anno_examp}.
+Because human brain can remember the past behaviors, and classified the current context into different behavior codes after considering their past behaviors.
+In this paper, we only consider those models that can capture the past events for the classification of behavior sequences.\\
 
@@
 \centering
-\begin{tabular}{|l|p{3.6cm}|l|p{8cm}|}
+\begin{tabular}{|l|p{3.2cm}|l|p{8cm}|}
 \hline
-Annotation  & Description & Speaker & Text \\\hline
-331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do?
-mm-hmm meaning is that food available for you?
-\\\hline
-117 &	Low Uptake, positive	& Adolescent &	Yes \\\hline
-301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\\hline
-112 &	Change Talk positive	& Adolescent &	my um aunt \\\hline
-301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\\hline
-112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\\hline
-309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\\hline
+\bf{Annotation}  & \bf{Behavior} & \bf{Speaker} & \bf{Text} \\\hline
+SS & Structure Session & Counselor & Okay. Can I meet with Xxxx alone for a few minutes? \\\hline
+OQO & Open-ended question, other & Counselor & So, Xxxx, how you doing? \\\hline
+HUPO &	High uptake, other	& Adolescent &	Fine \\\hline
+OQTBN &	Open-ended question, target behavior neutral & Counselor &	That's good.  So, tell me  how do you feel about your weight? \\\hline
+CHT+ &	Change Talk positive	& Adolescent &	It's not the best. \\\hline
+CQECHT+ & Closed question, elicit change talk positive & Counselor & It's not the best? \\\hline
+CHT+ &	Change Talk positive &	Adolescent & Yeah \\\hline
+CQTBN &	Closed question, target behavior neutral  & Counselor &	Okay, so have you tried to lose weight before? \\\hline
+HUPW &	High uptake, weight & Adolescent &	Yes \\\hline
 \end{tabular}
@@
 Then, we applied a softmax layer at each time step for predicting the target.
-This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
+This target replication (TR) strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
 As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
@@
 \centering
-\caption{Performance of MC, HMM, and RNN for predicting success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
+\caption{Performance of MC, HMM, and RNN for predicting the success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
 \label{tab:result_under_sampled}
@@
 \subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-We applied small learning rate 0.00005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used.
+We applied small learning rate 0.00005 and early stopping strategy with 150 epoch for training deep learning models when under-sampled data is used.
 Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}.
 Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
-First, recurrent neural networks outperforms over probabilistic models, and achieved 18.50\%-39.88\% higher F1-score.
+First, recurrent neural networks outperform over probabilistic models and achieved 18.50\%-39.88\% higher F1-score.
 Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858.
 Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication.
-Fourth, among probabilistic models, MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
-Fifth, it follows that second-order MC have lower acuracy, precision, recall and F-measure than first-order MC.
-In particular, acuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models.
+Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
+Fifth, it follows that second-order MC have lower accuracy, precision, recall, and F-measure than first-order MC.
+In particular, accuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models.
 At last, accuracy and recall illustrate exactly the same values for all models.
@@
 \centering
-\caption{Performance of MC, HMM, and RNN for predicting success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
+\caption{Performance of MC, HMM, and RNN for predicting the success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
 \label{tab:result_over_sampled}
@@
 Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences.
-These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still have 2.41\% lower F1-score than first-order Markov chain.
+These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still has 2.41\% lower F1-score than first-order Markov chain.
 Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models.
-However, different from under-sampled sequences, predictive performance of probabilistic models increases whereas recurrent neural networks performance decreases.
-We also observed that all models have largest value in precision compared to all other performance metrics.
+However, the predictive performance of recurrent neural networks decreases in over-sampled sequences whereas the performance of probabilistic models decreases.
+We also observed that all models have the largest value in precision compared to other performance metrics.
 

### 2017-09-01 paper-amia-summit-2018/amia_summit_2018.tex
update data collection

@@
 For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences.
-Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
+Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}.
 \\
@@
 On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
-Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end.
-It was noticed that same context represents different behavior codes which can be seen at row 7 and 9 in Table~\ref{tab:anno_examp}.
-Because human brain can remember the past behaviors, and classified the current context into different behavior codes after considering their past behaviors.
+A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
+Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive feedback at the end.
+It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively.
+Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors.
 In this paper, we only consider those models that can capture the past events for the classification of behavior sequences.\\
@@
 In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model.
-Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}.
+Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, won2004training, chai2001folk}.
 However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification.
@@
 \begin{itemize}
-\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
+\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUPW, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
 \item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
@@
 After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
-On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing positive tendency in their previous communication.
+On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
 This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation.

### 2017-09-01 paper-amia-summit-2018/amia_summit_2018.tex
prediction method

@@
 Generally, a sequence is a temporally ordered set of events.
-In this study, an event is a behavior code that also has a symbolic representation, such as 117 $LUP+$ (low uptake, positive), 331 $OQ-ECT+$ (open-ended question, elicit change talk positive), etc.
+In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive), $OQECHT+$ (open-ended question, elicit change talk positive), etc.
 Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some part of a motivational interview, the task of predicting interview success can be viewed as sequence classification.

### 2017-09-08 paper-amia-summit-2018/amia_summit_2018.tex
added T-SNE image and update paper

@@
 \subsection*{\textit{Data collection}}
-The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances.
+The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
 Each transcript consists of a counselor-adolescent and a counselor-caregiver session.
@@
 Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
-Out of 4169 observed sequences, 3427 were positive and 742 were negative.
+Out of 4169 observed sequences, 4225 were positive and 918 were negative.
 For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences.
@@
    \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
-Successful sequences & 3427 & 82.20\% & 10.05 \\\hline
-Unsuccessful sequences & 742 & 17.80\% & 9.89 \\\hline
+Successful sequences & 4225 & 82.20\% & 10.05 \\\hline
+Unsuccessful sequences & 918 & 17.80\% & 9.89 \\\hline
   \end{tabular}
@@
     \centering
-    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
-    \caption{\textbf{Proposed RNN model with target replication}.}
+    \includegraphics[width=0.70\textwidth]{figures/rnn.eps}
+    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
     \label{fig:rnn-model}
@@
 The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
-We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence.
+We used one hidden layer of 20 LSTM or GRU nodes after embedding the given myscope codes from the input sequence.
 Then, we applied a softmax layer at each time step for predicting the target.
-This target replication (TR) strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
+This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
 As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.

### 2017-09-08 most-likely-patterns/obesity-patterns-and-size.py

@@ -1,0 +1,99 @@
+# **************************
+# Author: Md Mehedi Hsan  *
+# Wayne State University  *
+# email: mehedi@wayne.edu *
+# **************************
+
+import utility
+from collections import Counter
+import operator
+import math
+
+successful_fp = open("successful_patterns.txt", "w")
+unsuccessful_fp = open("unsuccessful_patterns.txt", "w")
+
+map_successful_patterns = dict()
+map_unsuccessful_patterns = dict()
+
+successful_filename = "data/successful.txt"
+unsuccessful_filename = "data/unsuccessful.txt"
+
+total_utterances = 0
+successful_patterns = 0
+unsuccessful_patterns = 0
+
+with open(successful_filename, "r") as file_stream:
+    for line in file_stream:
+        codes = line.replace("\n", "").strip().split(",")
+        total_utterances += len(codes)
+        successful_patterns += 1
+
+with open(unsuccessful_filename, "r") as file_stream:
+    for line in file_stream:
+        codes = line.replace("\n", "").strip().split(",")
+        total_utterances += len(codes)
+        unsuccessful_patterns += 1
+
+print("Successful patterns: " + str(successful_patterns))
+print("Unsuccessful patterns: " + str(unsuccessful_patterns))
+print("Total utterances: " + str(total_utterances))
+
+# set dictionary for successful sequences
+successful_dict = utility.loadTransitionDictionary(successful_filename, 1, "500")
+
+# set dictionary for unsuccessful sequences
+unsuccessful_dict = utility.loadTransitionDictionary(unsuccessful_filename, 1, "400")
+
+with open(successful_filename, "r") as file_stream:
+    for line in file_stream:
+        words = line.replace("\n", "").split(",")
+
+        successful_prob = math.log(0.5)
+
+        # get probability of generating sequence from successful transcript
+        words[len(words) - 1] = "500"
+        for i in xrange(0, len(words) - 1 - 1):
+            current_tuple = tuple([words[j] for j in xrange(i, i + 1)])
+            if current_tuple in successful_dict.keys():
+                next_states = successful_dict[current_tuple]
+                next_states_counter = Counter(next_states)
+                transition_prob = -5.0
+                if words[i + 1] in next_states_counter:
+                    transition_prob = math.log(float(next_states_counter[words[i + 1]]) / len(next_states))
+                successful_prob += transition_prob
+            else:
+                successful_prob -= 5.0
[…]

### 2017-09-08 most-likely-patterns/utility.py

@@ -1,0 +1,373 @@
+import numpy as np
+import random
+import re
+import operator
+from hmmlearn import hmm
+from keras.preprocessing.sequence import pad_sequences
+from imblearn.over_sampling import SMOTE
+from collections import Counter
+from imblearn.under_sampling import ClusterCentroids
+
+
+###################################################################
+# read codebook from file
+def loadCodeBook(codebook_filename):
+    codebook = []
+    with open(codebook_filename, "r") as filestream:
+        for line in filestream:
+            codebook.append(line.replace("\n", ""))
+    return codebook
+
+
+###########################################################################################
+def loadCodeBookFromTrainingFile(fileLocation):
+    all_code_map = {}
+    success_code_map = {}
+    unsuccess_code_map = {}
+    with open(fileLocation, "r") as filestream:
+        for line in filestream:
+            l = re.sub(r"\s+", "", line).split(",")
+            if l[len(l) - 1] == "500":
+                for i in range(0, len(l) - 1):
+                    success_code_map[l[i]] = l[i]
+                    all_code_map[l[i]] = l[i]
+            elif l[len(l) - 1] == "400":
+                for i in range(0, len(l) - 1):
+                    unsuccess_code_map[l[i]] = l[i]
+                    all_code_map[l[i]] = l[i]
+
+    success_sorted_map = sorted(success_code_map.items(), key=operator.itemgetter(0))
+    unsuccess_sorted_map = sorted(unsuccess_code_map.items(), key=operator.itemgetter(0))
+    all_sorted_map = sorted(all_code_map.items(), key=operator.itemgetter(0))
+
+    success_map = []
+    unsuccess_map = []
+    all_map = []
+
+    for key, value in all_sorted_map:
+        all_map.append(key)
+    for key, value in success_sorted_map:
+        success_map.append(key)
+    for key, value in unsuccess_sorted_map:
+        unsuccess_map.append(key)
+
+    return all_map, success_map, unsuccess_map
+
+
+###########################################################################################
+def loadData(fileLocation, codebook, flag):
+    code_to_int = dict((c, i) for i, c in enumerate(codebook))
+    seq = []
+    seq_label = []
+    lengths = []
+    map = {}
+    with open(fileLocation, "r") as filestream:
+        for line in filestream:
+            l = re.sub(r"\s+", "", line).split(",")
+            if (flag == 1 and l[len(l) - 1] == "500") or (flag == 0 and l[len(l) - 1] == "400"):
[…]

### 2017-09-08 paper-amia-summit-2018/amia_summit_2018.tex
update most likely sequences and T-SNE diagram

@@
    \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
-Successful sequences & 4225 & 82.20\% & 10.05 \\\hline
-Unsuccessful sequences & 918 & 17.80\% & 9.89 \\\hline
+Successful sequences & 4225 & 82.15\% & 9.79 \\\hline
+Unsuccessful sequences & 918 & 17.85\% & 9.65 \\\hline
   \end{tabular}
@@
 
+\textbf {MYSCOPE Code Embedding}: is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
+In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.
+RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.
+Stochastic Neighbor Embedding for MYSCOPE code is shown in Figure~\ref{fig:code_embedding}.
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
+    \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}
+    \label{fig:code_embedding}
+\end{figure}
+
 \textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
@@
 2 x Precision x Recall / (Precision + Recall)).
-However, Accuracy is simply computed as the ratio of correctly classified sequences and a total number of sequences.
 
 \section*{Results}
-Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences. \\
-
+Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.
 \begin{table}[h]
 \centering
-\caption{Performance of MC, HMM, and RNN for predicting the success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
-\label{tab:result_under_sampled}
-  \begin{tabular}{|l|l|l|l|l|l|}
+\caption{Performance of MC, HMM, and RNN for predicting the success of under- and over-sampled patient-provider communication sequences.
+The highest value for each performance metric is highlighted in bold.}
+\label{tab:result_under_over_sampled}
+  \begin{tabular}{|l|l|l|l|l|l|l|}
   \hline
-   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
+   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under Sampling}} & \multicolumn{3}{|c|}{\textbf{Over Sampling}} \\\cline{2-7}
+   & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
 
- Markov Chain $1^{st}$ Order & 0.7013 & 0.7037 & 0.7013 & 0.7003\\ \hline
- Markov Chain $2^{nd}$ Order & 0.5905 & 0.5917 & 0.5905 & 0.5890\\ \hline
- Hidden Markov Model & 0.5439 & 0.5778 & 0.5439 & 0.4865\\ \hline
- LSTM RNN & 0.8770 & 0.8812 & 0.8770 & 0.8766\\ \hline
- LSTM RNN - TR & 0.8824 & 0.8890 & 0.8824 & 0.8817\\ \hline
- GRU RNN & 0.8844 & 0.8897 & 0.8844 & 0.8840\\ \hline
- GRU RNN - TR & \textbf{0.8858} & \textbf{0.8917} & \textbf{0.8858} & \textbf{0.8853}\\ \hline 
+ Markov Chain $1^{st}$ Order & 0.7060 & 0.7044 & 0.7038 & 0.7932 & 0.7799 & 0.7775 \\ \hline
+ Markov Chain $2^{nd}$ Order & 0.6395 & 0.6385 & 0.6379 & 0.7111 & 0.7029 & 0.7000\\ \hline
+ Hidden Markov Model & 0.6244 & 0.6143 & 0.6067 & 0.7775 & 0.7567 & 0.7520\\ \hline
+ LSTM RNN & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
+ LSTM RNN - TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
+ GRU RNN & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
+ GRU RNN - TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
   \end{tabular}
@@
 We applied small learning rate 0.00005 and early stopping strategy with 150 epoch for training deep learning models when under-sampled data is used.
-Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}.
-Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}.
+Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.
+Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}.
 First, recurrent neural networks outperform over probabilistic models and achieved 18.50\%-39.88\% higher F1-score.
@@
 At last, accuracy and recall illustrate exactly the same values for all models.
-\\
-
-\begin{table}[h]
-\centering
-\caption{Performance of MC, HMM, and RNN for predicting the success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
-\label{tab:result_over_sampled}
-  \begin{tabular}{|l|l|l|l|l|l|}
-  \hline
-   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
-
- Markov Chain $1^{st}$ Order & 0.7930 & 0.8127 & 0.7930 & 0.7896\\ \hline
- Markov Chain $2^{nd}$ Order & 0.7325 & 0.7461 & 0.7325 & 0.7288\\ \hline
- Hidden Markov Model & 0.7763 & 0.8062 & 0.7763 & 0.7706\\ \hline
- LSTM RNN & 0.8703 & 0.8782 & 0.8703 & 0.8696\\ \hline
- LSTM RNN - TR & 0.8713 & 0.8789 & 0.8713 & 0.8706\\ \hline
- GRU RNN & 0.8711 & 0.8785 & 0.8711 & 0.8705\\ \hline
- GRU RNN - TR & \textbf{0.8722} & \textbf{0.8790} & \textbf{0.8722} & \textbf{0.8716}\\ \hline
-  \end{tabular}
-\end{table}
 
@@
 For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy.
-Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences.
+Table~\ref{tab:result_under_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences.
 These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still has 2.41\% lower F1-score than first-order Markov chain.
@@
 
-\subsection*{\textit{Common patterns}}
+\subsection*{\textit{Most likely communication sequences}}
 Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question.
+It can be seen that most successful sequence start with a summary of the discussion or structured statement.
 After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
-On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
+On the other hand, providing additional information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
 This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation.
 Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews.
-\\
-
 \begin{table}[h]
 \centering
-\caption{Common patterns in successful and unsuccessful motivational interviews.}
+\caption{Most likely communication sequences in successful and unsuccessful motivational interviews.}
 \label{tab:common_patterns}
@@
   \hline
-   \textbf{Type} & \textbf{Pattern} \\ \hline      
-successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
-& commitment language positive $\rightarrow $ [positive commitment] \\\hline
-successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
-& commitment language positive $\rightarrow $ [positive commitment] \\\hline
-unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
-& General Information, positive $\rightarrow $ [negative commitment] \\\hline
-unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
-& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
+   \textbf{Type} & \textbf{Most likely communication sequences} \\ \hline      
+successful & GINFO+: General information, positive $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ OQTBN: \\ 
+& Open-ended question, target behavior neutral \\\hline
+successful & SS: Structure session $\rightarrow $ GINFO+: General information, positive $\rightarrow $ CQECHT+: Closed-ended \\
+& question, elicit change talk positive\\\hline
+successful & SO: Statement, other $\rightarrow $ LUP+: Low uptake, positive $\rightarrow $ AF: Affirm $\rightarrow $ HUPW: High uptake, \\
+& weight $\rightarrow $ OQECML+: Open-ended question, elicit commitment language positive. \\\hline
+unsuccessful & ADV+: Advise, positive $\rightarrow $ AMB-: Ambivalence negative $\rightarrow $ OQECHT-: Open-ended \\
+& question, elicit change talk negative\\\hline
+unsuccessful & CQECHT+: Open-ended question, elicit change talk positive $\rightarrow $ RCHT-S: Reflect, change talk \\
+& negative $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative \\\hline
+unsuccessful & SUP: Support $\rightarrow $ AF: Affirm $\rightarrow $ CQTBN: Closed-ended question, target behavior neutral \\
+& $\rightarrow $ OQECHT-: Open-ended question, elicit change talk negative $\rightarrow $ AMB-: Ambivalence negative\\\hline
   \end{tabular}
@@
 We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help with transcribing the recordings of motivational interviews.
-\pagebreak
+%\pagebreak
 

### 2017-09-08 most-likely-patterns/obesity-patterns-and-size.py

@@ -28,14 +28,20 @@ with open(successful_filename, "r") as file_stream:
         total_utterances += len(codes)
         successful_patterns += 1
 
+avg_length_successful = float(total_utterances)/successful_patterns
+
+unsuccessful_seq_length = 0
 with open(unsuccessful_filename, "r") as file_stream:
     for line in file_stream:
         codes = line.replace("\n", "").strip().split(",")
-        total_utterances += len(codes)
+        unsuccessful_seq_length += len(codes)
         unsuccessful_patterns += 1
 
-print("Successful patterns: " + str(successful_patterns))
-print("Unsuccessful patterns: " + str(unsuccessful_patterns))
+avg_length_unsuccessful = float(unsuccessful_seq_length)/unsuccessful_patterns
+total_utterances += unsuccessful_seq_length
+
+print("Successful patterns: " + str(successful_patterns), ("Avg. Length: " + str(avg_length_successful)))
+print("Unsuccessful patterns: " + str(unsuccessful_patterns), ("Avg. Length: " + str(avg_length_unsuccessful)))
 print("Total utterances: " + str(total_utterances))
 
 # set dictionary for successful sequences

### 2017-09-09 paper-amia-summit-2018/amia_summit_2018.tex
update word embedding part of the paper

@@
 In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.
-Our method achieved 70.03\%, 48.65\%, 88.53\% F1-score for under-sampled sequences and 78.96\%, 77.06\%, 87.16\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively.
+Our method achieved 70.38\%, 60.67\%, 86.77\% F1-score for under-sampled sequences and 77.75\%, 75.20\%, 83.81\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively.
 The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}
@@
 \subsection*{\textit{Data collection}}
-The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
+The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
 Each transcript consists of a counselor-adolescent and a counselor-caregiver session.
@@
 Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
-Out of 4169 observed sequences, 4225 were positive and 918 were negative.
+Out of 5143 observed sequences, 4225 were positive and 918 were negative.
 For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences.
@@
 
-\textbf {MYSCOPE Code Embedding}: is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
-In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.
-RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.
-Stochastic Neighbor Embedding for MYSCOPE code is shown in Figure~\ref{fig:code_embedding}.
+\textbf {MYSCOPE code embedding}: We utilize word embedding in place of one-hot vectors for the representation of behavior code.
+Because one-hot vectors are high-dimensional and sparse.
+On the other hand, word embeddings have the ability to represent semantically similar words with similar vectors in a lower dimensional space by retaining their internal relations.
+Figure~\ref{fig:code_embedding} illustrates the myscope code embedding visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
+It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc are nearest neighbors of CQECML+, which make sense in real world practice.
 \begin{figure}[!htb]

### 2017-09-09 paper-amia-summit-2018/amia_summit_2018.tex
update word embedding part of the paper

@@
 Figure~\ref{fig:code_embedding} illustrates the myscope code embedding visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
-It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc are nearest neighbors of CQECML+, which make sense in real world practice.
+It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc.
+are nearest neighbors of CQECML+, which make sense in real world practice.
 \begin{figure}[!htb]

### 2017-09-10 paper-amia-summit-2018/amia_summit_2018.tex
update word embedding part of the paper
update results

@@
 It was shown that RNNs are very effective in sequence modeling with temporal data.
-Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave superior performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
+Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave excellent performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
 
@@
 Each transcript consists of a counselor-adolescent and a counselor-caregiver session.
-Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions.
 The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups.
@@
 Out of 5143 observed sequences, 4225 were positive and 918 were negative.
-For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences.
+For each of the probabilistic models (Markov chain and HMM), two models were trained, one model was trained using successful sequences and another one was trained using unsuccessful sequences.
 Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}.
@@
 Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data.
-Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%.
+Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%.
 Therefore, it is important to handle imbalance data properly.
-Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels.
-In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set.
+Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level.
+In this study, we utilized under and over sampling methods at data processing level for balancing the adolescent obesity data set.
 Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
-In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}.
+We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}.
 On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
@@
 Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors.
-In this paper, we only consider those models that can capture the past events for the classification of behavior sequences.\\
+For this reason, we only consider those models that can capture the past events for the classification of behavior sequences.\\
 
@@
 OQTBN &	Open-ended question, target behavior neutral & Counselor &	That's good.  So, tell me  how do you feel about your weight? \\\hline
-CHT+ &	Change Talk positive	& Adolescent &	It's not the best. \\\hline
+CHT+ &	Change talk positive	& Adolescent &	It's not the best. \\\hline
 CQECHT+ & Closed question, elicit change talk positive & Counselor & It's not the best? \\\hline
-CHT+ &	Change Talk positive &	Adolescent & Yeah \\\hline
+CHT+ &	Change talk positive &	Adolescent & Yeah \\\hline
 CQTBN &	Closed question, target behavior neutral  & Counselor &	Okay, so have you tried to lose weight before? \\\hline
@@
 However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification.
-Therefore, we employed two special variants of RNN models in our experiemnts: long short-term memory (LSTM) and gated recurrent unit (GRU).
+Therefore, we employed two special variants of RNN models in our experiments: long short-term memory (LSTM) and gated recurrent unit (GRU).
 
-\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property).
-Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable.
+\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state and not on its past states (Markov property).
 For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
@@
 \begin{itemize}
-\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUPW, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
+\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
 \item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
@@
 \begin{equation}
-p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
+p(S\rightarrow successful) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
 \end{equation}
-If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.
+If $p(S\rightarrow successful) > 0 $, the interview is predicted to result in positive change talk or commitment language.
+Otherwise, it would be classified as a negative change talk or commitment language.
 
-The Markov model we have discussed so far is referred to as the first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
-A generalization of the first-order MC is the $k^{th}$ order MC, in which the transition probability to a particular state (symbol) $x_i$ is computed by looking at the k preceding states (symbols).
-Thus, $k^{th}$ order Markov chain will have $N^{k}$ states each associated with a sequence of k symbols.
-In our experiment, we also used second order Markov model.
-Hence, our model has $N^2$ states with transition probability matrix of size $N^2 \times N$.
+The above model also refered as first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
+In our experiment, we also considered second-order Markov model that compute transition probabilities based on previous two states.
 
-\textbf {Hidden Markov Model (HMM)} is a powerful tool for statistical modeling of processes varying in time.
+\textbf {Hidden Markov Model (HMM)} is another probabilistic model, which is used for modeling processes varying in time.
 HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations.
-An HMM can be fully specified by the following quintuple: $\lambda = (N, M, A, B, \pi)$
+Mathmatically, HMM can be defined as the following: $\lambda = (N, M, A, B, \pi)$
 \begin{itemize}
@@
 \end{itemize}
-We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$.
+HMM is also represented by a compact notation: $\lambda = (A, B, \pi)$ after excluding the structure parameters M and N.
 The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
@@
 
-\textbf {MYSCOPE code embedding}: We utilize word embedding in place of one-hot vectors for the representation of behavior code.
+\textbf {MYSCOPE code embedding}: We utilize word embedding technique in place of one-hot vectors for the representation of behavior code.
 Because one-hot vectors are high-dimensional and sparse.
@@
 It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc.
-are nearest neighbors of CQECML+, which make sense in real world practice.
+are nearest neighbors of CQECML+.
 \begin{figure}[!htb]
@@
 The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
-We used one hidden layer of 20 LSTM or GRU nodes after embedding the given myscope codes from the input sequence.
+We used one hidden layer of 15 LSTM or GRU nodes after embedding the given myscope codes from the input sequence.
 Then, we applied a softmax layer at each time step for predicting the target.
@@
 As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
-Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
+Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$.
+Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
 We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss.
@@
 Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds.
-However, LSTM and GRU are trained on 70\% of the data and validated on 20\%.
+However, LSTM and GRU are trained on 80\% of the data and validated on 10\%.
 The remaining 10\% of the data is used as a test set for reporting the performance of the model.
@@
 Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.
+Predictive performance summary of the proposed methods on under- and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.
 \begin{table}[h]
@@
   \hline
-   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under Sampling}} & \multicolumn{3}{|c|}{\textbf{Over Sampling}} \\\cline{2-7}
+   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under-sampled dataset}} & \multicolumn{3}{|c|}{\textbf{Over-sampled dataset}} \\\cline{2-7}
    & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
@@
 \subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-We applied small learning rate 0.00005 and early stopping strategy with 150 epoch for training deep learning models when under-sampled data is used.
-Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.
+We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used.
 Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}.
-First, recurrent neural networks outperform over probabilistic models and achieved 18.50\%-39.88\% higher F1-score.
-Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858.
-Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication.
+First, recurrent neural networks outperform over probabilistic models and achieved 16.39\%-26.1\% higher F1-score.
+Second, LSTM with target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681.
+Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than vanilla LSTM.
 Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
-Fifth, it follows that second-order MC have lower accuracy, precision, recall, and F-measure than first-order MC.
-In particular, accuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models.
-At last, accuracy and recall illustrate exactly the same values for all models.
+Fifth, it follows that second-order MC have lower precision, recall, and F-measure than first-order MC.
+In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first- to second-order models.
 
 \subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
-For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy.
-Table~\ref{tab:result_under_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences.
-These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still has 2.41\% lower F1-score than first-order Markov chain.
-Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models.
-However, the predictive performance of recurrent neural networks decreases in over-sampled sequences whereas the performance of probabilistic models decreases.
+Similar to the under-sampled dataset, early stopping strategy is also employed for the over-sampled dataset.
+For over-sampled data, RNN models are trained with learning rate 0.00010 and batch size 55.
+Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%, 2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain.
+Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with target replication has the highest F1-score among all models.
+However, the predictive performance of recurrent neural networks decreases in over-sampled sequences although the performance of probabilistic models increases.
 We also observed that all models have the largest value in precision compared to other performance metrics.
+\\
 
-\subsection*{\textit{Most likely communication sequences}}
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-It can be seen that most successful sequence start with a summary of the discussion or structured statement.
-After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
-On the other hand, providing additional information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
-This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation.
-Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews.
 \begin{table}[h]
@@
 \end{table}
+
+\subsection*{\textit{Most likely communication sequences}}
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
+After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
+On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
+This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation.
+Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
 
@@
 Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource intensive manual qualitative analysis of interview transcripts.
-Our proposed method can also help to identify specific patterns that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.
+Our proposed method can also help to identify most likely sequences that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.
 This work also has broad implications for qualitative public health research by providing a formal theoretically-grounded computational mechanism to facilitate the development of effective behavioral interventions.

### 2017-09-10 paper-amia-summit-2018/amia_summit_2018.tex
update word embedding

@@
 On the other hand, word embeddings have the ability to represent semantically similar words with similar vectors in a lower dimensional space by retaining their internal relations.
-Figure~\ref{fig:code_embedding} illustrates the myscope code embedding visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
+Figure~\ref{fig:code_embedding} illustrates the myscope code embedding in 44 dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
 It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc.

### 2017-09-11 paper-amia-summit-2018/amia_summit_2018.tex
update discussion section

@@
 By analyzing the experimental results of different communication sequence outcome prediction schemes proposed in this paper, we arrived at the following conclusions.
-First, using higher-order models results in better prediction accuracy compared to lower-order models, since higher-order models utilize larger context and can better capture different nuances of patient-provider communication.
-On the other hand, the number of states in higher-order Markov models grows exponentially with their order.
+First, the overall predictive performance of RNN models is substantially better than probabilistic models.
+In particular, RNN-based method achieves near-human accuracy for predicting the success of motivational interviews.
+This indicates that RNN are able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
+In addition to that word embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+
+Second, computing loss with target replication at each time step results in better performance for all configurations of the proposed RNN method.
+This indicates that mean of the losses over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters.
+Better estimates of parameters in RNN models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence.
+This allows to achieve an improvement in the prediction accuracy of the method.
+
+Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models.
+Because the number of states in higher-order Markov models grows exponentially with their order.
 Therefore, accurate estimation of transition probabilities requires much larger training set.
-Using smaller datasets will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates.
+Using smaller datasets such as under-sampled dataset will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates.
 Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing.
-In this study, we found out that using second-order Markov models is a reasonable trade-off between efficiency and accuracy.
+In this study, we found out that using first-order Markov models is a reasonable trade-off between efficiency and accuracy.
 
-Second, the overall predictive performance of HMM is substantially better than MC for both types of sequences.
-In particular, HMM-based method achieves near-human accuracy for predicting the success of motivational interviews.
-This indicates that hidden states in HMM are able to capture the structure of discourse in motivational interviews by grouping together the codes that correspond to different cognitive states of adolescents during motivational interviews, which reflect the overall progression of the interviews.
-This allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+Fourth, similar to traditional markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used.
+This indicates that sufficient training data is required to estimate optimal hyperparameters such as number of hidden states, intitial state distribution, transition probabilities, and emission probabilities.
 
-Third, converting natural PPC communication sequences to alternating ones results in better performance for all configurations of the proposed method.
-This indicates that alternating sequences emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the corresponding conditional probabilities.
-Better estimates of parameters in Markov models of successful and unsuccessful interviews are propagated to the next step, where they are aggregated into predictions for the entire sequence.
-This allows to achieve a dramatic improvement in the prediction accuracy of the method.
-
-Fourth, the proposed method can be used to identify the most effective communication strategies at eliciting a particular type of behavioral response.
+Fifth, the proposed method can be used to identify the most effective communication strategies at eliciting a particular type of behavioral response.
 Awareness of these strategies by researchers can significantly decrease the time and effort required to develop effective interventions to address many public health conditions, such as childhood obesity, and tailor these interventions to particular patient cohorts.

### 2017-09-11 paper-amia-summit-2018/amia_summit_2018.tex
update discussion section

@@
 First, the overall predictive performance of RNN models is substantially better than probabilistic models.
-In particular, RNN-based method achieves near-human accuracy for predicting the success of motivational interviews.
-This indicates that RNN are able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
-In addition to that word embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+In particular, the RNN-based method achieves near-human accuracy for predicting the success of motivational interviews.
+This indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
+In addition to that word-embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
 
@@
 Better estimates of parameters in RNN models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence.
-This allows to achieve an improvement in the prediction accuracy of the method.
+This allows achieving an improvement in the prediction accuracy of the method.
 
@@
 
-Fourth, similar to traditional markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used.
-This indicates that sufficient training data is required to estimate optimal hyperparameters such as number of hidden states, intitial state distribution, transition probabilities, and emission probabilities.
+Fourth, similar to traditional Markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used.
+This indicates that sufficient training data is required to estimate optimal hyper parameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
 
@@
 Awareness of these strategies by researchers can significantly decrease the time and effort required to develop effective interventions to address many public health conditions, such as childhood obesity, and tailor these interventions to particular patient cohorts.
-Awareness of these strategies by the counselors can lead to greater success rate of motivational interviews.
+Awareness of these strategies by the counselors can lead to a greater success rate of motivational interviews.
 

### 2017-09-12 paper-amia-summit-2018/amia_summit_2018.tex
update embedding

@@
 
-\textbf {MYSCOPE code embedding}: We utilize word embedding technique in place of one-hot vectors for the representation of behavior code.
+\textbf {Embedding}: We are inspired by the idea of word embedding to represent behavior code of the sequences.
+Word embedding is a representation of words in low-dimensional space by vectors which contain the feature of the words.
+In our study, we employed embedding in place of one-hot vectors for the representation of behavior code.
 Because one-hot vectors are high-dimensional and sparse.
-On the other hand, word embeddings have the ability to represent semantically similar words with similar vectors in a lower dimensional space by retaining their internal relations.
-Figure~\ref{fig:code_embedding} illustrates the myscope code embedding in 44 dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
-It can be seen that similar behavior codes are placed close to each other in high dimensional vector space such as OQECHT+, OQECML+, AF, ADV+, RCML+S, etc.
-are nearest neighbors of CQECML+.
+On the other hand, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
+Figure~\ref{fig:code_embedding} illustrates the myscope code embedding in 44-dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
+It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, ADV+, RCML+S, CQECML+, etc.
+formed a cluster which is highlighted in the left part of the figure.
+However, right part of the figure illustrates another cluster formed with negative behaviors including CQECML-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
+Because they are semantically similar in the low dimensional vector space.
+
 \begin{figure}[!htb]
@@
 We used one hidden layer of 15 LSTM or GRU nodes after embedding the given myscope codes from the input sequence.
-Then, we applied a softmax layer at each time step for predicting the target.
+Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful).
 This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
@@
   \hline
-   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under-sampled dataset}} & \multicolumn{3}{|c|}{\textbf{Over-sampled dataset}} \\\cline{2-7}
+   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under Sampling}} & \multicolumn{3}{|c|}{\textbf{Over Sampling}} \\\cline{2-7}
    & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
@@
 We also observed that all models have the largest value in precision compared to other performance metrics.
-\\
+
+\subsection*{\textit{Most likely communication sequences}}
+Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
+It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
+After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
+On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
+This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation.
+Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
 
@@
 \end{table}
-
-\subsection*{\textit{Most likely communication sequences}}
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
-After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
-On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
-This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation.
-Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
 

### 2017-09-12 paper-amia-summit-2018/amia_summit_2018.tex
update embedding section

@@
 
-\textbf {Embedding}: We are inspired by the idea of word embedding to represent behavior code of the sequences.
+\textbf {Embedding}: We are inspired by the idea of word embedding \cite{bengio2003neural} to represent behavior code of the sequences.
 Word embedding is a representation of words in low-dimensional space by vectors which contain the feature of the words.
@@
 formed a cluster which is highlighted in the left part of the figure.
-However, right part of the figure illustrates another cluster formed with negative behaviors including CQECML-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
+However, right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
 Because they are semantically similar in the low dimensional vector space.

### 2017-09-12 deeplearning/obesity-gru-lstm.py
add comments for deep learning code embedding

@@ -61,6 +61,7 @@ codebook = utility.loadCodeBook(codebook_filename)
 
 
 def RNN(x, weights, biases, step_size):
+    # layer to take the codes and convert them into vectors (embeddings)
     embed_x = tf.contrib.layers.embed_sequence(x, vocab_size=len(codebook), embed_dim=embed_dimension)
     cell = tf.contrib.rnn.LSTMCell(hidden_units)
     # cell = tf.contrib.rnn.GRUCell(hidden_units)

### 2017-09-12 paper-amia-summit-2018/amia_summit_2018.tex
update embedding
update code embedding figure

@@
 On the other hand, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
+The embedding was learned during the training of the RNN model.
 Figure~\ref{fig:code_embedding} illustrates the myscope code embedding in 44-dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
-It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, ADV+, RCML+S, CQECML+, etc.
+It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc.
 formed a cluster which is highlighted in the left part of the figure.
-However, right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
+However, right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
 Because they are semantically similar in the low dimensional vector space.
@@
     \centering
-    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
+    \includegraphics[width=0.50\textwidth]{figures/word_embed_highlighted.eps}
     \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}

### 2017-09-12 paper-amia-summit-2018/amia_summit_2018.tex
update code embedding figure

@@
 
-\textbf {Embedding}: We are inspired by the idea of word embedding \cite{bengio2003neural} to represent behavior code of the sequences.
+\textbf {Embedding}: We took the inspiration for the representation of behavior codes from the idea of word embedding \cite{bengio2003neural}.
 Word embedding is a representation of words in low-dimensional space by vectors which contain the feature of the words.
-In our study, we employed embedding in place of one-hot vectors for the representation of behavior code.
-Because one-hot vectors are high-dimensional and sparse.
+In our study, we employed embedding in place of one-hot vectors for the representation of behavior code since one-hot vectors are high-dimensional and sparse.
 On the other hand, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
@@
 However, right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
-Because they are semantically similar in the low dimensional vector space.
+because they are semantically similar in the low dimensional vector space.
 

### 2017-09-12 paper-amia-summit-2018/amia_summit_2018.tex
update embedding

@@
 In our study, we employed embedding in place of one-hot vectors for the representation of behavior code since one-hot vectors are high-dimensional and sparse.
-On the other hand, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
+Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
 The embedding was learned during the training of the RNN model.

### 2017-09-15 paper-amia-summit-2018/amia_summit_2018.tex
update embedding figure

@@
 It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc.
-formed a cluster which is highlighted in the left part of the figure.
+formed a cluster which is shown in the left part of the figure.
+The nearest neighbors of CQECML+ are also highlighted by their color intensity i.e.
+OQECML+ is more purple means closer to CQECML+.
 However, right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
@@
     \centering
-    \includegraphics[width=0.50\textwidth]{figures/word_embed_highlighted.eps}
+    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
     \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}

### 2017-09-16 paper-amia-summit-2018/amia_summit_2018.tex
modified paper based on feedback from DMC

@@
 Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics.
-The order of observations in a sequence is challenging to capture using features, which makes sequence classification a more challenging task than traditional classification.
-In general, sequence classification methods can be divided into three categories: feature-based, distance metric-based and model-based method.
-Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
-Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
-Distance-based methods measure the similarity between sequences to determine the quality of classification.
-The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
-The third type of sequence classification methods, first creates a probabilistic model of a sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
-
-However, Markov chain cannot handle long-term dependencies in sequences.
-Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step.
-Therefore, Markov chain and HMMs are only capable of passing short-term dependency among sequence steps.
-On the other hand, Recurrent Neural Networks (RNNs) have no such limitations.
-RNNs are capable to capture long-term dependencies with their unlimited memory.
-It was shown that RNNs are very effective in sequence modeling with temporal data.
-Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and Gated Reccurent Unit\cite{chung2014empirical} (GRU), which gave excellent performance in a wide range of sequence modeling problems \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
-
+For classifying temporally ordered sequenced, sequence classification methods are utilized.
 Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection.
@@
 
-In this paper, we direct our focus towards patient-provider communication sequences in the context of clinical interviews.
+In general, there are three categories of classification methods for teporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method.
+Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
+Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
+Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification.
+The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
+The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation.
+At the same time, distance-based methods are more appropriate for time series data.
+Model-based classification methods train a model by data and then applies it for the classification of new datasets.
+
+%This makes sequence classification a more challenging task than traditional classification.
+
+One of the model-based classification methods first creates a probabilistic model of a behavioral code sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences.
+Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step.
+Therefore, Markov chain and HMMs are only capable of passing upto five preceding steps among sequence steps[ref]\cite{}.
+On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory.
+It was shown that RNNs are very effective in sequence modeling with temporal data.
+Excellent performance in a wide range of sequence modeling problems with temporal data has been observed using RNNs \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
+A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem}.LSTM showed excellent performance in several domain, including speech recognition \cite{graves2013speech}.
+A further simplification and improvement of the LSTM model, called the Gated Reccurent Unit\cite{chung2014empirical} (GRU), was later proposed.
+%Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and an enhanced version of LSTM . 
+These variants show markedly improved performance among all RNN variants and as such, this work focuses on these varians of RNN specifically.
+
+In this paper, we test the applicability of state of the art models for the classification of patient-provider communication sequences in the context of clinical interviews.
 Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
@@
 One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
-The goal of MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
-These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.
+The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
+These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
 However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
-Since manual annotation and examination of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years.
+Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years.
 Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.
 
-Our research addresses this problem from multiple angles.
+%This work is the next step in a series of experiments to develop the [informatics based methods described above].
+
 In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}.
-In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
+In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
 change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
@@
 The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
-Each transcript consists of a counselor-adolescent and a counselor-caregiver session.
-The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups.
+Each transcript consists of a MI interview session involving counselor, adolescent, and caregiver.
+The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into client codes (adolescent and caregiver) group and counselor codes group.
 Utterances were divided into successful and unsuccessful communication sequences.
-Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
+Successful communication sequences result in positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
 Out of 5143 observed sequences, 4225 were positive and 918 were negative.
+Here, the successful sequences had an average length of 9.79 while that of unsuccessful sequences was 9.65.
 For each of the probabilistic models (Markov chain and HMM), two models were trained, one model was trained using successful sequences and another one was trained using unsuccessful sequences.
-Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}.
-\\
 
-\begin{table}[h]
-\centering
-\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
-\label{tab:data_dist}
-  \begin{tabular}{|l|l|l|l|l|}
-  \hline
-   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
-Successful sequences & 4225 & 82.15\% & 9.79 \\\hline
-Unsuccessful sequences & 918 & 17.85\% & 9.65 \\\hline
-  \end{tabular}
-\end{table}
+%Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\
+%
+%\begin{table}[h]
+%\centering
+%\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
+%\label{tab:data_dist}
+%  \begin{tabular}{|l|l|l|l|l|}
+%  \hline
+%   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
+%Successful sequences & 4225 & 82.15\% & 9.79 \\\hline
+%Unsuccessful sequences & 918 & 17.85\% & 9.65 \\\hline
+%  \end{tabular}
+%\end{table}
 
-As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences.
+Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences.
 Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data.
@@
 A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
-Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive feedback at the end.
+Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end.
 It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively.
@@
 Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
-The embedding was learned during the training of the RNN model.
-Figure~\ref{fig:code_embedding} illustrates the myscope code embedding in 44-dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
+The embedding was learned during the training of the classification model.
+Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
 It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc.
 formed a cluster which is shown in the left part of the figure.
-The nearest neighbors of CQECML+ are also highlighted by their color intensity i.e.
+The nearest neighbors of CQECML+ are highlighted by their color intensity i.e.
 OQECML+ is more purple means closer to CQECML+.
@@
 because they are semantically similar in the low dimensional vector space.
+It is interesting that the behaviors intended to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.
 
@@
 The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
-We used one hidden layer of 15 LSTM or GRU nodes after embedding the given myscope codes from the input sequence.
+We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence.
 Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful).
@@
 \section*{Results}
-Experimental evaluation of the proposed method is conducted on both under- and over-sampled sequences.
-Predictive performance summary of the proposed methods on under- and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.
+Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences.
+Predictive performance summary of the proposed methods on under- and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\
 \begin{table}[h]
@@
 Second, LSTM with target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681.
-Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than vanilla LSTM.
+Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM.
 Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
@@
 We also observed that all models have the largest value in precision compared to other performance metrics.
-
-\subsection*{\textit{Most likely communication sequences}}
-Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews.
-It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
-After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
-On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
-This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation.
-Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
 
@@
 
+\subsection*{\textit{Most likely communication sequences}}
+Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews.
+It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
+After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.
+On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication.
+This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation.
+Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
+
 \section*{Discussion}
@@
 This indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
-In addition to that word-embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+This provides evidence that the RNN is able to successfully replicate human cognitive processes to integrate previous information when formulating higher level thinking.
+In addition to that, word-embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
 
-Second, computing loss with target replication at each time step results in better performance for all configurations of the proposed RNN method.
+Second, using target replication to compute the losses at each time step results in better performance for all configurations of the proposed RNN method.
 This indicates that mean of the losses over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters.
@@
 \section*{Acknowledgments}
-This study was supported by an R21 grant DK108071 from the NIH.
-We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help with transcribing the recordings of motivational interviews.
+This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
+We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training data set by manually annotating the data set using the MYSCOPE codebook.
 %\pagebreak

### 2017-09-18 paper-amia-summit-2018/amia_summit_2018.tex
update reference
update introduction

@@
 
-In general, there are three categories of classification methods for teporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method.
+In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method.
 Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
@@
 At the same time, distance-based methods are more appropriate for time series data.
-Model-based classification methods train a model by data and then applies it for the classification of new datasets.
+Model-based classification methods involve building a model based on the dataset.
+In other words, information was learned from the dataset, and use that as a "model" to classify sequences.
 
@@
 Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences.
-Similarly, the hidden states of HMMs follow a first-order markov chain and can select only one hidden state at each time step.
-Therefore, Markov chain and HMMs are only capable of passing upto five preceding steps among sequence steps[ref]\cite{}.
+Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step.
+Therefore, Markov chain and HMMs are only used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}.
 On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory.
@@
 Excellent performance in a wide range of sequence modeling problems with temporal data has been observed using RNNs \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
-A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem}.LSTM showed excellent performance in several domain, including speech recognition \cite{graves2013speech}.
-A further simplification and improvement of the LSTM model, called the Gated Reccurent Unit\cite{chung2014empirical} (GRU), was later proposed.
-%Recently, two variant of RNNs is proposed: Long Short-Term Memory\cite{graves2013speech} (LSTM) and an enhanced version of LSTM . 
-These variants show markedly improved performance among all RNN variants and as such, this work focuses on these varians of RNN specifically.
+A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem}.
+LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}.
+In Lipton et.
+al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc.
+LSTM was also effectively used, recently, for designing a model, called Doctor AI \cite{choi2016doctor}, to predict the diagnosis and medication categories for a subsequent visit.
+A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed.
+These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.
 
@@
 Childhood obesity is a serious public health concern in the United States and worldwide.
-Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
+Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
 Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
@@
 The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
-Each transcript consists of a MI interview session involving counselor, adolescent, and caregiver.
+Each transcript consists of an MI interview session involving counselor, adolescent, and caregiver.
 The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into client codes (adolescent and caregiver) group and counselor codes group.
@@
 Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level.
-In this study, we utilized under and over sampling methods at data processing level for balancing the adolescent obesity data set.
-Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
+In this study, we utilized under and oversampling methods at data processing level for balancing the adolescent obesity data set.
+Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
 We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}.
-On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
+On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
 A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
@@
 
-The above model also refered as first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
+The above model also referred as first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
 In our experiment, we also considered second-order Markov model that compute transition probabilities based on previous two states.
@@
 HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-Mathmatically, HMM can be defined as the following: $\lambda = (N, M, A, B, \pi)$
+Mathematically, HMM can be defined as the following: $\lambda = (N, M, A, B, \pi)$
 \begin{itemize}
@@
 Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
-We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss.
+We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss.
 Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization.
@@
 We found out that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
-Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource intensive manual qualitative analysis of interview transcripts.
+Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource-intensive manual qualitative analysis of interview transcripts.
 Our proposed method can also help to identify most likely sequences that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.

### 2017-09-18 paper-amia-summit-2018/amia_summit_2018.tex
update abstract and introduction

@@
 Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics.
-For classifying temporally ordered sequenced, sequence classification methods are utilized.
-Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection.
+For classifying temporally ordered sequences, sequence classification methods are utilized.
+Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}.
+In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins.
+In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease.
+Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.
+
+%Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection.
 In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}.
@@
 Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
-Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
+%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
 Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification.
-The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
+The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
+%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
 The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation.
@@
 
-One of the model-based classification methods first creates a probabilistic model of a behavioral code sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+One of the model-based classification methods is probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
 Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences.
 Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step.
-Therefore, Markov chain and HMMs are only used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}.
+Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}.
 On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory.
-It was shown that RNNs are very effective in sequence modeling with temporal data.
-Excellent performance in a wide range of sequence modeling problems with temporal data has been observed using RNNs \cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
-A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem}.
+It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
+A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN.
 LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}.
@@
 al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc.
-LSTM was also effectively used, recently, for designing a model, called Doctor AI \cite{choi2016doctor}, to predict the diagnosis and medication categories for a subsequent visit.
+LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit.
 A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed.
@@
 
-In this paper, we test the applicability of state of the art models for the classification of patient-provider communication sequences in the context of clinical interviews.
+In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews.
 Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
@@
 The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
-These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
+These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
 However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
@@
     \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
-    \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}
+    \caption{\textbf{T-SNE diagram of behavior codes drawn after 4000 steps with perplexity 30}.}
     \label{fig:code_embedding}
@@
 Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews.
-It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
+In successful sequences, we observed the provision of information using patient-centered communication (GINFO+) and structure session which is the counselor either explaining the therapeutic agenda or attempting to transition to a new topic or session content (SS).
+Sometimes, counselor acknowledged the clients' communication or an off topic comment (SO).
+We also observed that the affirm and OQECML+ are consistent with MI theory as an effect for eliciting positive change talk or commitment language.
+For unsuccessful sequences, It can be seen that providing advice using non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-) which is heading in the wrong direction therapeutically.
+Questions phased to elicit negative change talk or commitment language leads to CHT- or CML-, which is consistent with the manual coding approach or analysis, or AMB-.
+
+%It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
 After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change.

### 2017-09-18 paper-amia-summit-2018/amia_summit_2018.tex
update method, discussion and conclusion

@@
 In this study, we utilized under and oversampling methods at data processing level for balancing the adolescent obesity data set.
-Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
+Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
 We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}.
@@
 Our proposed baseline prediction method consists of two steps.
-In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary.
+In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary.
 In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model.
-Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, won2004training, chai2001folk}.
+Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, won2004training}.
 However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification.
-Therefore, we employed two special variants of RNN models in our experiments: long short-term memory (LSTM) and gated recurrent unit (GRU).
+Therefore, we employed two special variants of RNN models in our experiments: long short-term memory and gated recurrent unit.
 
-\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state and not on its past states (Markov property).
+\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property).
 For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
@@
 \begin{itemize}
-\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
+\item $V = \{CML+, CHT+, CHT-, AMB-, LUP+, LUP-, HUPW, OQO, CQTBN, CQECHT+,...\}$ is a set of vertices, consisting of adolescent, caregiver and counselor MI behavior codes;
 \item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
@@
 
-\textbf {Hidden Markov Model (HMM)} is another probabilistic model, which is used for modeling processes varying in time.
+\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time.
 HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
@@
 This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
-As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
+As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
 Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$.
@@
 Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences.
-Predictive performance summary of the proposed methods on under- and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\
+Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\
 \begin{table}[h]
 \centering
-\caption{Performance of MC, HMM, and RNN for predicting the success of under- and over-sampled patient-provider communication sequences.
+\caption{Performance of MC, HMM, and RNN for predicting the success of under and over-sampled patient-provider communication sequences.
 The highest value for each performance metric is highlighted in bold.}
@@
   \hline
-   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under Sampling}} & \multicolumn{3}{|c|}{\textbf{Over Sampling}} \\\cline{2-7}
+   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Undersampling}} & \multicolumn{3}{|c|}{\textbf{Oversampling}} \\\cline{2-7}
    & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
@@
 Fifth, it follows that second-order MC have lower precision, recall, and F-measure than first-order MC.
-In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first- to second-order models.
+In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first to second-order models.
 
 \subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
-Similar to the under-sampled dataset, early stopping strategy is also employed for the over-sampled dataset.
-For over-sampled data, RNN models are trained with learning rate 0.00010 and batch size 55.
+Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset.
+For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size 55.
 Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%, 2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain.
@@
 First, the overall predictive performance of RNN models is substantially better than probabilistic models.
-In particular, the RNN-based method achieves near-human accuracy for predicting the success of motivational interviews.
+In particular, the RNN-based method achieves near-human accuracy for predicting the label of motivational interviews.
 This indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
@@
 Fourth, similar to traditional Markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used.
-This indicates that sufficient training data is required to estimate optimal hyper parameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
+This indicates that sufficient training data is required to estimate optimal hyperparameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
 

### 2017-09-19 paper-amia-summit-2018/amia_summit_2018.tex
update introduction

@@
 Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics.
-For classifying temporally ordered sequences, sequence classification methods are utilized.
+Sequence classification methods are utilized for classifying temporally ordered sequences.
 Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}.
@@
 Model-based classification methods involve building a model based on the dataset.
-In other words, information was learned from the dataset, and use that as a "model" to classify sequences.
+In other words, information is learned from the dataset, and used as a ``model'' to classify sequences.
 
@@
 
-In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews.
-Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
 Childhood obesity is a serious public health concern in the United States and worldwide.
@@
 %This work is the next step in a series of experiments to develop the [informatics based methods described above]. 
-
 In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}.
-In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
+In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews.
+Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
+We propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
 change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
@@
 On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
+
 A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
@@
 HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-Mathematically, HMM can be defined as the following: $\lambda = (N, M, A, B, \pi)$
+Mathematically, HMM can be defined as the following: $\lambda = (A, B, \pi)$.
 \begin{itemize}
-\item N is a number of hidden states in the model
-\item M is a number of distinct observations symbols per state, i.e. the discrete vocabulary size
 \item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
@@
 \end{itemize}
-HMM is also represented by a compact notation: $\lambda = (A, B, \pi)$ after excluding the structure parameters M and N.
+Hence, N is a number of hidden states in the model and M is a number of distinct observations symbols per state, i.e.
+the discrete vocabulary size.
 The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
@@
 \textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
-In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM.
+That RNN can capture long-term dependencies for predicting the future events is its main advantage of RNN over feedforward Neural Network, MC, and HMM.
 The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed.
@@
 Questions phased to elicit negative change talk or commitment language leads to CHT- or CML-, which is consistent with the manual coding approach or analysis, or AMB-.
-
 %It can be seen that most successful sequence starts with a summary of the discussion or structured statement.

### 2017-09-19 paper-amia-summit-2018/amia_summit_2018.tex
update introduction
update HMM model

@@
 Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level.
-In this study, we utilized under and oversampling methods at data processing level for balancing the adolescent obesity data set.
+In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity data set.
 Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
@@
 \end{itemize}
-Hence, N is a number of hidden states in the model and M is a number of distinct observations symbols per state, i.e.
+Hence, $N$ is a number of hidden states in the model and $M$ is a number of distinct observations symbols per state, i.e.
 the discrete vocabulary size.

### 2017-09-19 paper-amia-summit-2018/amia_summit_2018.tex
update HMM model

@@
 the discrete vocabulary size.
-The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state.
+The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols (emission probabilities) for each state.
 The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations.

### 2017-09-20 paper-amia-summit-2018/amia_summit_2018.tex
update RNN figure
fixed spelling mistakes

@@
 Model-based classification methods involve building a model based on the dataset.
-In other words, information is learned from the dataset, and used as a ``model'' to classify sequences.
+In other words, information is learned from the dataset and used as a ``model'' to classify sequences.
 
@@
 
-One of the model-based classification methods is probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
+One of the model-based classification methods is a probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
 Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences.
@@
 
-Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences.
+Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences.
 Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data.
@@
 Therefore, it is important to handle imbalance data properly.
-Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level.
-In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity data set.
+Many solutions are proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level.
+In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity dataset.
 Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
@@
 \textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
-That RNN can capture long-term dependencies for predicting the future events is its main advantage of RNN over feedforward Neural Network, MC, and HMM.
+That RNN can capture long-term dependencies for predicting the future events is its main advantage over feedforward Neural Network, MC, and HMM.
 The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed.
@@
     \centering
-    \includegraphics[width=0.70\textwidth]{figures/rnn.eps}
+    \includegraphics[width=0.70\textwidth]{figures/rnn_small.eps}
     \caption{\textbf{Proposed RNN model with target replication (TR)}.}
@@
 This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
-We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training data set by manually annotating the data set using the MYSCOPE codebook.
+We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset by manually annotating the dataset using the MYSCOPE codebook.
 %\pagebreak

### 2017-09-22 paper-amia-summit-2018/amia_summit_2018.tex
update description of the RNN model

@@
 It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc.
-formed a cluster which is shown in the left part of the figure.
+formed a cluster at the left of the figure.
 The nearest neighbors of CQECML+ are highlighted by their color intensity i.e.
 OQECML+ is more purple means closer to CQECML+.
-However, right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
-because they are semantically similar in the low dimensional vector space.
+The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
 It is interesting that the behaviors intended to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.
@@
 r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
+\label{eq:resetgru}
 \end{equation}
@@
 \tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
+\label{eq:candidategru}
 \end{equation}
@@
 \end{equation}  
+As these equations, $\sigma$ stands for an element-wise application of the sigmoid function and $\odot$ is an element-wise product.
+The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W$ and $U$ represent weights and $b$ represents biases.
+The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
 \begin{equation}
@@
 We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence.
-Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful).
+Then, we applied a softmax layer at each time-step for predicting the target which is the predicted label of the sequence.
 This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
@@
 The remaining 10\% of the data is used as a test set for reporting the performance of the model.
-A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
+
+%A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class.
 The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)).
@@
 \end{table}  
-
 \subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}

### 2017-10-10 biosketch/mehedi-nihbiosketch.tex
update biosketch

@@
 I have the expertise and motivation necessary to successfully carry out the proposed research project.
-I have a background in computer science, with specific expertise in health informatics, machine learning, and textual data analytics.
+I have a background in computer science, with specific expertise in health informatics, machine learning, and textual
+data analytics.
 My research includes automatic annotation of clinical text, information extraction from electronic health records for improving the study of cancer risk and outcomes.
-As a graduate reserach assistant on several university- and NIH-funded grants, I successfully collaborated with other researchers, and produced the following publications.
+As a graduate research assistant at several university and NIH-funded grants, I successfully collaborated with other researchers and produced the following publications.
 
@@
 Manual encoding has been used to annotate clinical text of motivational interview transcripts for many years, which is time consuming and resource intensive.
-However, my reserach shown that state-of-the-art machine learning methods can be used as a cheap alternate for the annotation that can facilitate researchers to establish a causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
+However, my research has shown that state-of-the-art machine learning methods can be used as a cheap alternative for the annotation that can facilitate researchers to establish a causal relationship between different communication strategies and desired behavioral outcomes without having to repeatedly wade through pages of interview transcripts.
 
@@
 {Provider Communication Behaviors that Predict Motivation to Change in Black Adolescents with Obesity.}
-{The goal of this research was to identify communication behaviors used by weight loss counselors that mostly strongly predicted Black adolescents’ motivational statements.}
+{The goal of this research was to identify communication behaviors used by weight loss counselors that most strongly predicted Black adolescents’ motivational statements.}
 {Role: Graduate Research Assistant}

### 2017-10-10 spm-dmc/src/MainTestCMClaSP_saveToFile.java
added new project for sequential pattern mining

@@ -1,0 +1,65 @@
+
+
+import java.io.IOException;
+import java.io.UnsupportedEncodingException;
+import java.net.URL;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.clasp_AGP.AlgoCM_ClaSP;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clasp_AGP.dataStructures.creators.AbstractionCreator;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clasp_AGP.dataStructures.creators.AbstractionCreator_Qualitative;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clasp_AGP.dataStructures.database.SequenceDatabase;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clasp_AGP.idlists.creators.IdListCreator;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clasp_AGP.idlists.creators.IdListCreatorStandard_Map;
+
+/**
+ * Example of how to use the algorithm ClaSP, saving the results in a given file
+ *
+ * @author agomariz
+ */
+public class MainTestCMClaSP_saveToFile {
+
+    /**
+     * @param args the command line arguments
+     */
+    public static void main(String[] args) throws IOException {
+        // Load a sequence database
+        double support = 0.5;
+
+        boolean keepPatterns = true;
+        boolean verbose = true;
+        boolean findClosedPatterns = true;
+        boolean executePruningMethods = true;
+     // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+        boolean outputSequenceIdentifiers = false;
+
+        AbstractionCreator abstractionCreator = AbstractionCreator_Qualitative.getInstance();
+        IdListCreator idListCreator = IdListCreatorStandard_Map.getInstance();
+
+        SequenceDatabase sequenceDatabase = new SequenceDatabase(abstractionCreator, idListCreator);
+
+        //double relativeSupport = sequenceDatabase.loadFile(fileToPath("contextClaSP.txt"), support);
+        double relativeSupport = sequenceDatabase.loadFile("input/contextPrefixSpan.txt", support);
+        //double relativeSupport = sequenceDatabase.loadFile(fileToPath("gazelle.txt"), support);
+
+        AlgoCM_ClaSP algorithm = new AlgoCM_ClaSP(relativeSupport, abstractionCreator, findClosedPatterns, executePruningMethods);
+
+
[…]

### 2017-10-10 spm-dmc/src/MainTestCMSPADE_saveToFile.java

@@ -1,0 +1,56 @@
+import java.io.IOException;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.AlgoCMSPADE;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.candidatePatternsGeneration.CandidateGenerator;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.candidatePatternsGeneration.CandidateGenerator_Qualitative;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.dataStructures.creators.AbstractionCreator;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.dataStructures.creators.AbstractionCreator_Qualitative;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.dataStructures.database.SequenceDatabase;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.idLists.creators.IdListCreator;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spade_spam_AGP.idLists.creators.IdListCreator_FatBitmap;
+
+/**
+ * Example of how to use the algorithm SPADE, saving the results in a given
+ * file
+ * @author agomariz
+ */
+public class MainTestCMSPADE_saveToFile {
+
+    /**
+     * @param args the command line arguments
+     */
+    public static void main(String[] args) throws IOException {
+    	String outputPath ="output/output_cmspade.txt";
+        // Load a sequence database
+        double support = 0.5;
+
+        boolean keepPatterns = true;
+        boolean verbose = false;
+
+        AbstractionCreator abstractionCreator = AbstractionCreator_Qualitative.getInstance();
+        boolean dfs=true;
+        
+        // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+        boolean outputSequenceIdentifiers = false; 
+
+        IdListCreator idListCreator = IdListCreator_FatBitmap.getInstance();
+                
+        CandidateGenerator candidateGenerator = CandidateGenerator_Qualitative.getInstance();
+        
+        SequenceDatabase sequenceDatabase = new SequenceDatabase(abstractionCreator, idListCreator);
+
+        sequenceDatabase.loadFile("input/contextPrefixSpan.txt", support);
+        
[…]

### 2017-10-10 spm-dmc/src/MainTestCMSPAM_saveToFile.java

@@ -1,0 +1,48 @@
+
+
+import java.io.IOException;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.spam.AlgoCMSPAM;
+
+
+/**
+ * Example of how to use the SPAM algorithm in source code.
+ * @author Philippe Fournier-Viger
+ */
+public class MainTestCMSPAM_saveToFile {
+
+	public static void main(String [] arg) throws IOException{    
+		// Load a sequence database
+		//String input = fileToPath("D1C20T20N0.5S6I5_SPMF.txt");
+       String input = "input/contextPrefixSpan.txt";
+       String output = "output/output_cmspam.txt";
+            
+		// Create an instance of the algorithm 
+		AlgoCMSPAM algo = new AlgoCMSPAM(); 
+		
+		// This optional parameter allows to specify the minimum pattern length:
+		algo.setMinimumPatternLength(3);  // optional
+
+		// This optional parameter allows to specify the maximum pattern length:
+//		algo.setMaximumPatternLength(4);  // optional
+		
+		// This optional parameter allows to specify constraints that some
+		// items MUST appear in the patterns found by TKS
+		// E.g.: This requires that items 1 and 3 appears in every patterns found
+//		algo.setMustAppearItems(new int[] {1, 3});
+		
+		// This optional parameter allows to specify the max gap between two
+		// itemsets in a pattern. If set to 1, only patterns of contiguous itemsets
+		// will be found (no gap).
+		//algo.setMaxGap(1);
+		
+        // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+        boolean outputSequenceIdentifiers = true;
+		
+		// execute the algorithm with minsup = 2 sequences  (50 %)
+		algo.runAlgorithm(input, output, 0.5, outputSequenceIdentifiers);     // minsup = 106   k = 1000   BMS
+		algo.printStatistics();
+	}
+	
+}

### 2017-10-10 spm-dmc/src/MainTestCloSpan_saveToFile.java

@@ -1,0 +1,60 @@
+
+
+
+import java.io.IOException;
+import java.io.UnsupportedEncodingException;
+import java.net.URL;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.clospan_AGP.AlgoCloSpan;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clospan_AGP.items.SequenceDatabase;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clospan_AGP.items.creators.AbstractionCreator;
+import ca.pfv.spmf.algorithms.sequentialpatterns.clospan_AGP.items.creators.AbstractionCreator_Qualitative;
+
+/**
+ * Example of how to use the algorithm CloSpan, saving the results in a given
+ * file
+ * @author agomariz
+ */
+public class MainTestCloSpan_saveToFile {
+
+    /**
+     * @param args the command line arguments
+     */
+    public static void main(String[] args) throws IOException {
+        // Load a sequence database
+        double support = (double) 180 / 360;
+
+        boolean keepPatterns = true;
+        boolean verbose = false;
+        boolean findClosedPatterns = true;
+        boolean executePruningMethods = true;
+        
+        // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+        boolean outputSequenceIdentifiers = false;
+
+        AbstractionCreator abstractionCreator = AbstractionCreator_Qualitative.getInstance();
+
+        SequenceDatabase sequenceDatabase = new SequenceDatabase();
+
+        sequenceDatabase.loadFile("input/contextPrefixSpan.txt", support);
+        //sequenceDatabase.loadFile(fileToPath("contextCloSpan.txt"), support);
+        //sequenceDatabase.loadFile(fileToPath("gazelle.txt"), support);
+
+        //System.out.println(sequenceDatabase.toString());
+
+        AlgoCloSpan algorithm = new AlgoCloSpan(support, abstractionCreator, findClosedPatterns,executePruningMethods);
+
+        algorithm.runAlgorithm(sequenceDatabase, keepPatterns, verbose, "output/output_clospan.txt", outputSequenceIdentifiers);
+        System.out.println(algorithm.getNumberOfFrequentPatterns() + " pattern found.");
+
+        if (keepPatterns) {
+            System.out.println(algorithm.printStatistics());
+        }
+    }
+
[…]

### 2017-10-10 spm-dmc/src/MainTestMaxSP_saveToFile.java

@@ -1,0 +1,30 @@
+
+
+import java.io.IOException;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.BIDE_and_prefixspan.AlgoMaxSP;
+import ca.pfv.spmf.input.sequence_database_list_integers.SequenceDatabase;
+/*
+ * Example of how to use the BIDE+ algorithm, from the source code.
+ */
+public class MainTestMaxSP_saveToFile {
+
+	public static void main(String [] arg) throws IOException{    
+		// Load a sequence database
+		SequenceDatabase sequenceDatabase = new SequenceDatabase(); 
+		sequenceDatabase.loadFile("input/contextPrefixSpan.txt");
+//		sequenceDatabase.print();
+		
+		int minsup = 2; // we use a minsup of 2 sequences (50 % of the database size)
+		
+		AlgoMaxSP algo  = new AlgoMaxSP();  //
+		
+        // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+        algo.setShowSequenceIdentifiers(false);
+		
+		// execute the algorithm
+		algo.runAlgorithm(sequenceDatabase, "output/output_maxsp.txt", minsup);    
+		algo.printStatistics(sequenceDatabase.size());
+	}
+}

### 2017-10-10 spm-dmc/src/MainTestTKS.java

@@ -1,0 +1,52 @@
+import java.io.IOException;
+import java.util.PriorityQueue;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.spam.AlgoTKS;
+import ca.pfv.spmf.algorithms.sequentialpatterns.spam.PatternTKS;
+
+
+/**
+ * Example of how to use the TKS algorithm in source code.
+ * @author Philippe Fournier-Viger
+ */
+public class MainTestTKS {
+
+	public static void main(String [] arg) throws IOException{ 
+
+		// Load a sequence database
+		String input = "input/contextPrefixSpan.txt";
+		String output = "output/output_tks.txt";
+		
+		int k=5;
+		
+		// Create an instance of the algorithm 
+		AlgoTKS algo = new AlgoTKS(); 
+		
+		// This optional parameter allows to specify the minimum pattern length:
+//		algo.setMinimumPatternLength(0);  // optional
+
+		// This optional parameter allows to specify the maximum pattern length:
+//		algo.setMaximumPatternLength(4);  // optional
+		
+		// This optional parameter allows to specify constraints that some
+		// items MUST appear in the patterns found by TKS
+		// E.g.: This requires that items 1 and 3 appears in every patterns found
+//		algo.setMustAppearItems(new int[] {1,3});
+		
+		// This optional parameter allows to specify the max gap between two
+		// itemsets in a pattern. If set to 1, only patterns of contiguous itemsets
+		// will be found (no gap).
+		//algo.setMaxGap(1);
+		
+	    // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+//		algo.showSequenceIdentifiersInOutput(true);
+		
+		// execute the algorithm, which returns some patterns
+		PriorityQueue<PatternTKS> patterns = algo.runAlgorithm(input, output, k);  
+		// save results to file
+		algo.writeResultTofile(output);   
+		algo.printStatistics();
+
+	}
+}

### 2017-10-10 spm-dmc/src/MainTestTSP_nonClosed.java

@@ -1,0 +1,37 @@
+
+
+import java.io.IOException;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.BIDE_and_prefixspan.AlgoTSP_nonClosed;
+import ca.pfv.spmf.input.sequence_database_list_integers.SequenceDatabase;
+
+
+/**
+ * Example of how to use the PrefixSpanWithSupportRising algorithm in source code.
+ * @author Philippe Fournier-Viger
+ */
+public class MainTestTSP_nonClosed {
+
+	public static void main(String [] arg) throws IOException{    
+		// Load a sequence database
+		long startTime = System.currentTimeMillis();
+		SequenceDatabase sequenceDatabase = new SequenceDatabase(); 
+		sequenceDatabase.loadFile("input/contextPrefixSpan.txt");
+		System.out.println(System.currentTimeMillis() - startTime + " ms (database load time)");
+		// print the database to console
+//		sequenceDatabase.print();
+			
+		AlgoTSP_nonClosed algo = new AlgoTSP_nonClosed(); 
+		
+        // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+        algo.setShowSequenceIdentifiers(false);
+		
+		int k = 2; // we use a k of 2 sequences.
+		
+		// execute the algorithm
+		algo.runAlgorithm(sequenceDatabase, k);    
+		algo.writeResultTofile("output/output_tsp.txt");
+		algo.printStatistics(sequenceDatabase.size());
+	}
+}

### 2017-10-10 spm-dmc/src/MainTestVMSP_saveToFile.java

@@ -1,0 +1,38 @@
+
+
+import java.io.IOException;
+
+import ca.pfv.spmf.algorithms.sequentialpatterns.spam.AlgoVMSP;
+
+
+/**
+ * Example of how to use the VMSP algorithm in source code.
+ * @author Philippe Fournier-Viger
+ */
+public class MainTestVMSP_saveToFile {
+
+	public static void main(String [] arg) throws IOException{    
+		// Load a sequence database
+		String input = "input/contextPrefixSpan.txt";
+		String output = "output/output_vmsp.txt";
+		
+		// Create an instance of the algorithm 
+		AlgoVMSP algo = new AlgoVMSP(); 
+
+		// This optional parameter allows to specify the maximum pattern length:
+//		algo.setMaximumPatternLength(4);  // optional
+		
+		// This optional parameter allows to specify the max gap between two
+		// itemsets in a pattern. If set to 1, only patterns of contiguous itemsets
+		// will be found (no gap).
+		//algo.setMaxGap(1);
+		
+        // if you set the following parameter to true, the sequence ids of the sequences where
+        // each pattern appears will be shown in the result
+//		algo.showSequenceIdentifiersInOutput(true);
+		
+		// execute the algorithm with minsup = 2 sequences  (50 %)
+		algo.runAlgorithm(input, output, 0.5);    
+		algo.printStatistics();
+	}
+}

### 2017-10-20 spm-dmc/src/MainTestTKS.java
preprocessed data and produce frequent sequential patterns

@@ -14,16 +14,16 @@ public class MainTestTKS {
 	public static void main(String [] arg) throws IOException{ 
 
 		// Load a sequence database
-		String input = "input/contextPrefixSpan.txt";
+		String input = "input/formatted_sequence/successful.txt";
 		String output = "output/output_tks.txt";
 		
-		int k=5;
+		int k = 1000;
 		
 		// Create an instance of the algorithm 
 		AlgoTKS algo = new AlgoTKS(); 
 		
 		// This optional parameter allows to specify the minimum pattern length:
-//		algo.setMinimumPatternLength(0);  // optional
+		algo.setMinimumPatternLength(3);  // optional
 
 		// This optional parameter allows to specify the maximum pattern length:
 //		algo.setMaximumPatternLength(4);  // optional
@@ -36,7 +36,7 @@ public class MainTestTKS {
 		// This optional parameter allows to specify the max gap between two
 		// itemsets in a pattern. If set to 1, only patterns of contiguous itemsets
 		// will be found (no gap).
-		//algo.setMaxGap(1);
+//		algo.setMaxGap(1);
 		
 	    // if you set the following parameter to true, the sequence ids of the sequences where
         // each pattern appears will be shown in the result

### 2017-10-20 spm-dmc/src/MainTestTSP_nonClosed.java

@@ -14,9 +14,10 @@ public class MainTestTSP_nonClosed {
 
 	public static void main(String [] arg) throws IOException{    
 		// Load a sequence database
+		String input = "input/formatted_sequence/successful.txt";
 		long startTime = System.currentTimeMillis();
 		SequenceDatabase sequenceDatabase = new SequenceDatabase(); 
-		sequenceDatabase.loadFile("input/contextPrefixSpan.txt");
+		sequenceDatabase.loadFile(input);
 		System.out.println(System.currentTimeMillis() - startTime + " ms (database load time)");
 		// print the database to console
 //		sequenceDatabase.print();

### 2017-10-20 spm-dmc/src/MainTestVMSP_saveToFile.java

@@ -13,7 +13,7 @@ public class MainTestVMSP_saveToFile {
 
 	public static void main(String [] arg) throws IOException{    
 		// Load a sequence database
-		String input = "input/contextPrefixSpan.txt";
+		String input = "input/formatted_sequence/successful.txt";
 		String output = "output/output_vmsp.txt";
 		
 		// Create an instance of the algorithm 
@@ -32,7 +32,9 @@ public class MainTestVMSP_saveToFile {
 //		algo.showSequenceIdentifiersInOutput(true);
 		
 		// execute the algorithm with minsup = 2 sequences  (50 %)
-		algo.runAlgorithm(input, output, 0.5);    
+		algo.runAlgorithm(input, output, 0.02);    
 		algo.printStatistics();
+		
+		System.out.println("Executed successfully");
 	}
 }

### 2017-10-20 spm-dmc/src/preprocess/DriverProgram.java

@@ -1,0 +1,14 @@
+package preprocess;
+
+public class DriverProgram {
+
+	public static void main(String[] args) {
+		String formFile = "input/code_sequence/successful.txt";
+		String toFile = "input/formatted_sequence/successful.txt";
+		
+		SequenceData seqData = new SequenceData(formFile);
+		seqData.transformToExpectedFormat(formFile, toFile);
+
+	}
+
+}

### 2017-10-20 spm-dmc/src/preprocess/SequenceData.java

@@ -1,0 +1,77 @@
+package preprocess;
+
+import java.io.BufferedReader;
+import java.io.BufferedWriter;
+import java.io.FileReader;
+import java.io.FileWriter;
+import java.io.IOException;
+import java.util.HashMap;
+
+public class SequenceData {
+	
+	private HashMap<String, Integer> dict = new HashMap<>();
+	
+	public SequenceData() {}
+	
+	public SequenceData(String filePath) {
+		try {
+			BufferedReader br = new BufferedReader(new FileReader(filePath));
+			
+			String line = null;
+			int x = 1;
+			
+			while((line = br.readLine()) != null) {
+				
+				String[] codes = line.split(",");				
+				
+				for (String code : codes) {
+					if (code.equalsIgnoreCase("400") || code.equalsIgnoreCase("500")) {
+						dict.put(code, -2);
+					}
+					else {
+						if (!dict.containsKey(code)) {
+							dict.put(code, x++);
+						}
+					}
+				}
+			}
+			
+			br.close();
+			
+		} catch (IOException e) {
+			e.printStackTrace();
+		}
+	}
+	
+	public void transformToExpectedFormat(String fromFilePath, String toFilePath) {
+		try {
+			BufferedReader br = new BufferedReader(new FileReader(fromFilePath));
+			BufferedWriter writer = new BufferedWriter(new FileWriter(toFilePath));			
+			String line = null;
+			
+			while((line = br.readLine()) != null) {
+				
+				String[] codes = line.split(",");
+				StringBuilder sb = new StringBuilder();
+				
+				for (String code : codes) {
+					String newCode = "";
+					if (code.equalsIgnoreCase("400") || code.equalsIgnoreCase("500")) 
+						newCode = "" + dict.get(code);
+					else
+						newCode = dict.get(code) + " -1 ";
+						
+					sb.append(newCode);
+				}
+				
+				writer.write(sb.toString() + "\n");
+			}
+			
+			br.close();
+			writer.close();
+			
+		} catch (IOException e) {
+			e.printStackTrace();
+		}
+	}
+}

### 2018-01-07 paper-amia-summit-2018/amia_summit_2018-submitted.tex
fixed typo and expand PPC term

--- 2017-09-22 paper-amia-summit-2018/amia_summit_2018.tex
@@
 
-\title{Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences}
+\title{Predicting the Outcome of Patient-Provider Communication Sequences using Recurrent Neural Networks and Probabilistic Models}
 
@@
 \institutes{
-$^1$Department of Computer Science,  
+$^1$Department of Computer Science, Wayne State University, Detroit, Michigan \\  
 $^2$Department of Family Medicine and Public Health Sciences, School of Medicine, Wayne State University, Detroit, Michigan\\
@@
 \noindent{\bf Abstract}
-\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics.
-In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.
-Our method achieved 70.38\%, 60.67\%, 86.77\% F1-score for under-sampled sequences and 77.75\%, 75.20\%, 83.81\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively.
-The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}
+\textit{The problem of analyzing temporally ordered observation sequences generated by a physiological, genomic or psychological process to make predictions about the outcome of that process arises in
+many domains of clinical informatics. In this paper, we focus on the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviewing.
+Specifically, we formulate the problem of predicting the success of a motivational interview (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of
+coded patient-provider communication exchanges as binary sequence classification.
+%Severe imbalance between positive and negative training sequences is an additional challenge that arises in the context of motivational interviewing.
+We propose two methods that are based on deep learning, such as Recurrent Neural Networks (RNNs), and probabilistic models, such as Markov Chain (MC) and Hidden Markov Model (HMM), and compare the
+accuracy of these methods using communication sequences annotated with behavior codes from real-life motivational interviews. Our experiments indicate that the deep learning-based method is
+significantly more accurate than the methods based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM,
+respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate
+that the proposed method can be used for real-time monitoring of the progression of clinical interviews and identification of the most effective communication strategies, which significantly decreases
+the effort required to develop psychological interventions and increases their effectiveness.}
 
 \section*{Introduction}
-Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics.
-Sequence classification methods are utilized for classifying temporally ordered sequences.
-Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}.
-In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins.
-In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease.
-Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.
+Temporally ordered sequences of discrete or continuous observations generated by genomic, psychological or psychological processes arise in many different domains of clinical informatics (e.g.,
+DNA sequences, gene activations, ECG measurements, laboratory results, diagnostic codes, utterances during clinical interviews). Classification (or categorization) is a type of analysis of those
+sequences that has a broad range of important practical applications, from protein classification \cite{yakhnenko2005discriminatively} and function prediction \cite{deshpande2002evaluation} to
+detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as their temporal order, makes sequence classification a
+more challenging task than traditional classification. Predicting the outcome of those sequences (e.g. physiological or behavior response) can also be formulated as sequence classification.
 
-%Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection.
-In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}.
-In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}.
-Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.
+In general, sequence classification methods fall into one of the three major categories: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature
+vector and apply a standard supervised learning methods, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
+Feature-based methods for sequence classification have had limited success, since traditional bag-of-words feature representation methods cannot easily account for the order of observations in a
+sequence.
+%Shapelet \cite{ye2009time} and pattern based techniques \cite{kudenko1998feature, lesh1999mining} as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of
+%standard classifiers as well.
+Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is
+Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by time points.
+The third type of sequence classification methods first creates a probabilistic model, such as Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class
+using the training data and then classifies new sequences by applying the constructed models. MC and HMM explicitly capture short-term temporal dependencies (e.g. first- and second-order MC and HMM
+\cite{kundu1988recognition}) between adjacent observations in a sequence. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short-
+and long-term dependencies and were shown to be effective for modeling different types of sequential data. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which
+successfully addresses the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and
+handwriting \cite{nion2013handwritten} recognition to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to
+recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting
+the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
+Unit\cite{chung2014empirical} (GRU), was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.
 
-In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method.
-Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
-%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
-Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification.
-The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
-%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
-The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation.
-At the same time, distance-based methods are more appropriate for time series data.
-Model-based classification methods involve building a model based on the dataset.
-In other words, information is learned from the dataset and used as a ``model'' to classify sequences.
+In this paper, we address the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of
+Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that
+approximately one-third (31.8\%) of US children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and
+have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI),
+an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires,
+ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior
+change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of
+patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since
+manual examination and sequential pattern analysis of transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
+Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the
+mechanisms of MI's efficacy, in particular.
 
-%This makes sequence classification a more challenging task than traditional classification.
-
-One of the model-based classification methods is a probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
-Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences.
-Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step.
-Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}.
-On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory.
-It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
-A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN.
-LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}.
-In Lipton et.
-al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc.
-LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit.
-A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed.
-These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.
-
-Childhood obesity is a serious public health concern in the United States and worldwide.
-Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
-Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
-One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
-The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
-These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
-However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
-Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years.
-Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.
-
-%This work is the next step in a series of experiments to develop the [informatics based methods described above]. 
-In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}.
-In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews.
-Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
-We propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
-change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
-While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.
+Our goal is to test the applicability of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for predicting the success of clinical interviews (i.e. eliciting a
+particular type of patient behavioral response, such as CHT) and dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on a sequence of
+coded previous patient-provider communication exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored
+several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook
+\cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability
+of state-of-the-art methods for sequence modeling to analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavior in a context of motivational
+interviews, in particular.
 
@@
 \subsection*{\textit{Data collection}}
-The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
-Each transcript consists of an MI interview session involving counselor, adolescent, and caregiver.
-The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into client codes (adolescent and caregiver) group and counselor codes group.
-Utterances were divided into successful and unsuccessful communication sequences.
-Successful communication sequences result in positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
-Out of 5143 observed sequences, 4225 were positive and 918 were negative.
-Here, the successful sequences had an average length of 9.79 while that of unsuccessful sequences was 9.65.
-For each of the probabilistic models (Markov chain and HMM), two models were trained, one model was trained using successful sequences and another one was trained using unsuccessful sequences.
+The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists
+of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, in which the codes are grouped into
+patient (adolescent and caregiver) codes and counselor codes. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in
+positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the
+sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Successful sequences had an average length of
+9.79 utterances, while unsuccessful sequences had on average 9.65 utterances. For each of the probabilistic models (MC and HMM), two models were trained, one model was trained using successful
+sequences and another one was trained using unsuccessful sequences.
 
@@
 
-Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences.
-Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data.
-Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%.
-Therefore, it is important to handle imbalance data properly.
-Many solutions are proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level.
-In this study, we utilized undersampling and oversampling methods at data processing level for balancing the adolescent obesity dataset.
-Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}.
-We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}.
-On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.
+Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the
+performance of an employed algorithm for an imbalanced data, since simply predicting the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to
+handle class imbalance properly. Many solutions have been proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level. In this study, we
+utilized under-sampling and over-sampling for balancing the samples in different classes in our dataset. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used
+method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between
+minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the
+clusters of samples by the cluster centroid in a $k$-means clustering algorithm.
 
-A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
-Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end.
-It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively.
-Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors.
-For this reason, we only consider those models that can capture the past events for the classification of behavior sequences.\\
+A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Annotation column shows successful sequences of behavior codes from top to bottom. For example,
+($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. As
+can be seen, similar utterances, such as  ``Yeah" and ``Yes'', can have different behavior codes (CHT+ and HUPW, respectively) depending on the context.\\
 
@@
 \subsection*{\textit{Prediction method}}
-Generally, a sequence is a temporally ordered set of events.
-In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive), $OQECHT+$ (open-ended question, elicit change talk positive), etc.
-Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some part of a motivational interview, the task of predicting interview success can be viewed as sequence classification.
-Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are ``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.
+Generally, a sequence can be viewed as a temporally ordered set of events. In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive),
+$OQECHT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some
+part of a motivational interview, the task of predicting interview success can be viewed as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are
+``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.
 
-Our proposed baseline prediction method consists of two steps.
-In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary.
-In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model.
-Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, won2004training}.
-However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification.
-Therefore, we employed two special variants of RNN models in our experiments: long short-term memory and gated recurrent unit.
+Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first and second-order Markov Chain and
+Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classify each test sequence based on the maximum
+likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence
+modeling \cite{mutsam2016maximum, won2004training}. However, the latest advances in deep learning indicate that RNNs provide better results than conventional machine learning methods for the task of
+sequence classification. Therefore, we employed two special variants of RNN in our experiments: Long Short-Term Memory (LSTM) \cite{hochreiter1997long} and Gated Recurrent Unit (GRU)
+\cite{chung2014empirical}.
 
-\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property).
-For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
-A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
+\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property). For the sequential analysis,
+we built two Markov models $M$ and $\overline{M}$, summarizing counselor strategies and patient responses, in the cases of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A
+Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
 \begin{itemize}
@@
 \end{equation}
-In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
+In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point as:
 \begin{equation}
@@
 \end{equation}
-If $p(S\rightarrow successful) > 0 $, the interview is predicted to result in positive change talk or commitment language.
-Otherwise, it would be classified as a negative change talk or commitment language.
+If $p(S\rightarrow successful) > 0 $, a communication sequence is predicted to result in positive change talk or commitment language. Otherwise, it would be classified as a negative change talk or
+commitment language.
 
-The above model also referred as first-order MC, since it only considers immediately preceding behavior code when computing the state transition probabilities.
-In our experiment, we also considered second-order Markov model that compute transition probabilities based on previous two states.
+The above model also referred as first-order MC, since it only considers immediately preceding behavior code, when computing the state transition probabilities. In our experiment, we also considered
+second-order Markov model that computes transition probabilities based on previous two states.
 
@@
 
-\textbf {Embedding}: We took the inspiration for the representation of behavior codes from the idea of word embedding \cite{bengio2003neural}.
-Word embedding is a representation of words in low-dimensional space by vectors which contain the feature of the words.
-In our study, we employed embedding in place of one-hot vectors for the representation of behavior code since one-hot vectors are high-dimensional and sparse.
-Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by retaining their internal relations.
-The embedding was learned during the training of the classification model.
-Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional vector space visualize with t-SNE, a sophisticated technique for visualizing high-dimensional data.
-It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc.
-formed a cluster at the left of the figure.
-The nearest neighbors of CQECML+ are highlighted by their color intensity i.e.
-OQECML+ is more purple means closer to CQECML+.
-The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
-It is interesting that the behaviors intended to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.
+\textbf {Embeddings}: We took the inspiration for the representation of behavior codes from the idea of word embeddings\cite{bengio2003neural}. Word embedding is a representation of words in
+low-dimensional space by vectors, which contain the features of the words. In our study, we employed embedding in place of one-hot vectors for representation of behavior codes as input to LSTM and
+GRU, since one-hot vectors are high-dimensional and sparse. Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by
+retaining their internal relations. The embedding was learned during the training of the classification model. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional
+vector space visualized by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left
+part of the figure. The nearest neighbors of CQECML+ are highlighted by their color intensity (i.e. OQECML+ being more purple means closer to CQECML+). The right part of the figure demonstrates
+another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended
+to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.
 
@@
 
-\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step.
-That RNN can capture long-term dependencies for predicting the future events is its main advantage over feedforward Neural Network, MC, and HMM.
-The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed.
-However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}.
-In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM.
-Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}.
-There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU.
-GRUs are simpler than LSTM units and experimentally outperform than other models in many cases.
-We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
+\textbf {Recurrent Neural Networks (RNN)} are a type of Neural Networks, which can send feedback from the current hidden state to the hidden state of the next time step. The ability of RNN to capture
+long-term dependencies for predicting the future events is its main advantage over MC and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior
+observed at some point in the interview is very informative for the future behaviors, that will be observed later, but not immediately. However, it was observed that RNN fails to capture long-term
+dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short
+Term Memory networks or simply LSTM. There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are
+simpler than LSTM units and have been shown to experimentally outperform other models\cite{cho2014properties}. GRU is formally defined as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
 \begin{equation}
@@
 \end{figure}
-The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
-We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence.
-Then, we applied a softmax layer at each time-step for predicting the target which is the predicted label of the sequence.
-This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
-As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
-Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$.
-Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
-We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss.
-Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization.
-The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
+The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes and embedding of MYSCOPE codes in an observed
+sequence of communication behaviors as input. Then, we applied a softmax layer at each time-step for predicting the target which is the predicted label of the sequence. As can be seen from
+Eq.~\ref{eq:loss}, the total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where $T$ is the total number of elements in a sequence,
+$\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that
+best performance achieved when we use $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which
+were determined on the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges
+to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
 
@@
 Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 \subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used.
-Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}.
-First, recurrent neural networks outperform over probabilistic models and achieved 16.39\%-26.1\% higher F1-score.
-Second, LSTM with target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681.
-Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM.
-Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
-Fifth, it follows that second-order MC have lower precision, recall, and F-measure than first-order MC.
-In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first to second-order models.
+We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used. Five major conclusions can be drawn
+from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform probabilistic models and achieve 16.39\%-26.1\% higher F1-score. Second, LSTM with
+target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681. Third, target replication strategy improves the
+performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics
+for under-sampled sequences. Fifth, second-order MC has lower precision, recall, and F-measure than first-order MC. In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 
+9.36\%, when going from first to second-order MC model.
 
 \subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
-Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset.
-For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size 55.
-Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%, 2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain.
-Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with target replication has the highest F1-score among all models.
-However, the predictive performance of recurrent neural networks decreases in over-sampled sequences although the performance of probabilistic models increases.
-We also observed that all models have the largest value in precision compared to other performance metrics.
+Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset. For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size
+55. Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%,
+2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain. Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with
+target replication has the highest F1-score among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences, while the performance of
+probabilistic models increases. We also observed that all models have the largest value in precision compared to other performance metrics.
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 \subsection*{\textit{Most likely communication sequences}}
-Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews.
-In successful sequences, we observed the provision of information using patient-centered communication (GINFO+) and structure session which is the counselor either explaining the therapeutic agenda or attempting to transition to a new topic or session content (SS).
-Sometimes, counselor acknowledged the clients' communication or an off topic comment (SO).
-We also observed that the affirm and OQECML+ are consistent with MI theory as an effect for eliciting positive change talk or commitment language.
-For unsuccessful sequences, It can be seen that providing advice using non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-) which is heading in the wrong direction therapeutically.
-Questions phased to elicit negative change talk or commitment language leads to CHT- or CML-, which is consistent with the manual coding approach or analysis, or AMB-.
+Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews. In successful
+sequences, we observed that information is frequently provided using patient-centered communication (GINFO+) and structure session, when the counselor either explains the therapeutic agenda or
+attempts to transition to a new topic or session content (SS). Sometimes, counselor acknowledges the clients' communication or an off topic comment (SO). We also observed that the affirm and OQECML+
+are consistent with MI theory, which claims their strong effect for eliciting positive change talk or commitment language. For unsuccessful sequences, it can be seen that providing advice using
+non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-), which is heading in the wrong direction therapeutically. Questions phased to elicit negative
+change talk or commitment language leads to CHT- or CML-, which is  consistent with the manual coding approach or analysis, or AMB-.
 %It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
@@
 \section*{Discussion}
-By analyzing the experimental results of different communication sequence outcome prediction schemes proposed in this paper, we arrived at the following conclusions.
-First, the overall predictive performance of RNN models is substantially better than probabilistic models.
-In particular, the RNN-based method achieves near-human accuracy for predicting the label of motivational interviews.
-This indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
-This provides evidence that the RNN is able to successfully replicate human cognitive processes to integrate previous information when formulating higher level thinking.
-In addition to that, word-embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+By analyzing the experimental results of different communication sequence outcome prediction methods proposed in this paper, we arrived at the following conclusions. First, the overall predictive
+performance of RNN models is substantially better than probabilistic models. In particular, the RNN-based method achieves near-human accuracy for predicting the label of motivational interviews. This
+indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the
+interviews. This provides evidence that the RNN is able to successfully replicate human cognitive processes to integrate previous information when formulating higher level thinking. In addition to
+that, embeddings allow to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
 
-Second, using target replication to compute the losses at each time step results in better performance for all configurations of the proposed RNN method.
-This indicates that mean of the losses over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters.
-Better estimates of parameters in RNN models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence.
-This allows achieving an improvement in the prediction accuracy of the method.
+Second, using target replication to compute the loss at each time step results in better performance for all configurations of the proposed RNN-based methods. This indicates that mean of the losses
+over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters. Better estimates of parameters in RNN
+models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence. This
+allows achieving an improvement in the prediction accuracy of the method.
 
-Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models.
-Because the number of states in higher-order Markov models grows exponentially with their order.
-Therefore, accurate estimation of transition probabilities requires much larger training set.
-Using smaller datasets such as under-sampled dataset will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates.
-Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing.
-In this study, we found out that using first-order Markov models is a reasonable trade-off between efficiency and accuracy.
+Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models, which we attribute to the fact that the number of states in higher-order
+Markov models grows exponentially with their order. As a result, accurate estimation of transition probabilities requires much larger training set. Using smaller datasets such as under-sampled dataset
+will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability
+estimates. Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing. In this study, we found out that using first-order Markov models is a
+reasonable trade-off between efficiency and accuracy.
 
-Fourth, similar to traditional Markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used.
-This indicates that sufficient training data is required to estimate optimal hyperparameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
+Fourth, similar to traditional Markov model, HMM achieves a dramatic improvement in the prediction accuracy when larger training set is used. This indicates that sufficient training data is required
+to estimate optimal hyperparameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
 
@@
 \section*{Conclusion}
-In this paper, we proposed Recurrent Neural Networks with two baseline methods Markov Chain and Hidden Markov Model for predicting the success of motivational interviews.
-We found out that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
-Our proposed method can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource-intensive manual qualitative analysis of interview transcripts.
-Our proposed method can also help to identify most likely sequences that are common to successful and unsuccessful motivational interviews, which can also directly inform clinical practice.
-This work also has broad implications for qualitative public health research by providing a formal theoretically-grounded computational mechanism to facilitate the development of effective behavioral interventions.
+In this paper, we compared the accuracy of Recurrent Neural Networks with Markov Chain and Hidden Markov Model for the task of predicting the success of motivational interviews. We found out
+that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
+Our proposed methods can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the
+interviews without resource-intensive manual qualitative analysis of interview transcripts, which can significantly decrease the time and effort required to develop behavioral interventions. Our
+proposed methods can also help to identify most likely sequences that are common to successful and unsuccessful motivational interviews, which can directly inform clinical practice and increase
+the effectiveness of behavioral interventions. Our experimental results indicate that the proposed methods can also be used for real-time monitoring of the progression of clinical interviews. This
+work also has broad implications for public health research by providing a formal theoretically-grounded computational mechanism for qualitative data analysis.
 
@@
 We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset by manually annotating the dataset using the MYSCOPE codebook.
+
 %\pagebreak

### 2018-01-08 paper-amia-summit-2018/amia_summit_2018.tex
update paper with reviewrs comments

@@
 \noindent{\bf Abstract}
-\textit{The problem of analyzing temporally ordered observation sequences to make predictions related to the outcome of the processes that generated these sequences arises in many domains of healthcare informatics.
-In this paper, we focus on patient-provider communication sequences in the context of clinical interviews and propose Recurrent Neural Network (RNN) with two baseline methods Markov chain and Hidden Markov Model (HMM), for predicting the likelihood of eliciting a particular type of patient behavioral response based on an observed sequence of patient-provider exchanges.
-Our method achieved 70.38\%, 60.67\%, 86.77\% F1-score for under-sampled sequences and 77.75\%, 75.20\%, 83.81\% F1-score for over-sampled sequences in predicting the outcome of motivational interviews with obese adolescents using Markov Chain, HMM, and RNN, respectively.
-The proposed method can be used to automatically identify the most effective communication strategies in motivational interviews, which significantly decreases the effort required to develop effective interventions to address many public health conditions.}
+\textit{The problem of analyzing temporally ordered observation sequences generated by a physiological, genomic or psychological process to make predictions about the outcome of that process arises in
+many domains of clinical informatics. In this paper, we focus on the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviewing.
+Specifically, we formulate the problem of predicting the success of a motivational interview (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of
+coded patient-provider communication exchanges as binary sequence classification.
+%Severe imbalance between positive and negative training sequences is an additional challenge that arises in the context of motivational interviewing.
+We propose two methods that are based on deep learning, such as Recurrent Neural Networks (RNNs), and probabilistic models, such as Markov Chain (MC) and Hidden Markov Model (HMM), and compare the
+accuracy of these methods using communication sequences annotated with behavior codes from real-life motivational interviews. Our experiments indicate that the deep learning-based method is
+significantly more accurate than the methods based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM,
+respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate
+that the proposed method can be used for real-time monitoring of the progression of clinical interviews and identification of the most effective communication strategies, which significantly decreases
+the effort required to develop psychological interventions and increases their effectiveness.}
 
 \section*{Introduction}
-Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics.
-Sequence classification methods are utilized for classifying temporally ordered sequences.
-Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}.
-In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins.
-In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease.
-Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.
+Temporally ordered sequences of discrete or continuous observations generated by genomic, psychological or psychological processes arise in many different domains of clinical informatics (e.g.,
+DNA sequences, gene activations, ECG measurements, laboratory results, diagnostic codes, utterances during clinical interviews). Classification (or categorization) is a type of analysis of those
+sequences that has a broad range of important practical applications, from protein classification \cite{yakhnenko2005discriminatively} and function prediction \cite{deshpande2002evaluation} to
+detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as their temporal order, makes sequence classification a
+more challenging task than traditional classification. Predicting the outcome of those sequences (e.g. physiological or behavior response) can also be formulated as sequence classification.
 
-%Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection.
-In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}.
-In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}.
-Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.
+In general, sequence classification methods fall into one of the three major categories: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature
+vector and apply a standard supervised learning methods, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
+Feature-based methods for sequence classification have had limited success since traditional bag-of-words feature representation methods cannot easily account for the order of observations in a
+sequence.
+%Shapelet \cite{ye2009time} and pattern based techniques \cite{kudenko1998feature, lesh1999mining} as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of
+%standard classifiers as well.
+Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is
+Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by time points.
+The third type of sequence classification methods first creates a probabilistic model, such as Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class
+using the training data and then classifies new sequences by applying the constructed models. MC and HMM explicitly capture short-term temporal dependencies (e.g. first- and second-order MC and HMM
+\cite{kundu1988recognition}) between adjacent observations in a sequence. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short-
+and long-term dependencies and were shown to be effective for modeling different types of sequential data. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which
+successfully addresses the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and
+handwriting \cite{nion2013handwritten} recognition to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to
+recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting
+the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
+Unit\cite{chung2014empirical} (GRU), was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.
 
-In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method.
-Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}.
-%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well.
-Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification.
-The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.
-%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
-The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation.
-At the same time, distance-based methods are more appropriate for time series data.
-Model-based classification methods involve building a model based on the dataset.
-In other words, information is learned from the dataset and used as a ``model'' to classify sequences.
+In this paper, we address the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of
+Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that
+approximately one-third (31.8\%) of US children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and
+have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI),
+an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires,
+ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior
+change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of
+patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since
+manual examination and sequential pattern analysis of transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
+Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the
+mechanisms of MI's efficacy, in particular.
 
-%This makes sequence classification a more challenging task than traditional classification.
-
-One of the model-based classification methods is a probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM).
-Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences.
-Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step.
-Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}.
-On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory.
-It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}.
-A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN.
-LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}.
-In Lipton et.
-al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc.
-LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit.
-A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed.
-These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.
-
-Childhood obesity is a serious public health concern in the United States and worldwide.
-Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
-Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
-One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
-The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
-These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
-However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
-Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years.
-Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.
-
-%This work is the next step in a series of experiments to develop the [informatics based methods described above]. 
-In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}.
-In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews.
-Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
-We propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e.
-change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview.
-While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.
+Our goal is to test the applicability of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for predicting the success of clinical interviews (i.e. eliciting a
+particular type of patient behavioral response, such as CHT) and dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on a sequence of
+coded previous patient-provider communication exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored
+several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook
+\cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability
+of state-of-the-art methods for sequence modeling to analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavior in a context of motivational
+interviews, in particular.
 
@@
 \subsection*{\textit{Data collection}}
-The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances.
-Each transcript consists of an MI interview session involving counselor, adolescent, and caregiver.
-The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into client codes (adolescent and caregiver) group and counselor codes group.
-Utterances were divided into successful and unsuccessful communication sequences.
-Successful communication sequences result in positive change talk and commitment language statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur.
-Out of 5143 observed sequences, 4225 were positive and 918 were negative.
-Here, the successful sequences had an average length of 9.79 while that of unsuccessful sequences was 9.65.
-For each of the probabilistic models (Markov chain and HMM), two models were trained, one model was trained using successful sequences and another one was trained using unsuccessful sequences.
+The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists
+of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, in which the codes are grouped into
+patient (adolescent and caregiver) codes and counselor codes. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in
+positive change talk (CHT+) and commitment language (CML+) statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk (CHT-) or commitment language (CML-) and the
+sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Successful sequences had an average length of
+9.79 utterances, while unsuccessful sequences had on average 9.65 utterances. For each of the probabilistic models (MC and HMM), two models were trained, one model was trained using successful
+sequences and another one was trained using unsuccessful sequences.
 
@@
 \end{figure}
-The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}.
-We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence.
-Then, we applied a softmax layer at each time-step for predicting the target which is the predicted label of the sequence.
-This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks.
-As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs.
-Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$.
-Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set.
-We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss.
-Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization.
-The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
+The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes and embedding of MYSCOPE codes in an observed
+sequence of communication behaviors as input.
+Then, we applied a softmax layer at each time-step for predicting the label of the sequence.
+Since the sequence label was predicted at each time step, the proposed model is also known as Recurrent Neural Network with Target Replication (RNN-TR).
+As can be seen from
+Eq.~\ref{eq:loss}, the total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where $T$ is the total number of elements in a sequence,
+$\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that
+best performance achieved when we use $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which
+were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges
+to optimal after 100 epochs and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
 
@@
 \end{table}  
-\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
-We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used.
-Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}.
-First, recurrent neural networks outperform over probabilistic models and achieved 16.39\%-26.1\% higher F1-score.
-Second, LSTM with target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681.
-Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM.
-Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences.
-Fifth, it follows that second-order MC have lower precision, recall, and F-measure than first-order MC.
-In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first to second-order models.
+\subsection*{\textit{Predictive performance for under-sampled Patient-Provider Communication code sequences}}
+We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used. Five major conclusions can be drawn
+from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform probabilistic models and achieve 16.39\%-26.1\% higher F1-score. Second, LSTM with
+target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681. Third, target replication strategy improves the
+performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics
+for under-sampled sequences. Fifth, second-order MC has lower precision, recall, and F-measure than first-order MC. In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 
+9.36\%, when going from first to second-order MC model.
 
-\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
-Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset.
-For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size 55.
-Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%, 2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain.
-Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with target replication has the highest F1-score among all models.
-However, the predictive performance of recurrent neural networks decreases in over-sampled sequences although the performance of probabilistic models increases.
-We also observed that all models have the largest value in precision compared to other performance metrics.
+\subsection*{\textit{Predictive performance for over-sampled Patient-Provider Communication code sequences}}
+Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset. For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size
+55. Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%,
+2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain. Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with
+target replication has the highest F1-score among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences, while the performance of
+probabilistic models increases. We also observed that all models have the largest value in precision compared to other performance metrics.
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
 \section*{Discussion}
-By analyzing the experimental results of different communication sequence outcome prediction schemes proposed in this paper, we arrived at the following conclusions.
-First, the overall predictive performance of RNN models is substantially better than probabilistic models.
-In particular, the RNN-based method achieves near-human accuracy for predicting the label of motivational interviews.
-This indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
-This provides evidence that the RNN is able to successfully replicate human cognitive processes to integrate previous information when formulating higher level thinking.
-In addition to that, word-embedding allows to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+By analyzing the experimental results of different communication sequence outcome prediction methods proposed in this paper, we arrived at the following conclusions. First, the overall predictive
+performance of RNN models is substantially better than probabilistic models. In particular, the RNN-based method achieves near-human accuracy for predicting the label of motivational interviews. This
+indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the
+interviews. This provides evidence that the RNN is able to successfully replicate human cognitive processes to integrate previous information when formulating higher level thinking. In addition to
+that, embeddings allow to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
 
-Second, using target replication to compute the losses at each time step results in better performance for all configurations of the proposed RNN method.
-This indicates that mean of the losses over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters.
-Better estimates of parameters in RNN models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence.
-This allows achieving an improvement in the prediction accuracy of the method.
+Second, using target replication to compute the loss at each time step results in better performance for all configurations of the proposed RNN-based methods. This indicates that mean of the losses
+over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters. Better estimates of parameters in RNN
+models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence. This
+allows achieving an improvement in the prediction accuracy of the method.
 
-Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models.
-Because the number of states in higher-order Markov models grows exponentially with their order.
-Therefore, accurate estimation of transition probabilities requires much larger training set.
-Using smaller datasets such as under-sampled dataset will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates.
-Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing.
-In this study, we found out that using first-order Markov models is a reasonable trade-off between efficiency and accuracy.
+Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models, which we attribute to the fact that the number of states in higher-order
+Markov models grows exponentially with their order. As a result, accurate estimation of transition probabilities requires much larger training set. Using smaller datasets such as under-sampled dataset
+will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability
+estimates. Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing. In this study, we found out that using first-order Markov models is a
+reasonable trade-off between efficiency and accuracy.
 
-Fourth, similar to traditional Markov model, HMM achieve a dramatic improvement in the prediction accuracy when larger training set is used.
-This indicates that sufficient training data is required to estimate optimal hyperparameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
+Fourth, similar to traditional Markov model, HMM achieves a dramatic improvement in the prediction accuracy when larger training set is used. This indicates that sufficient training data is required
+to estimate optimal hyperparameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
 
-Fifth, the proposed method can be used to identify the most effective communication strategies at eliciting a particular type of behavioral response.
+Fifth, the proposed method can be used to identify the most effective communication strategies for eliciting a particular type of behavioral response.
 Awareness of these strategies by researchers can significantly decrease the time and effort required to develop effective interventions to address many public health conditions, such as childhood obesity, and tailor these interventions to particular patient cohorts.

### 2018-01-10 paper-amia-summit-2018/amia_summit_2018.tex
camera ready version

@@
 \noindent{\bf Abstract}
-\textit{The problem of analyzing temporally ordered observation sequences generated by a physiological, genomic or psychological process to make predictions about the outcome of that process arises in
-many domains of clinical informatics. In this paper, we focus on the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviewing.
-Specifically, we formulate the problem of predicting the success of a motivational interview (i.e. eliciting a particular type of patient behavioral response) based on an observed sequence of
-coded patient-provider communication exchanges as binary sequence classification.
-%Severe imbalance between positive and negative training sequences is an additional challenge that arises in the context of motivational interviewing.
-We propose two methods that are based on deep learning, such as Recurrent Neural Networks (RNNs), and probabilistic models, such as Markov Chain (MC) and Hidden Markov Model (HMM), and compare the
-accuracy of these methods using communication sequences annotated with behavior codes from real-life motivational interviews. Our experiments indicate that the deep learning-based method is
-significantly more accurate than the methods based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM,
-respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling). These results indicate
-that the proposed method can be used for real-time monitoring of the progression of clinical interviews and identification of the most effective communication strategies, which significantly decreases
-the effort required to develop psychological interventions and increases their effectiveness.}
+\textit{The problem of analyzing temporally ordered sequences of observations generated by a molecular, physiological or psychological process to make predictions about the outcome of that process arises in many domains of clinical informatics.
+In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog.
+Specifically, we consider prediction of the motivational interview success (i.e.
+eliciting a particular type of patient behavioral response) based on an observed sequence of coded patient-provider communication exchanges as a sequence classification problem.
+We propose two solutions to this problem, one that is based on Recurrent Neural Networks (RNNs) and another that is based on Markov Chain (MC) and Hidden Markov Model (HMM), and compare the accuracy of these solutions using communication sequences annotated with behavior codes from the real-life motivational interviews.
+Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling).
+These results indicate that the proposed method can be used for real-time monitoring of clinical interviews progression and more efficient identification of effective provider communication strategies, which can significantly decrease the effort required to develop behavioral interventions and increases their effectiveness.}
 
 \section*{Introduction}
-Temporally ordered sequences of discrete or continuous observations generated by genomic, psychological or psychological processes arise in many different domains of clinical informatics (e.g.,
-DNA sequences, gene activations, ECG measurements, laboratory results, diagnostic codes, utterances during clinical interviews). Classification (or categorization) is a type of analysis of those
-sequences that has a broad range of important practical applications, from protein classification \cite{yakhnenko2005discriminatively} and function prediction \cite{deshpande2002evaluation} to
-detecting individuals with a heart disease \cite{wei2006semi}. Taking into account both the entire set of observations in a sequence, as well as their temporal order, makes sequence classification a
-more challenging task than traditional classification. Predicting the outcome of those sequences (e.g. physiological or behavior response) can also be formulated as sequence classification.
+Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances  in clinical dialog).
+Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}.
+Taking into account both the entire set of observations in a sequence, as well as temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations.
+Predicting the outcome of those sequences (e.g.
+physiological or behavioral response) can also be viewed as a sequence classification problem.
 
-In general, sequence classification methods fall into one of the three major categories: feature-based, distance-based and model-based. Feature-based methods transform a sequence into a feature
-vector and apply a standard supervised learning methods, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
-Feature-based methods for sequence classification have had limited success since traditional bag-of-words feature representation methods cannot easily account for the order of observations in a
+In general, sequence classification methods fall into one of the three major classes: feature-based, distance-based and model-based.
+Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
+The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
 sequence.
-%Shapelet \cite{ye2009time} and pattern based techniques \cite{kudenko1998feature, lesh1999mining} as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of
-%standard classifiers as well.
-Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric. The most commonly used distance metric is
-Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}. However, distance metrics are primarily designed for time series data, in which the observations are discretized by time points.
-The third type of sequence classification methods first creates a probabilistic model, such as Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class
-using the training data and then classifies new sequences by applying the constructed models. MC and HMM explicitly capture short-term temporal dependencies (e.g. first- and second-order MC and HMM
-\cite{kundu1988recognition}) between adjacent observations in a sequence. By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short-
-and long-term dependencies and were shown to be effective for modeling different types of sequential data. Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which
-successfully addresses the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and
-handwriting \cite{nion2013handwritten} recognition to health informatics \cite{lipton2015learning, choi2016doctor}. Specifically, LSTM was used as part of a multi-label classification method to
-recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}. LSTM was also effectively used for predicting
-the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}. A further simplification and improvement of LSTM model, called the Gated Recurrent
-Unit\cite{chung2014empirical} (GRU), was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.
 
-In this paper, we address the problem of predicting the outcome of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of
-Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that
-approximately one-third (31.8\%) of US children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents, who are obese, are likely to be obese in adulthood and
-have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing (MI),
-an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires,
-ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior
-change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of
-patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since
-manual examination and sequential pattern analysis of transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
-Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the
-mechanisms of MI's efficacy, in particular.
+Distance-based methods classify a sequence by finding the most similar sequences with known classes based on a distance metric.
+The most commonly used distance metric is Euclidean distance with Dynamic Time Wrapping \cite{keogh2000scaling}.
+However, distance metrics are primarily designed for time series data, in which the observations are discretized by timestamps.
+The third type of sequence classification methods first creates a probabilistic model, such as the Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class based on the training data and then classifies new sequences by applying the created models.
+While MCs and HMMs can capture first- and second-order dependencies between adjacent observations in a sequence, learning higher-order dependencies with these models requires prohibitively large amounts of data.
+By encoding sequences into low-dimensional representations, Recurrent Neural Networks (RNNs) are able to capture both short- and long-term dependencies and were shown to be effective at modeling different types of sequential data \cite{lipton2015critical}.
+Long Short-Term Memory (LSTM) \cite{hochreiter1997long} is a variant of RNNs, which successfully addressed the vanishing gradient problem \cite{bengio1993problem} of traditional RNN.
+LSTM demonstrated excellent performance in different domains, from speech \cite{graves2013speech} and handwriting recognition\cite{nion2013handwritten} to health informatics \cite{lipton2015learning, choi2016doctor}.
+Specifically, LSTM was used as part of a multi-label classification method to recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}.
+LSTM was also effectively used for predicting the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}.
+A further simplification and improvement of LSTM model, called the Gated Recurrent
+Unit (GRU)\cite{chung2014empirical}, was later proposed. LSTM and GRU demonstrated markedly better performance among all other RNN variants for a variety of tasks in different domains.
 
-Our goal is to test the applicability of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for predicting the success of clinical interviews (i.e. eliciting a
-particular type of patient behavioral response, such as CHT) and dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on a sequence of
-coded previous patient-provider communication exchanges in the same interview. This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored
-several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook
-\cite{carcone2013provider}. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability
-of state-of-the-art methods for sequence modeling to analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavior in a context of motivational
-interviews, in particular.
+In this paper, we address the problem of predicting the outcome of coded patient-provider communication sequences in the context of clinical dialog.
+Specifically, we focus on predicting the success (i.e.
+eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded patient-provider communication exchanges during those interviews.
+Childhood obesity is a serious public health concern in the United States.
+Recent estimates indicate that approximately one-third (31.8\%) of U.S.
+children 2-19 years of age are overweight and 16.9\% are obese \cite{ogden2012prevalence}.
+Adolescents, who are obese, are likely to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}.
+One approach to effective obesity intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change.
+The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
+These statements, collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
+However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
+Since manual examination and analysis of MI interview transcripts is a very time-consuming process, designing effective MI interventions and tailoring them to particular populations can take years.
+Therefore, there is a need for informatics-based methods to facilitate the development of effective behavioral interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.
+
+Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e.
+eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous patient-provider communication exchanges in the same interview, which we consider as a sequence classification problem.
+This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}.
+While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.
 
@@
 \subsection*{\textit{Data collection}}
-The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists
-of an MI interview session involving counselor, adolescent, and caregiver. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, in which the codes are grouped into
-patient (adolescent and caregiver) codes and counselor codes. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in
-positive change talk (CHT+) and commitment language (CML+) statements by an adolescent or caregiver, while unsuccessful sequences are the ones that result in negative change talk (CHT-) or commitment language (CML-) and the
-sequences, in which no change talk or commitment language statements occur. Out of 5143 observed sequences, 4225 were positive and 918 were negative. Successful sequences had an average length of
-9.79 utterances, while unsuccessful sequences had on average 9.65 utterances. For each of the probabilistic models (MC and HMM), two models were trained, one model was trained using successful
-sequences and another one was trained using unsuccessful sequences.
 
-%Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\
-%
-%\begin{table}[h]
-%\centering
-%\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
-%\label{tab:data_dist}
-%  \begin{tabular}{|l|l|l|l|l|}
-%  \hline
-%   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
-%Successful sequences & 4225 & 82.15\% & 9.79 \\\hline
-%Unsuccessful sequences & 918 & 17.85\% & 9.65 \\\hline
-%  \end{tabular}
-%\end{table}
+The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which consist of a total of 50,239 segmented and annotated utterances.
+Each transcript consists of an MI interview session, which typically involves a counselor, an adolescent and a caregiver.
+The utterances were annotated based on the MYSCOPE codebook \cite{carcone2013provider}, in which the behavior codes are grouped into the patient (adolescent and caregiver) codes and the counselor codes.
+Annotated utterances were divided into successful and unsuccessful communication sequences.
+Successful communication sequences are the ones, which resulted in positive change talk (CHT+) or commitment language (CML+) statements by an adolescent or a caregiver, while unsuccessful sequences are the ones, which resulted in negative change talk (CHT-) or commitment language (CML-), or the ones, in which no change talk or commitment language statements were made.
 
-Our dataset is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the
-performance of an employed algorithm for an imbalanced data, since simply predicting the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to
-handle class imbalance properly. Many solutions have been proposed to deal with the imbalanced datasets, which can be divided into two categories: data level and algorithmic level. In this study, we
-utilized under-sampling and over-sampling for balancing the samples in different classes in our dataset. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used
-method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between
-minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the
-clusters of samples by the cluster centroid in a $k$-means clustering algorithm.
-
-A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Annotation column shows successful sequences of behavior codes from top to bottom. For example,
-($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. As
-can be seen, similar utterances, such as  ``Yeah" and ``Yes'', can have different behavior codes (CHT+ and HUPW, respectively) depending on the context.\\
+A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}.
+In this example, $SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$ is a successful sequence, in which a counselor starts with an open-ended question and ultimately is able to elicit a positive change talk statement.
+As follows from this example, similar utterances, such as ``Yeah'' and ``Yes'', can be assigned different behavior codes (CHT+ and HUPW, respectively), depending on the context.\\
 
 \begin{table}[h]
-\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
+\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent. MYSCOPE codes assigned to the utterances and their meaning are shown in the first two columns.}    
 \label{tab:anno_examp}
@@
 \hline
-\bf{Annotation}  & \bf{Behavior} & \bf{Speaker} & \bf{Text} \\\hline
+\bf{Code}  & \bf{Behavior} & \bf{Speaker} & \bf{Utterance} \\\hline
 SS & Structure Session & Counselor & Okay. Can I meet with Xxxx alone for a few minutes? \\\hline
@@
 \end{table}
+
+The resulting experimental dataset was highly imbalanced.
+Out of 5143 observed sequences, 4225 or 82.15\% were positive and only 918 or 17.85\% were negative.
+No major differences were observed in the average length of successful (9.79 utterances) and unsuccessful (9.65 utterances) sequences.
+
+Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g.
+simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset).
+Therefore, it is important to
+properly address class imbalance.
+We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes.
+Synthetic Minority Over Sampling Technique (SMOTE) \cite{chawla2002smote} is a widely used oversampling method for imbalanced datasets, in which new synthetic examples are generated for minority classes.
+Specifically, we generated synthetic examples at the borderline between the majority and minority classes \cite{nguyen2011borderline}.
+On the other hand, the under-sampling method reduces the number of samples in majority class by replacing the clusters of samples identified by the $k$-means clustering algorithm with the cluster centroids.
+
+Two MC and HMM models were trained, one model was trained on successful sequences, while another was trained on unsuccessful sequences.
 

### 2018-01-10 paper-amia-summit-2018/amia_summit_2018.tex
camera ready version

@@
 \noindent{\bf Abstract}
-\textit{The problem of analyzing temporally ordered sequences of observations generated by a molecular, physiological or psychological process to make predictions about the outcome of that process arises in many domains of clinical informatics.
+\textit{The problem of analyzing temporally ordered sequences of observations generated by molecular, physiological or psychological processes to make predictions about the outcome of these processes arises in many domains of clinical informatics.
 In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog.
@@
 Our experiments indicate that the deep learning-based approach is significantly more accurate than the approach based on probabilistic models in predicting the success of motivational interviews (0.8677 versus 0.7038 and 0.6067 F1-score by RNN, MC and HMM, respectively, when using under-sampling to correct for class imbalance, and 0.8381 versus 0.7775 and 0.7520 F1-score by RNN, MC and HMM, respectively, when using over-sampling).
-These results indicate that the proposed method can be used for real-time monitoring of clinical interviews progression and more efficient identification of effective provider communication strategies, which can significantly decrease the effort required to develop behavioral interventions and increases their effectiveness.}
+These results indicate that the proposed method can be used for real-time monitoring of progression of clinical interviews and more efficient identification of effective provider communication strategies, which in turn can significantly decrease the effort required to develop behavioral interventions and increase their effectiveness.}
 
@@
 Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}.
-Taking into account both the entire set of observations in a sequence, as well as temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations.
+Taking into account both the entire set of observations in a sequence, as well as the temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations.
 Predicting the outcome of those sequences (e.g.
@@
 
-In general, sequence classification methods fall into one of the three major classes: feature-based, distance-based and model-based.
-Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at a classification decision.
+In general, sequence classification methods fall into one of three major classes: feature-based, distance-based and model-based.
+Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at classification decision.
 The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
@@
 
-In this paper, we address the problem of predicting the outcome of coded patient-provider communication sequences in the context of clinical dialog.
+In this paper, we address the problem of predicting the outcome of coded patient-provider communication (PPC) sequences in the context of clinical dialog.
 Specifically, we focus on predicting the success (i.e.
-eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded patient-provider communication exchanges during those interviews.
+eliciting a particular type of patient behavioral response) of motivational interviews with obese adolescents and their caregivers based on an observed sequence of coded PPC exchanges during those interviews.
 Childhood obesity is a serious public health concern in the United States.
@@
 Our goal is to compare the accuracy of probabilistic models, such as MC and HMM, and deep learning methods, such as LSTM and GRU, for the task of predicting the success of clinical interviews (i.e.
-eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous patient-provider communication exchanges in the same interview, which we consider as a sequence classification problem.
+eliciting a particular type of patient behavioral response, such as CHT) at any point during a clinical interview based on a sequence of coded previous PPC exchanges in the same interview.
 This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}.
-While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of patient-provider communication (PPC) exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.
+While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of PPC exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.
 
@@
 The experimental dataset for this work was constructed from the transcripts of 129 motivational interviews, which consist of a total of 50,239 segmented and annotated utterances.
-Each transcript consists of an MI interview session, which typically involves a counselor, an adolescent and a caregiver.
+Each transcript corresponds to an MI interview session, which typically involves a counselor, an adolescent and a caregiver.
 The utterances were annotated based on the MYSCOPE codebook \cite{carcone2013provider}, in which the behavior codes are grouped into the patient (adolescent and caregiver) codes and the counselor codes.
@@
 In this example, $SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$ is a successful sequence, in which a counselor starts with an open-ended question and ultimately is able to elicit a positive change talk statement.
-As follows from this example, similar utterances, such as ``Yeah'' and ``Yes'', can be assigned different behavior codes (CHT+ and HUPW, respectively), depending on the context.\\
+As follows from this example, similar utterances, such as ``Yeah'' and ``Yes'', can be assigned different behavior codes (CHT+ and HUPW), depending on the context.\\
 
@@
 Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g.
-simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset).
-Therefore, it is important to
-properly address class imbalance.
+simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset), it is important to properly address class imbalance.
 We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes.
@@
 
-Two MC and HMM models were trained, one model was trained on successful sequences, while another was trained on unsuccessful sequences.
+%Two MC and HMM models were trained, one model was trained on successful sequences, while another was trained on %unsuccessful sequences.
 
-\subsection*{\textit{Prediction method}}
-Generally, a sequence can be viewed as a temporally ordered set of events. In this study, an event is a behavior code that also has a symbolic representation, such as $LUP+$ (low uptake, positive),
-$OQECHT+$ (open-ended question, elicit change talk positive), etc.  Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing patient-provider communication exchanges during some
-part of a motivational interview, the task of predicting interview success can be viewed as sequence classification. Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are
+\subsection*{\textit{Sequence classification methods}}
+In general, a sequence can be viewed as a temporally ordered set of observations.
+In this study, an observation corresponds to a behavior code, which has a symbolic representation, such as $LUP+$ (low uptake, positive),
+$OQECHT+$ (open-ended question, elicit change talk positive), etc.
+Given a sequence of behavior codes $S_i = \{c_1, c_2,...,c_n\}$ representing PPC exchanges during some part of a motivational interview, the task of predicting interview success can be considered as sequence classification.
+Given a set of class labels $L = \{l_1, l_2,...,l_m\}$ (in our case, the labels are
 ``successful'' and ``unsuccessful'' motivational interview), a sequence classifier $C$ learns a function $S_i \to l_i, l_i \in L$ that maps a sequence $S_i$ into a class label $l_i \in L$.
@@
 likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence
-modeling \cite{mutsam2016maximum, won2004training}. However, the latest advances in deep learning indicate that RNNs provide better results than conventional machine learning methods for the task of
-sequence classification. Therefore, we employed two special variants of RNN in our experiments: Long Short-Term Memory (LSTM) \cite{hochreiter1997long} and Gated Recurrent Unit (GRU)
-\cite{chung2014empirical}.
+modeling \cite{mutsam2016maximum, won2004training}.
+However, the latest advances in deep learning suggest that RNNs may provide better results than conventional machine learning methods for the task of
+sequence classification.
+To verify this hypothesis, we employed two state-of-the-art variants of RNN in our experiments: Long Short-Term Memory (LSTM) \cite{hochreiter1997long} and Gated Recurrent Unit (GRU) \cite{chung2014empirical}.
 
-\textbf {Markov Chain (MC)} is a probabilistic model that predicts the probability of next state based on its current state but not on its past states (Markov property). For the sequential analysis,
-we built two Markov models $M$ and $\overline{M}$, summarizing counselor strategies and patient responses, in the cases of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A
+\textbf {Markov Chain (MC)} is a probabilistic model that conditions each observation in a sequence only on preceding observation and not on any other past observation.
+First, we estimated two Markov models $M$ and $\overline{M}$, summarizing counselor strategies and patient responses, in the cases of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews.
+A
 Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
@@
 \item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
-\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
+\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on the maximum likelihood estimation:
 \begin{equation}
@@
 \end{itemize}
-where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data.
+where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the number of times the code $c_i$ have been observed in the training data, respectively.
 Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
@@
 \end{equation}
-If $p(S\rightarrow successful) > 0 $, a communication sequence is predicted to result in positive change talk or commitment language. Otherwise, it would be classified as a negative change talk or
-commitment language.
+If $p(S\rightarrow successful) > 0 $, a communication sequence is predicted to be successful (i.e.
+result in positive change talk or commitment language).
+Otherwise, it is predicted to be unsuccessful.
 
-The above model also referred as first-order MC, since it only considers immediately preceding behavior code, when computing the state transition probabilities. In our experiment, we also considered
-second-order Markov model that computes transition probabilities based on previous two states.
+The above model is also referred as first-order MC, since it only considers immediately preceding behavior code, when computing the state transition probabilities.
+In our experiment, we also considered second-order Markov model, which conditions each observation on the preceding two observations.
 
 \textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time.
-HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence.
-Mathematically, HMM can be defined as the following: $\lambda = (A, B, \pi)$.
+HMMs are widely used for sequence analysis because of their ability to identify hidden states, corresponding to clusters of observations.
+Mathematically, HMM can be defined as $\lambda = (A, B, \pi)$, where:
 \begin{itemize}
@@
 \end{itemize}
-Hence, $N$ is a number of hidden states in the model and $M$ is a number of distinct observations symbols per state, i.e.
+Hence, $N$ is a number of hidden states in the model and $M$ is a number of distinct observations per hidden state, i.e.
 the discrete vocabulary size.
-The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols (emission probabilities) for each state.
-The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations.
-After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.
+The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter.
+HMM deduces a sequence of hidden states that best explains the observations along with the state transition probabilities and the distributions of observations (emission probabilities) per each hidden state.
+The Baum-Welch algorithm \cite{rabiner1989tutorial} is used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm \cite{rabiner1989tutorial} is used to determine the most likely sequence of hidden states for a given sequence of observations.
+After assignment of hidden states, the log-likelihood of success for an interview can be estimated using Eq.~\ref{eq:class} as well.
 
-\textbf {Embeddings}: We took the inspiration for the representation of behavior codes from the idea of word embeddings\cite{bengio2003neural}. Word embedding is a representation of words in
-low-dimensional space by vectors, which contain the features of the words. In our study, we employed embedding in place of one-hot vectors for representation of behavior codes as input to LSTM and
-GRU, since one-hot vectors are high-dimensional and sparse. Moreover, code embeddings have the ability to represent semantically similar codes with similar vectors in a low-dimensional space by
-retaining their internal relations. The embedding was learned during the training of the classification model. Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embedding in 44-dimensional
-vector space visualized by t-SNE \cite{maaten2008visualizing}. It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc. formed a cluster in the left
-part of the figure. The nearest neighbors of CQECML+ are highlighted by their color intensity (i.e. OQECML+ being more purple means closer to CQECML+). The right part of the figure demonstrates
-another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc. It is interesting that the behaviors intended
-to elicit CHT+/CML+ group together whereas the ones intended to elicit CHT-/CML- also group together and are located on opposite ends of the space.
+\textbf {Behavior code embeddings}.
+Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural}.
+Embedding is a representation of an object in low-dimensional space using a real-valued vector.
+In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as representation of behavior codes as input to these RNNs.
+Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space.
+Figure~\ref{fig:code_embedding} illustrates the MYSCOPE code embeddings visualized in 2-dimensional space by t-SNE \cite{maaten2008visualizing}.
+It can be seen that positive behavior codes such as OQECHT+, OQECML+, AF, AFL, SUP, RCML+S, CQECML+, etc.
+formed a cluster in the left part of Figure~\ref{fig:code_embedding}.
+The nearest neighbors of CQECML+ are highlighted by different color intensity (i.e.
+OQECML+ being more purple indicates that it is more similar to CQECML+).
+The right part of the figure demonstrates another cluster formed with negative behavior codes including CQECML-, AMB-, RCHT-C, OQECHT-, GINFO-, RBAC, LUP-, RCHT-S, RPTBC, RAMBC, AMB-, RCML-S, etc.
+It is interesting that the behaviors intended to elicit CHT+/CML+ group together, whereas the ones intended to elicit CHT-/CML- also group together and are located in the opposite regions of semantic space.
 
@@
     \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
-    \caption{\textbf{T-SNE diagram of behavior codes drawn after 4000 steps with perplexity 30}.}
+    \caption{\textbf{2-D representation of behavior code embeddings}.}
     \label{fig:code_embedding}
@@
 
-\textbf {Recurrent Neural Networks (RNN)} are a type of Neural Networks, which can send feedback from the current hidden state to the hidden state of the next time step. The ability of RNN to capture
-long-term dependencies for predicting the future events is its main advantage over MC and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior
-observed at some point in the interview is very informative for the future behaviors, that will be observed later, but not immediately. However, it was observed that RNN fails to capture long-term
-dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short
-Term Memory networks or simply LSTM. There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU). GRUs are
-simpler than LSTM units and have been shown to experimentally outperform other models\cite{cho2014properties}. GRU is formally defined as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
+\textbf {Recurrent Neural Networks (RNN)} are a class of neural networks that have an internal memory, which makes them particularly suitable for processing sequences of observations.
+The ability of RNNs to capture long-term dependencies and remember past observations for predicting future observations is their main advantage over MCs and HMMs.
+These features are very useful in analysis of motivational interviews, in which any behavior observed at a particular point in the interview may be indicative of other behaviors that are observed later.
+In order to mitigate the vanishing gradient problem of earlier versions of RNN \cite{bengio1993problem}, Hochreiter et al.\cite{hochreiter1997long} proposed Long Short Term Memory networks (LSTM).
+There are several variants of LSTM model, among which the most notable one is the Gated Recurrent Unit\cite{cho2014properties} (GRU).
+GRUs are simpler than LSTMs and have been shown to be effective for a variety of Natural Language Processing tasks \cite{cho2014properties}.
+GRU is formally defined as follows:
 \begin{equation}
@@
 \end{equation}  
-As these equations, $\sigma$ stands for an element-wise application of the sigmoid function and $\odot$ is an element-wise product.
-The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W$ and $U$ represent weights and $b$ represents biases.
+In Eq.~\ref{eq:firstgru}-\ref{eq:lastgru}, $\sigma$ corresponds to sigmoid function and $\odot$ designates an element-wise product.
+The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors.
 The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
-\begin{equation}
-total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
-\label{eq:loss}
-\end{equation}  
+
 \begin{figure}[!htb]
@@
 \end{figure}
-The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes and embedding of MYSCOPE codes in an observed
-sequence of communication behaviors as input.
-Then, we applied a softmax layer at each time-step for predicting the label of the sequence.
-Since the sequence label was predicted at each time step, the proposed model is also known as Recurrent Neural Network with Target Replication (RNN-TR).
-As can be seen from
-Eq.~\ref{eq:loss}, the total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where $T$ is the total number of elements in a sequence,
-$\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that
-best performance achieved when we use $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which
-were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges
-to optimal after 100 epochs and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
+
+The RNN architecture employed for sequence classification is shown in Figure~\ref{fig:rnn-model}.
+As can be seen from Figure~\ref{fig:rnn-model}, softmax is used at each time step to predict the class of a sequence observed so far.
+Since the sequence label is predicted at each observation, the proposed architecture is referred to as Recurrent Neural Network with Target Replication (TR).
+It is trained by minimizing the following hybrid loss function:
+
+\begin{equation}
+\tilde{\mathcal{L}} = \alpha \cdot \frac{1}{T}\sum_{t=1}^T \mathcal{L}(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot \mathcal{L}(\bar y^{(T)},y^{(T)})
+\label{eq:loss}
+\end{equation}
+
+As follows from Eq.~\ref{eq:loss}, the total loss $\tilde{\mathcal{L}}$ is a convex combination of the final loss $\mathcal{L}(\bar y^{(T)},y^{(T)})$ and the average loss over all observations in a sequence, where $T$ is the total number of observations, $\bar y^{(t)}$ is the output at step $t$, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter controlling the relative importance of each loss type.
+We experimentally determined that the best performance is achieved when $\alpha=0.5$.
+Our model also contains several other hyperparameters, such as the number of embedding dimensions, the number of hidden units, learning rate, batch size, etc., which were optimized on the validation set.
+We implemented our models in Tensorflow with Adam optimizer as well as early stopping based on the validation loss and observed that our model converges after 100 epochs.
 
 \subsection*{\textit{Evaluation metrics}}
-Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds.
-However, LSTM and GRU are trained on 80\% of the data and validated on 10\%.
-The remaining 10\% of the data is used as a test set for reporting the performance of the model.
+Performance of probabilistic and deep learning methods\footnote{the code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis} was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds.
+However, LSTM and GRU were trained on 80\% of the data and validated on 10\%, with the remaining 10\% of the data used for testing.
 
@@
 \section*{Results}
-Experimental evaluation of the proposed method is conducted on both under and over-sampled sequences.
-Predictive performance summary of the proposed methods on under and over-sampled sequences is presented in Table~\ref{tab:result_under_over_sampled}.\\
+All sequence classification methods were evaluated in the case of both under and over-sampling.
+Predictive performance summary of all methods is summarized in Table~\ref{tab:result_under_over_sampled}.\\
+
 \begin{table}[ht]
 \centering
-\caption{Performance of MC, HMM, and RNN for predicting the success of under and over-sampled patient-provider communication sequences.
+\caption{Performance of MC, HMM, LSTM and GRU with and without target replication (TR) for predicting the success of patient-provider communication sequences when under- and over-sampling were used to balance the dataset.
 The highest value for each performance metric is highlighted in bold.}
@@
   \hline
-   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Undersampling}} & \multicolumn{3}{|c|}{\textbf{Oversampling}} \\\cline{2-7}
+   \multirow{2}{*}{\textbf{Method}} & \multicolumn{3}{|c|}{\textbf{Under-sampling}} & \multicolumn{3}{|c|}{\textbf{Over-sampling}} \\\cline{2-7}
    & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score} & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline
@@
  Hidden Markov Model & 0.6244 & 0.6143 & 0.6067 & 0.7775 & 0.7567 & 0.7520\\ \hline
- LSTM RNN & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
- LSTM RNN - TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
- GRU RNN & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
- GRU RNN - TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
+ LSTM & 0.8672 & 0.8626 & 0.8622 & 0.8411 & 0.8372 & 0.8368\\ \hline
+ LSTM-TR & \textbf{0.8733} & \textbf{0.8681} & \textbf{0.8677} & \textbf{0.8424} & \textbf{0.8385} & \textbf{0.8381}\\ \hline
+ GRU & 0.8674 & 0.8648 & 0.8646 & 0.8379 & 0.8342 & 0.8337\\ \hline
+ GRU-TR & 0.8705 & 0.8676 & 0.8673 & 0.8412 & 0.8377 & 0.8373\\ \hline 
   \end{tabular}
 \end{table}
-\subsection*{\textit{Predictive performance for under-sampled Patient-Provider Communication code sequences}}
-We applied small learning rate (0.00005) and batch size (8) with early stopping strategy for training deep learning models when the under-sampled dataset is used. Five major conclusions can be drawn
-from the results in Table~\ref{tab:result_under_over_sampled}. First, recurrent neural networks outperform probabilistic models and achieve 16.39\%-26.1\% higher F1-score. Second, LSTM with
-target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681. Third, target replication strategy improves the
-performance of GRU and LSTM, while conventional GRU shows better performance than traditional LSTM. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics
-for under-sampled sequences. Fifth, second-order MC has lower precision, recall, and F-measure than first-order MC. In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 
-9.36\%, when going from first to second-order MC model.
 
-\subsection*{\textit{Predictive performance for over-sampled Patient-Provider Communication code sequences}}
-Similar to the under-sampled dataset, early stopping strategy was also employed for the over-sampled dataset. For over-sampled data, RNN models were trained with learning rate 0.00010 and batch size
-55. Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still has 1.98\%,
-2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order Markov chain. Similar to under-sampled sequences, target replication improves the performance of RNN models, and LSTM with
-target replication has the highest F1-score among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences, while the performance of
-probabilistic models increases. We also observed that all models have the largest value in precision compared to other performance metrics.
+\subsection*{\textit{Predictive performance in the case of under-sampling}}
+We used a small learning rate of 0.00005 and the batch size of 8 along with early stopping strategy for training deep learning models on the dataset balanced with under-sampling.
+Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_over_sampled}.
+First, recurrent neural networks outperform probabilistic models and achieve 16.39\%-26.1\% higher F1-score.
+Second, LSTM with target replication has the best performance over all other RNN methods, and achieved F1-score 0.8677 with precision 0.8733 and recall 0.8681.
+Third, target replication strategy improves the performance of GRU and LSTM, with conventional GRU showing better performance than traditional LSTM.
+Fourth, among probabilistic models, the MC based method generally outperforms HMM across all metrics for under-sampled sequences.
+Fifth, second-order MC has lower precision, recall, and F-measure than first-order MC.
+In particular, precision, recall and F-measure decrease by 9.42\%, 9.36\% and 9.36\%, when going from first to second-order MC model.
+
+\subsection*{\textit{Predictive performance in the case of over-sampling}}
+Similar to the under-sampling scenario, early stopping strategy was also employed for training deep learning models on the dataset balanced with over-sampling.
+However, in this case, RNN models were trained with the learning rate of 0.00010 and the batch size of 55.
+Experimental results indicate that HMM had better performance than second-order MC, achieving 9.34\%, 7.65\%, and 7.43\% higher precision, recall, and F-measure, while HMM still had 1.98\%,
+2.97\%, and 3.28\% lower precision, recall, and F-measure than first-order MC.
+Also similar to the under-sampling scenario, target replication improves the performance of RNN models and LSTM with target replication has the highest F1-score among all models.
+However, the predictive performance of LSTM and RNN decreases when over-sampling is used, while the performance of probabilistic models increases.
 
@@
 \subsection*{\textit{Most likely communication sequences}}
-Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews. In successful
-sequences, we observed that information is frequently provided using patient-centered communication (GINFO+) and structure session, when the counselor either explains the therapeutic agenda or
-attempts to transition to a new topic or session content (SS). Sometimes, counselor acknowledges the clients' communication or an off topic comment (SO). We also observed that the affirm and OQECML+
-are consistent with MI theory, which claims their strong effect for eliciting positive change talk or commitment language. For unsuccessful sequences, it can be seen that providing advice using
-non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-), which is heading in the wrong direction therapeutically. Questions phased to elicit negative
-change talk or commitment language leads to CHT- or CML-, which is  consistent with the manual coding approach or analysis, or AMB-.
+Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews.
+We observed that in successful motivational interviews information is frequently provided using patient-centered communication (GINFO+) and structure session (SS) utterances, in which the counselor either explains the therapeutic agenda or attempts to transition to a new topic or session content.
+Sometimes, counselors also acknowledge the clients' communication or an off topic comment (SO).
+We also observed that affirmations (AF) and open-ended questions (OQECML+) have a strong effect on eliciting positive change talk or commitment language, which is consistent with MI theory.
+It can also be seen that providing advice using non-patient centered strategies (ADV-) leads to negative ambivalence (AMB-), which results in the interview heading in therapeutically wrong direction.
+Questions posed to elicit negative change talk or commitment language lead to CHT-, CML- or AMB-, which is  consistent with the manual analysis by clinicians.
 %It can be seen that most successful sequence starts with a summary of the discussion or structured statement.
@@
 \section*{Discussion}
-By analyzing the experimental results of different communication sequence outcome prediction methods proposed in this paper, we arrived at the following conclusions. First, the overall predictive
-performance of RNN models is substantially better than probabilistic models. In particular, the RNN-based method achieves near-human accuracy for predicting the label of motivational interviews. This
-indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the
-interviews. This provides evidence that the RNN is able to successfully replicate human cognitive processes to integrate previous information when formulating higher level thinking. In addition to
-that, embeddings allow to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of the prediction method.
+By analyzing the experimental results of different communication sequence outcome prediction methods proposed in this paper, we arrived at the following conclusions.
+First, the overall predictive performance of RNN based methods is substantially higher than that of probabilistic models.
+In particular, the RNN-based methods achieve near-human accuracy for predicting the success of motivational interviews.
+This indicates that RNN is able to capture the structure of discourse in motivational interviews by preserving long-term dependencies among the behavior codes, which reflect the overall progression of the interviews.
+This provides evidence that RNNs are able to successfully replicate human cognitive processes to integrate previous information when making decisions.
+In addition to that, embeddings allow to reduce the dimensionality of codes in PPC sequences and consequently improve both precision and recall of  prediction.
 
-Second, using target replication to compute the loss at each time step results in better performance for all configurations of the proposed RNN-based methods. This indicates that mean of the losses
-over all steps emphasize the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters. Better estimates of parameters in RNN
-models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence. This
-allows achieving an improvement in the prediction accuracy of the method.
+Second, using target replication to compute the loss at each time step results in better performance for all configurations of the proposed RNN-based methods.
+This indicates that the average of the losses over all steps emphasizes the dependencies between the pairs of patient and provider codes, which results in more accurate estimates of the model parameters.
+Better estimates of parameters in RNN models of motivational interviews are propagated to the next step based on the relative importance of intermediate output, where they are aggregated into predictions for the entire sequence.
+This allows to achieve an improvement in prediction accuracy.
 
-Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models, which we attribute to the fact that the number of states in higher-order
-Markov models grows exponentially with their order. As a result, accurate estimation of transition probabilities requires much larger training set. Using smaller datasets such as under-sampled dataset
-will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability
-estimates. Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing. In this study, we found out that using first-order Markov models is a
-reasonable trade-off between efficiency and accuracy.
+Third, using first-order Markov model results in better prediction accuracy compared to higher-order Markov models, which we attribute to the fact that the number of states in higher-order Markov models may grow exponentially with their order.
+As a result, accurate estimation of transition probabilities requires much larger training data.
+Using smaller datasets, which is the case when under-sampling is employed, will result in a sparsity problem, when many transitions are either not observed in the training set at all or observed only a few times, leading to missing or potentially inaccurate probability estimates.
+Obtaining large training sets cannot be easily accomplished in many domains, including motivational interviewing.
+In this study, we found out that using first-order Markov models is a reasonable trade-off between efficiency and accuracy.
 
-Fourth, similar to traditional Markov model, HMM achieves a dramatic improvement in the prediction accuracy when larger training set is used. This indicates that sufficient training data is required
-to estimate optimal hyperparameters such as a number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
+Fourth, similar to traditional Markov model, HMM achieves a dramatic improvement in the prediction accuracy when larger training set is used.
+This indicates that sufficient training data is required to find the optimal settings of hyperparameters, such as the number of hidden states, initial state distribution, transition probabilities, and emission probabilities.
 
@@
 \section*{Conclusion}
-In this paper, we compared the accuracy of Recurrent Neural Networks with Markov Chain and Hidden Markov Model for the task of predicting the success of motivational interviews. We found out
-that individual patient-provider communication exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
-Our proposed methods can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the
-interviews without resource-intensive manual qualitative analysis of interview transcripts, which can significantly decrease the time and effort required to develop behavioral interventions. Our
-proposed methods can also help to identify most likely sequences that are common to successful and unsuccessful motivational interviews, which can directly inform clinical practice and increase
-the effectiveness of behavioral interventions. Our experimental results indicate that the proposed methods can also be used for real-time monitoring of the progression of clinical interviews. This
-work also has broad implications for public health research by providing a formal theoretically-grounded computational mechanism for qualitative data analysis.
+In this paper, we compared the accuracy of Recurrent Neural Networks with Markov Chain and Hidden Markov Model for the task of predicting the success of motivational interviews.
+We found out that individual PPC exchanges are highly indicative of the overall progression and future trajectory of clinical interviews and can be used to predict their overall success.
+Our proposed methods can facilitate motivational interviewing researchers in establishing causal relationships between different communication strategies and the desired behavioral outcomes during the interviews without resource-intensive manual qualitative analysis of interview transcripts, which can significantly decrease the time and effort required to develop behavioral interventions.
+Our proposed methods can also help to identify the most likely sequences in successful and unsuccessful motivational interviews, which can directly inform clinical practice and increase the effectiveness of behavioral interventions.
+Our experimental results also indicate that the proposed methods can be used for real-time monitoring of the progression of clinical interviews.
+This work also has broad implications for public health research by providing a theoretically-grounded computational approach to qualitative data analysis.
 
@@
 This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
-We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset by manually annotating the dataset using the MYSCOPE codebook.
+We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their manually annotation of experimental dataset with MYSCOPE codes.
 

### 2018-01-10 paper-amia-summit-2018/amia_summit_2018.tex
update camera-ready version after fixing all the spelling and grammatical errors

@@
 
-\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution.
+\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided an equal contribution.
 \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}
@@
 \textit{The problem of analyzing temporally ordered sequences of observations generated by molecular, physiological or psychological processes to make predictions about the outcome of these processes arises in many domains of clinical informatics.
-In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of clinical dialog.
+In this paper, we focus on predicting the outcome of patient-provider communication sequences in the context of the clinical dialog.
 Specifically, we consider prediction of the motivational interview success (i.e.
@@
 \section*{Introduction}
-Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances  in clinical dialog).
+Temporally ordered sequences of discrete or continuous observations generated by molecular, psychological or psychological process(es) arise in many different areas of biology and medicine (e.g., DNA base-pairs, protein sequences, ECG measurements, laboratory results, diagnostic codes, utterances in the clinical dialog).
 Classification (or categorization) is a type of analysis of those sequences that has a broad range of important practical applications, from protein function \cite{yakhnenko2005discriminatively} or structure \cite{deshpande2002evaluation} prediction to detecting individuals with a heart disease \cite{wei2006semi}.
-Taking into account both the entire set of observations in a sequence, as well as the temporal order and potential dependencies between observations, makes sequence classification a more challenging task than classification of independent observations.
+Taking into account both the entire set of observations in a sequence, as well as the temporal order and potential dependencies between observations, makes sequence classification a more challenging task than a classification of independent observations.
 Predicting the outcome of those sequences (e.g.
@@
 Feature-based methods transform a sequence into a feature vector and apply a standard supervised machine learning method, such as Support Vector Machine \cite{leslie2004fast} or Decision Tree \cite{chuzhanova1998feature} to arrive at classification decision.
-The methods in this class have had limited success, since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
+The methods in this class have had limited success since traditional feature representation methods cannot easily account for the order of and dependencies between observations in a
 sequence.
@@
 However, distance metrics are primarily designed for time series data, in which the observations are discretized by timestamps.
-The third type of sequence classification methods first creates a probabilistic model, such as the Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class based on the training data and then classifies new sequences by applying the created models.
+The third type of sequence classification methods first creates a probabilistic model, such as the Markov Chain (MC) or Hidden Markov Model \cite{rabiner1989tutorial} (HMM), for sequences in each class based on the training data and then, classifies new sequences by applying the created models.
 While MCs and HMMs can capture first- and second-order dependencies between adjacent observations in a sequence, learning higher-order dependencies with these models requires prohibitively large amounts of data.
@@
 Specifically, LSTM was used as part of a multi-label classification method to recognize patterns in multivariate time series of clinical measurements, such as body temperature, heart rate and blood pressure\cite{lipton2015learning}.
-LSTM was also effectively used for predicting the diagnosis and medication codes, given a sequence of codes from previous patient visits \cite{choi2016doctor}.
+LSTM was also effectively used for predicting the diagnosis and medication codes, given a sequence of codes from the previous patient visits \cite{choi2016doctor}.
 A further simplification and improvement of LSTM model, called the Gated Recurrent
@@
 
-In this paper, we address the problem of predicting the outcome of coded patient-provider communication (PPC) sequences in the context of clinical dialog.
+In this paper, we address the problem of predicting the outcome of coded patient-provider communication (PPC) sequences in the context of the clinical dialog.
 Specifically, we focus on predicting the success (i.e.
@@
 The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change.
-These statements, collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
+These statements collectively referred to as ``change talk'' (CHT), consistently predict the actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview.
 However, the ability of providers to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews.
@@
 This study is a continuation of our previous work \cite{kotov2015interpretable, hasan2016study}, in which we explored several machine learning methods for automatic annotation of clinical interview fragments with a large number of patient and provider behavior codes from a specialized codebook \cite{carcone2013provider}.
-While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored applicability of state-of-the-art methods for sequence modeling to the analysis of PPC exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.
+While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, no previous work explored the applicability of state-of-the-art methods for sequence modeling to the analysis of PPC exchanges, in general, and predicting the desired patient behavioral response in the context of motivational interviews, in particular.
 
@@
 Since severely imbalanced datasets often distort the true performance of a classification method relative to a simple ``majority vote'' baseline (e.g.
-simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset), it is important to properly address class imbalance.
+simply classifying every communication sequence as successful would result in 82.15\% accuracy on our dataset), it is important to properly address the class imbalance.
 We evaluated the performance of probabilistic and deep learning methods using both under-sampling and over-sampling for balancing the number of samples in different classes.
@@
 Embedding is a representation of an object in low-dimensional space using a real-valued vector.
-In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as representation of behavior codes as input to these RNNs.
+In our study, embeddings of behavior codes were obtained as a by-product of training LSTM and GRU after feeding one-hot vectors as a representation of behavior codes as input to these RNNs.
 Behavior code embeddings have the property of representing similar codes with the vectors that are close to each other in low-dimensional space.
@@
 This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
-We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their manually annotation of experimental dataset with MYSCOPE codes.
+We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their manual annotation of experimental dataset with MYSCOPE codes.
 

### 2018-01-11 paper-amia-summit-2018/amia_summit_2018.tex
camera ready version
update reviewer's comments and manuscript for camera-ready version

@@
 \textbf {Behavior code embeddings}.
-Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural}.
+Representation of behavior codes was inspired by the recent success of word embeddings\cite{bengio2003neural,mikolov2013distributed,pennington2014glove}.
 Embedding is a representation of an object in low-dimensional space using a real-valued vector.
@@
 The update gate $z_t$ and reset gate $r_t$ at time step $t$ are computed by the Eq.~(\ref{eq:firstgru}) and~(\ref{eq:resetgru}), where $W_z$, $W_r$, $W_h$, $U_z$, $U_r$, $U_h$ are the weight matrices and $b_z$, $b_h$ and $b_r$ are bias vectors.
-The activation $h_t$ of the GRU at time $t$ is a linear combination of previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
+The activation $h_t$ of the GRU at time $t$ is a linear combination of the previous activation $h_{t-1}$ and the candidate activation $\tilde h_t$, which is represented by Eq.~(\ref{eq:lastgru}) and~(\ref{eq:candidategru}).
 

### 2018-01-11 paper-amia-summit-2018/amia_summit_2018.tex
update reviewer's comments and manuscript for camera-ready version

@@
 The ability of RNNs to capture long-term dependencies and remember past observations for predicting future observations is their main advantage over MCs and HMMs.
-These features are very useful in analysis of motivational interviews, in which any behavior observed at a particular point in the interview may be indicative of other behaviors that are observed later.
+These features are very useful in the analysis of motivational interviews, in which any behavior observed at a particular point in the interview may be indicative of other behaviors that are observed later.
 In order to mitigate the vanishing gradient problem of earlier versions of RNN \cite{bengio1993problem}, Hochreiter et al.\cite{hochreiter1997long} proposed Long Short Term Memory networks (LSTM).

### 2018-04-04 jhir/latex-manuscript/manuscript.tex
prepare revised version of manuscript with LaTeX

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{geometry}
\usepackage{pdflscape}
%
\usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
\journalname{J Healthc Inform Res}
%
\begin{document}

\title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
\thanks{This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Mehedi Hasan, PhDc$^1$\and 
April Idalski Carcone, PhD$^2$\and 
Sylvie Naar, PhD$^3$\and 
Susan Eggly, PhD$^4$\and  
Gwen L. Alexander, PhD$^5$\and 
Kathryn E Brogan Hartlieb, PhD$^6$\and 
Alexander Kotov, PhD$^1$
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Alexander Kotov (corresponding author) \at
              5057 Woodward Suite 14001.6 \\
              Detroit, MI 48202\\
              Tel.: +1(313) 577-9307\\
              \email{kotov@wayne.edu} 
           \and
           $^1$Department of Computer Science, College of Engineering, Wayne State University, Detroit, MI 48202\\
$^2$Division of Behavioral Health Sciences, Department of Family Medicine and Public Health Sciences, Wayne State University School of Medicine, Detroit, MI 48202\\
$^3$Director, Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, FL 32306\\  
$^4$Department of Oncology, Wayne State University/Karmanos Cancer Institute, Detroit, MI 48201\\
$^5$Department of Public Health Sciences, Henry Ford Health System, Detroit, MI 48202\\
$^6$Department of Humanities, Health and Society, Wertheim College of Medicine, Florida International University, Miami, FL 33199    
}

\date{Received: 12/03/2018 Revised: 04/15/2018 Accepted: }
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Motivational Interviewing (MI) is a communication technique to increase intrinsic motivation for behavior change. Although there is a strong empirical evidence linking “MI-consistent” counselor behaviors and patient statements of intrinsic motivation (i.e., “change talk”) for behavior change, the specific patterns of patient- counselor communication that are effective for eliciting patient change talk are a subject of ongoing research in behavioral sciences. A significant barrier to this research is the time and resource-intensive cognitive tasks, such as manual coding and sequential analysis of coded motivational interview transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had previously been annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify meaningful and interpretable counselor communication strategies to guide clinical practice.
\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
%\PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change [129]. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change which is accomplished through the exercise of relational and technical skills. [2]CITE] The relational hypothesis suggests that counselors’ use accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change. MI’s technical hypothesis [3]cite] states counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”. Cto increase intrinsic motivation (engaging in an activity for reasons of satisfaction rather than external consequences) atients“change talk”, that is, is the  statements patients make during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Change talk expressed during treatment sessions consistently predicts behavior change [440] with results persisting as long as 34 months post-intervention [541]. MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) will lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk. [3] These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal sequencing of patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

[6][3]Sequential analysis is an analytic approach to examine the temporal sequencing of behavioral events. [7,8].ADD] In a study of adults seeking or mandated to treatment for alcohol abuse, Moyers and Martin [9]ADD] found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. In a follow up study with the same population, [10]ADD] change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk. MIIN was unrelated to sustain talk but decreased the likelihood of change talk. Among young adults in brief motivational interviewing for hazardous alcohol consumption, Gaume and colleagues [11]ADD] also found a MICO-change talk and MICO-sustain talk sequential pattern; a MIIN-sustain talk pattern was not observed. A second study with the same population confirmed MICO led to significantly more change talk and sustain talk [12]ADD]. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior that led to increased change talk; reflections and other MICO behaviors but not questions were related to increased sustain talk. Glynn and colleagues [13]ADD] found reflections of change talk were more likely to elicit change talk among incarcerated adolescents with high rates of alcohol and marijuana use. Similarly, reflections of sustain talk were more likely to elicit more sustain talk. In contrast, Carcone and colleagues [14]ADD] found open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy were the only counselor behaviors likely to result in change talk among adolescents engaged in weight loss treatment. In a parallel study of the adolescents’ caregivers, Jacques-Tiura et al. [15]ADD] found the same; questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors’ use of reflections was the only MICO behavior consistently linked to change talk suggesting a need for additional research to understand in what contexts the various MICO strategies are effective, i.e., lead to change talk.

The sequential analysis procedure used in the above MI process studies [16-19] is based on first-order Markov Chain models [9, 10, 12]. Markov Chain is a discrete-time stochastic process with the property that the system state or condition changes over time and only depends upon previous event. Markov chain models have two main drawbacks. The first is their inability to preserve the long-term dependencies between observations in a sequence (they consider each observation to be dependent only on the immediately preceding one). In MI and other behavioral interventions, the antecedents of a given observed behavior could be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes, although in MI multiple similar behaviors may have the same outcome. Thus, first-order Markov models may be insufficient for understanding the associations among behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of behavior codes and long-term dependencies among behaviors, to identify such causal relationships. The goal of the current research is to develop a such a method to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting topic models [20-24], classification models [25-28] and neural networks [25, 29, 30] to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al. [27] developed a system for the evaluation of counselor fidelity to the MI framework by observing counselor language during the motivational interview. This system employed a support vector machine (SVM), which constructs a hyperplane in a high dimensional space to separate data points. In their approach, Perez-Rosas et al. proposed SVM model based on n-gram, syntactic, and semantic features, which achieved near 90\% accuracy for predicting specific MI counselor behaviors, such as reflections. Patient’s engagement, verbal and non-verbal accommodation as well as topics discussed during the MI session, were analyzed to identify linguistic and acoustic features of counselor empathy. These features were then used to build a counselor empathy classifier that achieved an accuracy of up to 80\%. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and neural networks in conjunction with lexical, contextual, and semantic features for the task of automated annotation of MI transcripts with different numbers of behavior codes (codebooks) [25].  In a follow up study [31], we formulated the problem of predicting the likelihood of eliciting a certain type of behavioral response during motivational interview as a sequence classification problem. Markov chains, HMMs and recurrent neural networks were applied to monitor the progression of motivational interviews and predicting the likelihood of eliciting “change talk”. This paper continues this line of research by examining the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the antecedent counselor communication strategies leading to patient change talk.

HMMs are widely used for analysis of sequence data due of their ability to model dependencies between coherent clusters of discrete observations in a sequence. HMM associates a “hidden” state with each observation in a sequence, where each state corresponds to a different distribution over observations, and model’s sequences of observations as transitions between hidden states and sampling observations from each hidden state. HMMs were originally proposed for speech recognition [3263], in which the states were used to represent all sounds of English. In biomedical informatics, HMMs were employed for diagnosis of diseases and biological sequence modeling [3366, 3467]. For example, in [3366] the Doppler ultrasound was used for feature extraction and then an HMM based classifier was applied to distinguish healthy patients from the patients with a heart disease, while in [3467], HMM was applied to capture important statistical characteristics of protein families. HMM and adjusted regression models were employed in [35] to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention.  In the case of applying HMM to modeling patient-counselor communication, hidden states correspond to a set of related behavior codes, such as a patient’s underlying emotional state, during a motivational interview. Although MI literature clearly identifies patient change talk and commitment language as antecedents of patient’s behavior change [40], there is less clarity on which counselor communication strategies influence the articulation of change talk. HMM modeling of successful and unsuccessful MI sessions could provide additional evidence to identify counselor communication strategies that contribute to patient change talk.

Frequent Pattern Mining [3668] is a class of data mining methods to identify sets of items (referred to as itemsets) which frequently appear together in a collection of sequences. Frequent Pattern Mining was first introduced by Agrawal and Srikant with the Apriori algorithm [3769], which was designed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied in several other domains including health informatics [3870-4173], medical imaging [39] and chemical and biological analysis [4274-4476], web mining [4577], and outlier analysis [4678]. Abdullah et. al. applied frequent pattern mining on medical billing data to analyze associations between diagnosis and treatments [3870]. In particular, the Apriori algorithm was used to identify most frequent itemsets1 that consist of a minimum of one diagnosis code (DxCode) and one current procedural terminology (CPT) code. Another study [3971] was conducted on medical image data for mining association rules to provide diagnosis support. At first, important features were extracted and then domain knowledge was integrated to generate rules for the interpretation of frequent patterns among important image features, such as object size, noise level, contrast, and texture, etc. However, to the best of our knowledge, nNo publishedrior research has yet examined the utility of frequent pattern mining to study patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which are difficult to interpret, is a major challenge in applying frequent pattern mining methods to patient- communication sequences. To address this problem, in this study, we utilized the closed frequent itemset mining method [4779], which produces more compact patterns that are easier to interpret. Specifically, we leveraged FPClose [4880], an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that most frequently lead to patient change talk. FPClose is the best algorithm in terms of running time and memory consumption proposed to date for mining closed frequent itemset, which works particularly well for sparse data.

This paper is one of the first empirical evaluations of the utility of HMM and closed frequent pattern mining for the study of patient-counselor communication during informing MI practice via automated analysis of patient- communication. Bertholet and colleagues published the only other study using this approach…. The goal of this study is to use HMM and FPM to inform clinical practitioners of the novel type of analytical results that these models can provide to help them better understand the antecedent counselor communication strategies leading to patient change talk during Motivational Interviewing sessions. In particular, tThese two models offerprovide the following advances over the sequential analysis computational methods currently used for MI research:. 
As opposed to fFirst-order Markov chain models, which identify the influence of counselor statements immediately preceding on patient behavior only in the form of statements using a transition matrix between of individual behavior codes. In contrast, HMMs identifies summarizeidentify transitions between clusters of related behavior codes, allowing theo capture identification to examine which of the antecedent behavior cluster is responsible for the elicitation of change talk and sustain talk in successful and unsuccessful patient-counselor communication.of more abstract patterns in patient-counselor communication;. 
As opposed to the first-order Markov models and HMMs, fFrequent pattern mining can capture the can identify patterns that involveing long-range dependencies between patient and counselor behaviors. Long-range dependency is important because human behavior is informed by all the antecedent behaviors not just the immediately preceding behavior.

In this paper, we focus on computational methods to facilitate analysis of transcripts to identify patterns of patient-counselor communication, in the form of sequences of behavior codes, in both successful and unsuccessful motivational interviews. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge can inform clinical practice by facilitating the development of more effective and tailored behavioral interventions, including the ones focused on weight loss among minority adolescents.
\section{Section title}
\label{sec:1}
Text with citations \citep{wei2006semi} and \citep{keogh2000scaling}.
\subsection{Subsection title}
\label{sec:2}
as required. Don't forget to give each section
and subsection a unique label (see Sect.~\ref{sec:1}).
\paragraph{Paragraph headings} Use paragraph headings as needed.
\begin{equation}
a^2+b^2=c^2
\end{equation}

% For one-column wide figures use
\begin{figure}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics{example.eps}
% figure caption is below the figure
\caption{Please write your figure caption here}
\label{fig:1}       % Give a unique label
\end{figure}
%
% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=0.75\textwidth]{example.eps}
% figure caption is below the figure
\caption{Please write your figure caption here}
\label{fig:2}       % Give a unique label
\end{figure*}
%
%\newgeometry{margin=2cm} % modify this if you need even more space
\begin{landscape}

% For tables use
\begin{table}
% table caption is above the table
\caption{Please write your table caption here}
\label{tab:1}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

% For tables use
\begin{table}
% table caption is above the table
\caption{Please write your table caption here}
\label{tab:2}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\end{landscape}
\restoregeometry

\begin{acknowledgements}
If you'd like to thank anyone, place your comments here and remove the percent signs.
\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
\bibliographystyle{natbib}       % JHIR format
%\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{references}   % name your BibTeX data base

\end{document}
% end of file template.tex

### 2018-04-05 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 
-\author{Mehedi Hasan, PhDc$^1$\and 
+\author{Mehedi Hasan, BS$^1$\and 
 April Idalski Carcone, PhD$^2$\and 
@@
 \begin{abstract}
-Motivational Interviewing (MI) is a communication technique to increase intrinsic motivation for behavior change.
-Although there is a strong empirical evidence linking “MI-consistent” counselor behaviors and patient statements of intrinsic motivation (i.e., “change talk”) for behavior change, the specific patterns of patient- counselor communication that are effective for eliciting patient change talk are a subject of ongoing research in behavioral sciences.
-A significant barrier to this research is the time and resource-intensive cognitive tasks, such as manual coding and sequential analysis of coded motivational interview transcripts, that are an integral part of the qualitative analysis of patient-counselor communication.
+Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change.
+Although there is strong empirical evidence linking “MI-consistent” counselor behaviors and patient motivational statements (i.e., “change talk”), the specific counselor communication behaviors effective for eliciting patient change talk are a subject of ongoing research.
+A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication.
 Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks.
-In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining to inform MI practice.
+In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining, an algorithm used to mine the closed frequent patterns in MI communication sequences to inform MI practice.
 We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers.
@@
 Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify meaningful and interpretable counselor communication strategies to guide clinical practice.
+
 \keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
@@
 \section{Introduction}
-\label{intro}
-Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change [129].
-The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change which is accomplished through the exercise of relational and technical skills.
-[2]CITE] The relational hypothesis suggests that counselors’ use accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change.
-MI’s technical hypothesis [3]cite] states counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”.
-Cto increase intrinsic motivation (engaging in an activity for reasons of satisfaction rather than external consequences) atients“change talk”, that is, is the  statements patients make during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
-Change talk expressed during treatment sessions consistently predicts behavior change [440] with results persisting as long as 34 months post-intervention [541].
-MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) will lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
-Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk.
-[3] These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal sequencing of patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.
+\label{sec:intro}
+Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \citep{miller2013motivational}.
+The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \citep{miller2013motivational}.
+The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change.
+MI’s technical hypothesis \citep{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”.
+Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
+Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4].
+In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
+Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5].
+These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.
 
-[6][3]Sequential analysis is an analytic approach to examine the temporal sequencing of behavioral events.
-[7,8].ADD] In a study of adults seeking or mandated to treatment for alcohol abuse, Moyers and Martin [9]ADD] found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
-In a follow up study with the same population, [10]ADD] change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk.
-MIIN was unrelated to sustain talk but decreased the likelihood of change talk.
-Among young adults in brief motivational interviewing for hazardous alcohol consumption, Gaume and colleagues [11]ADD] also found a MICO-change talk and MICO-sustain talk sequential pattern; a MIIN-sustain talk pattern was not observed.
-A second study with the same population confirmed MICO led to significantly more change talk and sustain talk [12]ADD].
+Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations [6, 7].
+Moyers and Martin [8] applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
+A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk [9].
+Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
+Gaume et al.
+[10] used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
+They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
+A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk [11].
 In this sample, MIIN led to greater sustain talk, but was unrelated to change talk.
-Further analyses revealed that reflections were the only MICO behavior that led to increased change talk; reflections and other MICO behaviors but not questions were related to increased sustain talk.
-Glynn and colleagues [13]ADD] found reflections of change talk were more likely to elicit change talk among incarcerated adolescents with high rates of alcohol and marijuana use.
-Similarly, reflections of sustain talk were more likely to elicit more sustain talk.
-In contrast, Carcone and colleagues [14]ADD] found open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy were the only counselor behaviors likely to result in change talk among adolescents engaged in weight loss treatment.
-In a parallel study of the adolescents’ caregivers, Jacques-Tiura et al.
-[15]ADD] found the same; questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
-Across these studies, counselors’ use of reflections was the only MICO behavior consistently linked to change talk suggesting a need for additional research to understand in what contexts the various MICO strategies are effective, i.e., lead to change talk.
+Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk.
+Glynn and colleagues [12] linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
+In a study of adolescents engaged in weight loss treatment, Carcone et al.
+[13] used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
+A parallel study of the adolescents’ caregivers [14] drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
+Across these studies, counselors’ use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.
 
-The sequential analysis procedure used in the above MI process studies [16-19] is based on first-order Markov Chain models [9, 10, 12].
-Markov Chain is a discrete-time stochastic process with the property that the system state or condition changes over time and only depends upon previous event.
-Markov chain models have two main drawbacks.
-The first is their inability to preserve the long-term dependencies between observations in a sequence (they consider each observation to be dependent only on the immediately preceding one).
-In MI and other behavioral interventions, the antecedents of a given observed behavior could be influenced by any of the preceding behaviors.
-The second drawback is their inability to consider similarities between behavior codes, although in MI multiple similar behaviors may have the same outcome.
-Thus, first-order Markov models may be insufficient for understanding the associations among behaviors in patient-counselor communication sequences.
-Therefore, there is a need for more powerful computational methods, which consider clusters of behavior codes and long-term dependencies among behaviors, to identify such causal relationships.
-The goal of the current research is to develop a such a method to identify effective patterns of patient-counselor communication.
+The sequential analysis procedure used in the above MI process studies [8, 15-17] is based on the first-order Markov Chain model [8, 9, 11].
+The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event.
+Hence, Markov chain models have two main drawbacks.
+The first is their inability to preserve the long-range dependencies between observations in a sequence.
+However, in MI, an observed behavior can be influenced by any of the preceding behaviors.
+The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome.
+Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences.
+Therefore, there is a need for more powerful computational methods, which consider clusters of behavior codes and long-term dependencies between behaviors, to identify causal relationships.
+The goal of the current research is to test the applicability of data mining methods to identify effective patterns of patient-counselor communication.
 
-Several prior works have reported the results of adopting topic models [20-24], classification models [25-28] and neural networks [25, 29, 30] to the tasks of annotating MI transcripts and assessing intervention fidelity.
+Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and primarily for assessing intervention fidelity.
 Perez-Rosas et al.
-[27] developed a system for the evaluation of counselor fidelity to the MI framework by observing counselor language during the motivational interview.
-This system employed a support vector machine (SVM), which constructs a hyperplane in a high dimensional space to separate data points.
-In their approach, Perez-Rosas et al.
-proposed SVM model based on n-gram, syntactic, and semantic features, which achieved near 90\% accuracy for predicting specific MI counselor behaviors, such as reflections.
+[25] developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
+This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, which achieved nearly 90\% accuracy for predicting counselor questions and reflections.
 Patient’s engagement, verbal and non-verbal accommodation as well as topics discussed during the MI session, were analyzed to identify linguistic and acoustic features of counselor empathy.
 These features were then used to build a counselor empathy classifier that achieved an accuracy of up to 80\%.
-In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and neural networks in conjunction with lexical, contextual, and semantic features for the task of automated annotation of MI transcripts with different numbers of behavior codes (codebooks) [25].
-In a follow up study [31], we formulated the problem of predicting the likelihood of eliciting a certain type of behavioral response during motivational interview as a sequence classification problem.
-Markov chains, HMMs and recurrent neural networks were applied to monitor the progression of motivational interviews and predicting the likelihood of eliciting “change talk”.
-This paper continues this line of research by examining the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the antecedent counselor communication strategies leading to patient change talk.
+In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes [23].
+The SVM model with the proposed features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders.
+In a follow up study [29], we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
+RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews.
+This paper represents the next phase in this line of research.
+We examine the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the counselor communication strategies antecedent to patient change talk.
 
-HMMs are widely used for analysis of sequence data due of their ability to model dependencies between coherent clusters of discrete observations in a sequence.
-HMM associates a “hidden” state with each observation in a sequence, where each state corresponds to a different distribution over observations, and model’s sequences of observations as transitions between hidden states and sampling observations from each hidden state.
-HMMs were originally proposed for speech recognition [3263], in which the states were used to represent all sounds of English.
-In biomedical informatics, HMMs were employed for diagnosis of diseases and biological sequence modeling [3366, 3467].
-For example, in [3366] the Doppler ultrasound was used for feature extraction and then an HMM based classifier was applied to distinguish healthy patients from the patients with a heart disease, while in [3467], HMM was applied to capture important statistical characteristics of protein families.
-HMM and adjusted regression models were employed in [35] to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention.
-In the case of applying HMM to modeling patient-counselor communication, hidden states correspond to a set of related behavior codes, such as a patient’s underlying emotional state, during a motivational interview.
-Although MI literature clearly identifies patient change talk and commitment language as antecedents of patient’s behavior change [40], there is less clarity on which counselor communication strategies influence the articulation of change talk.
-HMM modeling of successful and unsuccessful MI sessions could provide additional evidence to identify counselor communication strategies that contribute to patient change talk.
+HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence.
+The HMM associates each observation in a sequence with a “hidden” state, where the hidden state corresponds to a different distribution over observations (i.e., distribution over behavior codes).
+Sequences of observations are modeled as transitions between different hidden states and sampling observations from each hidden state.
+HMMs were originally proposed for speech recognition [30], in which the states were used to represent all English language sounds.
+In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling [31, 32].
+For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease [31], while HMM was used to capture important characteristics of protein families [32].
+In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter.
+Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change [33]) as the antecedents of patient’s behavior change [3], there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
+Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.
 
-Frequent Pattern Mining [3668] is a class of data mining methods to identify sets of items (referred to as itemsets) which frequently appear together in a collection of sequences.
-Frequent Pattern Mining was first introduced by Agrawal and Srikant with the Apriori algorithm [3769], which was designed to identify patterns in customer purchases.
-Since its introduction, frequent pattern mining has been applied in several other domains including health informatics [3870-4173], medical imaging [39] and chemical and biological analysis [4274-4476], web mining [4577], and outlier analysis [4678].
-Abdullah et.
-al.
-applied frequent pattern mining on medical billing data to analyze associations between diagnosis and treatments [3870].
-In particular, the Apriori algorithm was used to identify most frequent itemsets1 that consist of a minimum of one diagnosis code (DxCode) and one current procedural terminology (CPT) code.
-Another study [3971] was conducted on medical image data for mining association rules to provide diagnosis support.
-At first, important features were extracted and then domain knowledge was integrated to generate rules for the interpretation of frequent patterns among important image features, such as object size, noise level, contrast, and texture, etc.
-However, to the best of our knowledge, nNo publishedrior research has yet examined the utility of frequent pattern mining to study patient-counselor communication.
-A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which are difficult to interpret, is a major challenge in applying frequent pattern mining methods to patient- communication sequences.
-To address this problem, in this study, we utilized the closed frequent itemset mining method [4779], which produces more compact patterns that are easier to interpret.
-Specifically, we leveraged FPClose [4880], an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that most frequently lead to patient change talk.
-FPClose is the best algorithm in terms of running time and memory consumption proposed to date for mining closed frequent itemset, which works particularly well for sparse data.
+Frequent pattern mining [34] is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together.
+Agrawal and Srikant [35] first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
+Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics [36-39], medical imaging [37], chemical and biological analysis [40-42], web mining [43], and outlier analysis [44].
+No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
+A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
+To address this problem, we utilized the closed frequent itemset mining method [45], which produces less number of patterns in more compact form that are easier to interpret.
+In this study, we leveraged FPClose [46], an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
+FPClose is the best closed frequent itemset mining algorithm, which shown highest performance in terms of running time and memory consumption.
 
-This paper is one of the first empirical evaluations of the utility of HMM and closed frequent pattern mining for the study of patient-counselor communication during informing MI practice via automated analysis of patient- communication.
-Bertholet and colleagues published the only other study using this approach….
-The goal of this study is to use HMM and FPM to inform clinical practitioners of the novel type of analytical results that these models can provide to help them better understand the antecedent counselor communication strategies leading to patient change talk during Motivational Interviewing sessions.
-In particular, tThese two models offerprovide the following advances over the sequential analysis computational methods currently used for MI research:.
-As opposed to fFirst-order Markov chain models, which identify the influence of counselor statements immediately preceding on patient behavior only in the form of statements using a transition matrix between of individual behavior codes.
-In contrast, HMMs identifies summarizeidentify transitions between clusters of related behavior codes, allowing theo capture identification to examine which of the antecedent behavior cluster is responsible for the elicitation of change talk and sustain talk in successful and unsuccessful patient-counselor communication.of more abstract patterns in patient-counselor communication;.
-As opposed to the first-order Markov models and HMMs, fFrequent pattern mining can capture the can identify patterns that involveing long-range dependencies between patient and counselor behaviors.
-Long-range dependency is important because human behavior is informed by all the antecedent behaviors not just the immediately preceding behavior.
+In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
+Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
+This knowledge will inform MI theory by providing additional evidence to support MI’s technical hypothesis.
+It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication.
+This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions.
+Bertholet et al.
+[47] used HMM and adjusted regression models to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention.
+Their HMM model limited hidden states to three, which are "towards change", "away from change" and "non-determined".
+First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention.
+In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication.
+The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
+These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research.
+First-order Markov chain models identify the likely transitions between individual behaviors.
+In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications.
+Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors.
+Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communication during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.
 
-In this paper, we focus on computational methods to facilitate analysis of transcripts to identify patterns of patient-counselor communication, in the form of sequences of behavior codes, in both successful and unsuccessful motivational interviews.
-Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
-This knowledge can inform clinical practice by facilitating the development of more effective and tailored behavioral interventions, including the ones focused on weight loss among minority adolescents.
-\section{Section title}
-\label{sec:1}
-Text with citations \citep{wei2006semi} and \citep{keogh2000scaling}.
-\subsection{Subsection title}
-\label{sec:2}
-as required. Don't forget to give each section
-and subsection a unique label (see Sect.~\ref{sec:1}).
-\paragraph{Paragraph headings} Use paragraph headings as needed.
-\begin{equation}
-a^2+b^2=c^2
-\end{equation}
+\section{Methods}
+\label{sec:methods}
+\subsection{Dataset}
+\label{subsec:dataset}
+The data set for this study consists of 37 transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss treatment.
+Each transcript represented a one-on-one interaction between the adolescent patient and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
+These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE) [13].
+The MYSCOPE is an adaptation of the original MI-SCOPE [48], a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI) [49], the MI Skill Code (MISC) [50], and Amrhein’s conceptualization of change talk and commitment [51].
+A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% (n = 7) of the sessions for inter-rater reliability which was good (k = .696).
+The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
 
-% For one-column wide figures use
-\begin{figure}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
-  \includegraphics{example.eps}
-% figure caption is below the figure
-\caption{Please write your figure caption here}
-\label{fig:1}       % Give a unique label
-\end{figure}
-%
+% For tables use
+\begin{table}
+% table caption is above the table
+\caption{MYSCOPE codebook.}
+\label{tab:codebook}       % Give a unique label
+% For LaTeX tables use
+\begin{tabular}{lll}
+\hline\noalign{\smallskip}
+first & second & third  \\
+\noalign{\smallskip}\hline\noalign{\smallskip}
+number & number & number \\
+number & number & number \\
+\noalign{\smallskip}\hline
+\end{tabular}
+\end{table}
+
+\subsection{Data preprocessing}
+\label{subsec:datapreprocessing}
+Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences.
+For each MI transcript, the stream of behavior codes from beginning to end was analyzed.
+Successful sequences were those that resulted in a patient change talk or commitment language statement and all preceding behavior codes were included.
+Unsuccessful sequences were similarly created for sequences resulting in sustain talk.
+A total of 1,360 sequences were generated.
+The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high.
+Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.
+
+\subsection{Data Analysis}
+\label{subsec:dataanalysis}
+The Hidden Markov Model (HMM; implemented in the hmmlearn package which is publicly available at http://hmmlearn.readthedocs.io/en/latest/index.html) analysis aimed to identify clusters of related behavior codes within successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
+Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence.
+The emission probability matrix describes the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions among the hidden states.
+
+Training an HMM with a given number of hidden states (N) involves estimating the following parameters using the Baum-Welch algorithm:
+M is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table 1).
+T is an  state transition probability matrix where  
+E is an  matrix where , with observation symbol probability distribution for each state (emission probability matrix), and
+p is the initial state distribution vector where .
+
+In this study, two HMM models were trained, one using successful sequences and the other using unsuccessful sequences, with the objective of maximizing log-likelihood on the observed sequence.
+The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest BIC value, which indicates an optimal fit of HMM model.
+BIC minimizes negative log-likelihood but penalizes complexity (i.e.
+\# of hidden states), which is important for avoiding overfitting problem.
+HMM analyses suggested 5 hidden states was optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).
+
 % For two-column wide figures use
@@
 % For example, with the graphicx package use
-  \includegraphics[width=0.75\textwidth]{example.eps}
+  \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
 % figure caption is below the figure
-\caption{Please write your figure caption here}
-\label{fig:2}       % Give a unique label
+\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states.}
+\label{fig:bic}       % Give a unique label
 \end{figure*}
-%
+
+A second experiment was conducted with Frequent Pattern Mining.
+Orange3 and arules are two available pattern mining libraries implemented in python and r-programming, which support only a few pattern mining algorithms including frequent itemsets and association rules mining.
+However, we conducted our experiment using SPMF [52, 53], which offered more than 150 data mining algorithms.
+Moreover, SPMF is an open-source, fast and lightweight, a most popular and widely used Java-based data mining library.
+A frequent itemset is determined by surpassing a user-specified threshold called minimum support count\footnote{Support count: \# of times an itemset appears in the database}.
+For example, \{A\}, \{C\}, {D}, and {C, D} are frequent itemset in the database illustrated in Figure~\ref{fig:fpm}, because these itemsets appear at least 2 times which was the minimum support count identified in the example database.
+We reported closed frequent itemsets among all sequences of behavior codes in successful and unsuccessful sequences.
+A frequent itemset is closed if none of its supersets have the same support count [45], where a set X is a superset of another set Y if X contains all elements of the set Y.
+For example, the itemsets {A}, {C}, {D}, and {C, D} are not closed frequent itemsets because their supersets \{A, B\}, {B, C}, {B, D}, and {B, C, D} have the same number of support.
+Therefore, {B}, {A, B}, {A, D}, {B, C}, {B, D}, and {B, C, D} are closed frequent itemsets because none of their supersets have same the number of support count.
+On the other hand, {A, C}, {A, B, C}, {A, B, D}, {A, C, D} and {A, B, C, D} have support count 1, 0, 1, 0, and 1, respectively, which are less than minimum support count and identified as non-frequent itemsets.
+This study applied FPClose [46], an efficient close frequent pattern mining algorithm in SPMF, to identify frequent patient-counselor behaviors in MI communication sequences.
+To improve the model’s ability to identify frequent patient behaviors, FPClose [46], a FP-tree based algorithm [54, 55] was used.
+FPClose identifies closed frequent itemsets by utilizing an FP-tree data structure in combination with the FP-array technique to avoid the frequent traversal of FP-trees.
+Before applying the FPClose method, the dataset was transformed into the expected format used by the algorithm in the SPMF library.
+Minimum support threshold is a certain proportion having a meaningful representation of patterns determined by the domain expert.
+Since there is no standard for the selection of minimum support threshold, by following the studies [56-57], the minimum support count considered was 10\% of the total number of sequences which is 110 for successful and 25 for unsuccessful sequences.
+In our study, support counts representing ≥10\% of the patterns was the threshold for the result interpretation.
+
+% For two-column wide figures use
+\begin{figure*}
+% Use the relevant command to insert your figure file.
+% For example, with the graphicx package use
+  \includegraphics[width=0.95\textwidth]{figures/fpm.eps}
+% figure caption is below the figure
+\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
+\label{fig:fpm}       % Give a unique label
+\end{figure*}
+
+\section{Results}
+\label{sec:results}
+The transition and emission probability matrices of HMM model, are reported in Tables~\ref{tab:transition} and~\ref{tab:emission}.
+The three behaviors with the highest frequency, representing 45-69\% of each state’s emission transitions, were used to interpret the emission matrix and label the hidden states.
+Most (84\%) successful sequences began in a state characterized as “high motivation” as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
+Successful sequences began in a state of “high receptivity” 11\% of the time.
+“High receptivity” sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%).
+Few successful sequences began in states of “moderate receptivity” and “low receptivity” (2\% and 3\% of the time, respectively).
+These two states were characterized by different proportions of the same behaviors.
+Moderate receptivity sequences were distinguished from low receptivity sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%).
+No (0\%) successful sequence began in the “active feedback” state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%).
+Successful sequences transitioned from “high motivation” to “active feedback” most often (41\%).
+“active feedback”, in turn, most frequently transitioned to “high receptivity” (39\%).
+The “moderate receptivity” state transitioned to most often to “active feedback” (27\%) and automatically transitioned back to “moderate receptivity” (24\%) or to “low receptivity” (23\%) with similar frequency.
+The full transition matrix is presented in Table~\ref{tab:transition}.
+
+In contrast, the majority of unsuccessful sequences (98\%) began in a state of “ambivalence” as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
+About 2\% of the time unsuccessful sequences started in a state of “avoidance”.
+Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished “avoidant” sequences.
+Both “ambivalent” (84\%) and “avoidant” (61\%) states most frequently transitioned to the “avoidant” state.
+
 %\newgeometry{margin=2cm} % modify this if you need even more space
 \begin{landscape}
-
 % For tables use
@@
 % table caption is above the table
-\caption{Please write your table caption here}
-\label{tab:1}       % Give a unique label
+\caption{Hidden Markov Model Emission Matrices}
+\label{tab:emission}       % Give a unique label
 % For LaTeX tables use
@@
 % table caption is above the table
-\caption{Please write your table caption here}
-\label{tab:2}       % Give a unique label
+\caption{Hidden Markov Model Transition Matrices}
+\label{tab:transition}       % Give a unique label
 % For LaTeX tables use
@@
 
+Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}.
+Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
+For each pattern, the significance of difference between successful and unsuccessful communications was reported by computing p-value using Pearson chi-square test.
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient’s decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
+In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= <0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p=<0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p=<0.0001).
+In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences.
+In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.
+
+% For tables use
+\begin{table}
+% table caption is above the table
+\caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences.}
+\label{tab:patterns}       % Give a unique label
+% For LaTeX tables use
+\begin{tabular}{lll}
+\hline\noalign{\smallskip}
+first & second & third  \\
+\noalign{\smallskip}\hline\noalign{\smallskip}
+number & number & number \\
+number & number & number \\
+\noalign{\smallskip}\hline
+\end{tabular}
+\end{table}
+
+\section{Discussion}
+\label{sec:discussion}
+We present the application of HMM and frequent pattern mining to test the technical hypothesis guiding Motivational Interviewing which suggests that counselors use of “MI-consistent” communication strategies (MICO) will lead to patient change talk [2].
+Previous studies have empirically linked counselors’ use of MICO to higher rates of patient change talk in first-order Markov Chain models [8, 9, 11].
+Evidence from the current study using machine-learning methods of pattern analysis provides stronger evidence for MI’s technical hypothesis by considering longer-term dependencies in the data.
+Both HMM and frequent pattern mining consider behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement offered by simple first-order Markov chain models.
+The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing which has previously relied upon simple Markov chain models [8-14].
+
+The states identified in HMM model, were given labels to describe the underlying “state” of the MI session.
+These labels were based on an analysis and interpretation of the three behaviors with the highest frequency according to the emission matrix.
+Three behaviors were interpreted because together they represented 45-69\% of the behaviors associated with a given state and all represented at least 10\%.
+These labels made it easier to understand types of interactions occurred between counselors and patients.
+
+In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement).
+In HMM, the majority successful sequences began in the high motivation state where counselors frequently used reflections of change talk or commitment language and affirmations.
+Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients’ decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies.
+The frequent pattern mining results were similar.
+Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient’s decision-making autonomy, and sensitively provided information.
+Previous MI sequential analysis studies using first-order Markov chain models to analyze communication sequences have linked patients’ expression of change talk to counselor reflections of change talk, [9, 11-14] open questions phrased to elicit change talk, [9, 13, 14], and statements emphasizing the patient’s decision-making autonomy [13, 14].
+These studies did not find a link between change talk and counselors’ use of affirmations or the provision of information when examined independently of the MICO index.
+Thus, the current study is the first to provide empirical evidence for these causal linkages.
+One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
+Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change [58].
+Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
+
+The analysis of unsuccessful sequences, i.e., those resulting in a patient sustain talk statement, was typified by a combination of MICO and MI-inconsistent communication strategies (MIIN).
+Specifically, the majority of unsuccessful sequences in the HMM analysis began in a state of Ambivalence which was characterized by large proportions of counselor reflections of both change talk and sustain talk.
+Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed.
+These results are consistent with those of Gaume and colleagues [11] who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
+Specifically, counselors’ use of simple and complex reflections and “other MICO” behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
+Carcone et al [13] found counselors’ questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
+In contrast, Moyers et al.
+[9] found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
+These variable findings provide further support for the need to tailor the communication strategies MI counselors use to the treatment context.
+
+This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns.
+We have previously reported on the development of a model to annotate clinical transcripts [18, 21, 23].
+Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits).
+In this study, we present the next phase of this research, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
+Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
+We are also experimenting with machine learning methods to segment (parse) the stream of communication behavior into chunks representing a patient or counselor utterance to which the annotation model can then assign a behavior code.
+Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions.
+An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanism of Motivational Interviewing which informs theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies with discrete patient populations.
+
+This study is limited by the use of one dataset composed of 37 Motivational Interviewing transcripts of counseling sessions with African American adolescents in weight loss treatment.
+Thus, there is a need to replicate these findings with larger and more diverse data samples as the findings may not be representative of communication patterns in other contexts employing the Motivational Interviewing framework.
+In fact, when interpreted in light of the published literature, the results obtained in these experiments suggest that communication patterns are likely to vary given the treatment context.
+There are, however, consistencies with previous Motivational Interviewing process studies providing support for the validity of our findings and suggesting some counselor communication strategies may cut across treatment contexts.
+Next steps for this work include enhancing the performance and utility of machine learning models for sequential analysis by combining transcript annotation with sequential analysis.
+Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently.
+One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to one of success cannot be determined from these analyses.
+
+\section{Conclusion}
+\label{sec:conclusion}
+These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
+While counselors’ use of “MI-consistent” communication behaviors has been previously linked to higher rates of change talk in correlational studies [8, 15-17] and simple Markov chain models [8, 9, 11].
+The use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
+The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.
+
+\section{Conflict of Interest}
+\label{sec:conflictofInterest}
+On behalf of all authors, the corresponding author states that there is no conflict of interest.
+
+%Text with citations \citep{wei2006semi} and \citep{keogh2000scaling}.
+
+%\paragraph{Paragraph headings} Use paragraph headings as needed.
+%\begin{equation}
+%a^2+b^2=c^2
+%\end{equation}
+
 \begin{acknowledgements}
-If you'd like to thank anyone, place your comments here and remove the percent signs.
+We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset.
+The authors would like to thank Lisa Todd, JD, MA for her thoughtful feedback and clinical expertise in the interpretation of these data.
 \end{acknowledgements}

### 2018-04-11 jhir/latex-manuscript/manuscript.tex
rotate table

@@
 %
+\usepackage{rotating}
 % Insert the name of "your journal" with
@@
 \title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
-\thanks{This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.}
 }
@@
 %\newgeometry{margin=2cm} % modify this if you need even more space
-\begin{landscape}
-% For tables use
-\begin{table}
-% table caption is above the table
+%\newgeometry{left=4cm}
+\begin{sidewaystable}
 \caption{Hidden Markov Model Emission Matrices}
@@
 \end{tabular}
-\end{table}
 
-% For tables use
-\begin{table}
-% table caption is above the table
+\vspace{2\baselineskip}
+
 \caption{Hidden Markov Model Transition Matrices}
@@
 \end{tabular}
-\end{table}
+\end{sidewaystable}
 
-\end{landscape}
-\restoregeometry
+%\restoregeometry
 
@@
 \begin{acknowledgements}
+This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
 We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset.

### 2018-04-11 jhir/latex-manuscript/manuscript.tex
finalized manuscript layout

@@
 \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
+
 \usepackage{rotating}
+\usepackage{tabularx}
+\usepackage[flushleft]{threeparttable}
+
 % Insert the name of "your journal" with
@@
 }
-%\subtitle{Do you have a subtitle?\\ If so, write it here}
 
-%\titlerunning{Short form of title}        % if too long for running head
-
-\author{Mehedi Hasan, BS$^1$\and 
-April Idalski Carcone, PhD$^2$\and 
-Sylvie Naar, PhD$^3$\and 
-Susan Eggly, PhD$^4$\and  
-Gwen L. Alexander, PhD$^5$\and 
-Kathryn E Brogan Hartlieb, PhD$^6$\and 
-Alexander Kotov, PhD$^1$
+\author{Mehedi Hasan$^1$\and 
+April Idalski Carcone$^2$\and 
+Sylvie Naar$^3$\and 
+Susan Eggly$^4$\and  
+Gwen L. Alexander$^5$\and 
+Kathryn E Brogan Hartlieb$^6$\and 
+Alexander Kotov$^1$
 }
 
-%\authorrunning{Short form of author list} % if too long for running head
-
-\institute{Alexander Kotov (corresponding author) \at
-              5057 Woodward Suite 14001.6 \\
-              Detroit, MI 48202\\
-              Tel.: +1(313) 577-9307\\
+\institute{Mehedi Hasan \at
+              \email{mehedi@wayne.edu} 
+           \and
+           April Idalski Carcone \at
+              \email{acarcone@med.wayne.edu} 
+           \and
+           Sylvie Naar \at
+              \email{sylvie.naar@med.fsu.edu}
+           \and
+           Susan Eggly \at
+              \email{egglys@karmanos.org} 
+           \and
+           Gwen L. Alexander \at
+              \email{GALEXAN2@hfhs.org}
+           \and
+           Kathryn E Brogan Hartlieb \at
+              \email{hartlieb@fiu.edu} 
+           \and
+           Alexander Kotov (corresponding author) \at
+              Tel.: +1(313) 577-9307 \\
               \email{kotov@wayne.edu} 
@@
 \keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
-%\PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
@@
 \label{sec:intro}
-Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \citep{miller2013motivational}.
-The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \citep{miller2013motivational}.
+Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \cite{miller2013motivational}.
+The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \cite{miller2013motivational}.
 The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change.
-MI’s technical hypothesis \citep{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”.
+MI’s technical hypothesis \cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”.
 Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
-Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4].
+Previous studies \cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention \cite{walker2011influence}.
 In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
-Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5].
+Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk \cite{magill2014technical}.
 These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.
@@
 
-% For tables use
 \begin{table}
-% table caption is above the table
 \caption{MYSCOPE codebook.}
-\label{tab:codebook}       % Give a unique label
-% For LaTeX tables use
-\begin{tabular}{lll}
+\label{tab:codebook}   
+\begin{tabularx}{\textwidth}{lll}
 \hline\noalign{\smallskip}
@@
 \noalign{\smallskip}\hline
-\end{tabular}
+\end{tabularx}
 \end{table}
@@
 
-% For two-column wide figures use
 \begin{figure*}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
 \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
-% figure caption is below the figure
 \caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states.}
-\label{fig:bic}       % Give a unique label
+\label{fig:bic}
 \end{figure*}
@@
 
-% For two-column wide figures use
 \begin{figure*}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
 \includegraphics[width=0.95\textwidth]{figures/fpm.eps}
-% figure caption is below the figure
 \caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
-\label{fig:fpm}       % Give a unique label
+\label{fig:fpm}
 \end{figure*}
@@
 
-%\newgeometry{margin=2cm} % modify this if you need even more space
-%\newgeometry{left=4cm}
 \begin{sidewaystable}
+\vspace{15\baselineskip}
+
 \caption{Hidden Markov Model Emission Matrices}
-\label{tab:emission}       % Give a unique label
-% For LaTeX tables use
-\begin{tabular}{lll}
+\label{tab:emission}
+\begin{tabularx}{\textwidth}{lllllllll}
+\hline\noalign{\smallskip}
+first & second & third & first & second & third & first & second & third  \\
+\noalign{\smallskip}\hline\noalign{\smallskip}
+number & number & number & number & number & number & number & number & number\\
+number & number & number & number & number & number & number & number & number\\
+\noalign{\smallskip}\hline
+\end{tabularx}
+
+\vspace{2\baselineskip}
+
+\caption{Hidden Markov Model Transition Matrices}
+\label{tab:transition}  
+\begin{tabularx}{\textwidth}{lll}
 \hline\noalign{\smallskip}
@@
 \noalign{\smallskip}\hline
-\end{tabular}
-
-\vspace{2\baselineskip}
-
-\caption{Hidden Markov Model Transition Matrices}
-\label{tab:transition}       % Give a unique label
-% For LaTeX tables use
-\begin{tabular}{lll}
-\hline\noalign{\smallskip}
-first & second & third  \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number \\
-number & number & number \\
-\noalign{\smallskip}\hline
-\end{tabular}
+\end{tabularx}
 \end{sidewaystable}
-
-%\restoregeometry
 
@@
 
-% For tables use
 \begin{table}
-% table caption is above the table
+\begin{threeparttable}
 \caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences.}
-\label{tab:patterns}       % Give a unique label
-% For LaTeX tables use
-\begin{tabular}{lll}
+\label{tab:patterns}   
+\begin{tabularx}{\textwidth}{lll}
 \hline\noalign{\smallskip}
@@
 \noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number \\
-number & number & number \\
+number & number & number number number number number number \\
+number number number number & number number number & number number  number \\
 \noalign{\smallskip}\hline
-\end{tabular}
+\end{tabularx}
+\begin{tablenotes}
+      \small
+      \item This is where authors provide additional information about the data, including whatever notes are needed.
+\end{tablenotes}
+\end{threeparttable}
 \end{table}
@@
 
-%Text with citations \citep{wei2006semi} and \citep{keogh2000scaling}.
-
-%\paragraph{Paragraph headings} Use paragraph headings as needed.
-%\begin{equation}
-%a^2+b^2=c^2
-%\end{equation}
-
 \begin{acknowledgements}
@@
 
-% BibTeX users please use one of
-%\bibliographystyle{spbasic}      % basic style, author-year citations
-%\bibliographystyle{spmpsci}      % mathematics and physical sciences
 \bibliographystyle{natbib}       % JHIR format
-%\bibliographystyle{spphys}       % APS-like style for physics
 \bibliography{references}   % name your BibTeX data base
@@
 \end{document}
-% end of file template.tex

### 2018-04-11 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 
-\date{Received: 12/03/2018 Revised: 04/15/2018 Accepted: }
+\date{Received: 3 December 2018 /Revised: 15 April 2018 /Accepted: }
 % The correct dates will be entered by the editor
@@
 Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change.
-Although there is strong empirical evidence linking “MI-consistent” counselor behaviors and patient motivational statements (i.e., “change talk”), the specific counselor communication behaviors effective for eliciting patient change talk are a subject of ongoing research.
+Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
 A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication.
 Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks.
-In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, and closed frequent pattern mining, an algorithm used to mine the closed frequent patterns in MI communication sequences to inform MI practice.
+In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
 We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers.
-Transcripts had previously been annotated with patient-counselor behavior codes using a specialized codebook.
-Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify meaningful and interpretable counselor communication strategies to guide clinical practice.
+Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook.
+Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients’ motivational statements to guide clinical practice.
 
-\keywords{sequential analysis \and hidden markov model \and closed frequent pattern mining \and motivational interviewing \and weight loss \and adolescents}
+\keywords{hidden markov model \and closed frequent pattern mining \and motivational interviewing \and pediatric obesity}
 \end{abstract}
@@
 \label{sec:intro}
-Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change \cite{miller2013motivational}.
-The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills \cite{miller2013motivational}.
-The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the “spirit of MI”, an optimal therapeutic state to explore behavior change.
-MI’s technical hypothesis \cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient “change talk”.
-Change talk are patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
-Previous studies \cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention \cite{walker2011influence}.
+Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change [1].
+The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills [1].
+The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
+MI’s technical hypothesis [2] states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
+Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
+Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4].
 In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
-Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk \cite{magill2014technical}.
+Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5].
 These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.
@@
 
-Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and primarily for assessing intervention fidelity.
+Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and assessing intervention fidelity.
 Perez-Rosas et al.
 [25] developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
-This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, which achieved nearly 90\% accuracy for predicting counselor questions and reflections.
-Patient’s engagement, verbal and non-verbal accommodation as well as topics discussed during the MI session, were analyzed to identify linguistic and acoustic features of counselor empathy.
-These features were then used to build a counselor empathy classifier that achieved an accuracy of up to 80\%.
+This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, to predict counselor questions and reflections, which achieved nearly 90\% accuracy.
+The fidelity is assessed by computing a ratio of questions to reflections, since in MI sessions the goal is to have this ration at least 1:1.
 In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes [23].
@@
 HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence.
-The HMM associates each observation in a sequence with a “hidden” state, where the hidden state corresponds to a different distribution over observations (i.e., distribution over behavior codes).
-Sequences of observations are modeled as transitions between different hidden states and sampling observations from each hidden state.
+The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden'' corresponds to a different distribution.
+Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state.
 HMMs were originally proposed for speech recognition [30], in which the states were used to represent all English language sounds.
@@
 Bertholet et al.
-[47] used HMM and adjusted regression models to examine the association of different sequences of “change talk” utterances with drinking outcomes in a brief motivational intervention.
+[47] used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention.
 Their HMM model limited hidden states to three, which are "towards change", "away from change" and "non-determined".
@@
 Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors.
-Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communication during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.
+Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.
 
-\section{Methods}
-\label{sec:methods}
+\section{Materials and Methods}
+\label{sec:materialsAndMethods}
 \subsection{Dataset}
 \label{subsec:dataset}
-The data set for this study consists of 37 transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss treatment.
-Each transcript represented a one-on-one interaction between the adolescent patient and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
+The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
+Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
 These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE) [13].
@@
 The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI) [49], the MI Skill Code (MISC) [50], and Amrhein’s conceptualization of change talk and commitment [51].
-A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% (n = 7) of the sessions for inter-rater reliability which was good (k = .696).
+A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
 The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
@@
 For each MI transcript, the stream of behavior codes from beginning to end was analyzed.
-Successful sequences were those that resulted in a patient change talk or commitment language statement and all preceding behavior codes were included.
+Successful sequences are defined as those that resulted in a patient change talk or commitment language statement.
 Unsuccessful sequences were similarly created for sequences resulting in sustain talk.
-A total of 1,360 sequences were generated.
+A total of 1,360 sequences were generated using this approach.
 The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high.
@@
 
-\subsection{Data Analysis}
-\label{subsec:dataanalysis}
-The Hidden Markov Model (HMM; implemented in the hmmlearn package which is publicly available at http://hmmlearn.readthedocs.io/en/latest/index.html) analysis aimed to identify clusters of related behavior codes within successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
+\subsection{Data Modeling}
+\label{subsec:dataModeling}
+\subsubsection{Hidden Markov Model}
+\label{subsubsec:hiddenMarkovModel}
+We applied Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
 Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence.
-The emission probability matrix describes the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions among the hidden states.
+The rows in the emission probability matrix correspond to the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions between the hidden states.
 
-Training an HMM with a given number of hidden states (N) involves estimating the following parameters using the Baum-Welch algorithm:
-M is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table 1).
-T is an  state transition probability matrix where  
-E is an  matrix where , with observation symbol probability distribution for each state (emission probability matrix), and
-p is the initial state distribution vector where .
+Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:
 
-In this study, two HMM models were trained, one using successful sequences and the other using unsuccessful sequences, with the objective of maximizing log-likelihood on the observed sequence.
-The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest BIC value, which indicates an optimal fit of HMM model.
-BIC minimizes negative log-likelihood but penalizes complexity (i.e.
-\# of hidden states), which is important for avoiding overfitting problem.
-HMM analyses suggested 5 hidden states was optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).
+\begin{itemize}
+\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
+\item $T$ is an state transition probability matrix, in which is the probability of HMM transitioning from state  to state  
+\item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
+\item $p$ is the initial state distribution vector where  is the probability of the first HMM state
+\end{itemize}
+
+We trained two HMM models, one using all successful sequences and the other one using all unsuccessful sequences.
+Each model was trained with the objective of maximizing the log-likelihood of all observations in the corresponding set of sequences.
+The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest value of BIC, which takes into account both log-likelihood and a penalty term for the number of parameters in the model to avoid overfitting.
+Experiments with different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).
 
@@
 
-A second experiment was conducted with Frequent Pattern Mining.
-Orange3 and arules are two available pattern mining libraries implemented in python and r-programming, which support only a few pattern mining algorithms including frequent itemsets and association rules mining.
-However, we conducted our experiment using SPMF [52, 53], which offered more than 150 data mining algorithms.
-Moreover, SPMF is an open-source, fast and lightweight, a most popular and widely used Java-based data mining library.
-A frequent itemset is determined by surpassing a user-specified threshold called minimum support count\footnote{Support count: \# of times an itemset appears in the database}.
-For example, \{A\}, \{C\}, {D}, and {C, D} are frequent itemset in the database illustrated in Figure~\ref{fig:fpm}, because these itemsets appear at least 2 times which was the minimum support count identified in the example database.
-We reported closed frequent itemsets among all sequences of behavior codes in successful and unsuccessful sequences.
-A frequent itemset is closed if none of its supersets have the same support count [45], where a set X is a superset of another set Y if X contains all elements of the set Y.
-For example, the itemsets {A}, {C}, {D}, and {C, D} are not closed frequent itemsets because their supersets \{A, B\}, {B, C}, {B, D}, and {B, C, D} have the same number of support.
-Therefore, {B}, {A, B}, {A, D}, {B, C}, {B, D}, and {B, C, D} are closed frequent itemsets because none of their supersets have same the number of support count.
-On the other hand, {A, C}, {A, B, C}, {A, B, D}, {A, C, D} and {A, B, C, D} have support count 1, 0, 1, 0, and 1, respectively, which are less than minimum support count and identified as non-frequent itemsets.
-This study applied FPClose [46], an efficient close frequent pattern mining algorithm in SPMF, to identify frequent patient-counselor behaviors in MI communication sequences.
-To improve the model’s ability to identify frequent patient behaviors, FPClose [46], a FP-tree based algorithm [54, 55] was used.
-FPClose identifies closed frequent itemsets by utilizing an FP-tree data structure in combination with the FP-array technique to avoid the frequent traversal of FP-trees.
-Before applying the FPClose method, the dataset was transformed into the expected format used by the algorithm in the SPMF library.
-Minimum support threshold is a certain proportion having a meaningful representation of patterns determined by the domain expert.
-Since there is no standard for the selection of minimum support threshold, by following the studies [56-57], the minimum support count considered was 10\% of the total number of sequences which is 110 for successful and 25 for unsuccessful sequences.
-In our study, support counts representing ≥10\% of the patterns was the threshold for the result interpretation.
+\subsubsection{Frequent Pattern Mining}
+\label{subsubsec:frequentPatternMining}
+We applied Frequent Pattern Mining to identify frequently occurring patterns of temporally ordered patient-provider behavior codes in successful and unsuccessful communication sequences.
+Behavior codes in these patterns may be separated by one or more other codes.
+For this purpose, we utilized FPClose [46], an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF [52, 53], to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
+SPMF is an open-source library providing more than 150 data mining algorithms.
+Popular non-close frequent pattern mining algorithms include Apriori [35] and FP-Growth [54].
+A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
+For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure 2, since these patterns appear at least 2 times, which is the minimum support count in this example.
+In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences.
+A \textit{frequent pattern is closed} if none of its supersets have the same support count [45], where a set X is a superset of another set Y, if X contains all the elements of the set Y.
+For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure 2 are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count.
+Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\}, and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count.
+On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0, and 1, respectively, which is less than the minimum support count, and thus are identified as non-frequent itemsets.
+Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we follow the procedure in [55-56] and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
 
 \begin{figure*}
-\includegraphics[width=0.95\textwidth]{figures/fpm.eps}
+\includegraphics[width=1.0\textwidth]{figures/fpm.eps}
 \caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
@@
 \end{figure*}
-
-\section{Results}
-\label{sec:results}
-The transition and emission probability matrices of HMM model, are reported in Tables~\ref{tab:transition} and~\ref{tab:emission}.
-The three behaviors with the highest frequency, representing 45-69\% of each state’s emission transitions, were used to interpret the emission matrix and label the hidden states.
-Most (84\%) successful sequences began in a state characterized as “high motivation” as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
-Successful sequences began in a state of “high receptivity” 11\% of the time.
-“High receptivity” sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%).
-Few successful sequences began in states of “moderate receptivity” and “low receptivity” (2\% and 3\% of the time, respectively).
-These two states were characterized by different proportions of the same behaviors.
-Moderate receptivity sequences were distinguished from low receptivity sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%).
-No (0\%) successful sequence began in the “active feedback” state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%).
-Successful sequences transitioned from “high motivation” to “active feedback” most often (41\%).
-“active feedback”, in turn, most frequently transitioned to “high receptivity” (39\%).
-The “moderate receptivity” state transitioned to most often to “active feedback” (27\%) and automatically transitioned back to “moderate receptivity” (24\%) or to “low receptivity” (23\%) with similar frequency.
-The full transition matrix is presented in Table~\ref{tab:transition}.
-
-In contrast, the majority of unsuccessful sequences (98\%) began in a state of “ambivalence” as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
-About 2\% of the time unsuccessful sequences started in a state of “avoidance”.
-Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished “avoidant” sequences.
-Both “ambivalent” (84\%) and “avoidant” (61\%) states most frequently transitioned to the “avoidant” state.
 
@@
 
+\section{Results}
+\label{sec:results}
+The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}.
+The three behaviors with the highest frequency, representing 45-69\% of each state’s emission probability mass, were used to interpret the emission matrix and label the hidden states.
+Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
+Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time.
+\textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%).
+Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively).
+These two states were characterized by different proportions of the same behaviors.
+\textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%).
+No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%).
+Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%).
+\textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%).
+The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
+The full transition matrix is presented in Table~\ref{tab:transition}.
+
+In contrast, the majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
+About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}.
+Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences.
+Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.
+
 Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}.
 Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
-For each pattern, the significance of difference between successful and unsuccessful communications was reported by computing p-value using Pearson chi-square test.
+For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson’s chi-square test.
 Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient’s decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
@@
 \label{sec:discussion}
-We present the application of HMM and frequent pattern mining to test the technical hypothesis guiding Motivational Interviewing which suggests that counselors use of “MI-consistent” communication strategies (MICO) will lead to patient change talk [2].
+In summary, this study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk [2].
 Previous studies have empirically linked counselors’ use of MICO to higher rates of patient change talk in first-order Markov Chain models [8, 9, 11].
-Evidence from the current study using machine-learning methods of pattern analysis provides stronger evidence for MI’s technical hypothesis by considering longer-term dependencies in the data.
-Both HMM and frequent pattern mining consider behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement offered by simple first-order Markov chain models.
-The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing which has previously relied upon simple Markov chain models [8-14].
+The present study leveraged data mining methods to provide an even stronger evidence for MI’s fundamental hypothesis by considering longer-term dependencies in the data.
+Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM is able to identify relations between groups of communication behaviors.
+The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models [8-14].
 
-The states identified in HMM model, were given labels to describe the underlying “state” of the MI session.
-These labels were based on an analysis and interpretation of the three behaviors with the highest frequency according to the emission matrix.
-Three behaviors were interpreted because together they represented 45-69\% of the behaviors associated with a given state and all represented at least 10\%.
-These labels made it easier to understand types of interactions occurred between counselors and patients.
+The states identified in HMM model, were given labels to describe the underlying ``state'' of the MI session.
+These labels were based on the analysis and interpretation of the three behaviors with the highest frequency according to the emission matrix.
+Three behaviors were chosen since together they represented 45-69\% of the behaviors associated with a given state.
+These labels make it easier for MI practitioners to gain a deeper understanding of the dynamics of interactions occurring between counselors and patients.
 
 In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement).
-In HMM, the majority successful sequences began in the high motivation state where counselors frequently used reflections of change talk or commitment language and affirmations.
+In HMM, the majority of successful sequences began in the high motivation state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
 Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients’ decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies.
@@
 Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient’s decision-making autonomy, and sensitively provided information.
-Previous MI sequential analysis studies using first-order Markov chain models to analyze communication sequences have linked patients’ expression of change talk to counselor reflections of change talk, [9, 11-14] open questions phrased to elicit change talk, [9, 13, 14], and statements emphasizing the patient’s decision-making autonomy [13, 14].
-These studies did not find a link between change talk and counselors’ use of affirmations or the provision of information when examined independently of the MICO index.
-Thus, the current study is the first to provide empirical evidence for these causal linkages.
+Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients’ expression of change talk to counselor reflections of change talk, [9, 11-14] open questions phrased to elicit change talk, [9, 13, 14], and statements emphasizing the patient’s decision-making autonomy [13, 14].
+However, these studies did not find a link between change talk and counselors’ use of affirmations or the provision of information, when examined independently of the MICO index.
+Thus, the present study is the first to provide empirical evidence for these causal linkages.
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
-Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change [58].
+Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change [57].
 Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
@@
 Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed.
-These results are consistent with those of Gaume and colleagues [11] who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
-Specifically, counselors’ use of simple and complex reflections and “other MICO” behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
-Carcone et al [13] found counselors’ questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
+These results are consistent with those of Gaume et al.
+[11] who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
+Specifically, counselors’ use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
+Carcone et al.
+[13] found counselors’ questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
 In contrast, Moyers et al.
@@
 This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns.
-We have previously reported on the development of a model to annotate clinical transcripts [18, 21, 23].
-Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits).
+We have previously reported on the development of probabilistic generative models [18, 21] and application of novel features for maximum margin and deep learning classifiers [23] with the goal of automated annotation of MI session transcripts.
+Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits [58]).
+We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview [29].
 In this study, we present the next phase of this research, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
 Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
-We are also experimenting with machine learning methods to segment (parse) the stream of communication behavior into chunks representing a patient or counselor utterance to which the annotation model can then assign a behavior code.
+We are also experimenting with machine learning methods to segment (parse) the stream of communications in weight loss intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
 Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions.
-An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanism of Motivational Interviewing which informs theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies with discrete patient populations.
+An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.
 
@@
 Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently.
-One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to one of success cannot be determined from these analyses.
+One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined to be unsuccessful to successful one, cannot be determined from these analyses.
 
@@
 These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
-While counselors’ use of “MI-consistent” communication behaviors has been previously linked to higher rates of change talk in correlational studies [8, 15-17] and simple Markov chain models [8, 9, 11].
-The use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
+While counselors’ use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies [8, 15-17] and simple Markov chain models [8, 9, 11], the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
 The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.
@@
 \bibliographystyle{natbib}       % JHIR format
-\bibliography{references}   % name your BibTeX data base
-
+\bibliography{references}   
 \end{document}

### 2018-04-11 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 \usepackage{graphicx}
-\usepackage[numbers]{natbib}
+\usepackage[numbers,sort&compress]{natbib}
 \usepackage{geometry}
@@
 \label{sec:intro}
-Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change [1].
-The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills [1].
+Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}.
+The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}.
 The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
-MI’s technical hypothesis [2] states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
+MI’s technical hypothesis~\cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
 Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
-Previous studies [3] have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention [4].
+Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}.
 In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
-Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk [5].
+Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}.
 These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.
 
-Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations [6, 7].
-Moyers and Martin [8] applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
-A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk [9].
+Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}.
+Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
+A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
 Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
-Gaume et al.
-[10] used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
+Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
 They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
-A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk [11].
+A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}.
 In this sample, MIIN led to greater sustain talk, but was unrelated to change talk.
 Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk.
-Glynn and colleagues [12] linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
-In a study of adolescents engaged in weight loss treatment, Carcone et al.
-[13] used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents’ caregivers [14] drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
+Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
+In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
+A parallel study of the adolescents’ caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
 Across these studies, counselors’ use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.
 
-The sequential analysis procedure used in the above MI process studies [8, 15-17] is based on the first-order Markov Chain model [8, 9, 11].
+The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
 The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event.
@@
 
-Several prior works have reported the results of adopting machine learning methods, such as topic models [18-22], classification methods [23-26] and neural networks [23, 27, 28] to the tasks of annotating MI transcripts and assessing intervention fidelity.
-Perez-Rosas et al.
-[25] developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
+Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity.
+Perez-Rosas et al.~\cite{perez2017predicting} developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
 This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, to predict counselor questions and reflections, which achieved nearly 90\% accuracy.
 The fidelity is assessed by computing a ratio of questions to reflections, since in MI sessions the goal is to have this ration at least 1:1.
-In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes [23].
+In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
 The SVM model with the proposed features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders.
-In a follow up study [29], we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
+In a follow up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
 RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews.
@@
 Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state.
-HMMs were originally proposed for speech recognition [30], in which the states were used to represent all English language sounds.
-In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling [31, 32].
-For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease [31], while HMM was used to capture important characteristics of protein families [32].
+HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds.
+In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}.
+For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}, while HMM was used to capture important characteristics of protein families~\cite{amrhein2003client}.
 In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter.
-Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change [33]) as the antecedents of patient’s behavior change [3], there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
+Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patient’s behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
 Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.
 
-Frequent pattern mining [34] is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together.
-Agrawal and Srikant [35] first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
-Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics [36-39], medical imaging [37], chemical and biological analysis [40-42], web mining [43], and outlier analysis [44].
+Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together.
+Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
+Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web}, and outlier analysis~\cite{aggarwal2015outlier}.
 No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
 A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
-To address this problem, we utilized the closed frequent itemset mining method [45], which produces less number of patterns in more compact form that are easier to interpret.
-In this study, we leveraged FPClose [46], an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
+To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces less number of patterns in more compact form that are easier to interpret.
+In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
 FPClose is the best closed frequent itemset mining algorithm, which shown highest performance in terms of running time and memory consumption.
@@
 This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions.
-Bertholet et al.
-[47] used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention.
-Their HMM model limited hidden states to three, which are "towards change", "away from change" and "non-determined".
+Bertholet et al.~\cite{bertholet2010change} used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention.
+Their HMM model limited hidden states to three, which are ``towards change'', ``away from change'' and ``non-determined''.
 First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention.
@@
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE) [13].
-The MYSCOPE is an adaptation of the original MI-SCOPE [48], a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
-The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI) [49], the MI Skill Code (MISC) [50], and Amrhein’s conceptualization of change talk and commitment [51].
+These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}.
+The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein’s conceptualization of change talk and commitment~\cite{amrhein2004does}.
 A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
@@
 \begin{figure*}
+\centering
 \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
@@
 Behavior codes in these patterns may be separated by one or more other codes.
-For this purpose, we utilized FPClose [46], an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF [52, 53], to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
+For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
 SPMF is an open-source library providing more than 150 data mining algorithms.
-Popular non-close frequent pattern mining algorithms include Apriori [35] and FP-Growth [54].
+Popular non-close frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
 A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
-For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure 2, since these patterns appear at least 2 times, which is the minimum support count in this example.
+For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.
 In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences.
-A \textit{frequent pattern is closed} if none of its supersets have the same support count [45], where a set X is a superset of another set Y, if X contains all the elements of the set Y.
-For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure 2 are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count.
+A \textit{frequent pattern is closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y.
+For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count.
 Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\}, and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count.
 On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0, and 1, respectively, which is less than the minimum support count, and thus are identified as non-frequent itemsets.
-Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we follow the procedure in [55-56] and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
+Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we follow the procedure in~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
 
 \begin{figure*}
-\includegraphics[width=1.0\textwidth]{figures/fpm.eps}
-\caption{Different type of itemsets is illustrated with a frequent pattern mining method applied on transaction database by considering minimum support count 2, where TID represents the transaction ID of the corresponding items.}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/fpm.eps}
+\caption{A sample collection of sequences and different types of frequent patterns obtained by a frequent pattern mining method with the minimum support of 2.}
 \label{fig:fpm}
@@
 \label{sec:discussion}
-In summary, this study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk [2].
-Previous studies have empirically linked counselors’ use of MICO to higher rates of patient change talk in first-order Markov Chain models [8, 9, 11].
+In summary, this study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
+Previous studies have empirically linked counselors’ use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
 The present study leveraged data mining methods to provide an even stronger evidence for MI’s fundamental hypothesis by considering longer-term dependencies in the data.
 Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM is able to identify relations between groups of communication behaviors.
-The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models [8-14].
+The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
 
@@
 Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient’s decision-making autonomy, and sensitively provided information.
-Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients’ expression of change talk to counselor reflections of change talk, [9, 11-14] open questions phrased to elicit change talk, [9, 13, 14], and statements emphasizing the patient’s decision-making autonomy [13, 14].
+Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients’ expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient’s decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
 However, these studies did not find a link between change talk and counselors’ use of affirmations or the provision of information, when examined independently of the MICO index.
@@
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
-Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change [57].
+Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
 Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
@@
 Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed.
-These results are consistent with those of Gaume et al.
-[11] who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
+These results are consistent with those of Gaume et al.~\cite{gaume2010counselor} who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
 Specifically, counselors’ use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
-Carcone et al.
-[13] found counselors’ questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
-In contrast, Moyers et al.
-[9] found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
+Carcone et al.~\cite{carcone2013provider} found counselors’ questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
+In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
 These variable findings provide further support for the need to tailor the communication strategies MI counselors use to the treatment context.
@@
 This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns.
-We have previously reported on the development of probabilistic generative models [18, 21] and application of novel features for maximum margin and deep learning classifiers [23] with the goal of automated annotation of MI session transcripts.
-Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits [58]).
-We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview [29].
+We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts.
+Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}).
+We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
 In this study, we present the next phase of this research, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
@@
 These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
-While counselors’ use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies [8, 15-17] and simple Markov chain models [8, 9, 11], the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
+While counselors’ use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
 The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.

### 2018-04-13 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 \usepackage{tabularx}
+\usepackage{longtable,tabu}
 \usepackage[flushleft]{threeparttable}
+\usepackage{caption}
+\captionsetup{justification=raggedright,singlelinecheck=off,labelfont=bf}
+\usepackage{color, colortbl}
+\definecolor{lightgraycol}{gray}{0.95}
 
@@
 \label{subsec:dataset}
+
+\begin{small}
+\begin{longtable}{p{1.3cm}p{2cm}p{5cm}p{4.8cm}}
+\caption{MYSCOPE codebook.} 
+\label{tab:codebook} \\
+\hline\noalign{\smallskip}
+Annotation & Behavior & Description & Example \\ 
+\noalign{\smallskip}\hline\noalign{\smallskip}
+\endfirsthead
+\multicolumn{4}{@{}l}{\textbf{Table\ref{tab:codebook}} (continued)} \\ 
+\noalign{\smallskip}\hline\noalign{\smallskip}
+Annotation & Behavior & Description & Example \\
+\noalign{\smallskip}\hline\noalign{\smallskip}
+\endhead
+
+\rowcolor{lightgraycol}
+\multicolumn{4}{l}{Counselor} \\
+AF & Affirmation & Positive or complimentary statements that express appreciation, confidence, or reinforce the patient’s strengths or efforts.
+& ``You guys, as a family, are already doing a lot of really positive things.'' \\
+AR & Action reflection & Statements that reflect back the patient’s statement(s) while at the same time embedding a solution to a barrier or an action plan.
+& ``If you decide to follow a meal plan, it has to include occasional dessert.'' \\
+EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient’s freedom of choice, autonomy, personal responsibility, and so forth.
+& ``Okay.
+Well, it’s your plan, so whatever works best for you.
+If you feel like you want one that’s written down that you can refer back to, then let’s write it and if not then that’s fine.''\\
+GINFON & General information negative & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, or offers educational information in a non-patient-centered manner.
+& ``Healthy weight loss is about one to two pounds a week and once we get you set up and actually into the program you can look for that to happen for about one to two pounds a week to get you on that goal.''\\
+GINFOP & General information positive & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, expresses a concern, or offers educational information in a patient-centered manner (i.e., asking permission, using the third person, giving the opportunity to reject the information, and offering a menu of options).
+& ``Okay.
+Alright so I just wanted to tell you that I will be asking you a lot of questions.
+It may get redundant.
+So, if at any point in time you need a break or I’m asking too much go ahead and let me know.''\\
+QEB & Question to elicit barriers & Questions designed to initiate a discussion of barriers to change. & ``Alright.  So, are these ideas you feel you can put in place for this week?''\\
+QECHTP & Question to elicit change talk positive & Questions that ask about the patient’s desire, ability, reasons, or need for change or that reference past action toward behavior change or barriers to change.
+& ``Okay.
+And tell me a little bit more about that.
+Like what do you foresee your goal in this program?
+Like what do you want to happen out of this program?''\\
+QECMLP & Question to elicit commitment language positive & Questions that ask about current or future action toward behavior change or reference barriers to change.
+& ``Okay.
+Is there something else that you could do eat maybe instead of a Pop-Tart that’s a little bit healthier?''\\
+QEF & Question to elicit feedback & Statements that solicit the patient’s thoughts, ideas, or feelings about a specific recommendation or piece of information.
+& ``So, do you have any questions about that?''\\
+QEST & Question to elicit sustain talk & Questions designed to elicit negative change talk or negative commitment language.
+& ``And about how many hours would you say you watched TV for today?
+Or played video games or YouTube?''\\
+QO & Question other & Open- or close-ended questions unrelated to the target behavior.
+& ``Yup.
+What do you think might get in your way of being able to provide that kind of support for [your daughter]?''\\
+RCHTP & Reflect change talk positive & A reflective listening statement that captures and returns a patient’s statement or behavior from the current or a previous session that describes the patient’s desire, ability, reasons, or need for change or past action or barriers to change.
+& ``So, it sounds like you just want to be healthy and you want to be stylish.
+You want to fit into some different types of clothes.''\\
+RCMLP & Reflect commitment language positive & A reflective listening statement that captures and returns a patient’s statement or behavior from the current or a previous session that describes current or future action or references barriers to changing with the goal of problem-solving.
+& ``You are ready to start this plan today.''\\
+RO & Reflect other & A reflective listening statement that captures and returns a patient’s utterance or behavior from the current or previous session that is unrelated to the target behavior.
+& ``You are having a hard time at work.''\\
+RST & Reflect sustain talk & These statements reflect negative change talk or negative commitment language made by the patient. & ``Oh okay.  So, money influences your environment.''\\
+SO & Statement other & An utterance eliciting feedback, offering support, self-disclosure, or of some other form besides a strategy or reflection & ``You’re being pulled in a million directions''\\
+SPT & Support & These are generally supportive, understanding comments.
+They have the quality of commenting on a situation, or of agreeing or siding with the patient in a genuine way.
+& ``I'm concerned about you, given all these difficulties you've been having.''\\
+SS & Structure session & A communication strategy that suggests an attempt to describe what will happen in the session or to refocus a meandering conversation back to the target behaviors & ``Maybe when the three of us come together in a few minutes, that’s something that we could just clarify with her, like is that really what she wants.''\\
+SUM & Summary & A reflective listening statement that captures and returns at least 2 different ideas from a patient’s utterance or behavior from the current session & ``You have thought a lot about this.
+Sometimes it feels like losing weight is just too hard.
+Yet you have lots of reasons to lose weight.
+If you could find a program you could stick to, a program that would not have too many changes at once, you would consider it.''\\
+
+\rowcolor{lightgraycol}
+\multicolumn{4}{l}{Patient} \\
+CT & Change Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
+ST & Sustain Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) to maintain the status quo or not change their behavior  & ``I didn’t get to the gym this week.''\\
+HUPO	 & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
+HUPW & High uptake weight & A turn that does develop the topic of the conversation.
+High Uptake statements include: weight related statements about actions of commitment, change talk, and ambivalence that occurred in the past, patient questions to the counselor, and session interruptions by persons who are not an active part of the treatment session.
+& ``Support is always good.
+You know that’s a key factor.
+Mm-hmm.''\\
+LUP & Low uptake  & an utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
+\hline
+\end{longtable}
+\end{small}
+
 The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
@@
 The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
-
-\begin{table}
-\caption{MYSCOPE codebook.}
-\label{tab:codebook}   
-\begin{tabularx}{\textwidth}{lll}
-\hline\noalign{\smallskip}
-first & second & third  \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number \\
-number & number & number \\
-\noalign{\smallskip}\hline
-\end{tabularx}
-\end{table}
 

### 2018-04-13 jhir/latex-manuscript/manuscript.tex
update TABLES

@@
 \rowcolor{lightgraycol}
-\multicolumn{4}{l}{Counselor} \\
+\noalign{\smallskip}
+\multicolumn{4}{l}{\textbf{Counselor}} \\
+\noalign{\smallskip}
 AF & Affirmation & Positive or complimentary statements that express appreciation, confidence, or reinforce the patient’s strengths or efforts.
@@
 \rowcolor{lightgraycol}
-\multicolumn{4}{l}{Patient} \\
+\noalign{\smallskip}
+\multicolumn{4}{l}{\textbf{Patient}} \\
+\noalign{\smallskip}
 CT & Change Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
@@
 \begin{sidewaystable}
-\vspace{15\baselineskip}
+\vspace{32\baselineskip}
 
@@
 \label{tab:emission}
-\begin{tabularx}{\textwidth}{lllllllll}
+\begin{tabularx}{\textwidth}{lXXXp{0.4cm}XXXXXXXXXXXXXXXXXXX}
 \hline\noalign{\smallskip}
-first & second & third & first & second & third & first & second & third  \\
+State & AF & AR & EA & GIN- FON & GIN- FOP & QEB & QEC- HTP & QEC- MLP & QEF & QE- ST & QO & RC- HTP & RC- MLP & RO & RST & SO & SPT & SS & SUM & HU- PO & HU- PW & LUP \\
 \noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number & number & number & number & number & number & number\\
-number & number & number & number & number & number & number & number & number\\
+\rowcolor{lightgraycol}
+\multicolumn{23}{l}{\textbf{Successful  Sequences}} \\
+
+High Motivation & 12\% & 5\% & 4\% & 0\% & 3\% & 0\% & 0\% & 1\% & 0\% & 0\% & 1\% & 29\% & 17\% & 1\% & 7\% & 2\% & 5\% & 6\% & 6\% & 0\% & 0\% & 0\% \\
+
+High Receptivity & 15\% & 2\% & 5\% & 0\% & 18\% & 1\% & 16\% & 5\% & 3\% & 1\% & 0\% & 1\% & 0\% & 6\% & 5\% & 3\% & 3\% & 7\% & 4\% & 4\% & 1\% & 1\% \\
+
+Moderate Receptivity & 2\% & 1\% & 14\% & 0\% & 8\% & 1\% & 20\% & 9\% & 5\% & 1\% & 2\% & 2\% & 0\% & 1\% & 0\% & 1\% & 0\% & 5\% & 4\% & 5\% & 2\% & 16\% \\
+
+Low Receptivity & 4\% & 1\% & 17\% & 0\% & 3\% & 1\% & 11\% & 0\% & 2\% & 0\% & 0\% & 5\% & 3\% & 3\% & 3\% & 0\% & 1\% & 7\% & 1\% & 6\% & 0\% & 28\% \\
+
+Active Feedback & 1\% & 1\% & 3\% & 0\% & 7\% & 1\% & 0\% & 9\% & 3\% & 0\% & 1\% & 0\% & 0\% & 0\% & 3\% & 1\% & 0\% & 1\% & 1\% & 10\% & 12\% & 47\% \\
+
+\rowcolor{lightgraycol}
+\multicolumn{23}{l}{\textbf{Unsuccessful  Sequences}} \\
+
+Ambivalent & 13\% & 5\% & 4\% & 1\% & 6\% & 0\% & 1\% & 5\% & 0\% & 2\% & 2\% & 20\% & 7\% & 2\% & 12\% & 3\% & 4\% & 5\% & 6\% & 0\% & 2\% & 0\% \\
+
+Avoidant  & 3\% & 1\% & 6\% & 0\% & 11\% & 7\% & 8\% & 1\% & 3\% & 2\% & 2\% & 1\% & 1\% & 5\% & 4\% & 0\% & 1\% & 2\% & 0\% & 11\% & 4\% & 29\% \\
+
 \noalign{\smallskip}\hline
@@
 \label{tab:transition}  
-\begin{tabularx}{\textwidth}{lll}
+\begin{tabularx}{\textwidth}{XXXXXX}
 \hline\noalign{\smallskip}
-first & second & third  \\
+State & High Motivation & High Receptivity & Moderate Receptivity & Low Receptivity & Active Feedback \\
 \noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number \\
-number & number & number \\
+\rowcolor{lightgraycol}
+\multicolumn{6}{l}{\textbf{Successful  Sequences}} \\
+High Motivation & 2\% & 9\% & 24\% & 24\% & 41\% \\
+High Receptivity & 7\% & 9\% & 20\% & 25\% & 39\% \\
+Moderate Receptivity  & 8\% & 19\% & 24\% & 23\% & 27\% \\
+Low Receptivity  & 11\% & 26\% & 28\% & 18\% & 17\% \\
+Active Feedback  & 25\% & 34\% & 20\% & 15\% & 6\% \\
+\rowcolor{lightgraycol}
+\multicolumn{6}{l}{\textbf{Unsuccessful  Sequences}} \\
+& Ambivalent & Avoidant & & & \\
+Ambivalent & 16\% & 84\% & & & \\
+Avoidant & 39\% & 61\% & & & \\
 \noalign{\smallskip}\hline
@@
 \label{tab:patterns}   
-\begin{tabularx}{\textwidth}{lll}
+\begin{tabularx}{\textwidth}{p{3cm}XX@{\extracolsep{4pt}}p{3cm}XX}
 \hline\noalign{\smallskip}
-first & second & third  \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number number number number number number \\
-number number number number & number number number & number number  number \\
+\multicolumn{3}{c}{Successful} & \multicolumn{3}{c}{Unsuccessful} \\
+\noalign{\smallskip}\cline{1-3} \cline{4-6}\noalign{\smallskip}
+LUP & 573 & 52.0\% & LUP & 118 & 45.7\%  \\
+RCHTP & 398 & 36.1\% & RCHTP & 87 & 33.7\% \\ 
+\multicolumn{1}{r}{LUP, RCHTP} & 224 & 20.3\% & \multicolumn{1}{r}{LUP, RCHTP} & 45 & 17.4\% \\
+QECHTP & 339 & 30.8\% & GINFOP & 74 & 28.7\% \\
+\multicolumn{1}{r}{LUP, QECHTP} & 184 & 16.7\% & \multicolumn{1}{r}{LUP, GINFOP} & 44 & 17.1\% \\
+AF & 314 & 28.5\% & RST & 70 & 27.1\% \\
+\multicolumn{1}{r}{LUP, AF} & 166 & 15.1\% & \multicolumn{1}{r}{LUP, RST} & 30 & 11.6\% \\
+EA & 314 & 28.5\% & AF & 68 & 26.4\% \\
+\multicolumn{1}{r}{LUP, EA} & 188 & 17.1\% & EA & 48 & 18.6\% \\
+GINFOP & 244 & 22.1\% & \multicolumn{1}{r}{LUP, EA} & 28 & 10.9\% \\
+\multicolumn{1}{r}{LUP, GINFOP} & 143 & 13.0\% & QECHTP & 45 & 17.4\% \\
+RCMLP & 228 & 20.7\% & RCMLP & 39 & 15.1\% \\
+\multicolumn{1}{r}{LUP, RCMLP} & 121 & 11.0\% & QEB & 38 & 14.7\% \\
+QECMLP & 200 & 18.1\% & RO & 30 & 11.6\% \\
+RST & 174 & 15.8\% & QECMLP & 30 & 11.6\% \\
+\multicolumn{1}{r}{LUP, RST} & 114 & 10.3\% & SUM & 29 & 11.2\% \\
+RCHTP, QECHTP & 154 & 14.0\% & SS & 28 & 10.9\% \\
+SUM & 138 & 12.5\% & RCHTP, GINFOP & 27 & 10.5\% \\
+LUP, SS & 112 & 10.2\% & HUPO & 47 & 18.2\% \\
+HUPO & 173 & 15.7\% &  &  &  \\ 
 \noalign{\smallskip}\hline
@@
       \small
-      \item This is where authors provide additional information about the data, including whatever notes are needed.
+      \item \textit{Note:} Patterns that are aligned to the right are included in the immediately preceding pattern count.
+In these patterns, a counselor behavior was paired with a patient low uptake/facilitative comment, which is a marker of patient attention to the conversation and feedback suggesting the line of discussion may continue.
 \end{tablenotes}

### 2018-04-13 jhir/latex-manuscript/manuscript.tex
update TABLES
fixed some spelling errors

@@
 Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook.
-Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients’ motivational statements to guide clinical practice.
+Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.
 
@@
 Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}.
-The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}.
-The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
-MI’s technical hypothesis~\cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
+The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}.
+The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
+MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
 Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
@@
 Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}.
-These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.
+These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.
 
@@
 In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents’ caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
-Across these studies, counselors’ use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.
+A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
+Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.
 
@@
 For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}, while HMM was used to capture important characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter.
-Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patient’s behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
+In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient's underlying motivational state, during a patient-counselor encounter.
+Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patient's behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
 Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.
@@
 Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
-This knowledge will inform MI theory by providing additional evidence to support MI’s technical hypothesis.
+This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis.
 It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication.
@@
 \noalign{\smallskip}
-AF & Affirmation & Positive or complimentary statements that express appreciation, confidence, or reinforce the patient’s strengths or efforts.
+AF & Affirmation & Positive or complimentary statements that express appreciation, confidence, or reinforce the patient's strengths or efforts.
 & ``You guys, as a family, are already doing a lot of really positive things.'' \\
-AR & Action reflection & Statements that reflect back the patient’s statement(s) while at the same time embedding a solution to a barrier or an action plan.
+AR & Action reflection & Statements that reflect back the patient's statement(s) while at the same time embedding a solution to a barrier or an action plan.
 & ``If you decide to follow a meal plan, it has to include occasional dessert.'' \\
-EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient’s freedom of choice, autonomy, personal responsibility, and so forth.
+EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient's freedom of choice, autonomy, personal responsibility, and so forth.
 & ``Okay.
-Well, it’s your plan, so whatever works best for you.
-If you feel like you want one that’s written down that you can refer back to, then let’s write it and if not then that’s fine.''\\
+Well, it's your plan, so whatever works best for you.
+If you feel like you want one that's written down that you can refer back to, then let's write it and if not then that's fine.''\\
 GINFON & General information negative & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, or offers educational information in a non-patient-centered manner.
@@
 It may get redundant.
-So, if at any point in time you need a break or I’m asking too much go ahead and let me know.''\\
+So, if at any point in time you need a break or I'm asking too much go ahead and let me know.''\\
 QEB & Question to elicit barriers & Questions designed to initiate a discussion of barriers to change. & ``Alright.  So, are these ideas you feel you can put in place for this week?''\\
-QECHTP & Question to elicit change talk positive & Questions that ask about the patient’s desire, ability, reasons, or need for change or that reference past action toward behavior change or barriers to change.
+QECHTP & Question to elicit change talk positive & Questions that ask about the patient's desire, ability, reasons, or need for change or that reference past action toward behavior change or barriers to change.
 & ``Okay.
@@
 & ``Okay.
-Is there something else that you could do eat maybe instead of a Pop-Tart that’s a little bit healthier?''\\
-QEF & Question to elicit feedback & Statements that solicit the patient’s thoughts, ideas, or feelings about a specific recommendation or piece of information.
+Is there something else that you could do eat maybe instead of a Pop-Tart that's a little bit healthier?''\\
+QEF & Question to elicit feedback & Statements that solicit the patient's thoughts, ideas, or feelings about a specific recommendation or piece of information.
 & ``So, do you have any questions about that?''\\
@@
 What do you think might get in your way of being able to provide that kind of support for [your daughter]?''\\
-RCHTP & Reflect change talk positive & A reflective listening statement that captures and returns a patient’s statement or behavior from the current or a previous session that describes the patient’s desire, ability, reasons, or need for change or past action or barriers to change.
+RCHTP & Reflect change talk positive & A reflective listening statement that captures and returns a patient's statement or behavior from the current or a previous session that describes the patient's desire, ability, reasons, or need for change or past action or barriers to change.
 & ``So, it sounds like you just want to be healthy and you want to be stylish.
 You want to fit into some different types of clothes.''\\
-RCMLP & Reflect commitment language positive & A reflective listening statement that captures and returns a patient’s statement or behavior from the current or a previous session that describes current or future action or references barriers to changing with the goal of problem-solving.
+RCMLP & Reflect commitment language positive & A reflective listening statement that captures and returns a patient's statement or behavior from the current or a previous session that describes current or future action or references barriers to changing with the goal of problem-solving.
 & ``You are ready to start this plan today.''\\
-RO & Reflect other & A reflective listening statement that captures and returns a patient’s utterance or behavior from the current or previous session that is unrelated to the target behavior.
+RO & Reflect other & A reflective listening statement that captures and returns a patient's utterance or behavior from the current or previous session that is unrelated to the target behavior.
 & ``You are having a hard time at work.''\\
 RST & Reflect sustain talk & These statements reflect negative change talk or negative commitment language made by the patient. & ``Oh okay.  So, money influences your environment.''\\
-SO & Statement other & An utterance eliciting feedback, offering support, self-disclosure, or of some other form besides a strategy or reflection & ``You’re being pulled in a million directions''\\
+SO & Statement other & An utterance eliciting feedback, offering support, self-disclosure, or of some other form besides a strategy or reflection & ``You're being pulled in a million directions''\\
 SPT & Support & These are generally supportive, understanding comments.
@@
 & ``I'm concerned about you, given all these difficulties you've been having.''\\
-SS & Structure session & A communication strategy that suggests an attempt to describe what will happen in the session or to refocus a meandering conversation back to the target behaviors & ``Maybe when the three of us come together in a few minutes, that’s something that we could just clarify with her, like is that really what she wants.''\\
-SUM & Summary & A reflective listening statement that captures and returns at least 2 different ideas from a patient’s utterance or behavior from the current session & ``You have thought a lot about this.
+SS & Structure session & A communication strategy that suggests an attempt to describe what will happen in the session or to refocus a meandering conversation back to the target behaviors & ``Maybe when the three of us come together in a few minutes, that's something that we could just clarify with her, like is that really what she wants.''\\
+SUM & Summary & A reflective listening statement that captures and returns at least 2 different ideas from a patient's utterance or behavior from the current session & ``You have thought a lot about this.
 Sometimes it feels like losing weight is just too hard.
@@
 \noalign{\smallskip}
-CT & Change Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
-ST & Sustain Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) to maintain the status quo or not change their behavior  & ``I didn’t get to the gym this week.''\\
+CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
+ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
 HUPO	 & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
@@
 & ``Support is always good.
-You know that’s a key factor.
+You know that's a key factor.
 Mm-hmm.''\\
@@
 The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
-The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein’s conceptualization of change talk and commitment~\cite{amrhein2004does}.
+The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
 A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
@@
 \item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
-\item $T$ is an state transition probability matrix, in which is the probability of HMM transitioning from state to state  
+\item $T$ is a state transition probability matrix, in which is the probability of HMM transitioning from state to state  
 \item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
@@
 \begin{sidewaystable}
-\vspace{32\baselineskip}
+\vspace{43\baselineskip}
 
@@
 \hline\noalign{\smallskip}
-State & AF & AR & EA & GIN- FON & GIN- FOP & QEB & QEC- HTP & QEC- MLP & QEF & QE- ST & QO & RC- HTP & RC- MLP & RO & RST & SO & SPT & SS & SUM & HU- PO & HU- PW & LUP \\
+State & AF & AR & EA & GIN-FON & GIN-FOP & QEB & QEC-HTP & QEC-MLP & QEF & QE-ST & QO & RC-HTP & RC-MLP & RO & RST & SO & SPT & SS & SUM & HU-PO & HU-PW & LUP \\
 \noalign{\smallskip}\hline\noalign{\smallskip}
@@
 The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}.
-The three behaviors with the highest frequency, representing 45-69\% of each state’s emission probability mass, were used to interpret the emission matrix and label the hidden states.
+The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
 Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
@@
 These two states were characterized by different proportions of the same behaviors.
-\textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%).
+\textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%).
 No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%).
@@
 Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
-For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson’s chi-square test.
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient’s decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
+For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
 In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= <0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p=<0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p=<0.0001).
@@
 In summary, this study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
-Previous studies have empirically linked counselors’ use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
-The present study leveraged data mining methods to provide an even stronger evidence for MI’s fundamental hypothesis by considering longer-term dependencies in the data.
+Previous studies have empirically linked counselors' use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
+The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data.
 Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM is able to identify relations between groups of communication behaviors.
@@
 In HMM, the majority of successful sequences began in the high motivation state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
-Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients’ decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies.
+Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies.
 The frequent pattern mining results were similar.
-Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient’s decision-making autonomy, and sensitively provided information.
-Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients’ expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient’s decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
-However, these studies did not find a link between change talk and counselors’ use of affirmations or the provision of information, when examined independently of the MICO index.
+Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy, and sensitively provided information.
+Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
+However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index.
 Thus, the present study is the first to provide empirical evidence for these causal linkages.
@@
 These results are consistent with those of Gaume et al.~\cite{gaume2010counselor} who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
-Specifically, counselors’ use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
-Carcone et al.~\cite{carcone2013provider} found counselors’ questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
+Specifically, counselors' use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
+Carcone et al.~\cite{carcone2013provider} found counselors' questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
 In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
@@
 These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
-While counselors’ use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
+While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
 The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.

### 2018-04-14 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
-In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= <0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p=<0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p=<0.0001).
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
+In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001).
 In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences.
@@
 In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement).
-In HMM, the majority of successful sequences began in the high motivation state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
+In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
 Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies.
@@
 The analysis of unsuccessful sequences, i.e., those resulting in a patient sustain talk statement, was typified by a combination of MICO and MI-inconsistent communication strategies (MIIN).
-Specifically, the majority of unsuccessful sequences in the HMM analysis began in a state of Ambivalence which was characterized by large proportions of counselor reflections of both change talk and sustain talk.
+Specifically, the majority of unsuccessful sequences in the HMM analysis began in a state of \textit{``ambivalence''} which was characterized by large proportions of counselor reflections of both change talk and sustain talk.
 Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed.

### 2018-04-14 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 \item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
-\item $T$ is a state transition probability matrix, in which is the probability of HMM transitioning from state to state  
-\item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
-\item $p$ is the initial state distribution vector where is the probability of the first HMM state
+\item $T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
+\item $E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
+\item $p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state
 \end{itemize}
@@
 These two states were characterized by different proportions of the same behaviors.
-\textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%).
+\textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%).
 No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%).
@@
 Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
-In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001).
+In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p=0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001).
 In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences.

### 2018-04-14 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 
-\caption{Hidden Markov Model Emission Matrices}
+\caption{Hidden Markov Model Emission Matrices.}
 \label{tab:emission}
@@
 
-\caption{Hidden Markov Model Transition Matrices}
+\caption{Hidden Markov Model Transition Matrices.}
 \label{tab:transition}  
@@
 For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p=0.0423).
 In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p=0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001).

### 2018-04-16 jhir/latex-manuscript/manuscript.tex
finalized revised manuscript

--- 2018-04-14 jhir/latex-manuscript/manuscript.tex
@@
            $^1$Department of Computer Science, College of Engineering, Wayne State University, Detroit, MI 48202\\
-$^2$Division of Behavioral Health Sciences, Department of Family Medicine and Public Health Sciences, Wayne State University School of Medicine, Detroit, MI 48202\\
+$^2$Division of Behavioral Sciences, Department of Family Medicine and Public Health Sciences, Wayne State University School of Medicine, Detroit, MI 48202\\
 $^3$Director, Center for Translational Behavioral Research, Department of Behavioral Sciences and Social Medicine, Florida State University, FL 32306\\  
@@
 
-\date{Received: 3 December 2018 /Revised: 15 April 2018 /Accepted: }
+\date{Received: 3 December 2018 /Revised: 16 April 2018}
 % The correct dates will be entered by the editor
@@
 The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
-MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
+MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
 Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
@@
 Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk.
-Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
+Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
 In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
@@
 Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences.
-Therefore, there is a need for more powerful computational methods, which consider clusters of behavior codes and long-term dependencies between behaviors, to identify causal relationships.
-The goal of the current research is to test the applicability of data mining methods to identify effective patterns of patient-counselor communication.
+Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-term dependencies between behaviors, to identify causal relationships.
+The goal of the current research is to test the applicability of pattern data mining and machine-learning methods to identify effective patterns of patient-counselor communication.
 
 Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity.
-Perez-Rosas et al.~\cite{perez2017predicting} developed a system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
-This system employed a support vector machine (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features, to predict counselor questions and reflections, which achieved nearly 90\% accuracy.
-The fidelity is assessed by computing a ratio of questions to reflections, since in MI sessions the goal is to have this ration at least 1:1.
+Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
+Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features achieving nearly 90\% accuracy.
 In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
-The SVM model with the proposed features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders.
+The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders.
 In a follow up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
@@
 This paper represents the next phase in this line of research.
-We examine the efficacy of Hidden Markov Models (HMMs) and Frequent Pattern Mining in identifying the counselor communication strategies antecedent to patient change talk.
+We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
 
 HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence.
-The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden'' corresponds to a different distribution.
+The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution.
 Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state.
@@
 In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}.
-For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}, while HMM was used to capture important characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient's underlying motivational state, during a patient-counselor encounter.
-Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patient's behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
-Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.
+For example, a HMM-based classifier was applied Doppler ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
+In another study, HMM was used to capture important
+characteristics of protein families~\cite{amrhein2003client}.
+In the application of HMM to patient-counselor communication, hidden states, and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter.
+Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
+Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.
 
@@
 A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
-To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces less number of patterns in more compact form that are easier to interpret.
+To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in more compact form that are easier to interpret.
 In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
-FPClose is the best closed frequent itemset mining algorithm, which shown highest performance in terms of running time and memory consumption.
+FPClose is the best closed frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.
 
 In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
-Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
+Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
 This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis.
 It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication.
-This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions.
-Bertholet et al.~\cite{bertholet2010change} used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention.
-Their HMM model limited hidden states to three, which are ``towards change'', ``away from change'' and ``non-determined''.
-First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention.
+This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions.
+Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention.
+Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''.
+These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention.
 In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication.
-The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
+The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
 These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research.
@@
 \label{sec:materialsAndMethods}
+
 \subsection{Dataset}
 \label{subsec:dataset}
+The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
+Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
+These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}.
+The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
+A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
+The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
 
@@
 \begin{longtable}{p{1.3cm}p{2cm}p{5cm}p{4.8cm}}
-\caption{MYSCOPE codebook.} 
+\caption{MYSCOPE codebook} 
 \label{tab:codebook} \\
@@
 Mm-hmm.''\\
-LUP & Low uptake  & an utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
+LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
 \hline
@@
 \end{small}
-
-The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
-Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}.
-The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
-The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
-A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
-The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
 
@@
 For each MI transcript, the stream of behavior codes from beginning to end was analyzed.
-Successful sequences are defined as those that resulted in a patient change talk or commitment language statement.
+Successful sequences were defined as those that resulted in a patient change talk or commitment language statement.
 Unsuccessful sequences were similarly created for sequences resulting in sustain talk.
@@
 \label{subsubsec:hiddenMarkovModel}
-We applied Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
+We applied the Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
 Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence.
 The rows in the emission probability matrix correspond to the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions between the hidden states.
-
 Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:
@@
 The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest value of BIC, which takes into account both log-likelihood and a penalty term for the number of parameters in the model to avoid overfitting.
-Experiments with different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).
+Experiments with a different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).
 
@@
 \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
-\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states.}
+\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}
 \label{fig:bic}
@@
 \label{subsubsec:frequentPatternMining}
-We applied Frequent Pattern Mining to identify frequently occurring patterns of temporally ordered patient-provider behavior codes in successful and unsuccessful communication sequences.
+We applied frequent pattern mining to identify frequently occurring patterns of temporally ordered patient-counselor behavior codes in successful and unsuccessful communication sequences.
 Behavior codes in these patterns may be separated by one or more other codes.
@@
 On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0, and 1, respectively, which is less than the minimum support count, and thus are identified as non-frequent itemsets.
-Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we follow the procedure in~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
+Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed some prior works~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
+For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
 
@@
 \includegraphics[width=0.8\textwidth]{figures/fpm.eps}
-\caption{A sample collection of sequences and different types of frequent patterns obtained by a frequent pattern mining method with the minimum support of 2.}
+\caption{A sample collection of sequences and different types of frequent patterns obtained by a frequent pattern mining method with the minimum support of 2}
 \label{fig:fpm}
@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
-Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time.
+Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
+Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time.
 \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%).
@@
 Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%).
-\textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%).
+\textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%).
 The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
@@
 
-Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}.
-Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
-For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423).
-In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p=0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001).
-In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences.
-In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.
+Results from frequent pattern mining analysis is presented in Table~\ref{tab:patterns}.
+Reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011), and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
+In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.001), providing information (28.7\% versus 22.1\%, p=0.025) and other reflections (11.6\% versus 0\%, p= $<$0.001).
+In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in $>$10\% of the unsuccessful sequences.
+In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in $>$10\% of the successful sequences.
 
@@
 \begin{threeparttable}
-\caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences.}
+\caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences}
 \label{tab:patterns}   
@@
 \label{sec:discussion}
-In summary, this study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
+This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
 Previous studies have empirically linked counselors' use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
 The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data.
-Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM is able to identify relations between groups of communication behaviors.
+Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences.
 The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
-
-The states identified in HMM model, were given labels to describe the underlying ``state'' of the MI session.
-These labels were based on the analysis and interpretation of the three behaviors with the highest frequency according to the emission matrix.
-Three behaviors were chosen since together they represented 45-69\% of the behaviors associated with a given state.
-These labels make it easier for MI practitioners to gain a deeper understanding of the dynamics of interactions occurring between counselors and patients.
 
@@
 In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
-These variable findings provide further support for the need to tailor the communication strategies MI counselors use to the treatment context.
+These variable findings suggest a need to tailor the communication strategies MI counselors use to the treatment context.
 
@@
 Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
-We are also experimenting with machine learning methods to segment (parse) the stream of communications in weight loss intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
+We are also experimenting with machine learning methods to segment (parse) the stream of communications in healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
 Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions.
@@
 Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently.
-One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined to be unsuccessful to successful one, cannot be determined from these analyses.
+One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to success, cannot be determined from these analyses.
 

### 2018-04-17 jhir/latex-manuscript/manuscript.tex
finalized revised manuscript

@@
 In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
-The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders.
+The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders.
 In a follow up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).

### 2018-04-17 jhir/latex-manuscript/manuscript_highlight.tex
finalized revised manuscript

--- 2018-04-16 jhir/latex-manuscript/manuscript.tex
@@
 \usepackage{color, colortbl}
+\usepackage{soul}
 \definecolor{lightgraycol}{gray}{0.95}
@@
 
-\title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
+\title{\hl{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis}
 }
@@
 \begin{abstract}
-Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change.
+\hl{Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change.
 Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
@@
 Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook.
-Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.
+Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.}
 
-\keywords{hidden markov model \and closed frequent pattern mining \and motivational interviewing \and pediatric obesity}
+\keywords{hidden markov model \and closed frequent pattern mining \and motivational interviewing \and \hl{pediatric obesity}}
 \end{abstract}
@@
 \label{sec:intro}
-Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}.
-The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}.
-The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
-MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
+\hl{Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change}~\cite{miller2013motivational}.
+\hl{The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills}~\cite{miller2013motivational}.
+\hl{The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
+MI's technical hypothesis}~\cite{miller2009toward} \hl{states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
 Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
-Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}.
-In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
-Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}.
-These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.
+Previous studies~}\cite{apodaca2009mechanisms} \hl{have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention}~\cite{walker2011influence}.
+\hl{In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
+Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk}~\cite{magill2014technical}.
+\hl{These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.}
 
-Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}.
-Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
-A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
-Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
-Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
+\hl{Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations}~\cite{bakeman1997observing, bakeman2011sequential}.
+\hl{Moyers and Martin}~\cite{moyers2006therapist} \hl{applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
+A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk}~\cite{moyers2009session}.
+\hl{Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
+Gaume et al.}~\cite{gaume2008counsellor} \hl{used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
 They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
-A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}.
-In this sample, MIIN led to greater sustain talk, but was unrelated to change talk.
+A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk}~\cite{gaume2010counselor}.
+\hl{In this sample, MIIN led to greater sustain talk, but was unrelated to change talk.
 Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk.
-Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
-In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
-Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.
+Glynn and colleagues}~\cite{glynn2014change} \hl{linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
+In a study of adolescents engaged in weight loss treatment, Carcone et al.}~\cite{carcone2013provider} \hl{used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
+A parallel study of the adolescents' caregivers}~\cite{jacques2016building} \hl{drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
+Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.}
 
-The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
-The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event.
+\hl{The sequential analysis procedure used in the above MI process studies}~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} \hl{is based on the first-order Markov Chain model}~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
+\hl{The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event.
 Hence, Markov chain models have two main drawbacks.
@@
 Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-term dependencies between behaviors, to identify causal relationships.
-The goal of the current research is to test the applicability of pattern data mining and machine-learning methods to identify effective patterns of patient-counselor communication.
+The goal of the current research is to test the applicability of pattern data mining and machine-learning methods to identify effective patterns of patient-counselor communication.}
 
-Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity.
-Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
+\hl{Several prior works have reported the results of adopting machine learning methods, such as topic models}~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, \hl{classification methods}~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} \hl{and neural networks}~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} \hl{to the tasks of annotating MI transcripts and assessing intervention fidelity.
+Perez-Rosas et al.}~\cite{perez2017predicting} \hl{developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
 Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features achieving nearly 90\% accuracy.
-In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
-The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of 17 behavior codes, which is comparable to human coders.
-In a follow up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
+In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes}~\cite{hasan2016study}.
+\hl{The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders.
+In a follow up study}~\cite{hasan2018predicting}, \hl{we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
 RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews.
 This paper represents the next phase in this line of research.
-We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
+We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.}
 
-HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence.
+\hl{HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence.
 The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution.
 Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state.
-HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds.
-In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}.
-For example, a HMM-based classifier was applied Doppler ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
-In another study, HMM was used to capture important
-characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states, and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter.
-Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
-Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.
+HMMs were originally proposed for speech recognition}~\cite{rabiner1989tutorial}, \hl{in which the states were used to represent all English language sounds.
+In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling}~\cite{uuguz2007biomedical, amrhein2003client}.
+\hl{For example, a HMM-based classifier was applied Doppler ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease}~\cite{uuguz2007biomedical}.
+\hl{In another study, HMM was used to capture important
+characteristics of protein families}~\cite{amrhein2003client}.
+\hl{In the application of HMM to patient-counselor communication, hidden states, and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter.
+Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans, and action steps toward behavior change}~\cite{baldi1994hidden}) \hl{as the antecedents of patients' behavior change}~\cite{apodaca2009mechanisms}, \hl{there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
+Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.}
 
-Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together.
-Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
-Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web}, and outlier analysis~\cite{aggarwal2015outlier}.
-No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
+\hl{Frequent pattern mining}~\cite{aggarwal2014frequent} \hl{is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together.
+Agrawal and Srikant}~\cite{agrawal1994fast} \hl{first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
+Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics}~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, \hl{medical imaging}~\cite{olukunle2002fast}, \hl{chemical and biological analysis}~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, \hl{web mining}~\cite{srivastava2000web}, \hl{and outlier analysis}~\cite{aggarwal2015outlier}.
+\hl{No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
 A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
-To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in more compact form that are easier to interpret.
-In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
-FPClose is the best closed frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.
+To address this problem, we utilized the closed frequent itemset mining method}~\cite{pasquier1999discovering}, \hl{which produces fewer patterns in more compact form that are easier to interpret.
+In this study, we leveraged FPClose}~\cite{grahne2005fast}, \hl{an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
+FPClose is the best closed frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.}
 
-In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
+\hl{In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
 Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
@@
 This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions.
-Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention.
+Bertholet et al.}~\cite{bertholet2010change} \hl{used HMM to identify hidden states in a brief motivational intervention.
 Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''.
@@
 Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors.
-Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.
+Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.}
 
-\section{Materials and Methods}
+\section{\hl{Materials and Methods}}
 \label{sec:materialsAndMethods}
@@
 \label{subsec:dataset}
-The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
+\hl{The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}.
-The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
-The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
-A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
-The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
+These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)}~\cite{carcone2013provider}.
+\hl{The MYSCOPE is an adaptation of the original MI-SCOPE}~\cite{martin2005motivational}, \hl{a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)}~\cite{moyers2005assessing}, \hl{the MI Skill Code (MISC)}~\cite{apodaca2007motivational}, \hl{and Amrhein's conceptualization of change talk and commitment}~\cite{amrhein2004does}.
+\hl{A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
+The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table}~\ref{tab:codebook}.
 
@@
 \begin{longtable}{p{1.3cm}p{2cm}p{5cm}p{4.8cm}}
-\caption{MYSCOPE codebook} 
+\caption{\hl{MYSCOPE codebook}} 
 \label{tab:codebook} \\
@@
 \label{subsec:datapreprocessing}
-Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences.
+\hl{Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences.
 For each MI transcript, the stream of behavior codes from beginning to end was analyzed.
@@
 The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high.
-Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.
+Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.}
 
@@
 \label{subsubsec:hiddenMarkovModel}
-We applied the Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
+\hl{We applied the Hidden Markov Model (HMM)\footnote{we used the implementation in the hmmlearn package publicly available at http://hmmlearn.readthedocs.io/} to identify clusters of behavior codes corresponding to successful and unsuccessful communication sequences and to describe the relationships (transitions) between these clusters.
 Given a set of behavior code sequences, the posterior inference of HMM parameters involves the deduction of a temporal sequence of hidden states that best explains observations in each sequence.
 The rows in the emission probability matrix correspond to the distribution of observation symbols (i.e., the MYSCOPE behaviors displayed) for each hidden state and the transition probability matrix describes the transitions between the hidden states.
-Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:
+Training an HMM with a given number of hidden states ($N$) involves estimating the following parameters using the Baum-Welch algorithm:}
 
 \begin{itemize}
-\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
-\item $T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
-\item $E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
-\item $p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state
+\item \hl{$M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table}~\ref{tab:codebook})
+\item \hl{$T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$} 
+\item \hl{$E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ }
+\item \hl{$p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state}
 \end{itemize}
 
-We trained two HMM models, one using all successful sequences and the other one using all unsuccessful sequences.
+\hl{We trained two HMM models, one using all successful sequences and the other one using all unsuccessful sequences.
 Each model was trained with the objective of maximizing the log-likelihood of all observations in the corresponding set of sequences.
 The optimal number of hidden states was determined by estimating the Bayesian information criterion (BIC) of HMM models with a different number of hidden states and selecting the model with the smallest value of BIC, which takes into account both log-likelihood and a penalty term for the number of parameters in the model to avoid overfitting.
-Experiments with a different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure~\ref{fig:bic}).
+Experiments with a different number of hidden states in HMMs estimated on successful and unsuccessful sequences indicated that 5 hidden states were optimal for successful sequences and 2 hidden states were optimal for unsuccessful sequences (Figure}~\ref{fig:bic}).
 
@@
 \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
-\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}
+\caption{\hl{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}}
 \label{fig:bic}
@@
 \label{subsubsec:frequentPatternMining}
-We applied frequent pattern mining to identify frequently occurring patterns of temporally ordered patient-counselor behavior codes in successful and unsuccessful communication sequences.
+\hl{We applied frequent pattern mining to identify frequently occurring patterns of temporally ordered patient-counselor behavior codes in successful and unsuccessful communication sequences.
 Behavior codes in these patterns may be separated by one or more other codes.
-For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
+For this purpose, we utilized FPClose}~\cite{grahne2005fast}, \hl{an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF}~\cite{fournier2014spmf, fournier2016spmf}, \hl{to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
 SPMF is an open-source library providing more than 150 data mining algorithms.
-Popular non-close frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
-A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
-For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.
+Popular non-close frequent pattern mining algorithms include Apriori}~\cite{agrawal1994fast} \hl{and FP-Growth}~\cite{han2000mining}.
+\hl{A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
+For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure}~\ref{fig:fpm}, \hl{since these patterns appear at least 2 times, which is the minimum support count in this example.
 In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences.
-A \textit{frequent pattern is closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y.
-For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count.
+A \textit{frequent pattern is closed} if none of its supersets have the same support count}~\cite{pasquier1999discovering}, \hl{where a set X is a superset of another set Y, if X contains all the elements of the set Y.
+For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure}~\ref{fig:fpm} \hl{are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count.
 Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\}, and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count.
 On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0, and 1, respectively, which is less than the minimum support count, and thus are identified as non-frequent itemsets.
-Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed some prior works~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
+Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed some prior works}~\cite{moradi2017quantifying, liu1999mining} \hl{and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
 For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
+}
 
@@
 \includegraphics[width=0.8\textwidth]{figures/fpm.eps}
-\caption{A sample collection of sequences and different types of frequent patterns obtained by a frequent pattern mining method with the minimum support of 2}
+\caption{\hl{A sample collection of sequences and different types of frequent patterns obtained by a frequent pattern mining method with the minimum support of 2}}
 \label{fig:fpm}
@@
 
-\caption{Hidden Markov Model Emission Matrices}
+\caption{\hl{Hidden Markov Model Emission Matrices}}
 \label{tab:emission}
@@
 
-\caption{Hidden Markov Model Transition Matrices}
+\caption{\hl{Hidden Markov Model Transition Matrices}}
 \label{tab:transition}  
@@
 \label{sec:results}
-The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}.
-The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
+\hl{The transition and emission probability matrices of HMM model are reported in Tables}~\ref{tab:emission} \hl{and} \ref{tab:transition}.
+\hl{The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
 Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
@@
 The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
-The full transition matrix is presented in Table~\ref{tab:transition}.
+The full transition matrix is presented in Table}~\ref{tab:transition}.
 
-In contrast, the majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
+\hl{In contrast, the majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
 About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}.
@@
 
-Results from frequent pattern mining analysis is presented in Table~\ref{tab:patterns}.
-Reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
+Results from frequent pattern mining analysis is presented in Table}~\ref{tab:patterns}.
+\hl{Reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
 Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011), and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
@@
 In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in $>$10\% of the unsuccessful sequences.
-In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in $>$10\% of the successful sequences.
+In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in $>$10\% of the successful sequences.}
 
@@
 \begin{threeparttable}
-\caption{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences}
+\caption{\hl{Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences}}
 \label{tab:patterns}   
@@
 \label{sec:discussion}
-This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
-Previous studies have empirically linked counselors' use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
-The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data.
+\hl{This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk}~\cite{miller2009toward}.
+\hl{Previous studies have empirically linked counselors' use of MICO to higher rates of patient change talk in first-order Markov Chain models}~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
+\hl{The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data.
 Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences.
-The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
+The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models}~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
 
-In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement).
+\hl{In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement).
 In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
@@
 Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy, and sensitively provided information.
-Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
-However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index.
+Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,}~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} \hl{open questions phrased to elicit change talk,}~\cite{moyers2009session, carcone2013provider, jacques2016building}, \hl{and statements emphasizing the patient's decision-making autonomy}~\cite{carcone2013provider, jacques2016building}.
+\hl{However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index.
 Thus, the present study is the first to provide empirical evidence for these causal linkages.
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
-Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
-Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
+Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change}~\cite{douaihy2015motivational}.
+\hl{Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
+}
 
-The analysis of unsuccessful sequences, i.e., those resulting in a patient sustain talk statement, was typified by a combination of MICO and MI-inconsistent communication strategies (MIIN).
+\hl{The analysis of unsuccessful sequences, i.e., those resulting in a patient sustain talk statement, was typified by a combination of MICO and MI-inconsistent communication strategies (MIIN).
 Specifically, the majority of unsuccessful sequences in the HMM analysis began in a state of \textit{``ambivalence''} which was characterized by large proportions of counselor reflections of both change talk and sustain talk.
 Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed.
-These results are consistent with those of Gaume et al.~\cite{gaume2010counselor} who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
+These results are consistent with those of Gaume et al.}~\cite{gaume2010counselor} \hl{who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
 Specifically, counselors' use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
-Carcone et al.~\cite{carcone2013provider} found counselors' questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
-In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
+Carcone et al.}~\cite{carcone2013provider} \hl{found counselors' questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
+In contrast, Moyers et al.}~\cite{moyers2009session} \hl{found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
 These variable findings suggest a need to tailor the communication strategies MI counselors use to the treatment context.
+}
 
-This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns.
-We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts.
-Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}).
-We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
-In this study, we present the next phase of this research, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
+\hl{This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns.
+We have previously reported on the development of probabilistic generative models}~\cite{kotov2014towards, kotov2015interpretable} \hl{and application of novel features for maximum margin and deep learning classifiers}~\cite{hasan2016study} \hl{with the goal of automated annotation of MI session transcripts.
+Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits}~\cite{carcone2018using}).
+\hl{We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview}~\cite{hasan2018predicting}.
+\hl{In this study, we present the next phase of this research, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
 Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
@@
 An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.
+}
 
-This study is limited by the use of one dataset composed of 37 Motivational Interviewing transcripts of counseling sessions with African American adolescents in weight loss treatment.
+\hl{This study is limited by the use of one dataset composed of 37 Motivational Interviewing transcripts of counseling sessions with African American adolescents in weight loss treatment.
 Thus, there is a need to replicate these findings with larger and more diverse data samples as the findings may not be representative of communication patterns in other contexts employing the Motivational Interviewing framework.
@@
 Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently.
-One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to success, cannot be determined from these analyses.
+One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to success, cannot be determined from these analyses.}
 
@@
 \label{sec:conclusion}
-These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
-While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
+\hl{These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
+While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies}~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} \hl{and simple Markov chain models}~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, \hl{the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
 The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.
+}
 
@@
 \begin{acknowledgements}
-This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
+\hl{This study was supported by a grant from the National Institutes of Health, NIDDK R21DK108071, Carcone and Kotov, MPIs.
 We would like to thank the student assistants in the Department of Family Medicine and Public Health Sciences at Wayne State University School of Medicine for their help in developing the training dataset.
-The authors would like to thank Lisa Todd, JD, MA for her thoughtful feedback and clinical expertise in the interpretation of these data.
+The authors would like to thank Lisa Todd, JD, MA for her thoughtful feedback and clinical expertise in the interpretation of these data.}
 \end{acknowledgements}

### 2018-05-28 jhir/latex-manuscript/manuscript.tex
update manuscript to address reviewer's feedback

@@
 
-\date{Received: 3 December 2018 /Revised: 16 April 2018}
+\date{Received: 3 December 2017 /Revised: 16 April 2018, 29 June 2018 /Accepted: }
 % The correct dates will be entered by the editor
@@
 Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
-A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication.
-Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks.
-In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
+\hl{A significant barrier to this research is the time-consuming and resource-intensive cognitive task, the sequential analysis of pre-coded transcripts which is an integral part of the qualitative analysis of patient-counselor communication.
+Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating this cognitive task.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
 We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers.
@@
 The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}.
-The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
-MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''.
-Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change.
+The relational hypothesis suggests that counselors' use of accurate empathy, positive regard and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change.
+MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe and support) will lead to patient ``change talk''.
+Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for and/or commitment to behavior change.
 Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}.
-In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
+In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk).
 Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}.
@@
 Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
-A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
+A follow-up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
 Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
 Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
-They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
+They found \hl{that} MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
 A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}.
@@
 Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation of change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use.
-In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk.
-Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.
+In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy.
+A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{are} the counselor behaviors, \hl{which lead} to the elicitation of change talk.
+Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in \hl{which various} MICO strategies are effective.
 
@@
 Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-term dependencies between behaviors, to identify causal relationships.
-The goal of the current research is to test the applicability of pattern data mining and machine-learning methods to identify effective patterns of patient-counselor communication.
+The goal of the current research is to test the applicability \hl{of data mining and machine learning} methods to identify effective patterns of patient-counselor communication.
 
@@
 Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
-Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic, and semantic features achieving nearly 90\% accuracy.
-In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual, and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
+Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}.
+In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
 The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders.
-In a follow up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
+In a follow-up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
 RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews.
-This paper represents the next phase in this line of research.
-We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
+\hl{The current work builds on past work performed on automation of annotation.} We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
 
@@
 In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}.
-For example, a HMM-based classifier was applied Doppler ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
+For example, a HMM-based classifier was \hl{applied to Doppler} ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
 In another study, HMM was used to capture important
 characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states, and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter.
-Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
+In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter.
+Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
 Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.
@@
 Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
-Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web}, and outlier analysis~\cite{aggarwal2015outlier}.
+Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}.
 No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
@@
 In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
-FPClose is the best closed frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.
+\hl{FPClose is a state-of-the-art closed} frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.
 
@@
 The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
-The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
+The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
 A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$).
-The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
+The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.
 
@@
 & ``If you decide to follow a meal plan, it has to include occasional dessert.'' \\
-EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient's freedom of choice, autonomy, personal responsibility, and so forth.
+EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient's freedom of choice, autonomy, personal responsibility and so forth.
 & ``Okay.
@@
 & ``Healthy weight loss is about one to two pounds a week and once we get you set up and actually into the program you can look for that to happen for about one to two pounds a week to get you on that goal.''\\
-GINFOP & General information positive & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, expresses a concern, or offers educational information in a patient-centered manner (i.e., asking permission, using the third person, giving the opportunity to reject the information, and offering a menu of options).
+GINFOP & General information positive & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, expresses a concern, or offers educational information in a patient-centered manner (i.e., asking permission, using the third person, giving the opportunity to reject the information and offering a menu of options).
 & ``Okay.
@@
 \noalign{\smallskip}
-CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
-ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
-HUPO	 & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
-HUPW & High uptake weight & A turn that does develop the topic of the conversation.
-High Uptake statements include: weight related statements about actions of commitment, change talk, and ambivalence that occurred in the past, patient questions to the counselor, and session interruptions by persons who are not an active part of the treatment session.
-& ``Support is always good.
+CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
+ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
+\hl{HUPW} & \hl{High uptake weight} & \hl{A turn that does develop the topic of the conversation.
+High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.} & \hl{``Support is always good.
 You know that's a key factor.
-Mm-hmm.''\\
+Mm-hmm.''}\\
+\hl{HUPO} & \hl{High uptake other} & \hl{An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions} & \hl{``Yeah because the mentor comes and they take off and they go someplace for a little while.''}\\
 LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
@@
 \item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
-\item $T$ is an $NXN$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
-\item $E$ is an $NXM$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
-\item $p$ is the initial state distribution vector where $\pi_i$ is the probability of the first HMM state
+\item $T$ is an \hl{$N \times N$} state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
+\item $E$ is an \hl{$N \times M$} emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
+\item \hl{$\pi$} is the initial state distribution vector where $\pi_i$ is the probability of \hl{the $i^{th}$ state to be the first state}
 \end{itemize}
@@
 A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
-For example, \{A\}, \{C\}, \{D\}, and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.
+For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.
 In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences.
 A \textit{frequent pattern is closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y.
-For example, the itemsets \{A\}, \{C\}, \{D\}, and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\}, and \{B, C, D\} have the same support count.
-Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\}, and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count.
-On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0, and 1, respectively, which is less than the minimum support count, and thus are identified as non-frequent itemsets.
-Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed some prior works~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
+For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count.
+Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count.
+On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets.
+Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed \hl{prior work}~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
 For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%).
+Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
 Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time.
-\textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%).
+\textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%).
 Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively).
@@
 \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%).
-No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%).
+No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%).
 Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%).
 \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%).
-The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
+The \textit{``moderate receptivity''} \hl{state most often transitioned to \textit{``active feedback''} (27\%) and back to} \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
 The full transition matrix is presented in Table~\ref{tab:transition}.
@@
 About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}.
-Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements, and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences.
+Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences.
 Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.
 
-Results from frequent pattern mining analysis is presented in Table~\ref{tab:patterns}.
+Results from frequent pattern mining analysis \hl{are} presented in Table~\ref{tab:patterns}.
 Reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011), and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change \hl{talk (30.8\% versus 17.4\%, Pearson's chi-square test p= $<$0.001)}, statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011) and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
 In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.001), providing information (28.7\% versus 22.1\%, p=0.025) and other reflections (11.6\% versus 0\%, p= $<$0.001).
@@
 In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
-Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies.
+Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies.
 The frequent pattern mining results were similar.
-Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy, and sensitively provided information.
-Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
+Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy and sensitively provided information.
+Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
 However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index.
@@
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
-Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
+Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
 Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
@@
 These results are consistent with those of Gaume et al.~\cite{gaume2010counselor} who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military.
-Specifically, counselors' use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
+Specifically, counselors' use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk.
 Carcone et al.~\cite{carcone2013provider} found counselors' questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial.
 In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
-These variable findings suggest a need to tailor the communication strategies MI counselors use to the treatment context.
+These variable findings suggest a need to tailor the \hl{MI communication strategies} to the treatment context.
 
@@
 We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
-In this study, we present the next phase of this research, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
-Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
+\hl{In this paper, we present the current work builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.} Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
 We are also experimenting with machine learning methods to segment (parse) the stream of communications in healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.

### 2018-05-28 jhir/latex-manuscript/manuscript.tex
update figure 1

@@
 
-\date{Received: 3 December 2017 /Revised: 16 April 2018, 29 June 2018 /Accepted: }
+\date{Received: 3 December 2017 /Revised: 16 April 2018 and 29 June 2018 /Accepted: }
 % The correct dates will be entered by the editor

### 2018-05-30 jhir/latex-manuscript/manuscript.tex
update results

@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
+\hl{We noticed that 4,724 state transitions occurred in the successful sequences including same state transitions.
+On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
 Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time.
@@
 
-In contrast, the majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
+\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences where} majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
 About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}.

### 2018-05-30 jhir/bic_plot.R

@@ -1,0 +1,23 @@
+
+#library
+library(ggplot2)
+library(reshape2)
+library(pROC)
+library(calibrate)
+
+par(mfrow=c(1, 2))
+
+# melt data
+r1 <- read.csv("C:/Users/Kotov's Lab-Mehedi/Desktop/succ.csv")
+r2 <- read.csv("C:/Users/Kotov's Lab-Mehedi/Desktop/unsucc.csv")
+
+# plot graph
+plot(range(1,15), range(25000, 28000), type="n", ylab = "BIC", xlab = "# of hidden states in successful communications")
+lines(r1$n, r1$bic, type="b", lwd=1.5, lty=1, col='black', pch=18)
+points(x = 5, y = 25098.3168027, col = "red", pch = 1, cex = 2)
+textxy(r1$n, r1$bic, labs=c("","25102","","","25098","",""), pos=3,  cex = 0.7)
+
+# plot graph
+plot(range(1,15), range(6000, 9000), type="n", ylab = "BIC", xlab = "# of hidden states in unsuccessful communications")
+lines(r2$n, r2$bic, type="b", lwd=1.5, lty=1, col='black', pch=18)
+points(x = 2, y = 6175.59727261, col = "red", pch = 1, cex = 2)

### 2018-05-30 jhir/hmm-python/bic plots.R

@@ -1,18 +1,0 @@
-
-#library
-library(ggplot2)
-library(reshape2)
-library(pROC)
-
-par(mfrow=c(1, 2))
-
-# melt data
-r1 <- read.csv("/home/mehedi/teana/projects/myscope-sequential-analysis/jhir/hmm-python/results_succ.csv")
-r2 <- read.csv("/home/mehedi/teana/projects/myscope-sequential-analysis/jhir/hmm-python/results_unsucc.csv")
-
-# plot graph
-plot(r1$n, r1$bic, col='red', type="l", ylab = "BIC", xlab = "# of hidden states", cex.axis=1, cex.lab=1, lwd=1)
-
-# plot graph
-plot(r2$n, r2$bic, col='red', type="l", ylab = "BIC", xlab = "# of hidden states", cex.axis=1, cex.lab=1, lwd=1)
-

### 2018-05-30 jhir/hmm-python/obesity-hmm-transition_emission.py

@@ -43,6 +43,19 @@ all_codebook, success_codebook, unsuccess_codebook = utility.loadCodeBookFromTra
 n_observations = len(success_codebook)
 sequences, seq_labels, seq_lengths = utility.loadData(training_filename, success_codebook, 1)
 success_model = utility.getHMMModel(succ_hidden_states, n_observations, sequences, seq_lengths)
+succ_logproba, succ_state_seq = success_model.decode(sequences, seq_lengths)
+
+# compute state transitions
+state_dict = {}
+for i in range(len(succ_state_seq)-1):
+    state_from_to = str(succ_state_seq[i]) + "-" + str(succ_state_seq[i+1])
+    if state_from_to in state_dict:
+	    state_dict[state_from_to] = state_dict[state_from_to] + 1
+    else:
+	    state_dict[state_from_to] = 1
+
+print("Total # of transaction in successfull: ", str(len(succ_state_seq)-1))
+print(state_dict)
 
 for i in np.arange(0, len(success_model.startprob_)):
     success_start_fp.write("state" + str(i + 1) + ",")
@@ -78,6 +91,19 @@ for line_emission_prob in success_model.emissionprob_:
 n_observations = len(unsuccess_codebook)
 sequences, seq_labels, seq_lengths = utility.loadData(training_filename, unsuccess_codebook, 0)
 unsuccess_model = utility.getHMMModel(unsucc_hidden_states, n_observations, sequences, seq_lengths)
+unsucc_logproba, unsucc_state_seq = unsuccess_model.decode(sequences, seq_lengths)
+
+# compute state transitions
+state_dict = {}
+for i in range(len(unsucc_state_seq)-1):
+    state_from_to = str(unsucc_state_seq[i]) + "-" + str(unsucc_state_seq[i+1])
+    if state_from_to in state_dict:
+	    state_dict[state_from_to] = state_dict[state_from_to] + 1
+    else:
+	    state_dict[state_from_to] = 1
+
+print("Total # of transaction in unsuccessfull: ", str(len(unsucc_state_seq)-1))
+print(state_dict)
 
 for i in np.arange(0, len(unsuccess_model.startprob_)):
     unsuccess_start_fp.write("state" + str(i + 1) + ",")

### 2018-05-30 jhir/latex-manuscript/manuscript.tex
update manuscript with reviewer's feedback

@@
 % your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
 %
@@
 
-HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence.
+HMMs are widely used for the analysis of sequence data due to their ability to model \hl{long-range dependencies} between clusters of discrete observations in a sequence.
 The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution.
@@
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}.
-The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding greetings, farewell and conversations for interview setup such as camera and table setup.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
 The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
@@
 \label{subsubsec:frequentPatternMining}
-We applied frequent pattern mining to identify frequently occurring patterns of temporally ordered patient-counselor behavior codes in successful and unsuccessful communication sequences.
+We applied frequent pattern mining to identify frequently occurring \hl{patterns of patient-counselor} behavior codes in successful and unsuccessful communication sequences.
 Behavior codes in these patterns may be separated by one or more other codes.
@@
 Popular non-close frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
-A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
+A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
 For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.

### 2018-06-01 jhir/latex-manuscript/manuscript.tex

@@
 
-In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
+In this paper, we focus on computational methods to facilitate the sequential analysis \hl{of pre-coded MI transcripts} to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
 Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.

### 2018-06-01 jhir/latex-manuscript/manuscript.tex
update state transitions results

@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-\hl{We noticed that 4,724 state transitions occurred in the successful sequences including same state transitions.
+\hl{We observed 4,724 state transitions in the successful sequences where 45 state transitions occured within same states.
 On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
@@
 
-\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences where} majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
+\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences where 409 state transitions happened within same states}.
+The majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
 About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}.

### 2018-06-05 jhir/latex-manuscript/manuscript.tex
update state transitions results

@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-\hl{We observed 4,724 state transitions in the successful sequences where 45 state transitions occured within same states.
+\hl{We observed 4,724 state transitions in the successful sequences, where 4679 and 45 state transitions occured between different and same states, respectively.
 On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
@@
 
-\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences where 409 state transitions happened within same states}.
+\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences, where 697 and 409 state transitions happened between different and same states, respectively}.
 The majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).

### 2018-06-08 jhir/latex-manuscript/manuscript.tex
update manuscript

@@
 Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
-\hl{A significant barrier to this research is the time-consuming and resource-intensive cognitive task, the sequential analysis of pre-coded transcripts which is an integral part of the qualitative analysis of patient-counselor communication.
+\hl{A significant barrier to this research is the time-consuming sequential analysis of pre-coded transcripts which is an integral part of patient-counselor communication research.
 Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating this cognitive task.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
@@
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding greetings, farewell and conversations for interview setup such as camera and table setup.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
 The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-\hl{We observed 4,724 state transitions in the successful sequences, where 4679 and 45 state transitions occured between different and same states, respectively.
+\hl{We observed 4,724 state transitions in the successful sequences, where 4679 and 45 state transitions occurred between different and same states, respectively.
 On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).

### 2018-06-08 jhir/latex-manuscript/manuscript.tex
update results

@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-\hl{We observed 4,724 state transitions in the successful sequences, where 4679 and 45 state transitions occurred between different and same states, respectively.
+\hl{We observed 4,724 state transitions in the successful sequences, where 4,679 and 45 state transitions occurred between different and same states, respectively.
 On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).

### 2018-06-15 jhir/latex-manuscript/manuscript.tex
update manuscript based on april's feedback

@@
 In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{are} the counselor behaviors, \hl{which lead} to the elicitation of change talk.
-Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in \hl{which various} MICO strategies are effective.
+A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk.
+Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts} but not others suggesting a need for additional research to understand the \hl{treatment contexts in which various} MICO strategies are effective.
+\hl{The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.}
 
@@
 This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
-Previous studies have empirically linked counselors' use of MICO to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
+Previous studies have empirically linked counselors' use of \hl{MICO communication strategies} to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
 The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data.
@@
 Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
-However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index.
-Thus, the present study is the first to provide empirical evidence for these causal linkages.
+However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, \hl{when examining specifically which of the MICO communication strategies were empirically linked to the elicitation of change talk.} Thus, the present study is the first to provide empirical evidence for these causal linkages.
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.

### 2018-06-15 jhir/bic_plot.R

@@ -8,16 +8,16 @@ library(calibrate)
 par(mfrow=c(1, 2))
 
 # melt data
-r1 <- read.csv("C:/Users/Kotov's Lab-Mehedi/Desktop/succ.csv")
-r2 <- read.csv("C:/Users/Kotov's Lab-Mehedi/Desktop/unsucc.csv")
+r1 <- read.csv("/home/mehedi/Desktop/bic results/succ.csv")
+r2 <- read.csv("/home/mehedi/Desktop/bic results/unsucc.csv")
 
 # plot graph
-plot(range(1,15), range(25000, 28000), type="n", ylab = "BIC", xlab = "# of hidden states in successful communications")
+plot(range(1,10), range(25000, 26500), type="n", ylab = "BIC", xlab = "# of hidden states in successful communications")
 lines(r1$n, r1$bic, type="b", lwd=1.5, lty=1, col='black', pch=18)
 points(x = 5, y = 25098.3168027, col = "red", pch = 1, cex = 2)
-textxy(r1$n, r1$bic, labs=c("","25102","","","25098","",""), pos=3,  cex = 0.7)
+textxy(r1$n, r1$bic, labs=c("","25102","25122","25204","25098","25120",""), pos=3,  cex = 0.7)
 
 # plot graph
-plot(range(1,15), range(6000, 9000), type="n", ylab = "BIC", xlab = "# of hidden states in unsuccessful communications")
+plot(range(1,10), range(6000, 8000), type="n", ylab = "BIC", xlab = "# of hidden states in unsuccessful communications")
 lines(r2$n, r2$bic, type="b", lwd=1.5, lty=1, col='black', pch=18)
 points(x = 2, y = 6175.59727261, col = "red", pch = 1, cex = 2)

### 2018-06-15 jhir/latex-manuscript/manuscript.tex
update abstract

@@
 Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
-\hl{A significant barrier to this research is the time-consuming sequential analysis of pre-coded transcripts which is an integral part of patient-counselor communication research.
-Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating this cognitive task.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
+\hl{An integral part of patient-counselor communication research is the sequential analysis of pre-coded MI transcripts.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
 We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers.

### 2018-06-15 jhir/latex-manuscript/manuscript.tex
update abstract

@@
 Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
-\hl{An integral part of patient-counselor communication research is the sequential analysis of pre-coded MI transcripts.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
+\hl{An integral part of this research is the sequential analysis of pre-coded MI transcripts.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
 We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers.

### 2018-06-22 jhir/latex-manuscript/manuscript.tex
incorporate co-authors feedback

@@
 We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
-\hl{In this paper, we present the current work builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.} Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
-We are also experimenting with machine learning methods to segment (parse) the stream of communications in healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
+\hl{In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.} Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
+We are also experimenting with machine learning methods to segment (parse) the stream of communications \hl{in a healthy} eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
 Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions.
@@
 Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently.
-One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to success, cannot be determined from these analyses.
+One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined \hl{for failure to success}, cannot be determined from these analyses.
 

### 2018-06-27 jhir/latex-manuscript/manuscript.tex
incorporate Alex feedback

@@
 
-\title{Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis
+\title{\hl{Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis}
 }
@@
 Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}.
-Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
-A follow-up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
+Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and \hl{found out that change talk} was significantly more likely after MICO and sustain talk more likely after MIIN.
+A follow-up study with the same population \hl{found that change talk} was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
 Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
@@
 A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk.
-Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts} but not others suggesting a need for additional research to understand the \hl{treatment contexts in which various} MICO strategies are effective.
+Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts, but not others,} suggesting a need for additional research to understand the \hl{treatment contexts, in which various} MICO strategies are effective.
 \hl{The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.}
@@
 The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event.
-Hence, Markov chain models have two main drawbacks.
+Hence, \hl{Markov Chain models} have two main drawbacks.
 The first is their inability to preserve the long-range dependencies between observations in a sequence.
@@
 Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity.
-Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
-Because MI fidelity is assessed in part by the ratio of counselor questions to reflection (a minimum goal is 1:1), their system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}.
+Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational \hl{interviews.
+Their} system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}.
 In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
@@
 characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter.
+In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states, \hl{may correspond to a patients' underlying} motivational state during a patient-counselor encounter.
 Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
@@
 Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}.
-No published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
+\hl{However, no published} research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
 A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
@@
 In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
-\hl{FPClose is a state-of-the-art closed} frequent itemset mining algorithm, which has shown highest performance in terms of running time and memory consumption.
+\hl{FPClose is a state-of-the-art closed} frequent itemset mining algorithm, which has \hl{demonstrated good performance} in terms of running time and memory consumption.
 
@@
 \label{subsec:dataset}
-The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
+The \hl{dataset} for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+These encounters \hl{had been previously} annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
 The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
@@
 Behavior codes in these patterns may be separated by one or more other codes.
-For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
+For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art \hl{closed} frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
 SPMF is an open-source library providing more than 150 data mining algorithms.
-Popular non-close frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
-A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}.
+Popular \hl{non-closed} frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
+A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}.
 For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.
@@
 Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed \hl{prior work}~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
-For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
+For each pattern, the statistical significance \hl{of the difference} between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{p-value of Pearson's} chi-square test.
 
@@
 Reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change \hl{talk (30.8\% versus 17.4\%, Pearson's chi-square test p= $<$0.001)}, statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011) and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
-In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.001), providing information (28.7\% versus 22.1\%, p=0.025) and other reflections (11.6\% versus 0\%, p= $<$0.001).
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change \hl{talk (30.8\% versus 17.4\%, Pearson's chi-square test p $<$ 0.001)}, statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011) and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
+In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, \hl{p $<$ 0.001}), reflections of sustain talk (27.1\% versus 15.8\%, \hl{p $<$ 0.001}), providing information (28.7\% versus 22.1\%, p=0.025) and other reflections (11.6\% versus 0\%, \hl{p $<$ 0.001}).
 In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in $>$10\% of the unsuccessful sequences.
@@
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
-Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
+Adaptations of MI for the \hl{healthcare} setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
 Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
@@
 \label{sec:conflictofInterest}
-On behalf of all authors, the corresponding author states that there is no conflict of interest.
+%The authors have no conflict of interest to declare.
+\hl{The authors declare that they have no conflict of interest.}
 

### 2018-06-28 jhir/latex-manuscript/manuscript.tex
finished proofread

@@
 In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and autonomy supportive statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk.
+A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and \hl{autonomy-supportive} statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk.
 Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts, but not others,} suggesting a need for additional research to understand the \hl{treatment contexts, in which various} MICO strategies are effective.
@@
 Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences.
-Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-term dependencies between behaviors, to identify causal relationships.
+Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and \hl{long-range} dependencies between behaviors, to identify causal relationships.
 The goal of the current research is to test the applicability \hl{of data mining and machine learning} methods to identify effective patterns of patient-counselor communication.
@@
 Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational \hl{interviews.
-Their} system employed a SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}.
+Their} system employed \hl{an SVM} classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}.
 In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
@@
 In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}.
-For example, a HMM-based classifier was \hl{applied to Doppler} ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
+For example, \hl{an HMM-based} classifier was \hl{applied to Doppler} ultrasound imaging data to extract features \hl{from} the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
 In another study, HMM was used to capture important
 characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states, \hl{may correspond to a patients' underlying} motivational state during a patient-counselor encounter.
+In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to a patients' underlying} motivational state during a patient-counselor encounter.
 Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
@@
 A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
-To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in more compact form that are easier to interpret.
+To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in \hl{a more} compact form that are easier to interpret.
 In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
@@
 The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
-These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research.
-First-order Markov chain models identify the likely transitions between individual behaviors.
+These two approaches offer the following advantages over the first-order \hl{Markov Chain-based} methods most typically used in MI research.
+First-order \hl{Markov Chain} models identify the likely transitions between individual behaviors.
 In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications.
@@
 In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences.
-A \textit{frequent pattern is closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y.
-For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count.
-Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count.
+A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y.
+For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent \hl{patterns since} their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count.
+Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent \hl{patterns since} none of their supersets have the same support count.
 On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets.
@@
 Previous studies have empirically linked counselors' use of \hl{MICO communication strategies} to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
-The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-term dependencies in the data.
-Unlike simple first-order Markov chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences.
-The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
+The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering \hl{longer-range} dependencies in the data.
+Unlike simple first-order \hl{Markov Chain} models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences.
+The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple \hl{Markov Chain} models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
 
@@
 In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
-Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies.
+Other \hl{high-frequency} counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies.
 The frequent pattern mining results were similar.
 Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy and sensitively provided information.
-Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
+Previous studies of MI behavior code sequences, which relied on first-order \hl{Markov Chain} models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
 However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, \hl{when examining specifically which of the MICO communication strategies were empirically linked to the elicitation of change talk.} Thus, the present study is the first to provide empirical evidence for these causal linkages.
@@
 These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
-While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data.
+While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple \hl{Markov Chain} models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering \hl{long-range} dependencies in the data.
 The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.

### 2018-06-28 jhir/latex-manuscript/manuscript.tex
prepared final copy for the minor revision

@@
 characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to a patients' underlying} motivational state during a patient-counselor encounter.
+In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to patients' underlying} motivational state during a patient-counselor encounter.
 Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.

### 2018-06-28 jhir/latex-manuscript/manuscript.tex
prepared final copy for the minor revision

@@
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters \hl{had been previously} annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+These encounters \hl{had been previously} annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations that correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
 The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
@@
 Popular \hl{non-closed} frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
-A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}.
+A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the \hl{\textit{minimum support count}}.
 For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.

### 2018-06-28 jhir/latex-manuscript/manuscript.tex
finalized minor revision

@@
 Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}.
-Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and \hl{found out that change talk} was significantly more likely after MICO and sustain talk more likely after MIIN.
+Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and \hl{found that change talk} was significantly more likely after MICO and sustain talk more likely after MIIN.
 A follow-up study with the same population \hl{found that change talk} was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.

### 2018-06-29 jhir/latex-manuscript/manuscript.tex
submitted to JHIR as minor revision

--- 2018-06-28 jhir/latex-manuscript/manuscript.tex
@@
 \usepackage{color, colortbl}
-\usepackage{soul}
 \definecolor{lightgraycol}{gray}{0.95}
@@
 
-\title{\hl{Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis}
+\title{Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis
 }
@@
 Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research.
-\hl{An integral part of this research is the sequential analysis of pre-coded MI transcripts.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
+An integral part of this research is the sequential analysis of pre-coded MI transcripts.
+In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice.
 We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers.
@@
 Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}.
-Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and \hl{found out that change talk} was significantly more likely after MICO and sustain talk more likely after MIIN.
-A follow-up study with the same population \hl{found that change talk} was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
+Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found that change talk was significantly more likely after MICO and sustain talk more likely after MIIN.
+A follow-up study with the same population found that change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}.
 Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk.
 Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service.
-They found \hl{that} MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
+They found that MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed.
 A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}.
@@
 In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk and statements emphasizing decision-making autonomy.
-A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion \hl{that} asking questions phrased to elicit change talk, reflections of change talk and \hl{autonomy-supportive} statements \hl{were} the counselor behaviors, \hl{which led} to the elicitation of change talk.
-Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors\hl{, however, }led to change talk in some \hl{treatment contexts, but not others,} suggesting a need for additional research to understand the \hl{treatment contexts, in which various} MICO strategies are effective.
-\hl{The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.}
+A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion that asking questions phrased to elicit change talk, reflections of change talk and autonomy-supportive statements were the counselor behaviors, which led to the elicitation of change talk.
+Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors, however, led to change talk in some treatment contexts, but not others, suggesting a need for additional research to understand the treatment contexts, in which various MICO strategies are effective.
+The current study contributes to existing knowledge by examining African American adolescents in weight loss treatment.
 
@@
 The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event.
-Hence, \hl{Markov Chain models} have two main drawbacks.
+Hence, Markov Chain models have two main drawbacks.
 The first is their inability to preserve the long-range dependencies between observations in a sequence.
@@
 Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences.
-Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and \hl{long-range} dependencies between behaviors, to identify causal relationships.
-The goal of the current research is to test the applicability \hl{of data mining and machine learning} methods to identify effective patterns of patient-counselor communication.
+Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships.
+The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.
 
 Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity.
-Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational \hl{interviews.
-Their} system employed \hl{an SVM} classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and \hl{semantic features}.
+Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
+Their system employed an SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and semantic features.
 In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
@@
 RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews.
-\hl{The current work builds on past work performed on automation of annotation.} We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
+The current work builds on past work performed on automation of annotation.
+We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
 
-HMMs are widely used for the analysis of sequence data due to their ability to model \hl{long-range dependencies} between clusters of discrete observations in a sequence.
+HMMs are widely used for the analysis of sequence data due to their ability to model long-range dependencies between clusters of discrete observations in a sequence.
 The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution.
@@
 In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}.
-For example, \hl{an HMM-based} classifier was \hl{applied to Doppler} ultrasound imaging data to extract features \hl{from} the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
+For example, an HMM-based classifier was applied to Doppler ultrasound imaging data to extract features from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}.
 In another study, HMM was used to capture important
 characteristics of protein families~\cite{amrhein2003client}.
-In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states \hl{may correspond to patients' underlying} motivational state during a patient-counselor encounter.
+In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states may correspond to patients' underlying motivational state during a patient-counselor encounter.
 Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk.
@@
 Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}.
-\hl{However, no published} research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
+However, no published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication.
 A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns.
-To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in \hl{a more} compact form that are easier to interpret.
+To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in a more compact form that are easier to interpret.
 In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk.
-\hl{FPClose is a state-of-the-art closed} frequent itemset mining algorithm, which has \hl{demonstrated good performance} in terms of running time and memory consumption.
+FPClose is a state-of-the-art closed frequent itemset mining algorithm, which has demonstrated good performance in terms of running time and memory consumption.
 
-In this paper, we focus on computational methods to facilitate the sequential analysis \hl{of pre-coded MI transcripts} to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
+In this paper, we focus on computational methods to facilitate the sequential analysis of pre-coded MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions.
 Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk.
@@
 The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
-These two approaches offer the following advantages over the first-order \hl{Markov Chain-based} methods most typically used in MI research.
-First-order \hl{Markov Chain} models identify the likely transitions between individual behaviors.
+These two approaches offer the following advantages over the first-order Markov Chain-based methods most typically used in MI research.
+First-order Markov Chain models identify the likely transitions between individual behaviors.
 In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications.
@@
 \label{subsec:dataset}
-The \hl{dataset} for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
+The dataset for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
-These encounters \hl{had been previously} annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
+These encounters had been previously annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} excluding conversations that correspond to greetings, farewell and interview setups such as table and camera settings.
+The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions.
 The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}.
@@
 ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
-\hl{HUPW} & \hl{High uptake weight} & \hl{A turn that does develop the topic of the conversation.
-High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.} & \hl{``Support is always good.
+HUPW & High uptake weight & A turn that does develop the topic of the conversation.
+High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.
+& ``Support is always good.
 You know that's a key factor.
-Mm-hmm.''}\\
-\hl{HUPO} & \hl{High uptake other} & \hl{An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions} & \hl{``Yeah because the mentor comes and they take off and they go someplace for a little while.''}\\
+Mm-hmm.''\\
+HUPO & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
 LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
@@
 \item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
-\item $T$ is an \hl{$N \times N$} state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
-\item $E$ is an \hl{$N \times M$} emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
-\item \hl{$\pi$} is the initial state distribution vector where $\pi_i$ is the probability of \hl{the $i^{th}$ state to be the first state}
+\item $T$ is an $N \times N$ state transition probability matrix, in which $t_{ij}$ is the probability of HMM transitioning from state $i$ to state $j$  
+\item $E$ is an $N \times M$ emission probability matrix, in which $e_{jk}$ is the probability of observing symbol $k$, when HMM is in state $j$ 
+\item $\pi$ is the initial state distribution vector where $\pi_i$ is the probability of the $i^{th}$ state to be the first state
 \end{itemize}
@@
 \includegraphics[width=1.0\textwidth]{figures/bic-vs-hidden-states.eps}
-\caption{\hl{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}}
+\caption{Bayesian information criterion (BIC) of HMM models of successful (left) and unsuccessful (right) interviews by varying the number of hidden states}
 \label{fig:bic}
@@
 \label{subsubsec:frequentPatternMining}
-We applied frequent pattern mining to identify frequently occurring \hl{patterns of patient-counselor} behavior codes in successful and unsuccessful communication sequences.
+We applied frequent pattern mining to identify frequently occurring patterns of patient-counselor behavior codes in successful and unsuccessful communication sequences.
 Behavior codes in these patterns may be separated by one or more other codes.
-For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art \hl{closed} frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
+For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences.
 SPMF is an open-source library providing more than 150 data mining algorithms.
-Popular \hl{non-closed} frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
-A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}.
+Popular non-closed frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}.
+A \textit{frequent pattern} is defined as a pattern of observations, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}.
 For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example.
@@
 A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y.
-For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent \hl{patterns since} their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count.
-Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent \hl{patterns since} none of their supersets have the same support count.
+For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count.
+Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns since none of their supersets have the same support count.
 On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets.
-Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed \hl{prior work}~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
-For each pattern, the statistical significance \hl{of the difference} between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{p-value of Pearson's} chi-square test.
+Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
+For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value of Pearson's chi-square test.
 
@@
 The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-\hl{We observed 4,724 state transitions in the successful sequences, where 4,679 and 45 state transitions occurred between different and same states, respectively.
-On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
+We observed 4,724 state transitions in the successful sequences, where 4,679 and 45 state transitions occurred between different and same states, respectively.
+On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.
+Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%).
 Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time.
@@
 \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%).
-The \textit{``moderate receptivity''} \hl{state most often transitioned to \textit{``active feedback''} (27\%) and back to} \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
+The \textit{``moderate receptivity''} state most often transitioned to \textit{``active feedback''} (27\%) and back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency.
 The full transition matrix is presented in Table~\ref{tab:transition}.
 
-\hl{In contrast, 1,106 state transitions occurred within the unsuccessful sequences, where 697 and 409 state transitions happened between different and same states, respectively}.
+In contrast, 1,106 state transitions occurred within the unsuccessful sequences, where 697 and 409 state transitions happened between different and same states, respectively.
 The majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
@@
 
-Results from frequent pattern mining analysis \hl{are} presented in Table~\ref{tab:patterns}.
+Results from frequent pattern mining analysis are presented in Table~\ref{tab:patterns}.
 Reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%).
-Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change \hl{talk (30.8\% versus 17.4\%, Pearson's chi-square test p $<$ 0.001)}, statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011) and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
-In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, \hl{p $<$ 0.001}), reflections of sustain talk (27.1\% versus 15.8\%, \hl{p $<$ 0.001}), providing information (28.7\% versus 22.1\%, p=0.025) and other reflections (11.6\% versus 0\%, \hl{p $<$ 0.001}).
+Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, Pearson's chi-square test p $<$ 0.001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.001), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.011) and reflections of commitment language (20.7\% versus 15.1\%, p=0.042).
+In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p $<$ 0.001), reflections of sustain talk (27.1\% versus 15.8\%, p $<$ 0.001), providing information (28.7\% versus 22.1\%, p=0.025) and other reflections (11.6\% versus 0\%, p $<$ 0.001).
 In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in $>$10\% of the unsuccessful sequences.
@@
 This study applied HMM and frequent pattern mining to test the fundamental hypothesis guiding Motivational Interviewing, which posits that counselors use of ``MI-consistent'' communication strategies (MICO) will lead to patient change talk~\cite{miller2009toward}.
-Previous studies have empirically linked counselors' use of \hl{MICO communication strategies} to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
-The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering \hl{longer-range} dependencies in the data.
-Unlike simple first-order \hl{Markov Chain} models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences.
-The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple \hl{Markov Chain} models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
+Previous studies have empirically linked counselors' use of MICO communication strategies to higher rates of patient change talk in first-order Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}.
+The present study leveraged data mining methods to provide an even stronger evidence for MI's fundamental hypothesis by considering longer-range dependencies in the data.
+Unlike simple first-order Markov Chain models, frequent pattern mining considers behavioral antecedents beyond the counselor behavior immediately preceding a patient change talk statement, while HMM identifies groups of communication behaviors occurring in successful and unsuccessful communication sequences.
+The ability of HMM and frequent pattern mining to identify critical patterns in patient-counselor communication sequences advances research in the field of Motivational Interviewing, which has previously relied upon simple Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2008counsellor, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building}.
 
@@
 In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations.
-Other \hl{high-frequency} counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies.
+Other high-frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk and the provision of information using patient-centered strategies.
 The frequent pattern mining results were similar.
 Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy and sensitively provided information.
-Previous studies of MI behavior code sequences, which relied on first-order \hl{Markov Chain} models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
-However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, \hl{when examining specifically which of the MICO communication strategies were empirically linked to the elicitation of change talk.} Thus, the present study is the first to provide empirical evidence for these causal linkages.
+Previous studies of MI behavior code sequences, which relied on first-order Markov Chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building} and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}.
+However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examining specifically which of the MICO communication strategies were empirically linked to the elicitation of change talk.
+Thus, the present study is the first to provide empirical evidence for these causal linkages.
 One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial.
-Adaptations of MI for the \hl{healthcare} setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
+Adaptations of MI for the healthcare setting suggest that asking questions, demonstrating active listening through reflections and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}.
 Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.
@@
 In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not.
-These variable findings suggest a need to tailor the \hl{MI communication strategies} to the treatment context.
+These variable findings suggest a need to tailor the MI communication strategies to the treatment context.
 
@@
 We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
-\hl{In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.} Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
-We are also experimenting with machine learning methods to segment (parse) the stream of communications \hl{in a healthy} eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
+In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
+Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
+We are also experimenting with machine learning methods to segment (parse) the stream of communications in a healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
 Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions.
@@
 Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently.
-One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined \hl{for failure to success}, cannot be determined from these analyses.
+One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for failure to success, cannot be determined from these analyses.
 
@@
 These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods.
-While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple \hl{Markov Chain} models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering \hl{long-range} dependencies in the data.
+While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov Chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-range dependencies in the data.
 The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.
@@
 \label{sec:conflictofInterest}
-%The authors have no conflict of interest to declare.
-\hl{The authors declare that they have no conflict of interest.}
+The authors declare that they have no conflict of interest.
 

### 2018-09-04 jhir/latex-manuscript/manuscript_publish_highlighted.tex
prepared final version of manuscript for publication

--- 2018-06-29 jhir/latex-manuscript/manuscript.tex
@@
 
-\date{Received: 3 December 2017 /Revised: 16 April 2018 and 29 June 2018 /Accepted: }
+\date{Received: 3 December 2017 /Revised: 16 April 2018 and 29 June 2018 /Accepted: 31 August 2018}
 % The correct dates will be entered by the editor
@@
 However, in MI, an observed behavior can be influenced by any of the preceding behaviors.
-The second drawback is their inability to consider similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome.
+The second drawback is their inability to consider \hl{the} similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome.
 Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences.
@@
 HUPW & High uptake weight & A turn that does develop the topic of the conversation.
-High Uptake statements include: weight related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.
+High Uptake statements include: \hl{weight-related} statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.
 & ``Support is always good.
@@
 Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
-For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value of Pearson's chi-square test.
+For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{the} p-value of Pearson's chi-square test.
 
@@
 Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}).
-We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
+We also developed and evaluated probabilistic and deep learning methods for the task of predicting \hl{the} success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
 In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.

### 2018-09-18 jhir/latex-manuscript/manuscript_publish.tex
integrate April's comments into the final manuscript

--- 2018-09-04 jhir/latex-manuscript/manuscript_publish_highlighted.tex
@@
 MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe and support) will lead to patient ``change talk''.
-Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for and/or commitment to behavior change.
+Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for and/or commitment to behavior change~\cite{amrhein2004does}.
 Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}.
@@
 The first is their inability to preserve the long-range dependencies between observations in a sequence.
-However, in MI, an observed behavior can be influenced by any of the preceding behaviors.
-The second drawback is their inability to consider \hl{the} similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome.
+In MI, an observed behavior can be influenced by any of the preceding behaviors.
+The second drawback is their inability to consider similarities between behavior codes and, consequently, first-order Markov chain models are unable to identify multiple similar behaviors that lead to the same outcome.
 Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences.
-Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships.
+There is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships.
 The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.
 
-Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity.
-Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews.
-Their system employed an SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and semantic features.
-In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
-The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders.
+Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts for the assessment of intervention fidelity.
+Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework.
+Their system employed a Support Vector Machines (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic (structure of the clinician statements) and semantic (cognitive state) features.
+In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical (words expressed), contextual (prior code) and semantic (inferred cognitive state based on Linguistic Inquiry and Word Count dictionaries~\cite{tausczik2010psychological}) features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}.
+An SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, accuracy comparable to human coders.
 In a follow-up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk).
 RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews.
-The current work builds on past work performed on automation of annotation.
-We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.
+The current work builds on our past work by examining the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining for the identification of the counselor communication strategies leading to patient change talk.
 
@@
 Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together.
-Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases.
+Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, developed to identify customer purchasing patterns.
 Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}.
@@
 Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention.
-Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''.
-These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention.
+Limiting their HMM model to three hidden states which were characterized as ``towards change'', ``away from change'' and ``non-determined'', these states were used to predict drinking outcomes 12 months post-intervention.
 In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication.
-The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
+The goal of this study was to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions.
 These two approaches offer the following advantages over the first-order Markov Chain-based methods most typically used in MI research.
@@
 \label{subsec:dataset}
-The dataset for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
+The dataset for this study consists of 37 audio-recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss.
 Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor.
@@
 HUPW & High uptake weight & A turn that does develop the topic of the conversation.
-High Uptake statements include: \hl{weight-related} statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.
+High Uptake statements include: weight-related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session.
 & ``Support is always good.
@@
 Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences.
-For each MI transcript, the stream of behavior codes from beginning to end was analyzed.
+For each MI transcript, the stream of behavior codes from the beginning of a session to the end of the session was analyzed.
 Successful sequences were defined as those that resulted in a patient change talk or commitment language statement.
@@
 A total of 1,360 sequences were generated using this approach.
-The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high.
+The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which patients initial motivation for behavior change is typically high.
 Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.
@@
 Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences.
-For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{the} p-value of Pearson's chi-square test.
+For each pattern, the statistical significance of the difference between successful and unsuccessful sequences was computed with Pearson's chi-square test.
 
@@
 \label{sec:results}
-The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}.
-The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states.
-We observed 4,724 state transitions in the successful sequences, where 4,679 and 45 state transitions occurred between different and same states, respectively.
+The transition and emission probability matrices of the HMM models are reported in Tables~\ref{tab:emission} and \ref{tab:transition}.
+Three behaviors represented 45-69\% of each state's emission probability mass and, thus, were used to interpret the emission matrix and label the hidden states.
+We observed 4,724 state transitions in the successful sequences; 4,679 and 45 state transitions occurred between different and same states, respectively.
 On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.
@@
 
-In contrast, 1,106 state transitions occurred within the unsuccessful sequences, where 697 and 409 state transitions happened between different and same states, respectively.
+In contrast, 1,106 state transitions occurred within the unsuccessful sequences; 697 and 409 state transitions happened between different and same states, respectively.
 The majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%).
@@
 Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}).
-We also developed and evaluated probabilistic and deep learning methods for the task of predicting \hl{the} success of motivational interviews at any point during the interview~\cite{hasan2018predicting}.
-In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.
-Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems.
+We also developed and evaluated probabilistic and deep learning methods for the task of predicting the change talk at any point during the motivational interview~\cite{hasan2018predicting}.
+Our current work builds on this past work to automatically annotate clinical encounters, specifically, the current study presents two approaches for the sequential analysis of patient-counselor communication data for the purpose of identifying the counselor communication strategies linked to the elicitation of change talk and sustain talk.
+Next steps include examining the performance of the HMM and frequent pattern mining models in diverse data sets representing different populations and behavioral problems.
 We are also experimenting with machine learning methods to segment (parse) the stream of communications in a healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code.
